### 2024-10-02 manuscript/manuscript.qmd
initial commit

---
title: ""

code-fold: false
echo: false
number-sections: false

format: 
  docx:
    fig-cap-location: top
    tbl-cap-location: top
    reference-doc: custom-reference-doc.docx
  pdf:
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
    fig-cap-location: top
    tbl-cap-location: top

editor: visual
bibliography: references.bib
#csl: american-journal-of-epidemiology.csl

filters:
  - docx-landscape.lua
---

```{r}
#| label: setup
#| include: false

library(here)
library(captioner)

table_cap <- captioner(prefix = "Table")
```

**Authors**: Janick Weberpals^1^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^

[Author affiliations:]{.underline}

^1^ Division of Pharmacoepidemiology and Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital, Harvard Medical School, Boston, MA, USA

^2^ Oncology Center of Excellence, US Food and Drug Administration, Silver Spring, MD, USA

[**Correspondence:**]{.underline}

Shirley V. Wang, PhD

Division of Pharmacoepidemiology and Pharmacoeconomics,

Department of Medicine, Brigham and Women's Hospital, Harvard Medical School,

1620 Tremont Street, Suite 3030-R, Boston, MA 02120, USA

Phone: +1 617-278-0932

Fax: + 1 617-232-8602

Email: [jweberpals\@bwh.harvard.edu](mailto:jweberpals@bwh.harvard.edu)

[**Article type:**]{.underline} Original Research Article

[**Word count:**]{.underline} 3,806 words / 4000 words

[**Tables:**]{.underline} 1

[**Figures:**]{.underline} 6

[**Supplementary material:**]{.underline} Supplementary tables and figures

[**Short running title**]{.underline}: High-dimensional multiple imputation (HDMI)

[**Keywords:**]{.underline} Missing data, Confounding, EHR, NLP, Real-World Evidence

[**Funding Statement:**]{.underline} This project was supported by Task Order 75F40119F19002 under Master Agreement 75F40119D10037 from the US Food and Drug Administration (FDA). Additional funding was provided by NIH RO1LM013204.

[**Competing Interests Statement:**]{.underline} Dr. Schneeweiss is participating in investigator-initiated grants to the Brigham and Women's Hospital from Boehringer Ingelheim, Takeda, and UCB unrelated to the topic of this study. He owns equity in Aetion Inc., a software manufacturer. He is an advisor to Temedica GmbH, a patient-oriented data generation company. His interests were declared, reviewed, and approved by the Brigham and Women's Hospital in accordance with their institutional compliance policies.

[**Data sharing statement:**]{.underline} ...

[**Analytic code sharing statement:**]{.underline} ...

*Manuscript last updated: `r Sys.time()`*

{{< pagebreak >}}

# Abstract {.unnumbered}

xxx words/xxx words

...

{{< pagebreak >}}

# Background

...

{{< pagebreak >}}

# Methods

...

{{< pagebreak >}}

# Results {#sec-results}

...

# Discussion

...

## Conclusions

...

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

\blandscape

# Tables {.unnumbered}

{{< pagebreak >}}

# Figures {.unnumbered}

\elandscape

### 2024-10-02 manuscript/supplement.qmd
initial commit

---
title: "Supplementary Material"
subtitle: "..."

echo: false
number-sections: true

format: 
  pdf:
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
    fig-cap-location: top
    crossref:
      fig-title: '**Supplementary Figure**'
      tbl-title: '**Supplementary Table**'

editor: visual
bibliography: references.bib
#csl: american-journal-of-epidemiology.csl
---

```{r}
#| label: setup_supplement
#| include: false

# table_cap <- captioner(prefix = "Supplementary Table")
```

**Authors**: ...

$\vspace{0.25cm}$

*Supplementary Material last updated: `r Sys.Date()`*

{{< pagebreak >}}

-->

# Supplementary Tables

...

{{< pagebreak >}}

# Supplementary Figures

...

{{< pagebreak >}}

# References {.unnumbered}

::: refs
:::

### 2024-10-02 manuscript/manuscript.qmd
added references.bib

@@
 
-Email: [jweberpals\@bwh.harvard.edu](mailto:jweberpals@bwh.harvard.edu)
+Email: [SWANG1\@BWH.HARVARD.EDU](SWANG1@BWH.HARVARD.EDU)
 
@@
 
-[**Word count:**]{.underline} 3,806 words / 4000 words
+[**Word count:**]{.underline} xxx words / xxx words
 
-[**Tables:**]{.underline} 1
+[**Tables:**]{.underline} x
 
-[**Figures:**]{.underline} 6
+[**Figures:**]{.underline} x
 
@@
 
-[**Short running title**]{.underline}: High-dimensional multiple imputation (HDMI)
+[**Short running title**]{.underline}: ...
 
-[**Keywords:**]{.underline} Missing data, Confounding, EHR, NLP, Real-World Evidence
+[**Keywords:**]{.underline} Real-World Evidence
 
-[**Funding Statement:**]{.underline} This project was supported by Task Order 75F40119F19002 under Master Agreement 75F40119D10037 from the US Food and Drug Administration (FDA).
-Additional funding was provided by NIH RO1LM013204.
+[**Funding Statement:**]{.underline} This project was supported by ...
 
@@
 
-...
+... [@weberpals2024]
 

### 2024-10-02 manuscript/manuscript.qmd
figure orientation

@@
 ---
-title: ""
+title: "A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)"
 
@@
 bibliography: references.bib
-#csl: american-journal-of-epidemiology.csl
+csl: journal-of-clinical-oncology.csl
 
@@
 library(here)
+library(knitr)
 library(captioner)
@@
 
-**Authors**: Janick Weberpals^1^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
+**Authors**: Janick Weberpals^1^, Donna R. Rivera^2^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
 
@@
 
-[**Article type:**]{.underline} Original Research Article
+[**Article type:**]{.underline} Review
 
@@
 
-[**Keywords:**]{.underline} Real-World Evidence
+[**Keywords:**]{.underline} Oncology, Real-World Evidence, Trial emulation, EHR
 
@@
 [**Competing Interests Statement:**]{.underline} Dr.
+Wang has consulted ad hoc for Exponent Inc.
+and MITRE a federally funded research center for the Centers for Medicare and Medicaid Services on unrelated work.
+Dr.
 Schneeweiss is participating in investigator-initiated grants to the Brigham and Women's Hospital from Boehringer Ingelheim, Takeda, and UCB unrelated to the topic of this study.
@@
 [**Analytic code sharing statement:**]{.underline} ...
+
+[***Proposed target journals:***]{.underline} *JCO CCI =\> JAMA Network Open =\> CPT*
 
@@
 
-...
+**Systematic process for understanding the validity of RWE for oncology submissions**
 
 {{< pagebreak >}}
-
-# Results {#sec-results}
-
-...
 
@@
 
-\blandscape
+::: landscape
 
@@
 
-\elandscape
+```{r}
+#| label: fig-process
+#| fig-cap: "Systematic process to understand effectiveness claims of oncology trials using real-world evidence."
+#| fig-cap-location: top
+#| fig-width: 12
+
+knitr::include_graphics(here("figures", "process.png"))
+```
+
+[View figure in higher resolution]()
+:::

### 2024-11-24 manuscript/manuscript.qmd
added some tables

@@
 library(here)
+library(dplyr)
 library(knitr)
 library(captioner)
+library(gt)
+library(readxl)
 
@@
 [**Competing Interests Statement:**]{.underline} Dr.
+Weberpals is now an employee of AstraZeneca and owns stocks in AstraZeneca.
+Dr.
 Wang has consulted ad hoc for Exponent Inc.
@@
 # Background
+
+With 21% of all drug approvals, oncology was the disease area with the most FDA drug approvals in 2023.[@senior2024fresh] Although randomized controlled trials (RCTs) are considered the gold standard, decision-makers increasingly rely on real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Especially in the field of precision oncology, RWE has many essential use cases and plays a critical role in complementing evidence like in patient populations that are underrepresented in RCTs, to construct external control arms in single-arm trials where active recruitment may not be feasible or in drug/biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic signatures.
+However, [to draw causal inferences from such comparisons in non-randomized data, it is pivotal that endpoints and prognostic information can be measured reliably and at scale]{.underline}, which remains a significant challenge in oncological comparative effectiveness research (CER).
 
@@
 ::: landscape
+# Tables {.unnumbered}
 
-# Tables {.unnumbered}
+```{r}
+#| label: tbl-criteria
+#| tbl-cap: "Criteria ."
+#| tbl-cap-location: top
+
+eligibility_criteria <- read_excel(
+  path = here("tables", "helper_tables.xlsx"),
+  sheet = "ctgov_criteria_defintions"
+  )
+
+eligibility_criteria |> 
+  select(-Link) |> 
+  filter(Criteria != "Study phase") |> 
+  gt()
+```
+
+```{r}
+#| label: tbl-rcts
+#| tbl-cap: "Tentative list of randomized controlled trials (RCTs) considered for emulation."
+#| tbl-cap-location: top
+
+rct_selection <- read_excel(
+  path = here("tables", "helper_tables.xlsx"),
+  sheet = "rct_selection"
+  )
+
+rct_selection |> 
+  group_by(cancer) |> 
+  gt() |> 
+  cols_label(
+    nctid = md("**NCTID**"),
+    acronym = md("**Acronym**"),
+    setting = md("**Clinical setting**"),
+    line = md("**Line of therapy**"),
+    exposures = md("**Treatment comparison**")
+    ) |> 
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_row_groups()
+    )
+```
 
@@
 knitr::include_graphics(here("figures", "process.png"))
+
+# copy figure to manuscript
+# file_copy(
+#   path = path, 
+#   new_path = here("manuscript", "Figure_2_attrition.png"), 
+#   overwrite = TRUE
+#   )
 ```

### 2024-11-28 manuscript/manuscript.qmd
more changes
more progress on introduction

@@
 
-**Authors**: Janick Weberpals^1^, Donna R. Rivera^2^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
+**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
 
@@
 
-^2^ Oncology Center of Excellence, US Food and Drug Administration, Silver Spring, MD, USA
+^2^ Dana-Farber, ..., USA
+
+^3^ Oncology Center of Excellence, US Food and Drug Administration, Silver Spring, MD, USA
 
@@
 
-With 21% of all drug approvals, oncology was the disease area with the most FDA drug approvals in 2023.[@senior2024fresh] Although randomized controlled trials (RCTs) are considered the gold standard, decision-makers increasingly rely on real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Especially in the field of precision oncology, RWE has many essential use cases and plays a critical role in complementing evidence like in patient populations that are underrepresented in RCTs, to construct external control arms in single-arm trials where active recruitment may not be feasible or in drug/biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic signatures.
-However, [to draw causal inferences from such comparisons in non-randomized data, it is pivotal that endpoints and prognostic information can be measured reliably and at scale]{.underline}, which remains a significant challenge in oncological comparative effectiveness research (CER).
+Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
+With the advent of the the 21^st^ Century Cures Act directive, the Food and Drug Administration (FDA) established a framework to increasingly include and consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all drug approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
+Examples comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the use of real-world data (RWD) for biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic and immunological signatures.
+
+However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and limitations include missing data, small sample sizes, data discontinuity [@lin2018identifying; @joshua2022longitudinal], rapid changes in temporal prescribing patterns and the inability to measure and emulate common eligibility criteria and prognostic factors.[@rider2024emulations] While there are examples of emulations of oncology trials[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials is necessary to gain confidence the accuracy of RWE studies and to provide an answer as to which questions can be validly answered with which non-interventional study designs and analysis methods, given the data that is available.
+
+The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by comparing results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
+In settings where the RCT designs could be emulated well, RWE studies came to the same conclusions.[@heyard2024design] However, prior work from RCT-DUPLICATE has focused primarily on emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas using claims databases.
+
+The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project aims to extend this work to the field of oncology which comes with its own unique challenges that are not necessarily comparable with learnings from other disease areas and which must be systematically explored and understood.
 
@@
   filter(Criteria != "Study phase") |> 
-  gt()
+  gt() #|> 
+  # tab_style(
+  #   style = cell_text(weight = "bold"),
+  #   locations = 
+  #   )
 ```
@@
 
-[View figure in higher resolution]()
+[View figure in higher resolution](https://github.com/janickweberpals/encore-process-manuscript/blob/main/figures/process.png)
 :::

### 2024-11-28 manuscript/manuscript.qmd
first draft of introduction

@@
 Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
-With the advent of the the 21^st^ Century Cures Act directive, the Food and Drug Administration (FDA) established a framework to increasingly include and consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all drug approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
+With the advent of the the 21^st^ Century Cures Act directive, the Food and Drug Administration (FDA) established a framework to increasingly include and additionally consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023\[@senior2024fresh\], and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
 Examples comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the use of real-world data (RWD) for biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic and immunological signatures.
 
-However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and limitations include missing data, small sample sizes, data discontinuity [@lin2018identifying; @joshua2022longitudinal], rapid changes in temporal prescribing patterns and the inability to measure and emulate common eligibility criteria and prognostic factors.[@rider2024emulations] While there are examples of emulations of oncology trials[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials is necessary to gain confidence the accuracy of RWE studies and to provide an answer as to which questions can be validly answered with which non-interventional study designs and analysis methods, given the data that is available.
+However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and frequently references limitations include missing data, small sample sizes, data discontinuity [@Merola2022;@joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in RWD.[@rider2024emulations] While there are publsihed examples of emulations of oncology trials[@rider2024emulations;@merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in various heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered with which non-interventional study designs and analysis methods.
 
@@
 
-The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project aims to extend this work to the field of oncology which comes with its own unique challenges that are not necessarily comparable with learnings from other disease areas and which must be systematically explored and understood.
+The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project [@encoreFDA] aims to extend this work to the field of oncology which comes with its own unique set of challenges that are not necessarily comparable with previous learnings from other disease areas and which must be systematically explored and understood.
+Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology is going to emulate 12 randomized clinical trials using multiple EHR data sources.
+The process includes an emphasis on transparency with documented assessment of data fitness of the RWD source for each trial [@rivera2024oncology;@gatto2022structured] and conducting extensive sensitivity analyses to assess robustness of findings and trial eligibility criteria.
 
-... [@weberpals2024]
+The objectives of this project are to develop state-of-the-art methodological approaches and apply those to create insights that may provide guidance on the potential use of RWE for regulatory science in oncology.
+This includes the systematic evaluation of the suitability of data in relation to the study design and statistical analysis by emulating 12 oncology trials across four cancers and assessing the agreement of treatment effect estimates between trial emulation and RCT.
+
+In this process paper, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development, study design and statistical analysis and final agreement metrics to evaluate the concordance between RCT and its emulation.
 
@@
 **Systematic process for understanding the validity of RWE for oncology submissions**
+
+Trial selection
+
+Data feasibility
+
+Protocol development
+
+Application of the target trial emulation framework to study design and statistical analysis
+
+... [@weberpals2024]
+
+Agreement metrics
 

### 2024-11-28 manuscript/manuscript.qmd
reference correction format

@@
 Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
-With the advent of the the 21^st^ Century Cures Act directive, the Food and Drug Administration (FDA) established a framework to increasingly include and additionally consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023\[@senior2024fresh\], and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
+With the advent of the the 21^st^ Century Cures Act directive, the Food and Drug Administration (FDA) established a framework to increasingly include and additionally consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023,\[@senior2024fresh\] and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
 Examples comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the use of real-world data (RWD) for biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic and immunological signatures.

### 2024-11-28 manuscript/manuscript.qmd
advanced with trial selectioin process

--- 2024-11-28 manuscript/manuscript.qmd
@@
 library(readxl)
+library(fs)
 
@@
 Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
-With the advent of the the 21^st^ Century Cures Act directive, the Food and Drug Administration (FDA) established a framework to increasingly include and consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all drug approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
-Examples comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the use of real-world data (RWD) for biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic and immunological signatures.
+With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly include and consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023,[@senior2024fresh] and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
+Potential use cases comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the use of real-world data (RWD) for biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
 
-However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and limitations include missing data, small sample sizes, data discontinuity [@lin2018identifying; @joshua2022longitudinal], rapid changes in temporal prescribing patterns and the inability to measure and emulate common eligibility criteria and prognostic factors.[@rider2024emulations] While there are examples of emulations of oncology trials[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials is necessary to gain confidence the accuracy of RWE studies and to provide an answer as to which questions can be validly answered with which non-interventional study designs and analysis methods, given the data that is available.
+However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in RWD.[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in various heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered with which non-interventional study designs and analysis methods.
 
@@
 
-The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project aims to extend this work to the field of oncology which comes with its own unique challenges that are not necessarily comparable with learnings from other disease areas and which must be systematically explored and understood.
+The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project [@encoreFDA] aims to extend this work to the field of oncology which comes with its own unique set of challenges which must be systematically explored and understood.
+Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology is going to emulate 12 randomized clinical trials using multiple EHR data sources.
+The process includes an emphasis on transparency with documented assessment of data fitness of the RWD source for each trial[@rivera2024oncology; @gatto2022structured] and conducting extensive sensitivity analyses to assess robustness of findings and trial eligibility criteria.
 
-... [@weberpals2024]
+The objectives of this project are to develop state-of-the-art methodological approaches and apply these to create insights that may provide guidance on the potential use of RWE for regulatory science in oncology.
+This includes the systematic evaluation of the suitability of data in relation to the study design and statistical analysis by emulating 12 oncology trials across four cancers and assessing the agreement of treatment effect estimates between RCTs and their respective emulations.
+
+In this process paper, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development, study design and statistical analysis and final agreement metrics to evaluate the concordance between RCTs and emulations.
 
@@
 
-**Systematic process for understanding the validity of RWE for oncology submissions**
+A visual summary of the entire systematic process from trial selection to final results is provided in @fig-process.
+
+## Trial selection
+
+The focus of ENCORE is to maximize potential learnings on when RWE studies can or cannot yield similar results compared to RCTs.
+To that end, the focus of the project is on trials of therapies for the most common cancers and/or cancers for which there has been substantial therapeutic development in recent years.
+After careful review exchange with clinical and regulatory experts, four cancer indications were identified including lung cancer, breast cancer, colorectal cancer and multiple myeloma.
+For each cancer we aim to conduct three trial emulations which will be implemented using multiple databases which will be accessible for the scope of this project (i.e.
+the total number of emulations will equal 12 trials x n databases which are found fit-for-purpose for each trial).
+
+The trial selection will follow a semi-automated process for which we will document the eligibility criteria and decisions, resulting in a CONSORT diagram showing reasons for excluding RCTs.
+The search will be conducted using the AACT database which is s a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) which contains all information (protocol and result data elements) about every study registered in ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we will use a combined search query strategy of the National Library of Medicine (NLM)-controlled *MeSH* term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.
+
+Eligible trials need to fulfill the following basic criteria:
+
+-   Interventional
+
+-   Randomized
+
+-   Intervention model: parallel assignment
+
+-   Industry-sponsored
+
+-   Trial start in 2011 or later
+
+-   Primary purpose was to study treatment effects
+
+-   Primary or secondary endpoint must be overall survival
+
+-   Recruitment status: ‘Completed’ or ‘Active, not recruiting’
+
+-   Feasibility and clinical relevance
+
+The rationale and operationalization of each criterion is listed in detail in @tbl-criteria.
+While all but the last criteria are highly objective and can be used to filter potential trials in an automated fashion, the last criterion on emulation feasibility and clinical relevance involves extensive human review, a thorough feasibility assessment in context of available databases and discussion with clinical and regulatory experts.
+The aim is to identify a broad and diverse set of potentially feasible RCTs for which there is reasonable believe that key study parameters can be emulated and there is a high enough clinical relevance for regulatory decision-making (e.g., paradigm-changing trials).
+
+## Databases
+
+## Protocol development
+
+## Application of the target trial emulation framework to study design and statistical analysis
+
+... [@weberpals2024]
+
+Agreement metrics
 
@@
 #| label: tbl-criteria
-#| tbl-cap: "Criteria ."
+#| tbl-cap: "Criteria to select eligible trials for emulation in ENCORE."
 #| tbl-cap-location: top
 
+# read excel blueprint
 eligibility_criteria <- read_excel(
@@
 
-eligibility_criteria |> 
+# convert to gt table
+tbl_eligibility_criteria <- eligibility_criteria |> 
   select(-Link) |> 
   filter(Criteria != "Study phase") |> 
-  gt() #|> 
-  # tab_style(
-  #   style = cell_text(weight = "bold"),
-  #   locations = 
-  #   )
+  gt() |> 
+  tab_style(
+    style = cell_text(weight = "bold"),
+      locations = cells_column_labels()
+    )
+
+# save table as .docx document
+tbl_eligibility_criteria |> 
+   gtsave(
+    filename = here("manuscript", "Table_1_trial_eligibility.docx")
+    )
+
+# save image to display within manuscript document
+tbl_eligibility_criteria |> 
+  gtsave(
+    filename = here("tables", "Table_1_trial_eligibility.png"),
+    expand = 10, # amount of whitespace,
+    zoom = 5 # higher resolution
+    )
+
+include_graphics(path = here("tables", "Table_1_trial_eligibility.png"))
 ```
@@
 
-rct_selection |> 
+tbl_rct_selection <- rct_selection |> 
   group_by(cancer) |> 
@@
     )
+
+# save table as .docx document
+tbl_rct_selection |> 
+   gtsave(
+    filename = here("manuscript", "Table_2_trial_selection.docx")
+    )
+
+# save image to display within manuscript document
+tbl_rct_selection |> 
+  gtsave(
+    filename = here("tables", "Table_2_trial_selection.png"),
+    expand = 10, # amount of whitespace,
+    zoom = 5 # higher resolution
+    )
+
+include_graphics(path = here("tables", "Table_2_trial_selection.png"))
 ```
@@
 # copy figure to manuscript
-# file_copy(
-#   path = path, 
-#   new_path = here("manuscript", "Figure_2_attrition.png"), 
-#   overwrite = TRUE
-#   )
+file_copy(
+  path = here("figures", "process.png"),
+  new_path = here("manuscript", "Figure_1_process.png"),
+  overwrite = TRUE
+  )
 ```
 
-[View figure in higher resolution](https://github.com/janickweberpals/encore-process-manuscript/blob/main/figures/process.png)
+[View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/figures/process.png)
 :::

### 2024-11-29 manuscript/manuscript.qmd
installed webshot2
removed landscape condition

@@
 library(fs)
+library(webshot2)
 
@@
 
-::: landscape
 # Tables {.unnumbered}
@@
 [View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/figures/process.png)
-:::

### 2024-12-01 manuscript/manuscript.qmd
drafted trial selection chapter and added consort diagrams to supplement

@@
 Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
-With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly include and consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023,[@senior2024fresh] and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
-Potential use cases comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the use of real-world data (RWD) for biomarker discovery and label extensions among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
+With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023,[@senior2024fresh] and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
+Potential use cases comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
 
-However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in RWD.[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in various heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered with which non-interventional study designs and analysis methods.
+However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in various heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered.
 
@@
 Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology is going to emulate 12 randomized clinical trials using multiple EHR data sources.
-The process includes an emphasis on transparency with documented assessment of data fitness of the RWD source for each trial[@rivera2024oncology; @gatto2022structured] and conducting extensive sensitivity analyses to assess robustness of findings and trial eligibility criteria.
+The process includes an emphasis on transparency with documented assessment of data fitness of the RWD source for each trial[@rivera2024oncology; @gatto2022structured] and the conduct of extensive sensitivity analyses to assess robustness of findings and trial eligibility criteria.
 
@@
 The focus of ENCORE is to maximize potential learnings on when RWE studies can or cannot yield similar results compared to RCTs.
-To that end, the focus of the project is on trials of therapies for the most common cancers and/or cancers for which there has been substantial therapeutic development in recent years.
-After careful review exchange with clinical and regulatory experts, four cancer indications were identified including lung cancer, breast cancer, colorectal cancer and multiple myeloma.
-For each cancer we aim to conduct three trial emulations which will be implemented using multiple databases which will be accessible for the scope of this project (i.e.
-the total number of emulations will equal 12 trials x n databases which are found fit-for-purpose for each trial).
+To that end, the emphasis of the project is on trials of therapies for the most common cancers and/or cancers for which there has been substantial therapeutic development in recent years.
+After careful review and exchange with clinical and regulatory experts, four cancer indications were identified including lung cancer, breast cancer, colorectal cancer and multiple myeloma.
+For each cancer we aim to conduct three trial emulations which will be implemented using multiple databases accessible for the scope of this project (i.e.
+the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).
 
-The trial selection will follow a semi-automated process for which we will document the eligibility criteria and decisions, resulting in a CONSORT diagram showing reasons for excluding RCTs.
-The search will be conducted using the AACT database which is s a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) which contains all information (protocol and result data elements) about every study registered in ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we will use a combined search query strategy of the National Library of Medicine (NLM)-controlled *MeSH* term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.
+The trial selection will follow a semi-automated process for which we will document the eligibility criteria resulting in a CONSORT diagram showing reasons for excluding RCTs.
+The search will be conducted using the AACT database which is s a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) which contains all information (protocol and result data elements) about every study registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we will use a combined search query strategy of the National Library of Medicine (NLM)-controlled *MeSH* term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.
 
@@
 
--   Primary or secondary endpoint must be overall survival
+-   Overall survival must be one of the endpoints reported
 
@@
 The rationale and operationalization of each criterion is listed in detail in @tbl-criteria.
-While all but the last criteria are highly objective and can be used to filter potential trials in an automated fashion, the last criterion on emulation feasibility and clinical relevance involves extensive human review, a thorough feasibility assessment in context of available databases and discussion with clinical and regulatory experts.
-The aim is to identify a broad and diverse set of potentially feasible RCTs for which there is reasonable believe that key study parameters can be emulated and there is a high enough clinical relevance for regulatory decision-making (e.g., paradigm-changing trials).
+We will mainly consider pivotal interventional, randomized trials after 2011 since many cancer indications have witnessed significant changes in treatment guidelines in recent years and due to the usually rapid adoption of new breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in the real-world.
+In parallel, trials readout should have also not occurred too recently in order to allow for enough data and follow-up time accrual in databases used for this project.
+We further focus only on trials that have reported overall survival (OS) as one pre-specified endpoint in the protocol.
+Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR;@mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST;@rivera2022friends] Given the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, the scope of this project was to focus on the emulation of OS as the endpoint of interest.
+
+While most trial-eligibility criteria can be operationalized in an automated fashion, the last criterion on emulation feasibility and clinical relevance involves extensive human review.
+The critical points considered in this step include a thorough feasibility assessment of the data fitness, sample size considerations and the assessment if critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and can be balanced using propensity score matching or weighting methods.[@brookhart2006variable]
+
+A list of tentative, shortlisted candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
+Naturally, the majority trials will cover advanced or metastatic cancer populations since a large proportion of drug development efforts have focused on these settings in recent years.
+A key learning that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex gene/biomarker signatures) can be emulated using RWD.
+If ongoing feasibility assessments indicate that these trials cannot be emulated with high enough confidence, runner-up will be considered instead.
 

### 2024-12-01 manuscript/supplement.qmd
drafted trial selection chapter and added consort diagrams to supplement

@@
 title: "Supplementary Material"
-subtitle: "..."
+subtitle: "A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)"
 
@@
 number-sections: true
+toc: true
 
@@
 
+library(here)
+library(knitr)
+
 # table_cap <- captioner(prefix = "Supplementary Table")
 ```
+{{< pagebreak >}}
 
-**Authors**: ...
+# Supplementary Figures
+
+```{r}
+#| label: fig-consort-nsclc
+#| fig-cap: "CONSORT diagram for non-small cell lung cancer (NSCLC) top candidate trials.
+Figure suggests four shortlisted candidates because both CheckMate017/057 have been both shortlisted and only differ in the squamous versus nonsquamous histological eligibility of the trial population."
+#| out-width: "115%"
+
+include_graphics(here("figures", "consort_nsclc.png"))
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-consort-breast
+#| fig-cap: "CONSORT diagram for breast cancer (BC) top candidate trials."
+#| out-width: "115%"
+
+include_graphics(here("figures", "consort_bc.png"))
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-consort-crc
+#| fig-cap: "CONSORT diagram for colorectal cancer (CRC) top candidate trials."
+#| out-width: "115%"
+
+include_graphics(here("figures", "consort_crc.png"))
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-consort-mm
+#| fig-cap: "CONSORT diagram for multiple myeloma (MM) top candidate trials."
+#| out-width: "115%"
+
+include_graphics(here("figures", "consort_mm.png"))
+```
+
+{{< pagebreak >}}
+
+# Supplementary Tables
 
@@
 
--->
-
-# Supplementary Tables
-
-...
-
-{{< pagebreak >}}
-
-# Supplementary Figures
-
-...
-
-{{< pagebreak >}}
-
 # References {.unnumbered}

### 2024-12-02 manuscript/supplement.qmd
not on nsclc trials

@@
 #| label: fig-consort-nsclc
-#| fig-cap: "CONSORT diagram for non-small cell lung cancer (NSCLC) top candidate trials.
-Figure suggests four shortlisted candidates because both CheckMate017/057 have been both shortlisted and only differ in the squamous versus nonsquamous histological eligibility of the trial population."
+#| fig-cap: "CONSORT diagram for non-small cell lung cancer (NSCLC) top candidate trials."
 #| out-width: "115%"
@@
 ```
+
+*Note that @fig-consort-nsclc results in four shortlisted candidates because both CheckMate017/057 have been both shortlisted and only differ in the squamous versus nonsquamous histological eligibility of the trial population.*
 
@@
 
+{{< pagebreak >}}
+
 $\vspace{0.25cm}$

### 2024-12-02 manuscript/manuscript.qmd
added protocol development chapter

@@
 library(knitr)
-library(captioner)
 library(gt)
@@
 library(webshot2)
-
-table_cap <- captioner(prefix = "Table")
 ```
@@
 
-{{< pagebreak >}}
-
 # Methods
@@
 The rationale and operationalization of each criterion is listed in detail in @tbl-criteria.
-We will mainly consider pivotal interventional, randomized trials after 2011 since many cancer indications have witnessed significant changes in treatment guidelines in recent years and due to the usually rapid adoption of new breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in the real-world.
-In parallel, trials readout should have also not occurred too recently in order to allow for enough data and follow-up time accrual in databases used for this project.
-We further focus only on trials that have reported overall survival (OS) as one pre-specified endpoint in the protocol.
+We will mainly consider pivotal interventional, randomized trials after 2011 since treatment guidelines among studied cancer indications have undergone significant changes in recent years.
+Due to the rapid adoption of new breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in the real-world.
+In parallel, trial results should have also not been published too recently in order to allow for enough data and follow-up time accrual in databases used for this project.
+An further focus is on trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.
 Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR;@mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST;@rivera2022friends] Given the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, the scope of this project was to focus on the emulation of OS as the endpoint of interest.
@@
 While most trial-eligibility criteria can be operationalized in an automated fashion, the last criterion on emulation feasibility and clinical relevance involves extensive human review.
-The critical points considered in this step include a thorough feasibility assessment of the data fitness, sample size considerations and the assessment if critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and can be balanced using propensity score matching or weighting methods.[@brookhart2006variable]
+The critical points considered in this step include a thorough feasibility assessment of the data fitness [@rivera2022friends], sample size considerations and the assessment if critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and can be balanced using propensity score matching or weighting methods.[@brookhart2006variable]
 
@@
 Naturally, the majority trials will cover advanced or metastatic cancer populations since a large proportion of drug development efforts have focused on these settings in recent years.
-A key learning that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex gene/biomarker signatures) can be emulated using RWD.
-If ongoing feasibility assessments indicate that these trials cannot be emulated with high enough confidence, runner-up will be considered instead.
+A key learning that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD.
+If ongoing feasibility assessments indicate that these trials cannot be emulated with high enough confidence, runner-up candidates will be considered instead.
 
@@
 
+The ENCORE project will utilize data from a total four different oncology-specific electronic health records (EHR)-derived data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada.
+All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology.
+For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results.
+If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach.
+
 ## Protocol development
+
+For each shortlisted and selected RCT, a detailed protocol will be developed following the HARPER protocol template[@wang2022harmonized] and registered on ClinicalTrials.gov.
+The protocol will pre-specify key elements of the trial emulation and following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will summarize each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to.
+All eligibility criteria will be extracted based on publicly available protocols and statistical analysis plans of the selcted RCT.
+Following the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], tables will outline how eligibility criteria will be measured and operationalized and a color-coded heatmap will indicate the level of confidence on how well each criterion can be emulated in each selected database.
+As there are general eligibility criteria in oncology trials which either won't be possible to emulate (e.g., physician-assessed survival prognosis of xy months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.
+We will additionally provide a definition on how exposure and outcomes are defined and ascertained in each respective database.
+For most considered databases, the OS endpoint is a composite that is derived from different sources including EHR flags, social security death index, obituary and other linkages.
+Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative censoring dates will be considered for each trial emulation to mitigate the impact of potential ghost-time bias.[@meyer2020open] Each final protocol draft will be reviewed by a clinical and FDA regulatory expert panel upon registration on ClinicalTrials.gov.
 
@@
 
-... [@weberpals2024]
+[@weberpals2024]
 
-Agreement metrics
+## Agreement metrics
 

### 2024-12-03 manuscript/manuscript.qmd
figure 2 added

@@
 number-sections: false
+warnings: false
+messages: false
 
@@
 library(webshot2)
+library(tibble)
+library(ggplot2)
+library(cowplot)
 ```
@@
 While most trial-eligibility criteria can be operationalized in an automated fashion, the last criterion on emulation feasibility and clinical relevance involves extensive human review.
-The critical points considered in this step include a thorough feasibility assessment of the data fitness [@rivera2022friends], sample size considerations and the assessment if critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and can be balanced using propensity score matching or weighting methods.[@brookhart2006variable]
+The critical points considered in this step include a thorough feasibility assessment of the data fitness [@rivera2022friends], sample size considerations and the assessment if critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and can be balanced using propensity score matching or weighting methods.[@brookhart2006variable] Lastly, trial candidates are ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.
 
-A list of tentative, shortlisted candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
+A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
 Naturally, the majority trials will cover advanced or metastatic cancer populations since a large proportion of drug development efforts have focused on these settings in recent years.
@@
 
-For each shortlisted and selected RCT, a detailed protocol will be developed following the HARPER protocol template[@wang2022harmonized] and registered on ClinicalTrials.gov.
-The protocol will pre-specify key elements of the trial emulation and following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will summarize each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to.
+For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed following the HARPER protocol template[@wang2022harmonized] and registered on ClinicalTrials.gov.
+Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to.
 All eligibility criteria will be extracted based on publicly available protocols and statistical analysis plans of the selcted RCT.
-Following the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], tables will outline how eligibility criteria will be measured and operationalized and a color-coded heatmap will indicate the level of confidence on how well each criterion can be emulated in each selected database.
+According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], tables will outline how eligibility criteria will be measured and operationalized and a color-coded heatmap will indicate the level of confidence on how well each criterion can be emulated in each selected database.
 As there are general eligibility criteria in oncology trials which either won't be possible to emulate (e.g., physician-assessed survival prognosis of xy months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.
 We will additionally provide a definition on how exposure and outcomes are defined and ascertained in each respective database.
-For most considered databases, the OS endpoint is a composite that is derived from different sources including EHR flags, social security death index, obituary and other linkages.
-Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative censoring dates will be considered for each trial emulation to mitigate the impact of potential ghost-time bias.[@meyer2020open] Each final protocol draft will be reviewed by a clinical and FDA regulatory expert panel upon registration on ClinicalTrials.gov.
+For the considered databases, the OS endpoint is a composite that is derived from different sources like EHR flags, social security death index, obituary and other linkages.
+Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open] Each final protocol draft will be reviewed by a clinical and FDA regulatory expert panel before registration on ClinicalTrials.gov.
 
-## Application of the target trial emulation framework to study design and statistical analysis
+Critical aspects when emulating oncology trials are the choice and estimation of the approriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the (historical) control cohort, the adoption rate of the novel intervention in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null.
+To that end, comprehensive diagnostics will be performed to contextualize these parameters and (if reported) draw comparisons to the emulated trial.
+
+## gn and statistical analysis
 
@@
 
+{{< pagebreak >}}
+
 ```{r}
@@
 [View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/figures/process.png)
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-initiators
+#| fig-cap: "Descriptive drug utilization analyses."
+#| fig-widh: 4
+#| fig-height: 7
+#| warnings: false
+
+data <- tibble::tibble(
+  drug = c(rep("Drug A", 5), rep("Drug B", 5)),
+  initiators = c(0, 220, 300, 399, 500, 456, 400, 278, 180, 35),
+  year = rep(c(2011, 2012, 2013, 2014, 2015), 2)
+  )
+
+initiators <- data |> 
+  ggplot(aes(x = year, y = initiators, group = drug, color = drug)) +
+  geom_line(linewidth = 1) +
+  labs(
+    x = "Calendar Year",
+    y = "# initiators",
+    color = "Exposure"
+    ) +
+  scale_color_manual(values=c('orange','blue')) +
+  theme_minimal(base_size = 10)
+
+# treatment switch
+data <- tibble::tibble(
+  drug = c(rep("Drug A", 6), rep("Drug B", 6)),
+  prop = c(4.1, 7.7, 17.7, 22.2, 24.4, 24.5, 4.6, 13.2, 25.1, 29, 29.3, 31.3),
+  time = rep(seq(1, 12, 2), 2)
+  )
+
+switchers <- data |> 
+  ggplot(aes(x = time, y = prop, group = drug, color = drug)) +
+  geom_line(linewidth = 1) +
+  labs(
+    x = "Follow-up time [Months]",
+    y = "Cumulative switch rate",
+    color = "Exposure"
+    ) +
+  scale_color_manual(values=c('orange','blue')) +
+  theme_minimal(base_size = 10)
+
+# combine plots with shared legend
+shared_legend <- get_legend(initiators)
+
+plot_utilization <- plot_grid(
+  plot_grid(initiators + theme(legend.position = "top")),
+  plot_grid(switchers + theme(legend.position = "none")),
+  ncol = 1,
+  nrow = 2,
+  labels = c("a)", "b)"),
+  shared_legend
+  )
+
+# save
+ggsave(
+  plot = plot_utilization,
+  filename = "Figure_2_utlization.png",
+  path = here("manuscript"),
+  dpi = 800)
+
+# print to manuscript
+include_graphics(path = here("manuscript", "Figure_2_utlization.png"))
+```

### 2024-12-08 manuscript/manuscript.qmd
finished emulation feasibility part

@@
 
-[***Proposed target journals:***]{.underline} *JCO CCI =\> JAMA Network Open =\> CPT*
+[***Proposed target journals:***]{.underline} *JCO CCI =\> CPT =\> ...*
 
@@
 
--   Overall survival must be one of the endpoints reported
+-   Overall survival must be one of the endpoints reported (either as hazard ratio or median overall survival time)
 
@@
 In parallel, trial results should have also not been published too recently in order to allow for enough data and follow-up time accrual in databases used for this project.
-An further focus is on trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.
-Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR;@mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST;@rivera2022friends] Given the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, the scope of this project was to focus on the emulation of OS as the endpoint of interest.
+A further focus is on trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.
+Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] Given the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, the scope of this project was to focus on the emulation of OS as the endpoint of interest.
 
@@
 
-The ENCORE project will utilize data from a total four different oncology-specific electronic health records (EHR)-derived data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada.
+The ENCORE project will utilize data from a total four different oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada.
 All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology.
@@
 
+<!--# maybe data partners have  1-2 sentences and/or publications to reference? -->
+
 ## Protocol development
 
-For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed following the HARPER protocol template[@wang2022harmonized] and registered on ClinicalTrials.gov.
-Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to.
-All eligibility criteria will be extracted based on publicly available protocols and statistical analysis plans of the selcted RCT.
-According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], tables will outline how eligibility criteria will be measured and operationalized and a color-coded heatmap will indicate the level of confidence on how well each criterion can be emulated in each selected database.
+For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed following the HARPER protocol template[@wang2022harmonized] and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel.
+Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics).
+All eligibility criteria will be extracted based on publicly available protocols and statistical analysis plans of the selected RCT.
+
+### Emulation feasibility
+
+**Fit-for-purpose data.** Real-world data fitness and emulation feasibility for a given shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care.
+Since a vast majority of oncological RCTs in recent years have focused on selected, biomarker-defined populations, subtleties in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population.
+For example, immunotherapies have significantly changed the cancer treatment landscape since the approval of the first PD-L1 inhibitor in 2015.
+With many trials that have followed thereafter, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has also evolved since then and PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.
+
+According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step will outline tables that describe how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database.
 As there are general eligibility criteria in oncology trials which either won't be possible to emulate (e.g., physician-assessed survival prognosis of xy months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.
-We will additionally provide a definition on how exposure and outcomes are defined and ascertained in each respective database.
-For the considered databases, the OS endpoint is a composite that is derived from different sources like EHR flags, social security death index, obituary and other linkages.
-Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open] Each final protocol draft will be reviewed by a clinical and FDA regulatory expert panel before registration on ClinicalTrials.gov.
 
-Critical aspects when emulating oncology trials are the choice and estimation of the approriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the (historical) control cohort, the adoption rate of the novel intervention in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null.
-To that end, comprehensive diagnostics will be performed to contextualize these parameters and (if reported) draw comparisons to the emulated trial.
+We will additionally provide a definition on how exposures, outcomes and covariates are exactly defined and operationalized in each respective database.
+There will be special emphasis on how exposure, in context of their respective disease and line of therapy settings, and the OS outcome will be emulated.
+For all considered databases, the OS endpoint is typically a composite that is derived from different sources comprising EHR abstractions, social security death index, obituary and other linkages.
+Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open]
 
-## gn and statistical analysis
+**Descriptives and data exploration.** Critical aspects when emulating oncology trials are the choice and estimation of the appropriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the (historical) control cohort, the adoption rate of the novel intervention in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null.
+To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial.
+Example for such standard diagnostics are visualized in @fig-initiators.
+
+The distribution of patient characteristics, stratified by exposure status, will be examined in Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT.
+Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, eposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting.
+At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design and analytic choices based on known outcome information.
+
+**Statistical power**.
+Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power than RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, statistical power is a critical to assess the feasibility that a comparison of agreement metrics of estimated parameters between RCT and RWD with sufficient precision can be made.
+Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
+To assess if the unstratified number of events is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic; @schoenfeld1983sample; @gsDesign]
+
+<!--# https://github.com/keaven/gsDesign/blob/master/R/nEvents.R#L22C12-L22C78 -->
+
+## Study design and statistical analysis
+
+The main parameters of interest in ENCORE are the marginal hazard ratio (HR) coefficient for the treatment comparison for time to all-cause mortality (OS)[@cox1972regression] and the corresponding Kaplan-Meier median survival time estimates.[@kaplan1958nonparametric] The median survival time is defined as the first time at which the the survival probability drops below 50%.[@therneau2015package]
+
+##
 
@@
 
-## Agreement metrics
+## Agreement metrics {#sec-agreement-metrics}
 
@@
 
+{{< pagebreak >}}
+
 # Figures {.unnumbered}

### 2024-12-14 manuscript/manuscript.qmd
added study design and statistics part

@@
 library(cowplot)
+library(mice)
+library(MatchThem)
+library(cobalt)
 ```
 
-**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
+**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
 
@@
 
-^2^ Dana-Farber, ..., USA
+^2^ Dana-Farber Cancer Institute, Boston, MA, USA
 
@@
 
-[***Proposed target journals:***]{.underline} *JCO CCI =\> CPT =\> ...*
+[***Proposed target journals:***]{.underline} *CPT =\> JCO CCI =\> ...*
 
@@
 
-However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in various heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered.
+However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered.
 
@@
 The rationale and operationalization of each criterion is listed in detail in @tbl-criteria.
-We will mainly consider pivotal interventional, randomized trials after 2011 since treatment guidelines among studied cancer indications have undergone significant changes in recent years.
+We will mainly consider pivotal interventional, randomized trials after 2011 since treatment guidelines among included cancer indications have undergone significant changes in recent years.
 Due to the rapid adoption of new breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in the real-world.
@@
 A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
-Naturally, the majority trials will cover advanced or metastatic cancer populations since a large proportion of drug development efforts have focused on these settings in recent years.
+Naturally, the majority of trials will cover advanced or metastatic cancer populations since a large proportion of drug development efforts have focused on these settings in recent years.
 A key learning that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD.
@@
 
-The ENCORE project will utilize data from a total four different oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada.
+The ENCORE project will utilize data from a total of four different oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada.
 All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology.
+A detailed description and methodology on how patients are sampled will be provided for each emulation protocol individually.
 For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results.
@@
 
-The main parameters of interest in ENCORE are the marginal hazard ratio (HR) coefficient for the treatment comparison for time to all-cause mortality (OS)[@cox1972regression] and the corresponding Kaplan-Meier median survival time estimates.[@kaplan1958nonparametric] The median survival time is defined as the first time at which the the survival probability drops below 50%.[@therneau2015package]
+The study design for each trial emulation will be visualized as part of the protocol using a graphical depiction of the exact measurement windows of eligibility criteria, washout periods and covariates relative to the cohort entry time.[@schneeweiss2019graphical]
 
-##
+### Missing data
 
-[@weberpals2024]
+To establish an analytic cohort, key eligibility criteria will be applied in which patients with missing values in eligibility criteria are considered eligible in the respective attrition steps to allow for thorough missing data investigations.
+These missing data investigations will empirically assess assumptions on potentially underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that was developed as part of a FDA Sentinel Innovation Center causal inference workstream that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] In brief, these diagnostic cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different.
+Together with expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams[@moreno2018canonical], this will inform decisions regarding the in- or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]
+
+While the MAR assumption is a strong assumption to hold across all considered covariates, it was shown that especially in the context of partially observed covariate data (as opposed to missing exposure and outcome data), only mechanisms in which a covariate causes its own missingness leads to critical bias (MNAR).[@moreno2018canonical] Hence, methodologies which retain patients and give the potential to adjust for a broader set of prognostic factors (e.g., multiple imputation[@weberpals2024hdmi] or doubly robust methods[@Shaw2024]) may be preferred over complete case analyses.
+
+### Endpoints and propensity score analyses
+
+Due to its ubiquity in oncology trials, the primary parameter of interest in ENCORE will be defined as the marginal hazard ratio (HR) coefficient for the treatment comparison for time to all-cause mortality (OS).[@cox1972regression] However, the HR has many limitations that can make the comparison of results across varying follow-up times between RCT and emulations challenging, including its non-collapsibility, in-built selection bias and the requirement of proportional hazards.[@hernan2010hazards; @stensrud2020test] For this reason, we will also consider alternative endpoints on an absolute risk scale such as median survival times, survival probabilities at pre-defined time points during follow-up[@kaplan1958nonparametric] and restricted mean survival times as secondary endpoints of interest.
+
+For the estimation of marginal treatment effects, we will employ propensity score analytics to adjust for measured confounding between treatment arms.
+The selection of important prognostic covariates will be based on expert clinical knowledge and published literature on prognostic scores in oncology.[@becker2020enhanced] The implementation of propensity scores in combination with multiple imputation will follow the '*within*' methodology as described by Leyrat et al.[@leyrat2019propensity; @pishgar2020matchthem] That is, propensity score matching or weighting will be applied to each imputed dataset.
+The marginal treatment effect will then be estimated in each imputed and matched or weighted dataset separately and pooled into a final estimate following Rubin's rule.[@rubin2018multiple; @van2011mice] This approach has been shown to lead to unbiased estimates across different simulated scenarios with a sufficient estimation of the variance.[@leyrat2019propensity]
+
+To asses the balance of pre-exposure covariates after matching or weighting on each imputed dataset, the average standardized mean difference (SMD) and corresponding minimum and maximum SMD range will be visualized (see example in @fig-balance).
+Covariate balance is typically considered at a SMD \< 0.1.[@austin2009balance] Further, we will compute the average post-matching or post-weighting C-statistics.[@franklin2014metrics] In addition, we will use a published prognostic score fro OS[@becker2020enhanced] as a balance measure to visually assess if the prognostic score is balanced between treatment arms after propensity score matching or weighting as this approach was described to show the highest correlations with bias compared with other balance measures and not affected by model misspecification.[@stuart2013prognostic]
+
+Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically don't follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given time during follow-up.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability.
+The median survival time can be finally determined by extracting the time point during follow-up at which the survival probability drops below 0.5 for the first time.
+
+### Reproducibility and consistency across emulation
 
@@
 
-...
+Target populations in international trials...genetic setup.
 
@@
 # save image to display within manuscript document
-tbl_eligibility_criteria |> 
+suppressMessages(tbl_eligibility_criteria |> 
   gtsave(
@@
     zoom = 5 # higher resolution
-    )
+    ))
 
@@
 # save image to display within manuscript document
-tbl_rct_selection |> 
+suppressMessages(tbl_rct_selection |> 
   gtsave(
@@
 #| fig-height: 7
-#| warnings: false
 
@@
 
-plot_utilization <- plot_grid(
+plot_utilization <- suppressMessages(plot_grid(
   plot_grid(initiators + theme(legend.position = "top")),
@@
   shared_legend
-  )
+  ))
 
 # save
-ggsave(
+suppressMessages(ggsave(
   plot = plot_utilization,
@@
   path = here("manuscript"),
-  dpi = 800)
+  dpi = 800
+  ))
 
@@
 ```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: fig-balance
+#| fig-cap: "Assessment of balance after propensity score matching or weighting across multiple imputed datasets."
+#| fig-widh: 4
+#| fig-height: 7
+
+source(here("functions", "simulate_flaura.R"))
+
+data <- simulate_flaura(
+  n_total = 2000, 
+  seed = 41, 
+  include_id = FALSE, 
+  imposeNA = TRUE,
+  propNA = .25
+  ) |> 
+  mutate(treat = ifelse(treat == 1, "Drug A", "Drug B"))
+
+# impute
+mids_data <- mice(data, m = 10, printFlag = F)
+
+# weight
+covariates <- data |> 
+  select(starts_with("dem_"), starts_with("c_"))
+
+# ps formula
+ps_formula <- as.formula(paste0("treat ~ ", paste0(colnames(covariates), collapse = " + ")))
+
+# weighting  
+wimids_data <- weightthem(
+  datasets = mids_data, 
+  formula = ps_formula, 
+  method = "glm",
+  estimand = "ATO"
+  )
+
+# Create the named character vector
+# named_covariates <- setNames(paste("Covariate", seq_len(ncol(covariates))), colnames(covariates))
+
+# covariate balance
+covariate_balance <- love.plot(
+  x = wimids_data,
+  abs = TRUE,
+  var.names = named_covariates,
+  thresholds = 0.1, 
+  drop.distance = TRUE,
+  var.order = "unadjusted",
+  colors = c("orange", "blue"), 
+  stars = "std",
+  shapes = 17, 
+  size = 4, 
+  grid = TRUE,
+  position = "top"
+  )
+
+# distributional balance
+score_balance <- bal.plot(
+  x = wimids_data,
+  var.name = "prop.score",
+  which = "both",
+  which.imp = .none,
+  colors = c("orange", "blue")
+  ) +
+  labs(
+    x = "Prognostic score",
+    title = "Distributional balance for prognostic score"
+    ) +
+  theme(legend.position = "top")
+
+plot_balance <- suppressMessages(plot_grid(
+  plot_grid(covariate_balance),
+  plot_grid(score_balance),
+  ncol = 2,
+  nrow = 1,
+  labels = c("a)", "b)"),
+  shared_legend
+  ))
+
+# save
+suppressMessages(ggsave(
+  plot = plot_balance,
+  filename = "Figure_3_balance.png",
+  path = here("manuscript"),
+  dpi = 800,
+  width = 15
+  ))
+
+# print to manuscript
+include_graphics(path = here("manuscript", "Figure_3_balance.png"))
+```

### 2024-12-14 manuscript/supplement.qmd
added protocol development chapter
figure 2 added
added study design and statistics part

@@
 echo: false
-number-sections: true
+number-sections: false
 toc: true
@@
         \usepackage{lscape}
+        \usepackage{pdfpages}
         \newcommand{\blandscape}{\begin{landscape}}
@@
 ```
+
 {{< pagebreak >}}
@@
 
+\addcontentsline{toc}{section}{\protect\numberline{}R package documentation}
+\includepdf[pages=-]{encore.io_0.2.0.pdf}
+
+The following documentation for the internal `encore.io` packages describes and details reproducible functions to query analytic cohorts consistently across trial emulations.
+
+::: callout-important
+The `encore.io` package will be continually developed throughout the ENCORE project and the following documentation is a version-controlled snapshot at the time of this publication.
+Updated versions will be published with the corresponding protocols for each trial emulation separately.
+:::
+
+{{< pagebreak >}}
+
 $\vspace{0.25cm}$
@@
 *Supplementary Material last updated: `r Sys.Date()`*
-
-{{< pagebreak >}}
 

### 2024-12-14 functions/simulate_data.R

@@ -1,0 +1,278 @@
+#' Simulates and artifical FLAURA EHR-derived dataset
+#'
+#' @description
+#' Parameterized function to quickly create an artificial FLAURA
+#' EHR-derived analytic cohort for analytic code development.
+#'
+#' @details
+#' ...
+#'
+#' @param n_total integer, number of total patients
+#' @param seed integer, seed for reproducibility
+#' @param include_id logical, include a generated patientid variable
+#' @param imposeNA logical, set covariates to missing
+#' @param propNA numeric, proportion of missingness, needs to be between 0 and 1
+#' 
+#' @return data frame with simulated analytic cohort
+#'
+#' @importFrom dplyr select rename mutate
+#' @importFrom tibble tibble
+#' @importFrom mice ampute
+#' @importFrom stats rbinom rnorm
+#'
+#' @export
+#'
+#' @examples
+#' \dontrun{
+#' library(encore.io)
+#'
+#' data_miss <- simulate_flaura(
+#'   n_total = 3500, 
+#'   seed = 41, 
+#'   include_id = FALSE, 
+#'   imposeNA = TRUE,
+#'   propNA = .33
+#'   )
+#'   
+#' head(data_miss)
+#' 
+#'}
+
+simulate_data <- function(n_total = 3500, 
+                          seed = 42,
+                          include_id = TRUE,
+                          imposeNA = TRUE,
+                          propNA = NULL
+                          ){
+  
+  # input checks
+  assertthat::assert_that(inherits(n_total, "numeric"), msg = "<n_total> needs to be an integer")
+  assertthat::assert_that(inherits(seed, "numeric"), msg = "<seed> needs to be an integer")
+  assertthat::assert_that(inherits(include_id, "logical"), msg = "<seed> needs to be a logical")
+  assertthat::assert_that(inherits(imposeNA, "logical"), msg = "<imposeNA> needs to be a logical")
+  if(!is.null(propNA)) assertthat::assert_that(inherits(propNA, "numeric"), msg = "<propNA> needs to be numeric")
[…]

### 2024-12-14 functions/simulate_data.R
renamed functioin

@@ -1,7 +1,7 @@
-#' Simulates and artifical FLAURA EHR-derived dataset
+#' Simulates and artifical EHR-derived dataset
 #'
 #' @description
-#' Parameterized function to quickly create an artificial FLAURA
+#' Parameterized function to quickly create an artificial
 #' EHR-derived analytic cohort for analytic code development.
 #'
 #' @details
@@ -26,7 +26,7 @@
 #' \dontrun{
 #' library(encore.io)
 #'
-#' data_miss <- simulate_flaura(
+#' data_miss <- simulate_data(
 #'   n_total = 3500, 
 #'   seed = 41, 
 #'   include_id = FALSE, 

### 2024-12-14 manuscript/manuscript.qmd
renamed functioin
modified figure 3

@@
 
+To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed and internal R package `encore.io`with parameterized functions
+
 ## Agreement metrics {#sec-agreement-metrics}
@@
 
-source(here("functions", "simulate_flaura.R"))
+source(here("functions", "simulate_data.R"))
 
-data <- simulate_flaura(
+data <- simulate_data(
   n_total = 2000, 
@@
 # impute
-mids_data <- mice(data, m = 10, printFlag = F)
+suppressMessages(mids_data <- mice(data, m = 10, printFlag = F))
 
@@
 # weighting  
-wimids_data <- weightthem(
+suppressMessages(wimids_data <- weightthem(
   datasets = mids_data, 
@@
   estimand = "ATO"
-  )
+  ))
 
@@
   abs = TRUE,
-  var.names = named_covariates,
   thresholds = 0.1, 
@@
 # distributional balance
-score_balance <- bal.plot(
-  x = wimids_data,
-  var.name = "prop.score",
-  which = "both",
-  which.imp = .none,
-  colors = c("orange", "blue")
+# Simulate data
+set.seed(42)  # For reproducibility
+n <- 2000  # Number of samples per group
+drugA_pre <- rnorm(n, mean = 0, sd = 2.5 / 2)  # Spread = 2.5
+drugB_pre <- rnorm(n, mean = -1, sd = 2.5 / 2)  # Spread = 2.5
+drugA_post <- rnorm(n, mean = 0, sd = 2.5 / 2)  # Spread = 2.5
+drugB_post <- rnorm(n, mean = 0, sd = 2.5 / 2)  # Spread = 2.5
+
+# Combine into a data frame
+data <- data.frame(
+  Value = c(drugA_pre, drugA_post, drugB_pre, drugB_post),
+  Group = rep(c("Drug A", "Drug A", "Drug B", "Drug B"), each = n),
+  PrePost = rep(c("Unadjusted sample", "Adjusted sample", "Unadjusted sample", "Adjusted sample"), each = n)
+  ) |> 
+  mutate(PrePost = factor(PrePost, levels = c("Unadjusted sample", "Adjusted sample")))
+
+# Create the plot
+score_balance <- ggplot(data, aes(x = Value, fill = Group)) +
+  geom_density(alpha = 0.5) +
+  scale_fill_manual(values = c("orange", "blue")) +
+  labs(
+    title = "Distributional balance for prognostic score",
+    x = "Value",
+    y = "Density",
+    fill = "Group"
   ) +
-  labs(
-    x = "Prognostic score",
-    title = "Distributional balance for prognostic score"
-    ) +
-  theme(legend.position = "top")
+  theme_minimal() +
+  theme(legend.position = "top") +
+  facet_wrap(~PrePost)
 

### 2024-12-14 manuscript/manuscript.qmd
added draft for agreement metrics

@@
 
+### Sensitivity analyses
+
+Given that a large number of factors could influence a difference between RCT results and emulated trial results, a broad range of sensitivity analyses will be conducted.
+This can comprise decisions on the considered databases, calendar time period, covariate measurement (e.g., measurement windows or trade-offs on sensitivity versus specificity in measurements), approaches to missing data, selection of covariates for imputation and propensity score models and decisions on when to censor patients.
+All sensitivity will be pre-specified in the study protocol and reported using appropriate visualizations such as forest plots.
+
 ### Reproducibility and consistency across emulation
 
-To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed and internal R package `encore.io`with parameterized functions
+To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed and internal R package `encore.io`with parameterized functions.
+A detailed documentation can be found in the Supplementary Material.
 
 ## Agreement metrics {#sec-agreement-metrics}
+
+To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive two qualitative agreement metrics (regulatory agreement, estimate agreement) and one quantitative agreement based on the SMD.
+
+-   **Regulatory agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.
+This agreement metric mimics the question if RWE would have lead to the same regulatory conclusion as the trial, based on standard regulatory decision rules.
+
+-   **Estimate agreement**: agreement that the estimated RWE treatment treatment effect is within the 95% CI of the RCT treatment effect estimate.
+Provided that for some emulations the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there us no regulatory agreement although the treatment effect estimates are highly overlapping but the RCT estimate crosses the null (or vice versa).
+
+-   **SMD agreement**: quantification of the agreement between the emulated RWE and RCT treatment effect estimate. The SMD is calculated as
+
+$$
+SMD = \frac{\hat{\theta}_{RCT} - \hat{\theta}_{RWE}}{\sqrt{\text{Var}(\hat{\theta}_{RCT}) + \text{Var}(\hat{\theta}_{RWE})}}
+$$
+
+where $\hat{\theta}$ are the treatment effect estimates (hazard ratios or median survival times) and $\text{Var}(\hat{\theta})$ are the corresponding variances.
+The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart.
+For an $\alpha$-level of 0.05, the null hypothesis of no difference would be rejected whenever $|Z| > 1.96$.
+
+We would like to note that only the SMD agreement metric will be applicable to absolute risk measures like median survival times or survival probabilities.
 
@@
 
-{{< pagebreak >}}
-
 # Figures {.unnumbered}
@@
 # combine plots with shared legend
-shared_legend <- get_legend(initiators)
+suppressWarnings(shared_legend <- get_legend(initiators))
 
-plot_utilization <- suppressMessages(plot_grid(
+plot_utilization <- plot_grid(
   plot_grid(initiators + theme(legend.position = "top")),
@@
   shared_legend
-  ))
+  )
 
@@
 # impute
-suppressMessages(mids_data <- mice(data, m = 10, printFlag = F))
+mids_data <- suppressWarnings(mice(data, m = 10, printFlag = F))
 
@@
   dpi = 800,
-  width = 15
+  width = 15,
+  height = 10
   ))

### 2024-12-15 manuscript/manuscript.qmd
working on plot

@@
   pdf:
-    include-in-header: 
-      text: |
-        \usepackage{lscape}
-        \newcommand{\blandscape}{\begin{landscape}}
-        \newcommand{\elandscape}{\end{landscape}}
     fig-cap-location: top
@@
 library(cobalt)
+library(gtExtras)
 ```
@@
 
-To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed and internal R package `encore.io`with parameterized functions.
+To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed and internal R package `encore.io` with parameterized functions.
 A detailed documentation can be found in the Supplementary Material.
@@
 -   **Estimate agreement**: agreement that the estimated RWE treatment treatment effect is within the 95% CI of the RCT treatment effect estimate.
-Provided that for some emulations the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there us no regulatory agreement although the treatment effect estimates are highly overlapping but the RCT estimate crosses the null (or vice versa).
+Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no regulatory agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).
 
@@
 
-We would like to note that only the SMD agreement metric will be applicable to absolute risk measures like median survival times or survival probabilities.
+For the secondary endpoints of interest (e.g., median survival time or survival probabilities) only the SMD agreement metric will be applicable.
 
@@
 
-Target populations in international trials...genetic setup.
+Building off an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to assess the potential use of RWE for regulatory science.
+The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations.
+
+-   Distributions of target populations in international trials...genetic setup.
+
+-   Bias towards the null, per protocol effect difficult
+
+-   Transportability =\> RWE estimates may not be the same as RCT but still informative given more generalizable/broader populations considered
 
@@
   width = 15,
-  height = 10
+  height = 8
   ))
@@
 ```
+
+```{r}
+#| label: fig-metrics
+#| fig-cap: "Visualization of agreement metrics example."
+#| fig-widh: 4
+#| fig-height: 7
+
+source(here("functions","smd_agreement.R"))
+
+data <- tribble(
+  ~group, ~estimate, ~lower, ~upper,
+  "RCT", 0.75, 0.61, 0.91,
+  "RWE", 0.80, 0.51, 1.1
+  )
+
+smd <- smd_agreement(
+  rct_estimate = data |> filter(group == "RCT") |> pull(estimate),
+  rct_lower = data |> filter(group == "RCT") |> pull(lower),
+  rct_upper = data |> filter(group == "RCT") |> pull(upper),
+  rwe_estimate = data |> filter(group == "RWE") |> pull(estimate),
+  rwe_lower = data |> filter(group == "RWE") |> pull(lower),
+  rwe_upper = data |> filter(group == "RWE") |> pull(upper)
+  )
+
+data |> 
+  gt() |> 
+  gt_plt_conf_int(
+    column = estimate, 
+    ci_columns = c(lower, upper),
+    ref_line = 1,
+    palette = c("red", "lightgrey", "black", "red")
+    ) |> 
+  cols_hide(c(lower, upper)) |> 
+  # add agreement metrics
+  cols_add(
+    metric = c(md("Regulatory agreement: <br><br> Estimate agreement:"), "SMD:"),
+    results = c(md("No <br><br> Yes"), ifelse(abs(smd) < 1.96, "Yes", "No"))
+    ) |> 
+  tab_options(
+    table_body.hlines.width = px(0)
+  ) |> 
+  cols_label(
+    group = "",
+    estimate = "HR (95% CI)",
+    metric = "Agreement metrics",
+    results = ""
+    ) |> 
+  fmt_markdown() |> 
+  tab_style(
+    style = cell_text(align = "center", v_align = "middle"),
+    locations = cells_column_labels()
+    ) |> 
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels()
+    )
+```

### 2024-12-15 manuscript/manuscript.qmd
added example table for agreement metrics

@@
 library(gtExtras)
+library(glue)
+library(tidyr)
 ```
@@
 
-To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive two qualitative agreement metrics (regulatory agreement, estimate agreement) and one quantitative agreement based on the SMD.
+To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive two qualitative agreement metrics (regulatory agreement, estimate agreement) and one quantitative agreement based on the SMD (examples are illustrated @tbl-metrics).
 
@@
 include_graphics(path = here("tables", "Table_2_trial_selection.png"))
+```
+
+{{< pagebreak >}}
+
+```{r}
+#| label: tbl-metrics
+#| tbl-cap: "Visualization of agreement metrics example."
+
+source(here("functions","smd_agreement.R"))
+
+data <- tribble(
+  ~trial, ~group, ~estimate, ~lower, ~upper,
+  "Trial 1", "RCT", 0.75, 0.61, 0.91,
+  "Trial 1", "RWE", 0.80, 0.51, 1.1,
+  "Trial 2", "RCT", 0.62, 0.51, 0.71,
+  "Trial 2", "RWE", 0.65, 0.55, 0.69,
+  "Trial 3", "RCT", 0.71, 0.67, 0.80,
+  "Trial 3", "RWE", 0.51, 0.41, 0.61,
+  "Trial 4", "RCT", 0.90, 0.81, 0.99,
+  "Trial 4", "RWE", 1.2, 1.09, 1.34,
+  ) |> 
+  pivot_wider(names_from = group, values_from = c(estimate, lower, upper)) |> 
+  mutate(
+    smd_value = smd_agreement(
+      rct_estimate = estimate_RCT, 
+      rct_lower = lower_RCT, 
+      rct_upper = upper_RCT,
+      rwe_estimate = estimate_RWE,
+      rwe_lower = lower_RWE, 
+      rwe_upper = upper_RWE
+      )
+    ) |> 
+  mutate(
+    regulatory = ifelse(
+      # superiority
+      estimate_RCT < 1 & upper_RCT < 1 & estimate_RWE < 1 & upper_RWE < 1 |
+        # non-inferiority
+        estimate_RCT < 1 & upper_RCT >= 1 & estimate_RWE < 1 & upper_RWE >= 1 |
+        # inferiority (made up)
+        estimate_RCT >= 1 & lower_RCT >= 1 & estimate_RWE >= 1 & upper_RWE >= 1,
+      "Yes", "No"
+      ),
+    estimate = ifelse(
+      estimate_RWE >= lower_RCT & estimate_RWE <= upper_RCT, "Yes", "No"
+      ),
+    smd = ifelse(
+      abs(smd_value) < 1.96, "Yes", "No"
+      )
+    )
+
+tbl_metrics <- data |> 
+  mutate(across(where(is.numeric), function(x) format(x, digits = 2, nsmall = 2))) |> 
+  mutate(
+    RCT = glue("{estimate_RCT} ({lower_RCT} - {upper_RCT})"),
+    RWE = glue("{estimate_RWE} ({lower_RWE} - {upper_RWE})"),
+    smd = glue("{smd} ({smd_value})")
+  ) |> 
+  select(trial, RCT, RWE, regulatory, estimate, smd) |> 
+  gt() |> 
+  tab_spanner(
+    label = "HR (95% CI)",
+    columns = c(RCT, RWE)
+  ) |> 
+  cols_label(
+    trial = "Trial",
+    regulatory = md("Regulatory <br> agreement"),
+    estimate = md("Estimate <br> agreement"),
+    smd = "SMD"
+    ) |> 
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_labels()
+    ) |> 
+    tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_column_spanners()
+    ) |> 
+  tab_style(
+    style = cell_text(color = "darkgreen"), # Pink for "Inactive"
+    locations = cells_body(
+      columns = regulatory,
+      rows = regulatory == "Yes"
+      )
+    ) |> 
+  tab_style(
+    style = cell_text(color = "darkred"), # Pink for "Inactive"
+    locations = cells_body(
+      columns = regulatory,
+      rows = regulatory == "No"
+      )
+    ) |> 
+  tab_style(
+    style = cell_text(color = "darkgreen"), # Pink for "Inactive"
+    locations = cells_body(
+      columns = estimate,
+      rows = estimate == "Yes"
+      )
+    ) |> 
+  tab_style(
+    style = cell_text(color = "darkred"), # Pink for "Inactive"
+    locations = cells_body(
+      columns = estimate,
+      rows = estimate == "No"
+      )
+    ) |> 
+  tab_style(
+    style = cell_text(color = "darkgreen"), # Pink for "Inactive"
+    locations = cells_body(
+      columns = smd,
+      rows = grepl("^Yes", smd)
+      )
+    ) |> 
+  tab_style(
+    style = cell_text(color = "darkred"), # Pink for "Inactive"
+    locations = cells_body(
+      columns = smd,
+      rows = estimate == "No"
+      )
+    )
+
+# save table as .docx document
+tbl_metrics |> 
+   gtsave(
+    filename = here("manuscript", "Table_3_agreement_metrics.docx")
+    )
+
+# save image to display within manuscript document
+suppressMessages(tbl_metrics |> 
+  gtsave(
+    filename = here("tables", "Table_3_agreement_metrics.png"),
+    expand = 10, # amount of whitespace,
+    zoom = 5 # higher resolution
+    ))
+
+include_graphics(path = here("tables", "Table_3_agreement_metrics.png"))
 ```
@@
 ```
-
-```{r}
-#| label: fig-metrics
-#| fig-cap: "Visualization of agreement metrics example."
-#| fig-widh: 4
-#| fig-height: 7
-
-source(here("functions","smd_agreement.R"))
-
-data <- tribble(
-  ~group, ~estimate, ~lower, ~upper,
-  "RCT", 0.75, 0.61, 0.91,
-  "RWE", 0.80, 0.51, 1.1
-  )
-
-smd <- smd_agreement(
-  rct_estimate = data |> filter(group == "RCT") |> pull(estimate),
-  rct_lower = data |> filter(group == "RCT") |> pull(lower),
-  rct_upper = data |> filter(group == "RCT") |> pull(upper),
-  rwe_estimate = data |> filter(group == "RWE") |> pull(estimate),
-  rwe_lower = data |> filter(group == "RWE") |> pull(lower),
-  rwe_upper = data |> filter(group == "RWE") |> pull(upper)
-  )
-
-data |> 
-  gt() |> 
-  gt_plt_conf_int(
-    column = estimate, 
-    ci_columns = c(lower, upper),
-    ref_line = 1,
-    palette = c("red", "lightgrey", "black", "red")
-    ) |> 
-  cols_hide(c(lower, upper)) |> 
-  # add agreement metrics
-  cols_add(
-    metric = c(md("Regulatory agreement: <br><br> Estimate agreement:"), "SMD:"),
-    results = c(md("No <br><br> Yes"), ifelse(abs(smd) < 1.96, "Yes", "No"))
-    ) |> 
-  tab_options(
-    table_body.hlines.width = px(0)
-  ) |> 
-  cols_label(
-    group = "",
-    estimate = "HR (95% CI)",
-    metric = "Agreement metrics",
-    results = ""
-    ) |> 
-  fmt_markdown() |> 
-  tab_style(
-    style = cell_text(align = "center", v_align = "middle"),
-    locations = cells_column_labels()
-    ) |> 
-  tab_style(
-    style = cell_text(weight = "bold"),
-    locations = cells_column_labels()
-    )
-```

### 2024-12-15 functions/smd_agreement.R

@@ -1,0 +1,52 @@
+smd_agreement <- function(rct_estimate = NULL,
+                          rct_lower = NULL,
+                          rct_upper = NULL,
+                          rwe_estimate = NULL,
+                          rwe_lower = NULL,
+                          rwe_upper = NULL
+                          ){
+  
+  # numerator
+  numerator <- (rct_estimate - rwe_estimate)
+  
+  # variances
+  var_rct <- (rct_upper - rct_lower)/(2*1.96)
+  var_rwe <- (rwe_upper - rwe_lower)/(2*1.96)
+  
+  # denominator
+  denominator <- sqrt(var_rct^2 + var_rwe^2)
+  
+  # smd
+  smd <- numerator/denominator
+  
+  return(smd)
+  
+}
+
+# examples (taken from DUPLICATE)
+# smd_agreement <- smd_agreement(
+#   rct_estimate = log(0.87),
+#   rct_lower = log(0.78),
+#   rct_upper = log(0.97),
+#   rwe_estimate = log(0.82),
+#   rwe_lower = log(0.87),
+#   rwe_upper = log(0.76)
+#   )
+# 
+# smd_agreement <- smd_agreement(
+#   rct_estimate = log(0.83),
+#   rct_lower = log(0.73),
+#   rct_upper = log(0.95),
+#   rwe_estimate = log(0.69),
+#   rwe_lower = log(0.59),
+#   rwe_upper = log(0.81)
+#   )
+# 
+# smd_agreement <- smd_agreement(
+#   rct_estimate = log(1),
+#   rct_lower = log(0.89),
+#   rct_upper = log(1.12),
+#   rwe_estimate = log(0.81),
+#   rwe_lower = log(0.76),
+#   rwe_upper = log(0.86)
+#   )

### 2024-12-15 manuscript/manuscript.qmd
first full draft

@@
 
-[**Word count:**]{.underline} xxx words / xxx words
+[**Manuscript word count:**]{.underline} xxx words / 8,000 words
 
-[**Tables:**]{.underline} x
+[**Abstract word count:**]{.underline} xxx words / 250 words
 
-[**Figures:**]{.underline} x
+[**Tables:**]{.underline} 3
+
+[**Figures:**]{.underline} 3
 
@@
 
-[**Data sharing statement:**]{.underline} ...
+[**Data sharing statement:**]{.underline} No data was analyzed as part of this project.
 
-[**Analytic code sharing statement:**]{.underline} ...
+[**Analytic code sharing statement:**]{.underline} Code to reproduce this manuscript, including all figures and tables, can be found at <https://gitlab-scm.partners.org/drugepi/encore/encore-process-paper>.
 
@@
 
-xxx words/xxx words
-
-...
+Real-world data (RWD) is increasingly utilized to complement randomized controlled trials (RCTs) and contextualize the efficacy and safety of medical interventions through real-world evidence (RWE) studies.
+Simultaneously, advancements in the curation and accessibility of electronic health record data (EHR) at scale have presented the opportunity to investigate disease domains that have previously been inaccessible through administrative healthcare claims databases, including the field of oncology.
+The RCT DUPLICATE initiative has previously enhanced our understanding of when RWE studies can draw causal conclusions by emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas.
+In this process paper, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to the field of oncology.
+This expansion will emulate 12 RCTs in multiple EHR databases and across four different cancer indications, including non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
+The objectives of this project are to develop state-of-the-art methodological approaches and apply these to create insights that may provide guidance on the potential use of RWE in oncology.
+The project will place a special emphasis on a systematic evaluation of the suitability of fit-for-purpose data in relation to the study design and statistical analysis, and assessing the agreement of treatment effect estimates between RCTs and their respective emulations.
+Through extensive sensitivity analyses, the ENCORE project will further assess as to which measurement, design, and analytic decisions influence bias and validity in oncological RWE studies by calibrating those against RCTs.
 
@@
 The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations.
+Historically, administrative health claims databases have been the backbone of most research in RWE.
+With increasing access to EHR and a maturing set of methodological approaches to draw causal inferences from such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to deliver similar causal conclusions compared to RCTs in the field of oncology.
 
--   Distributions of target populations in international trials...genetic setup.
+The RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges which we expect to also encounter in ENCORE.
+One particular aspect that we may not always be able emulate is the exact distribution of patient characteristics of the trial population.
+This can be due to the lack of data granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide, which we won't be able to mirror given that all considered databases reflect the US only.
+This may be a critical factor especially given that the pathophysiology, prognosis and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries.
+For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations) are much more prevalent in Asian countries compared to the US or Europe.
 
--   Bias towards the null, per protocol effect difficult
+Another common challenge in the emulation of oncology trials is the estimation of an "intention-to-treat" (ITT) analogous estimand which is usually the primary estimand reported in oncology RCTs.
+Due to intercurrent events, such as non-adherence or the crossover of a high proportion of patients from the control to the intervention arm, trial emulation are prone to be biased towards the null.
+Although this is also a common challenge in the analysis of RCTs[@rufibach2018], treatment protocols in routine care are often observed to be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
 
--   Transportability =\> RWE estimates may not be the same as RCT but still informative given more generalizable/broader populations considered
+This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD.
+As a result, exposure misclassification (due to too narrow ascertainment windows) and selection bias (due to too long ascertainment windows) are common trade-offs in such scenarios.
+Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design[@gaber2024mystifying], may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters and provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.
 
@@
 
-...
+RWE based on fit-for-purpose data and principled, well-designed and reproducible studies may complement evidence coming from RCTs.
+The ENCORE project will systematically assess and eventually increase our understanding as to which measurement, design and analytic decisions influence bias and validity in oncological RWE studies by calibrating those against RCTs.
 
@@
     smd_value = smd_agreement(
-      rct_estimate = estimate_RCT, 
-      rct_lower = lower_RCT, 
-      rct_upper = upper_RCT,
-      rwe_estimate = estimate_RWE,
-      rwe_lower = lower_RWE, 
-      rwe_upper = upper_RWE
+      rct_estimate = log(estimate_RCT), 
+      rct_lower = log(lower_RCT), 
+      rct_upper = log(upper_RCT),
+      rwe_estimate = log(estimate_RWE),
+      rwe_lower = log(lower_RWE), 
+      rwe_upper = log(upper_RWE)
       )
@@
       )
-    )
+    ) |> 
+  tab_footnote("Abbreviations: CI = Confidence interval, HR = Hazard ratio, RCT = Randomized controlled trial, RWE = Real-world evidence, SMD = standardized mean difference (based on log hazard ratios)")
 
@@
 #| label: fig-initiators
-#| fig-cap: "Descriptive drug utilization analyses."
+#| fig-cap: "Visualization of descriptive drug utilization analyses displaying a) initiation trends between compared regimens based on calendar time, b) cumulative rate of patients switching to another line of treatment."
 #| fig-widh: 4
@@
 #| label: fig-balance
-#| fig-cap: "Assessment of balance after propensity score matching or weighting across multiple imputed datasets."
+#| fig-cap: "Assessment of a) covariate bbalance and b) distributional balance of a prognostic score for overall survival before and after propensity score matching or weighting across multiple imputed datasets."
 #| fig-widh: 4

### 2024-12-15 manuscript/manuscript.qmd
wording

@@
 Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
-With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] With 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023,[@senior2024fresh] and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
+With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023,[@senior2024fresh] and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
 Potential use cases comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

### 2024-12-15 manuscript/manuscript.qmd
typo

@@
 
-However, the validity and transportability of results derived betwen RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered.
+However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered.
 

### 2024-12-15 manuscript/manuscript.qmd
added funding statement

@@
 
-[**Short running title**]{.underline}: ...
+[**Short running title**]{.underline}: Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)
 
@@
 
-[**Funding Statement:**]{.underline} This project was supported by ...
+[**Funding Statement:**]{.underline} This project was supported by an FDA BAA contract with contract number 75F40122C00181.
 

### 2024-12-15 manuscript/manuscript.qmd
minor formatting and added word count

@@
 
-[**Manuscript word count:**]{.underline} xxx words / 8,000 words
+[**Manuscript word count:**]{.underline} 4,131 words / 8,000 words
 
-[**Abstract word count:**]{.underline} xxx words / 250 words
+[**Abstract word count:**]{.underline} 248 words / 250 words
 
@@
 
-[**Supplementary material:**]{.underline} Supplementary tables and figures
+[**Supplementary material:**]{.underline} Supplementary figures and material
 
@@
 
-**Fit-for-purpose data.** Real-world data fitness and emulation feasibility for a given shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care.
+#### **Fit-for-purpose data**
+
+Real-world data fitness and emulation feasibility for a given shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care.
 Since a vast majority of oncological RCTs in recent years have focused on selected, biomarker-defined populations, subtleties in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population.
@@
 
-**Descriptives and data exploration.** Critical aspects when emulating oncology trials are the choice and estimation of the appropriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the (historical) control cohort, the adoption rate of the novel intervention in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null.
+#### **Descriptives and data exploration**
+
+Critical aspects when emulating oncology trials are the choice and estimation of the appropriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the (historical) control cohort, the adoption rate of the novel intervention in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null.
 To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial.
@@
 
--   **Regulatory agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.
+**Regulatory agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.
 This agreement metric mimics the question if RWE would have lead to the same regulatory conclusion as the trial, based on standard regulatory decision rules.
 
--   **Estimate agreement**: agreement that the estimated RWE treatment treatment effect is within the 95% CI of the RCT treatment effect estimate.
+**Estimate agreement**: agreement that the estimated RWE treatment treatment effect is within the 95% CI of the RCT treatment effect estimate.
 Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no regulatory agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).
 
--   **SMD agreement**: quantification of the agreement between the emulated RWE and RCT treatment effect estimate. The SMD is calculated as
+**SMD agreement**: quantification of the agreement between the emulated RWE and RCT treatment effect estimate. The SMD is calculated as
 

### 2024-12-15 manuscript/manuscript.qmd
minor formatting

@@
 After careful review and exchange with clinical and regulatory experts, four cancer indications were identified including lung cancer, breast cancer, colorectal cancer and multiple myeloma.
-For each cancer we aim to conduct three trial emulations which will be implemented using multiple databases accessible for the scope of this project (i.e.
-the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).
+For each cancer we aim to conduct three trial emulations which will be implemented using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).
 
@@
 
-**Statistical power**.
+#### **Statistical power**
+
 Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power than RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, statistical power is a critical to assess the feasibility that a comparison of agreement metrics of estimated parameters between RCT and RWD with sufficient precision can be made.
@@
 
-This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD.
+This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD.
 As a result, exposure misclassification (due to too narrow ascertainment windows) and selection bias (due to too long ascertainment windows) are common trade-offs in such scenarios.

### 2024-12-15 manuscript/manuscript.qmd
removed link
added links to higher resolution figures
corrected link

@@
 
-[View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/figures/process.png)
+[View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/manuscript/Figure_1_process.png)
 
@@
 ```
+
+[View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/manuscript/Figure_2_utlization.png)
 
@@
 ```
+
+[View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/manuscript/Figure_3_balance.png)

### 2024-12-15 manuscript/supplement.qmd
correction in supplement
added example table for agreement metrics
no references in supplement
minor correctioin

@@
 
-*Note that @fig-consort-nsclc results in four shortlisted candidates because both CheckMate017/057 have been both shortlisted and only differ in the squamous versus nonsquamous histological eligibility of the trial population.*
+*Note that Supplementary @fig-consort-nsclc results in four shortlisted candidates because both CheckMate017/057 have been both shortlisted and only differ in the squamous versus nonsquamous histological eligibility of the trial population.*
 
@@
 
-# Supplementary Tables
-
-{{< pagebreak >}}
-
-\addcontentsline{toc}{section}{\protect\numberline{}R package documentation}
-\includepdf[pages=-]{encore.io_0.2.0.pdf}
+# R package documentation
 
@@
 
+\includepdf[pages=-]{encore.io_0.2.0.pdf}
+
 {{< pagebreak >}}
@@
 *Supplementary Material last updated: `r Sys.Date()`*
-
-# References {.unnumbered}
-
-::: refs
-:::

### 2024-12-15 manuscript/manuscript.qmd
modified url repo

@@
 
-[**Analytic code sharing statement:**]{.underline} Code to reproduce this manuscript, including all figures and tables, can be found at <https://gitlab-scm.partners.org/drugepi/encore/encore-process-paper>.
+[**Analytic code sharing statement:**]{.underline} Code to reproduce this manuscript, including all figures and tables, can be found at <https://github.com/janickweberpals/encore-process-manuscript>.
 

### 2024-12-18 manuscript/manuscript.qmd
added PA as co-author
corrected string expression for coloring agreement metrics

@@
 
-**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
+**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, Priyanka Anand^1^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
 
@@
   tab_style(
-    style = cell_text(color = "darkgreen"), # Pink for "Inactive"
+    style = cell_text(color = "darkgreen"), 
     locations = cells_body(
@@
   tab_style(
-    style = cell_text(color = "darkred"), # Pink for "Inactive"
+    style = cell_text(color = "darkred"), 
     locations = cells_body(
@@
   tab_style(
-    style = cell_text(color = "darkgreen"), # Pink for "Inactive"
+    style = cell_text(color = "darkgreen"), 
     locations = cells_body(
       columns = smd,
-      rows = grepl("^Yes", smd)
+      rows = stringr::str_detect(smd, "Yes")
       )
@@
   tab_style(
-    style = cell_text(color = "darkred"), # Pink for "Inactive"
+    style = cell_text(color = "darkred"), 
     locations = cells_body(
       columns = smd,
-      rows = estimate == "No"
+      rows = stringr::str_detect(smd, "No")
       )

### 2025-01-07 manuscript/manuscript.qmd
implemented SVW comments

@@
 
-Real-world data (RWD) is increasingly utilized to complement randomized controlled trials (RCTs) and contextualize the efficacy and safety of medical interventions through real-world evidence (RWE) studies.
-Simultaneously, advancements in the curation and accessibility of electronic health record data (EHR) at scale have presented the opportunity to investigate disease domains that have previously been inaccessible through administrative healthcare claims databases, including the field of oncology.
-The RCT DUPLICATE initiative has previously enhanced our understanding of when RWE studies can draw causal conclusions by emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas.
-In this process paper, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to the field of oncology.
-This expansion will emulate 12 RCTs in multiple EHR databases and across four different cancer indications, including non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
-The objectives of this project are to develop state-of-the-art methodological approaches and apply these to create insights that may provide guidance on the potential use of RWE in oncology.
-The project will place a special emphasis on a systematic evaluation of the suitability of fit-for-purpose data in relation to the study design and statistical analysis, and assessing the agreement of treatment effect estimates between RCTs and their respective emulations.
-Through extensive sensitivity analyses, the ENCORE project will further assess as to which measurement, design, and analytic decisions influence bias and validity in oncological RWE studies by calibrating those against RCTs.
+Real-world evidence (RWE) studies are increasingly used to complement evidence from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in clinical practice.
+Advancements in the curation and accessibility of electronic health record data (EHR) have presented the opportunity to investigate disease domains such as oncology, where administrative healthcare claims databases alone are not fit-for-purpose.
+The RCT DUPLICATE initiative has previously enhanced understanding of when RWE studies can draw causal conclusions by emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas.
+In this process paper, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to oncology.
+This expansion will emulate 12 RCTs in multiple EHR databases across four different cancer indications, including non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
+The objectives of this project are to develop state-of-the-art methodological approaches and to apply them in a demonstration project designed to create insights for guiding the use of RWE in oncology.
+The project will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to analysis.
+Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria.
+Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence bias and validity in oncological RWE studies.
 
@@
 Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
-With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023,[@senior2024fresh] and especially in the field of precision oncology, RWE has a large potential to complement evidence coming from RCTs.
-Potential use cases comprise the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
+With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures..
 
-However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the accuracy of RWE studies and to provide an answer as to which questions can be validly answered.
+However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.
 
-The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by comparing results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
+The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
 In settings where the RCT designs could be emulated well, RWE studies came to the same conclusions.[@heyard2024design] However, prior work from RCT-DUPLICATE has focused primarily on emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas using claims databases.
 
-The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project [@encoreFDA] aims to extend this work to the field of oncology which comes with its own unique set of challenges which must be systematically explored and understood.
-Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology is going to emulate 12 randomized clinical trials using multiple EHR data sources.
-The process includes an emphasis on transparency with documented assessment of data fitness of the RWD source for each trial[@rivera2024oncology; @gatto2022structured] and the conduct of extensive sensitivity analyses to assess robustness of findings and trial eligibility criteria.
+The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project [@encoreFDA] aims to extend this work to the field of oncology.
+Studies in oncology come with their own unique set of challenges which must be systematically explored and understood.
+Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources.
+The process will emphasize transparency and include documented data fitness assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.
 
-The objectives of this project are to develop state-of-the-art methodological approaches and apply these to create insights that may provide guidance on the potential use of RWE for regulatory science in oncology.
-This includes the systematic evaluation of the suitability of data in relation to the study design and statistical analysis by emulating 12 oncology trials across four cancers and assessing the agreement of treatment effect estimates between RCTs and their respective emulations.
+The objectives of this project are to 1) develop state-of-the-art methodological approaches and 2) to apply them to create insights that may provide guidance on the potential use of RWE for regulatory science in oncology.
+To achieve these objectives, this demonstration project will systematically emulate 12 oncology trials across four cancers and assess the agreement of treatment effect estimates between RCTs and their respective emulations.
 
-In this process paper, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development, study design and statistical analysis and final agreement metrics to evaluate the concordance between RCTs and emulations.
+In this process paper, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development and pre-registration, study design and statistical analysis, and pre-specified agreement metrics to evaluate the concordance between RCTs and emulations.
 
@@
 The focus of ENCORE is to maximize potential learnings on when RWE studies can or cannot yield similar results compared to RCTs.
-To that end, the emphasis of the project is on trials of therapies for the most common cancers and/or cancers for which there has been substantial therapeutic development in recent years.
+To that end, the emphasis of the project is on trials of therapies for the most common cancers for which there has been substantial therapeutic development in recent years.
 After careful review and exchange with clinical and regulatory experts, four cancer indications were identified including lung cancer, breast cancer, colorectal cancer and multiple myeloma.
-For each cancer we aim to conduct three trial emulations which will be implemented using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).
+For each cancer type we aim to conduct three trial emulations using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).
 
-The trial selection will follow a semi-automated process for which we will document the eligibility criteria resulting in a CONSORT diagram showing reasons for excluding RCTs.
-The search will be conducted using the AACT database which is s a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) which contains all information (protocol and result data elements) about every study registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we will use a combined search query strategy of the National Library of Medicine (NLM)-controlled *MeSH* term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.
+We used a semi-automated process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs for each cancer type.
+The search was conducted using the AACT database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains all information (protocol and result data elements) about every study registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *MeSH* term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.
 
-Eligible trials need to fulfill the following basic criteria:
+Eligible trials had to fulfill the following basic criteria:
 
@@
 The rationale and operationalization of each criterion is listed in detail in @tbl-criteria.
-We will mainly consider pivotal interventional, randomized trials after 2011 since treatment guidelines among included cancer indications have undergone significant changes in recent years.
-Due to the rapid adoption of new breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in the real-world.
-In parallel, trial results should have also not been published too recently in order to allow for enough data and follow-up time accrual in databases used for this project.
-A further focus is on trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.
-Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] Given the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, the scope of this project was to focus on the emulation of OS as the endpoint of interest.
+We will mainly consider pivotal interventional, randomized trials after 2011 because treatment guidelines among included cancer indications have undergone significant changes in recent years.
+Due to the rapid adoption of new breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in current clinical practice.
+We also did not include trials with results published too recently.
+This decision was in order to allow for enough data and follow-up time accrual in the databases used for this project.
+Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] Given this, and the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
+Therefore, we include only trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.
 
-While most trial-eligibility criteria can be operationalized in an automated fashion, the last criterion on emulation feasibility and clinical relevance involves extensive human review.
-The critical points considered in this step include a thorough feasibility assessment of the data fitness [@rivera2022friends], sample size considerations and the assessment if critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and can be balanced using propensity score matching or weighting methods.[@brookhart2006variable] Lastly, trial candidates are ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.
+While most trial eligibility criteria could be operationalized in an automated fashion, our final criterion was an assessment of emulation feasibility and clinical relevance.
+This criterion involved extensive human review.
+The critical points considered in this step include an initial feasibility assessment of the data fitness[@rivera2022friends], including assessment of whether critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary sample size counts are reasonable.
+Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.
 
 A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
-Naturally, the majority of trials will cover advanced or metastatic cancer populations since a large proportion of drug development efforts have focused on these settings in recent years.
+Naturally, the majority of trials cover advanced or metastatic cancer populations because a large proportion of drug development efforts have focused on these settings in recent years.
 A key learning that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD.
-If ongoing feasibility assessments indicate that these trials cannot be emulated with high enough confidence, runner-up candidates will be considered instead.
+If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the sample size is not sufficient, then runner-up candidates will be considered instead.
 
@@
 
-The ENCORE project will utilize data from a total of four different oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada.
+The ENCORE project will use data from a four oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada.
 All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology.
-A detailed description and methodology on how patients are sampled will be provided for each emulation protocol individually.
+A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol.
 For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results.
-If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach.
+If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects models.
 
@@
 
-For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed following the HARPER protocol template[@wang2022harmonized] and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel.
+For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] which has been recommended for regulatory submissions of RWE studies[@guidelinegeneral], and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel.
 Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics).
-All eligibility criteria will be extracted based on publicly available protocols and statistical analysis plans of the selected RCT.
+All eligibility criteria will be extracted based on publications, publicly available protocols and statistical analysis plans of the selected RCT.
 
@@
 
-Real-world data fitness and emulation feasibility for a given shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care.
+Real-world data fitness and emulation feasibility for each shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care.
 Since a vast majority of oncological RCTs in recent years have focused on selected, biomarker-defined populations, subtleties in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population.
@@
 To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial.
-Example for such standard diagnostics are visualized in @fig-initiators.
+Examples for such standard diagnostics are visualized in @fig-initiators.
+All exploratory analyses will be conducted blinded towards the outcome to avoid influencing study design and analytic choices.
 
-The distribution of patient characteristics, stratified by exposure status, will be examined in Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT.
+he distribution of patient characteristics, stratified by exposure status, will be examined in Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT.
 Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, eposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting.
@@
 
-Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power than RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, statistical power is a critical to assess the feasibility that a comparison of agreement metrics of estimated parameters between RCT and RWD with sufficient precision can be made.
+Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power than RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results.
 Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
-To assess if the unstratified number of events is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic; @schoenfeld1983sample; @gsDesign]
+To assess if the unstratified number of events is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.\[@schoenfeld1981asymp\]
 
 <!--# https://github.com/keaven/gsDesign/blob/master/R/nEvents.R#L22C12-L22C78 -->
+
+## Agreement metrics {#sec-agreement-metrics}
+
+To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: regulatory agreement, estimate agreement and agreement based on the standardized mean difference (SMD).
+Examples are illustrated @tbl-metrics.
+
+**Regulatory agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.
+
+**Estimate agreement**: agreement that the estimated RWE treatment treatment effect is within the 95% CI of the RCT treatment effect estimate.
+Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no regulatory agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).
+
+**SMD agreement**: quantification of the agreement between the emulated RWE and RCT treatment effect estimate. The SMD is calculated as
+
+$$
+SMD = \frac{\hat{\theta}_{RCT} - \hat{\theta}_{RWE}}{\sqrt{\text{Var}(\hat{\theta}_{RCT}) + \text{Var}(\hat{\theta}_{RWE})}}
+$$
+
+where $\hat{\theta}$ are the treatment effect estimates (hazard ratios or median survival times) and $\text{Var}(\hat{\theta})$ are the corresponding variances.
+The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart.
+For an $\alpha$-level of 0.05, the null hypothesis of no difference would be rejected whenever $|Z| > 1.96$.
+
+For the secondary endpoints of interest (e.g., median survival time or survival probabilities) only the quantitative SMD agreement metric will be applicable.
 
@@
 
-For the estimation of marginal treatment effects, we will employ propensity score analytics to adjust for measured confounding between treatment arms.
+For the estimation of marginal treatment effects, we will employ propensity score methods to adjust for measured confounding between treatment arms.
 The selection of important prognostic covariates will be based on expert clinical knowledge and published literature on prognostic scores in oncology.[@becker2020enhanced] The implementation of propensity scores in combination with multiple imputation will follow the '*within*' methodology as described by Leyrat et al.[@leyrat2019propensity; @pishgar2020matchthem] That is, propensity score matching or weighting will be applied to each imputed dataset.
@@
 
-To asses the balance of pre-exposure covariates after matching or weighting on each imputed dataset, the average standardized mean difference (SMD) and corresponding minimum and maximum SMD range will be visualized (see example in @fig-balance).
-Covariate balance is typically considered at a SMD \< 0.1.[@austin2009balance] Further, we will compute the average post-matching or post-weighting C-statistics.[@franklin2014metrics] In addition, we will use a published prognostic score fro OS[@becker2020enhanced] as a balance measure to visually assess if the prognostic score is balanced between treatment arms after propensity score matching or weighting as this approach was described to show the highest correlations with bias compared with other balance measures and not affected by model misspecification.[@stuart2013prognostic]
+To asses the balance of pre-exposure covariates after matching or weighting on each imputed dataset, the average SMD and corresponding minimum and maximum SMD range will be visualized (see example in @fig-balance).
+Covariate balance is typically considered at a SMD \< 0.1.[@austin2009balance] Further, we will compute the average post-matching or post-weighting C-statistics.[@franklin2014metrics] In addition, we will use a published prognostic score for OS[@becker2020enhanced] as a balance measure to visually assess if the prognostic score is balanced between treatment arms after propensity score matching or weighting as this approach was described to show the highest correlations with bias compared with other balance measures and not affected by model misspecification.[@stuart2013prognostic]
 
@@
 
-## Agreement metrics {#sec-agreement-metrics}
-
-To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive two qualitative agreement metrics (regulatory agreement, estimate agreement) and one quantitative agreement based on the SMD (examples are illustrated @tbl-metrics).
-
-**Regulatory agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.
-This agreement metric mimics the question if RWE would have lead to the same regulatory conclusion as the trial, based on standard regulatory decision rules.
-
-**Estimate agreement**: agreement that the estimated RWE treatment treatment effect is within the 95% CI of the RCT treatment effect estimate.
-Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no regulatory agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).
-
-**SMD agreement**: quantification of the agreement between the emulated RWE and RCT treatment effect estimate. The SMD is calculated as
-
-$$
-SMD = \frac{\hat{\theta}_{RCT} - \hat{\theta}_{RWE}}{\sqrt{\text{Var}(\hat{\theta}_{RCT}) + \text{Var}(\hat{\theta}_{RWE})}}
-$$
-
-where $\hat{\theta}$ are the treatment effect estimates (hazard ratios or median survival times) and $\text{Var}(\hat{\theta})$ are the corresponding variances.
-The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart.
-For an $\alpha$-level of 0.05, the null hypothesis of no difference would be rejected whenever $|Z| > 1.96$.
-
-For the secondary endpoints of interest (e.g., median survival time or survival probabilities) only the SMD agreement metric will be applicable.
-
 {{< pagebreak >}}
@@
 
-Building off an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to assess the potential use of RWE for regulatory science.
+Building off an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to inform the potential use of RWE for regulatory science in oncology.
 The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations.
@@
 As a result, exposure misclassification (due to too narrow ascertainment windows) and selection bias (due to too long ascertainment windows) are common trade-offs in such scenarios.
-Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design[@gaber2024mystifying], may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters and provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.
+Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design[@gaber2024mystifying], may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters, provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.
 
@@
 RWE based on fit-for-purpose data and principled, well-designed and reproducible studies may complement evidence coming from RCTs.
-The ENCORE project will systematically assess and eventually increase our understanding as to which measurement, design and analytic decisions influence bias and validity in oncological RWE studies by calibrating those against RCTs.
+Through a systematic benchmarking approach, the ENCORE project will provide insights as to how measurement, design and analytic decisions influence bias and validity in oncological RWE studies.
 
@@
 :::
-
-{{< pagebreak >}}
 
@@
 #| fig-cap: "Visualization of descriptive drug utilization analyses displaying a) initiation trends between compared regimens based on calendar time, b) cumulative rate of patients switching to another line of treatment."
-#| fig-widh: 4
-#| fig-height: 7
+#| fig-widh: 3
+#| fig-height: 6
 

### 2025-01-07 manuscript/manuscript.qmd
formatting

@@
 Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
-To assess if the unstratified number of events is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.\[@schoenfeld1981asymp\]
+To assess if the unstratified number of events is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymp]
 

### 2025-01-07 manuscript/manuscript.qmd
both fixed and random effects?

@@
 For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results.
-If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects models.
+If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed and random effects models.
 

### 2025-01-07 manuscript/manuscript.qmd

@@
 
-For the secondary endpoints of interest (e.g., median survival time or survival probabilities) only the quantitative SMD agreement metric will be applicable.
+For the secondary endpoints of interest (e.g., median survival time or survival probabilities) only the SMD agreement metric will be applicable.
 

### 2025-01-07 manuscript/manuscript.qmd
corrected broken reference

@@
 Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
-To assess if the unstratified number of events is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymp]
+To assess if the unstratified number of events is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]
 

### 2025-01-08 manuscript/manuscript.qmd
round 2 SVW comments

--- 2025-01-07 manuscript/manuscript.qmd
@@
 Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
-To assess if the unstratified number of events is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymp]
+To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]
 
@@
 
-**Regulatory agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.
+Statistical significance agreement: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.
 
@@
 To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed and internal R package `encore.io` with parameterized functions.
-A detailed documentation can be found in the Supplementary Material.
+Detailed documentation can be found in the Supplementary Material.
 

### 2025-01-08 manuscript/manuscript.qmd
formatting minor

@@
 
-Statistical significance agreement: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.
+**Statistical significance agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.
 

### 2025-01-09 manuscript/manuscript.qmd
added GH and corrected agreement metrics table label

@@
 
-**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, Priyanka Anand^1^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
+**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
 
@@
     trial = "Trial",
-    regulatory = md("Regulatory <br> agreement"),
+    regulatory = md("Statistical <br> significance <br> agreement"),
     estimate = md("Estimate <br> agreement"),

### 2025-02-12 manuscript/manuscript.qmd
added pagebreak
implemented GAH comments

@@
 
-**Authors**: Janick Weberpals^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, ..., Sebastian Schneeweiss^1^, Shirley V. Wang^1^
+**Authors**: Janick Weberpals^1^, Kenneth L.
+Kehl^2^, Donna R.
+Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^4^, Janet Espirito^5^, Nicholas J Robert^5^, Andrew J.
+Belli^6^, Sebastian Schneeweiss^1^, Shirley V.
+Wang^1^
 
@@
 ^3^ Oncology Center of Excellence, US Food and Drug Administration, Silver Spring, MD, USA
+
+^4^ ConcertAI, Cambridge, MA, USA
+
+^5^ Ontada, Boston, MA, USA
+
+^6^ COTA, Inc., New York, NY, USA
 
@@
 Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
-With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures..
+With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
 
@@
 
-The *Emulation of Comparative Oncology Trials with Real-world Evidence* (**ENCORE**) project [@encoreFDA] aims to extend this work to the field of oncology.
+The *Emulation of Comparative Oncology Trials with Real-world Evidence* (ENCORE) project [@encoreFDA] aims to extend this work to the field of oncology.
 Studies in oncology come with their own unique set of challenges which must be systematically explored and understood.
@@
 For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results.
-If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed and random effects models.
+If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]
 
@@
 According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step will outline tables that describe how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database.
-As there are general eligibility criteria in oncology trials which either won't be possible to emulate (e.g., physician-assessed survival prognosis of xy months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.
+As there are general eligibility criteria in oncology trials which either will not be possible to emulate (e.g., physician-assessed survival prognosis of xy months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.
 
@@
 
-he distribution of patient characteristics, stratified by exposure status, will be examined in Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT.
-Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, eposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting.
+The distribution of patient characteristics, stratified by exposure status, will be examined in Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT.
+Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, exposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting.
 At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design and analytic choices based on known outcome information.
@@
 Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
-To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power $\beta$ will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]
+To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]
 
@@
 
-where $\hat{\theta}$ are the treatment effect estimates (hazard ratios or median survival times) and $\text{Var}(\hat{\theta})$ are the corresponding variances.
+where $\hat{\theta}_{RCT}$ and $\hat{\theta}_{RWE}$ are the treatment effect estimates (hazard ratios or median survival times) and $\text{Var}(\hat{\theta}_{RCT})$ and $\text{Var}(\hat{\theta}_{RWE})$ are the corresponding variances for RCT and RWE, respectively.
 The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart.
@@
 
-Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically don't follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given time during follow-up.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability.
+Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically do not follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given time during follow-up.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability.
 The median survival time can be finally determined by extracting the time point during follow-up at which the survival probability drops below 0.5 for the first time.
@@
 
-To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed and internal R package `encore.io` with parameterized functions.
+To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed an internal R package `encore.io` with parameterized functions.
 Detailed documentation can be found in the Supplementary Material.
@@
 
-Building off an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to inform the potential use of RWE for regulatory science in oncology.
+Building on an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to inform the potential use of RWE for regulatory science in oncology.
 The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations.
@@
 One particular aspect that we may not always be able emulate is the exact distribution of patient characteristics of the trial population.
-This can be due to the lack of data granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide, which we won't be able to mirror given that all considered databases reflect the US only.
+This can be due to the lack of data granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide, which we will not be able to mirror given that all considered databases reflect the US only.
 This may be a critical factor especially given that the pathophysiology, prognosis and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries.
@@
 :::
+
+{{< pagebreak >}}
 
@@
 #| label: fig-balance
-#| fig-cap: "Assessment of a) covariate bbalance and b) distributional balance of a prognostic score for overall survival before and after propensity score matching or weighting across multiple imputed datasets."
+#| fig-cap: "Assessment of a) covariate balance and b) distributional balance of a prognostic score for overall survival before and after propensity score matching or weighting across multiple imputed datasets."
 #| fig-widh: 4

### 2025-02-13 manuscript/manuscript.qmd
COTA edits

@@
 Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^4^, Janet Espirito^5^, Nicholas J Robert^5^, Andrew J.
-Belli^6^, Sebastian Schneeweiss^1^, Shirley V.
+Belli^6^, Ching-Kun Wang^6^, Sebastian Schneeweiss^1^, Shirley V.
 Wang^1^
@@
 Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
-With the advent of the the 21^st^ Century Cures Act directive[@RWEFDA], the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
+With the advent of the 21^st^ Century Cures Act directive,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
 
-However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity [@Merola2022; @joshua2022longitudinal], rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria and prognostic factors in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations[@rider2024emulations; @merola2023aetion; @merola2024calibrating], a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.
+However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria, prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.
 
@@
 Studies in oncology come with their own unique set of challenges which must be systematically explored and understood.
-Building on a process co-developed with the FDA through RCT DUPLICATE[@wang2023emulation], this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources.
+Building on a process co-developed with the FDA through RCT DUPLICATE,[@wang2023emulation] this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources.
 The process will emphasize transparency and include documented data fitness assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.
@@
 This criterion involved extensive human review.
-The critical points considered in this step include an initial feasibility assessment of the data fitness[@rivera2022friends], including assessment of whether critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary sample size counts are reasonable.
+The critical points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary sample size counts are reasonable.
 Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.
@@
 
-The ENCORE project will use data from a four oncology-specific electronic health records (EHR)-derived specialty oncology data sources: ConcertAI, COTA, Flatiron Health, McKesson/Ontada.
+The ENCORE project will use data from four oncology-specific electronic health records (EHR)-derived data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, McKesson/Ontada.
 All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology.
 A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol.
-For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results.
+For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (the numbering does not coincide with the above order of mention of the databases).
 If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]
@@
 
-For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] which has been recommended for regulatory submissions of RWE studies[@guidelinegeneral], and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel.
+For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] which has been recommended for regulatory submissions of RWE studies,[@guidelinegeneral] and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel.
 Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics).
@@
 
-To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: regulatory agreement, estimate agreement and agreement based on the standardized mean difference (SMD).
+To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD).
 Examples are illustrated @tbl-metrics.
@@
 
-**Estimate agreement**: agreement that the estimated RWE treatment treatment effect is within the 95% CI of the RCT treatment effect estimate.
+**Estimate agreement**: agreement that the estimated RWE treatment effect is within the 95% CI of the RCT treatment effect estimate.
 Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no regulatory agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).
@@
 These missing data investigations will empirically assess assumptions on potentially underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that was developed as part of a FDA Sentinel Innovation Center causal inference workstream that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] In brief, these diagnostic cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different.
-Together with expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams[@moreno2018canonical], this will inform decisions regarding the in- or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]
+Together with expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams,[@moreno2018canonical] this will inform decisions regarding the in- or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]
 
@@
 
-Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically do not follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given time during follow-up.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability.
+Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according to the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically do not follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given time during follow-up.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability.
 The median survival time can be finally determined by extracting the time point during follow-up at which the survival probability drops below 0.5 for the first time.
@@
 This can comprise decisions on the considered databases, calendar time period, covariate measurement (e.g., measurement windows or trade-offs on sensitivity versus specificity in measurements), approaches to missing data, selection of covariates for imputation and propensity score models and decisions on when to censor patients.
-All sensitivity will be pre-specified in the study protocol and reported using appropriate visualizations such as forest plots.
+All sensitivity analyses will be pre-specified in the study protocol and reported using appropriate visualizations such as forest plots.
 
@@
 Due to intercurrent events, such as non-adherence or the crossover of a high proportion of patients from the control to the intervention arm, trial emulation are prone to be biased towards the null.
-Although this is also a common challenge in the analysis of RCTs[@rufibach2018], treatment protocols in routine care are often observed to be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
+Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment protocols in routine care are often observed to be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
 
@@
 As a result, exposure misclassification (due to too narrow ascertainment windows) and selection bias (due to too long ascertainment windows) are common trade-offs in such scenarios.
-Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design[@gaber2024mystifying], may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters, provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.
+Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design,[@gaber2024mystifying] may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters, provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.
 

### 2025-02-16 manuscript/manuscript.qmd
changed phrasing to estimand differences

@@
 Another common challenge in the emulation of oncology trials is the estimation of an "intention-to-treat" (ITT) analogous estimand which is usually the primary estimand reported in oncology RCTs.
-Due to intercurrent events, such as non-adherence or the crossover of a high proportion of patients from the control to the intervention arm, trial emulation are prone to be biased towards the null.
-Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment protocols in routine care are often observed to be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
+Due to intercurrent events, such as non-adherence, crossover of a high proportion of patients from the control to the intervention arm, or differences in subsequent therapy lines there may be estimand differences between trial and emulation.[@manitz2022estimands] Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment protocols in routine care are often observed to be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
 

### 2025-02-16 manuscript/manuscript.qmd
added ConcertAI comments

@@
 
-However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] rapid changes in guideline treatment patterns and the inability to measure and emulate common eligibility criteria, prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.
+However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.
 
@@
 This decision was in order to allow for enough data and follow-up time accrual in the databases used for this project.
-Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] Given this, and the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
+Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] In addition, the timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
 Therefore, we include only trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.

### 2025-02-16 manuscript/manuscript.qmd
added Ontada comments

@@
 Dr.
+Kehl has received research funding from Meta, Inc.
+to his institution.
+Drs.
+Espirito and Robert are employees of McKesson and own McKesson stock.
+Dr.
 Wang has consulted ad hoc for Exponent Inc.
@@
 In this process paper, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to oncology.
-This expansion will emulate 12 RCTs in multiple EHR databases across four different cancer indications, including non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
+This expansion will emulate 12 RCTs in multiple EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
 The objectives of this project are to develop state-of-the-art methodological approaches and to apply them in a demonstration project designed to create insights for guiding the use of RWE in oncology.
@@
 Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
-With the advent of the 21^st^ Century Cures Act directive,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
+With the advent of the 21^st^ Century Cures Act directive,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023.[@senior2024fresh] RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology, where potential use cases include the assessment of effectiveness in rare patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment to a randomized trial may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
 
-However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.
+However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors.
+Frequently referenced limitations include missing data, small sample sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.
 
@@
 To that end, the emphasis of the project is on trials of therapies for the most common cancers for which there has been substantial therapeutic development in recent years.
-After careful review and exchange with clinical and regulatory experts, four cancer indications were identified including lung cancer, breast cancer, colorectal cancer and multiple myeloma.
-For each cancer type we aim to conduct three trial emulations using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).
+After careful review and exchange with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma.
+For each cancer type, we aim to conduct three trial emulations using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).
 
 We used a semi-automated process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs for each cancer type.
-The search was conducted using the AACT database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains all information (protocol and result data elements) about every study registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *MeSH* term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.
+The search was conducted using the *Aggregated Analysis of ClinicalTrials.gov* (AACT) database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains all information (protocol and result data elements) about every study registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *Medical Subject Headings **(***MeSH) term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.
 
@@
 
--   Feasibility and clinical relevance
+-   Feasibility and clinical relevance (latter was defined as treatment or paradigm-changing trials or such that challenged existing treatment policies)
 
@@
 A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
-Naturally, the majority of trials cover advanced or metastatic cancer populations because a large proportion of drug development efforts have focused on these settings in recent years.
-A key learning that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD.
+Naturally, the majority of trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts start in these settings.
+A key objective that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD.
 If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the sample size is not sufficient, then runner-up candidates will be considered instead.
@@
 
-The ENCORE project will use data from four oncology-specific electronic health records (EHR)-derived data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, McKesson/Ontada.
-All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology.
+The ENCORE project will use data from four oncology-specific electronic health records (EHR)-derived data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, and Ontada/McKesson.
+All available databases draw from a comprehensive national sample of patients with cancer in the US with the detailed EHR-based data necessary to study medication effectiveness in oncology.
 A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol.
@@
 
-For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] which has been recommended for regulatory submissions of RWE studies,[@guidelinegeneral] and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel.
+For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] recommended for regulatory submissions of RWE studies[@guidelinegeneral] and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel.
 Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics).
@@
 Real-world data fitness and emulation feasibility for each shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care.
-Since a vast majority of oncological RCTs in recent years have focused on selected, biomarker-defined populations, subtleties in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population.
-For example, immunotherapies have significantly changed the cancer treatment landscape since the approval of the first PD-L1 inhibitor in 2015.
-With many trials that have followed thereafter, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has also evolved since then and PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.
+Since many oncological RCTs in recent years have focused on selected, biomarker-defined populations, subtleties in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population.
+For example, immune checkpoint inhibitors have significantly changed the cancer treatment landscape since the first approvals of the CTLA-4 inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and nivolumab in 2014 in the United States.
+With many trials that have followed thereafter, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has also evolved, and PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.
 
 According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step will outline tables that describe how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database.
-As there are general eligibility criteria in oncology trials which either will not be possible to emulate (e.g., physician-assessed survival prognosis of xy months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.
+As there are general eligibility criteria in oncology trials which either will not be possible to emulate (e.g., physician-assessed survival prognosis of x months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.
 
@@
 There will be special emphasis on how exposure, in context of their respective disease and line of therapy settings, and the OS outcome will be emulated.
-For all considered databases, the OS endpoint is typically a composite that is derived from different sources comprising EHR abstractions, social security death index, obituary and other linkages.
+For all considered databases, the OS endpoint is typically a composite that is derived from different sources comprising EHR documentation, social security death index, obituary and other linkages.
 Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open]
@@
 
-Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power than RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results.
+Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power as RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results.
 Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
@@
 
-To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD).
+To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR for OS and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD).
 Examples are illustrated @tbl-metrics.
@@
 **Estimate agreement**: agreement that the estimated RWE treatment effect is within the 95% CI of the RCT treatment effect estimate.
-Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no regulatory agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).
+Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no statistical significance agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).
 
@@
 
-where $\hat{\theta}_{RCT}$ and $\hat{\theta}_{RWE}$ are the treatment effect estimates (hazard ratios or median survival times) and $\text{Var}(\hat{\theta}_{RCT})$ and $\text{Var}(\hat{\theta}_{RWE})$ are the corresponding variances for RCT and RWE, respectively.
+where $\hat{\theta}_{RCT}$ and $\hat{\theta}_{RWE}$ are the treatment effect estimates (hazard ratios or median survival time differences) and $\text{Var}(\hat{\theta}_{RCT})$ and $\text{Var}(\hat{\theta}_{RWE})$ are the corresponding variances for RCT and RWE, respectively.
 The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart.
@@
 To establish an analytic cohort, key eligibility criteria will be applied in which patients with missing values in eligibility criteria are considered eligible in the respective attrition steps to allow for thorough missing data investigations.
-These missing data investigations will empirically assess assumptions on potentially underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that was developed as part of a FDA Sentinel Innovation Center causal inference workstream that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] In brief, these diagnostic cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different.
+These missing data investigations will empirically assess assumptions on potential underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that was developed as part of a FDA Sentinel Innovation Center causal inference workstream that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] In brief, these diagnostics cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different.
 Together with expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams,[@moreno2018canonical] this will inform decisions regarding the in- or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]
@@
 Historically, administrative health claims databases have been the backbone of most research in RWE.
-With increasing access to EHR and a maturing set of methodological approaches to draw causal inferences from such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to deliver similar causal conclusions compared to RCTs in the field of oncology.
+With increasing access to EHR data and a maturing set of methodological approaches to draw causal inferences from such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to deliver similar causal conclusions compared to RCTs in the field of oncology.
 
 The RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges which we expect to also encounter in ENCORE.
-One particular aspect that we may not always be able emulate is the exact distribution of patient characteristics of the trial population.
+One particular aspect that we may not always be able to emulate is the exact distribution of patient characteristics of the trial population.
 This can be due to the lack of data granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide, which we will not be able to mirror given that all considered databases reflect the US only.
 This may be a critical factor especially given that the pathophysiology, prognosis and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries.
-For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations) are much more prevalent in Asian countries compared to the US or Europe.
+For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are much more prevalent in Asian countries compared to the US or Europe.
 
@@
 This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD.
-As a result, exposure misclassification (due to too narrow ascertainment windows) and selection bias (due to too long ascertainment windows) are common trade-offs in such scenarios.
+As a result, exposure misclassification (due to overly ascertainment windows) and selection bias (due to overly long ascertainment windows) are common trade-offs in such scenarios.
 Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design,[@gaber2024mystifying] may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters, provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.
@@
 #| label: fig-initiators
-#| fig-cap: "Visualization of descriptive drug utilization analyses displaying a) initiation trends between compared regimens based on calendar time, b) cumulative rate of patients switching to another line of treatment."
+#| fig-cap: "Example visualization of descriptive drug utilization analyses displaying a) initiation trends between compared regimens based on calendar time, b) cumulative rate of patients switching to another line of treatment."
 #| fig-widh: 3

### 2025-02-26 manuscript/manuscript.qmd
added RJG as co-author
minor edits to author list and affiliations

@@
 Kehl^2^, Donna R.
-Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^4^, Janet Espirito^5^, Nicholas J Robert^5^, Andrew J.
-Belli^6^, Ching-Kun Wang^6^, Sebastian Schneeweiss^1^, Shirley V.
+Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^4^, Andrew J.
+Belli^5^, Ching-Kun Wang^5^, Jenna Collins^6^**,** JonathanKish^6^**,** Janet Espirito^7^, Nicholas J Robert^7^, Robert J.
+Glynn^1^ ,Sebastian Schneeweiss^1^, Shirley V.
 Wang^1^
@@
 
-^5^ Ontada, Boston, MA, USA
+^5^ COTA, Inc., New York, NY, USA
 
-^6^ COTA, Inc., New York, NY, USA
+^6^ Flatiron Health, Inc., New York, NY, USA
+
+^7^ Ontada, Boston, MA, USA
 

### 2025-02-26 manuscript/manuscript.qmd
prelim render to send to RJG for review

@@
     tbl-cap-location: top
+    extra_dependencies: ["float"]
+    keep-tex: true
+    fig-pos: 'h'
+    tbl-pos: 'h'
 
@@
 #| tbl-cap-location: top
+#| fig-pos: 'h'
+#| tbl-pos: 'h'
 
@@
 #| tbl-cap: "Visualization of agreement metrics example."
+#| fig-pos: 'h'
+#| tbl-pos: 'h'
 
@@
 #| fig-width: 12
+#| fig-pos: 'h'
+#| tbl-pos: 'h'
 
@@
 
+::: {.content-visible unless-format="pdf"}
 [View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/manuscript/Figure_1_process.png)
+:::
 
@@
 #| fig-height: 6
+#| fig-pos: 'h'
+#| tbl-pos: 'h'
 
@@
 
+::: {.content-visible unless-format="pdf"}
 [View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/manuscript/Figure_2_utlization.png)
+:::
 
@@
 #| fig-height: 7
+#| fig-pos: 'h'
+#| tbl-pos: 'h'
 
@@
 
+::: {.content-visible unless-format="pdf"}
 [View figure in higher resolution here](https://github.com/janickweberpals/encore-process-manuscript/blob/main/manuscript/Figure_3_balance.png)
+:::

### 2025-02-26 manuscript/manuscript.tex
prelim render to send to RJG for review

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{luacolor}
  \usepackage[soul]{lua-ul}
\else
  \usepackage{soul}

\fi
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading,figureheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{A Process for the Emulation of Comparative Oncology Trials with
Real-world Evidence (ENCORE)}
\author{}
\date{}

\begin{document}
\maketitle

\textbf{Authors}: Janick Weberpals\textsuperscript{1}, Kenneth L.
Kehl\textsuperscript{2}, Donna R. Rivera\textsuperscript{3}, Pallavi
Mishra-Kalyani\textsuperscript{3}, Georg Hahn\textsuperscript{1},
Priyanka Anand\textsuperscript{1}, Yanina Natanzon\textsuperscript{4},
Andrew J. Belli\textsuperscript{5}, Ching-Kun Wang\textsuperscript{5},
Jenna Collins\textsuperscript{6}\textbf{,}
JonathanKish\textsuperscript{6}\textbf{,} Janet
Espirito\textsuperscript{7}, Nicholas J Robert\textsuperscript{7},
Robert J. Glynn\textsuperscript{1} ,Sebastian
Schneeweiss\textsuperscript{1}, Shirley V. Wang\textsuperscript{1}

\ul{Author affiliations:}

\textsuperscript{1} Division of Pharmacoepidemiology and
Pharmacoeconomics, Department of Medicine, Brigham and Women's Hospital,
Harvard Medical School, Boston, MA, USA

\textsuperscript{2} Dana-Farber Cancer Institute, Boston, MA, USA

\textsuperscript{3} Oncology Center of Excellence, US Food and Drug
Administration, Silver Spring, MD, USA

\textsuperscript{4} ConcertAI, Cambridge, MA, USA

\textsuperscript{5} COTA, Inc., New York, NY, USA

\textsuperscript{6} Flatiron Health, Inc., New York, NY, USA

\textsuperscript{7} Ontada, Boston, MA, USA

\ul{\textbf{Correspondence:}}

Shirley V. Wang, PhD

Division of Pharmacoepidemiology and Pharmacoeconomics,

Department of Medicine, Brigham and Women's Hospital, Harvard Medical
School,

1620 Tremont Street, Suite 3030-R, Boston, MA 02120, USA

Phone: +1 617-278-0932

Fax: + 1 617-232-8602

Email: \url{SWANG1@BWH.HARVARD.EDU}

\ul{\textbf{Article type:}} Review

\ul{\textbf{Manuscript word count:}} 4,131 words / 8,000 words

\ul{\textbf{Abstract word count:}} 248 words / 250 words

\ul{\textbf{Tables:}} 3

\ul{\textbf{Figures:}} 3

\ul{\textbf{Supplementary material:}} Supplementary figures and material

\ul{\textbf{Short running title}}: Emulation of Comparative Oncology
Trials with Real-world Evidence (ENCORE)

\ul{\textbf{Keywords:}} Oncology, Real-World Evidence, Trial emulation,
EHR

\ul{\textbf{Funding Statement:}} This project was supported by an FDA
BAA contract with contract number 75F40122C00181.

\ul{\textbf{Competing Interests Statement:}} Dr.~Weberpals is now an
employee of AstraZeneca and owns stocks in AstraZeneca. Dr.~Kehl has
received research funding from Meta, Inc.~to his institution. Drs.
Espirito and Robert are employees of McKesson and own McKesson stock.
Dr.~Wang has consulted ad hoc for Exponent Inc.~and MITRE a federally
funded research center for the Centers for Medicare and Medicaid
Services on unrelated work. Dr.~Schneeweiss is participating in
investigator-initiated grants to the Brigham and Women's Hospital from
Boehringer Ingelheim, Takeda, and UCB unrelated to the topic of this
study. He owns equity in Aetion Inc., a software manufacturer. He is an
advisor to Temedica GmbH, a patient-oriented data generation company.
His interests were declared, reviewed, and approved by the Brigham and
Women's Hospital in accordance with their institutional compliance
policies.

\ul{\textbf{Data sharing statement:}} No data was analyzed as part of
this project.

\ul{\textbf{Analytic code sharing statement:}} Code to reproduce this
manuscript, including all figures and tables, can be found at
\url{https://github.com/janickweberpals/encore-process-manuscript}.

\ul{\textbf{\emph{Proposed target journals:}}} \emph{CPT =\textgreater{}
JCO CCI =\textgreater{} \ldots{}}

\emph{Manuscript last updated: 2025-02-26 21:44:25.38931}

\newpage{}

\section*{Abstract}\label{abstract}
\addcontentsline{toc}{section}{Abstract}

Real-world evidence (RWE) studies are increasingly used to complement
evidence from randomized controlled trials (RCTs), contextualizing the
effectiveness and safety of medical interventions as delivered in
clinical practice. Advancements in the curation and accessibility of
electronic health record data (EHR) have presented the opportunity to
investigate disease domains such as oncology, where administrative
healthcare claims databases alone are not fit-for-purpose. The RCT
DUPLICATE initiative has previously enhanced understanding of when RWE
studies can draw causal conclusions by emulating trials in the
cardio-metabolic, renal, and pulmonary clinical areas. In this process
paper, we present the Emulation of Comparative Oncology Trials with
Real-world Evidence (ENCORE) project, which aims to extend this work to
oncology. This expansion will emulate 12 RCTs in multiple EHR databases
across four different cancer indications, specifically non-small cell
lung cancer, breast cancer, colorectal cancer, and multiple myeloma. The
objectives of this project are to develop state-of-the-art
methodological approaches and to apply them in a demonstration project
designed to create insights for guiding the use of RWE in oncology. The
project will place a special emphasis on systematic evaluation of
fitness of data in relation to the study design and statistical analysis
for a particular research question, and pre-registration of study
protocols prior to analysis. Agreement of treatment effect estimates
between RCTs and their respective emulations will be assessed using
pre-specified criteria. Through extensive sensitivity analyses
benchmarked against RCT results, the ENCORE project will inform
understanding of how measurement, design, and analytic decisions
influence bias and validity in oncological RWE studies.

\newpage{}

\section{Background}\label{background}

Randomized controlled trials (RCTs) have been the gold standard for
establishing the efficacy and safety of medical products. With the
advent of the 21\textsuperscript{st} Century Cures Act
directive,\textsuperscript{1} the Food and Drug Administration (FDA)
established a framework to increasingly consider real-world evidence
(RWE) generated from routine-care health data such as electronic health
records (EHR) to evaluate and contextualize the comparative safety and
effectiveness of novel cancer therapies.\textsuperscript{2} Accounting
for 21\% of all approvals, oncology was the disease area with the most
FDA drug approvals in 2023.\textsuperscript{3} RWE has particularly
important potential to complement evidence coming from RCTs in the field
of precision oncology, where potential use cases include the assessment
of effectiveness in rare patient populations that are underrepresented
in RCTs, the construction of external control arms in single-arm trials
where active recruitment to a randomized trial may not be feasible, or
the discovery of biomarkers among pan-tumor populations that harbor
specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between
RWE studies and RCTs can depend on many factors. Frequently referenced
limitations include missing data, small sample sizes, data
discontinuity,\textsuperscript{4, 5} adoption of changes in guidelines
in real-world care and the inability to measure and emulate common
eligibility criteria, including prognostic factors, and standardized
response assessments in real-world data (RWD).\textsuperscript{6} While
there are already published examples of oncology trial
emulations,\textsuperscript{6--8} a systematic and scaled approach to
emulate a diverse set of different oncology trials with multiple
heterogeneous databases is necessary to gain confidence in the validity
of RWE studies and to provide context as to which questions can be
validly answered.

The RCT DUPLICATE initiative\textsuperscript{9} increased our
understanding of when RWE studies can come to causal conclusions on
treatment effects by benchmarking results against RCTs under the
assumption that each RCT finding reflects a causal treatment effect. In
settings where the RCT designs could be emulated well, RWE studies came
to the same conclusions.\textsuperscript{10} However, prior work from
RCT-DUPLICATE has focused primarily on emulating trials in the
cardio-metabolic, renal, and pulmonary clinical areas using claims
databases.

The \emph{Emulation of Comparative Oncology Trials with Real-world
Evidence} (ENCORE) project\textsuperscript{11} aims to extend this work
to the field of oncology. Studies in oncology come with their own unique
set of challenges which must be systematically explored and understood.
Building on a process co-developed with the FDA through RCT
DUPLICATE,\textsuperscript{9} this expansion to oncology will emulate 12
randomized clinical trials using multiple specialty oncology EHR data
sources. The process will emphasize transparency and include documented
data fitness assessment\textsuperscript{12, 13} for each RWD source with
respect to each trial emulation as well as extensive sensitivity
analyses to assess robustness of findings.

The objectives of this project are to 1) develop state-of-the-art
methodological approaches and 2) to apply them to create insights that
may provide guidance on the potential use of RWE for regulatory science
in oncology. To achieve these objectives, this demonstration project
will systematically emulate 12 oncology trials across four cancers and
assess the agreement of treatment effect estimates between RCTs and
their respective emulations.

In this process paper, we describe the design and process for the
selection of the 12 oncology RCTs, the assessment of the database
quality and selection, protocol development and pre-registration, study
design and statistical analysis, and pre-specified agreement metrics to
evaluate the concordance between RCTs and emulations.

\section{Methods}\label{methods}

A visual summary of the entire systematic process from trial selection
to final results is provided in Figure~\ref{fig-process}.

\subsection{Trial selection}\label{trial-selection}

The focus of ENCORE is to maximize potential learnings on when RWE
studies can or cannot yield similar results compared to RCTs. To that
end, the emphasis of the project is on trials of therapies for the most
common cancers for which there has been substantial therapeutic
development in recent years. After careful review and exchange with
clinical and regulatory experts, four cancer indications were
identified: non-small cell lung cancer, breast cancer, colorectal cancer
and multiple myeloma. For each cancer type, we aim to conduct three
trial emulations using multiple databases accessible for the scope of
this project (i.e., the total number of emulations will equal 12 trials
x \emph{n} databases which are found fit-for-purpose for each trial).

We used a semi-automated process for trial selection where the
eligibility criteria are documented in CONSORT diagrams showing reasons
for excluding RCTs for each cancer type. The search was conducted using
the \emph{Aggregated Analysis of ClinicalTrials.gov} (AACT) database
which is a publicly available relational database developed and
maintained by the Clinical Trials Transformation Initiative (CTTI) that
contains all information (protocol and result data elements) about every
study registered on ClinicalTrials.gov.\textsuperscript{14} To identify
eligible trials, we used a combined search query strategy of the
National Library of Medicine (NLM)-controlled \emph{Medical Subject
Headings \textbf{(}}MeSH) term and a free keyword search for the
respective cancer indication in the \emph{conditions}, \emph{studies}
and \emph{detailed\_descriptions} fields of each trial entry on
ClinicalTrials.gov.

Eligible trials had to fulfill the following basic criteria:

\begin{itemize}
\item
  Interventional
\item
  Randomized
\item
  Intervention model: parallel assignment
\item
  Industry-sponsored
\item
  Trial start in 2011 or later
\item
  Primary purpose was to study treatment effects
\item
  Overall survival must be one of the endpoints reported (either as
  hazard ratio or median overall survival time)
\item
  Recruitment status: `Completed' or `Active, not recruiting'
\item
  Feasibility and clinical relevance (latter was defined as treatment or
  paradigm-changing trials or such that challenged existing treatment
  policies)
\end{itemize}

The rationale and operationalization of each criterion is listed in
detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
interventional, randomized trials after 2011 because treatment
guidelines among included cancer indications have undergone significant
changes in recent years. Due to the rapid adoption of new breakthrough
therapies in routine care, it is unlikely to find patients who may be
still treated with outdated treatment regimens in current clinical
practice. We also did not include trials with results published too
recently. This decision was in order to allow for enough data and
follow-up time accrual in the databases used for this project. Although
there have been substantial methodological advancements to increase our
understanding on the emulation and comparison of real-world
progression-free survival (PFS) and objective response rates (ORR) to a
RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
level of granularity which may not be necessarily reflected in
chart-abstracted assessments of a patient's progression in routine
care.\textsuperscript{18, 19} In addition, the timing and cadence of
intervals between progression assessments can differ between RCTs and
routine care which may result in measurement error and
bias.\textsuperscript{20} Given this, and the large number of other
methodological challenges like missing data, small sample sizes, data
discontinuity and rapidly changing guideline treatments, we focus on the
emulation of overall survival (OS) as the endpoint of interest.
Therefore, we include only trials that have reported overall survival
(OS) as one of the pre-specified endpoints in the protocol.

While most trial eligibility criteria could be operationalized in an
automated fashion, our final criterion was an assessment of emulation
feasibility and clinical relevance. This criterion involved extensive
human review. The critical points considered in this step include an
initial feasibility assessment of the data fitness,\textsuperscript{19}
including assessment of whether critical eligibility criteria (e.g.,
biomarker status) and prognostic factors (e.g., ECOG performance score)
are measurable and whether preliminary sample size counts are
reasonable. Lastly, trial candidates were ranked and shortlisted into
primary and runner-up candidates based on their clinical and regulatory
relevance.

A list of tentative, shortlisted primary candidates is presented in
Table~\ref{tbl-rcts} and the corresponding selection process is
illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
Naturally, the majority of trials cover advanced (locally advanced,
inoperable, recurrent/progressive disease) or metastatic cancer
populations because a large proportion of drug development efforts start
in these settings. A key objective that we aim to foster with the
shortlisted trials is to achieve a better understanding how different
disease settings (early, late), line settings ({[}neo{]}adjuvant, first
line, advanced lines of therapy), therapy protocols (monotherapy,
combination therapy) and population characteristics (simple versus
complex genetic or immunological signatures) can be emulated using RWD.
If more thorough feasibility assessments suggest that the threat of bias
from mis-measurement of key study parameters or residual confounding
remains high, or that the sample size is not sufficient, then runner-up
candidates will be considered instead.

\subsection{Databases}\label{databases}

The ENCORE project will use data from four oncology-specific electronic
health records (EHR)-derived data sources (in alphabetical order):
ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. All available
databases draw from a comprehensive national sample of patients with
cancer in the US with the detailed EHR-based data necessary to study
medication effectiveness in oncology. A detailed description and
methodology on how patients are sampled will be provided with each trial
emulation protocol. For ENCORE, not all databases will be available for
each cancer indication and the names of the databases will be blinded
and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final
reporting of results (the numbering does not coincide with the above
order of mention of the databases). If more than one database is
considered fit-for-purpose for a respective trial emulation, the best
possible analytic model will be employed for each database separately
and final treatment effect estimates will be pooled using a
meta-analytic approach using fixed effects (weights reflect the inverse
of the variance of each database study) and random effects (weights
reflect the inverse of the variance of each database study plus an
additional variance term that quantifies the assumed heterogeneity
between databases) models.\textsuperscript{21}

\subsection{Protocol development}\label{protocol-development}

For each shortlisted and selected RCT, a detailed protocol,
pre-specifying key elements of the trial emulation, will be developed
using the HARPER protocol template\textsuperscript{22} recommended for
regulatory submissions of RWE studies\textsuperscript{23} and will be
registered on ClinicalTrials.gov after careful review by a clinical and
FDA regulatory expert panel. Following the target trial emulation
framework, we will provide an explicit statement and rationale on how
each element will be emulated including database selection, covariate
measurement, operationalization of key eligibility criteria, study
design, data analysis and causal contrasts of
interest.\textsuperscript{24, 25} Since it is common that oncology RCTs
update OS estimates periodically based on accrued follow-up time, the
protocol will give a brief summary of each emulated RCT and specify
which target OS estimates will be used to compare agreement metrics to
(see Section~\ref{sec-agreement-metrics}). All eligibility criteria will
be extracted based on publications, publicly available protocols and
statistical analysis plans of the selected RCT.

\subsubsection{Emulation feasibility}\label{emulation-feasibility}

\paragraph{\texorpdfstring{\textbf{Fit-for-purpose
data}}{Fit-for-purpose data}}\label{fit-for-purpose-data}

Real-world data fitness and emulation feasibility for each shortlisted
candidate trial will be assessed in multiple steps based on guidance of
the oncology quality, characterization, and assessment of real-world
data (Oncology QCARD) Initiative.\textsuperscript{19} The first step
assesses if relevant variables like exposure/line of therapy, outcomes,
and covariates are generally available, measured and operationalizable
in routine-care. Since many oncological RCTs in recent years have
focused on selected, biomarker-defined populations, subtleties in
measurement and operationalizability of specific biomarkers must be
reflected to ensure a representative and large enough study population.
For example, immune checkpoint inhibitors have significantly changed the
cancer treatment landscape since the first approvals of the CTLA-4
inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and
nivolumab in 2014 in the United States. With many trials that have
followed thereafter, the operationalization of the expression of the
PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion
score or combined positive score) has also evolved, and PD-L1
'\emph{positivity'} may have different definitions across calendar years
based on different cut-off values.

According to the \emph{Structured Process to Identify Fit-For-Purpose
Data} (SPIFD) framework\textsuperscript{13}, the next step will outline
tables that describe how eligibility criteria will be ascertained using
a color-coded heatmap that will indicate the level of confidence on how
well each criterion can be emulated in each selected database. As there
are general eligibility criteria in oncology trials which either will
not be possible to emulate (e.g., physician-assessed survival prognosis
of x months) or that are clinically not relevant for the emulation of
the trial (e.g., male patients should be willing to use barrier
contraception), the study team will decide on key eligibility criteria
for the emulation of the trial.

We will additionally provide a definition on how exposures, outcomes and
covariates are exactly defined and operationalized in each respective
database. There will be special emphasis on how exposure, in context of
their respective disease and line of therapy settings, and the OS
outcome will be emulated. For all considered databases, the OS endpoint
is typically a composite that is derived from different sources
comprising EHR documentation, social security death index, obituary and
other linkages. Given that not all relevant sources that provide
mortality data are synchronized and updated uniformly, sensitivity
analyses with more conservative (i.e., earlier) censoring dates will be
considered for each trial emulation to mitigate the potential impact of
ghost-time bias.\textsuperscript{26}

\paragraph{\texorpdfstring{\textbf{Descriptives and data
exploration}}{Descriptives and data exploration}}\label{descriptives-and-data-exploration}

Critical aspects when emulating oncology trials are the choice and
estimation of the appropriate estimand of interest.\textsuperscript{27}
Particularly when emulating pivotal trials of paradigm-changing
treatments, multiple aspects need to be considered such as the
contemporaneity of the (historical) control cohort, the adoption rate of
the novel intervention in routine care, the magnitude of the clinical
treatment benefit and the rate in which (particularly patients in the
control arm) discontinue or cross-over to the interventional treatment,
which could lastly bias emulated treatment effects towards the null. To
that end, comprehensive data explorations will be performed as part of
the protocol development to contextualize these parameters and (if
reported) draw comparisons to the emulated trial. Examples for such
standard diagnostics are visualized in Figure~\ref{fig-initiators}. All
exploratory analyses will be conducted blinded towards the outcome to
avoid influencing study design and analytic choices.

The distribution of patient characteristics, stratified by exposure
status, will be examined in Table 1's before and after applying
eligibility criteria and contrasted with the distributions of patient
characteristics of the original RCT. Initial propensity score matching
or weighting methods will be applied to ensure that measured
pre-exposure covariates can be balanced, exposure cohorts are
conditionally exchangeable at baseline and resulting sample sizes are
still sufficient after matching or weighting. At this stage, all
exploratory analyses will be conducted blinded towards the outcome to
not bias any study design and analytic choices based on known outcome
information.

\paragraph{\texorpdfstring{\textbf{Statistical
power}}{Statistical power}}\label{statistical-power}

Causal analyses of observational data may not have the same
pre-requisites in terms of formal hypothesis testing and statistical
power as RCTs since the number of `recruited' patients is given and
cannot be influenced.\textsuperscript{28} For this project, however, the
emulations in EHR data must have at least equal power to the relevant
trial for interpretation of the pre-specified agreements metrics between
RCT and RWD results. Since the main outcome of interest is defined as
time to all-cause mortality (OS), the estimation of the statistical
power is driven by the number of events rather than the number of
patients. To assess if the overall number of events, unstratified by
exposure, is sufficient such that a significant difference can be
detected based on the original RCT-reported hazard ratio (HR), the
statistical power (1-\(\beta\)) will be estimated using Schoenfeld's
sample-size formula for the proportional-hazards regression
model.\textsuperscript{29}

\subsection{Agreement metrics}\label{sec-agreement-metrics}

To formally compare treatment effects between RCTs and their respective
emulations, we will adapt the approach of the RCT-DUPLICATE
project.\textsuperscript{9, 30} That is, for the primary endpoint of
interest (HR for OS and corresponding 95\% confidence intervals), we
will derive three qualitative agreement metrics: statistical agreement,
estimate agreement and agreement based on the standardized mean
difference (SMD). Examples are illustrated Table~\ref{tbl-metrics}.

\textbf{Statistical significance agreement}: agreement between RCT and
emulated trial treatment effect with regards to directionality and
statistical significance.

\textbf{Estimate agreement}: agreement that the estimated RWE treatment
effect is within the 95\% CI of the RCT treatment effect estimate.
Provided that for some emulations, the power of the RWE study may be
larger than that of the original RCT, this could lead to situations
where there is no statistical significance agreement although the
treatment effect estimates are highly overlapping but with the RCT
estimate crossing the null (or vice versa in case the RCT has a larger
power than the RWE emulation).

\textbf{SMD agreement}: quantification of the agreement between the
emulated RWE and RCT treatment effect estimate. The SMD is calculated as

\[
SMD = \frac{\hat{\theta}_{RCT} - \hat{\theta}_{RWE}}{\sqrt{\text{Var}(\hat{\theta}_{RCT}) + \text{Var}(\hat{\theta}_{RWE})}}
\]

where \(\hat{\theta}_{RCT}\) and \(\hat{\theta}_{RWE}\) are the
treatment effect estimates (hazard ratios or median survival time
differences) and \(\text{Var}(\hat{\theta}_{RCT})\) and
\(\text{Var}(\hat{\theta}_{RWE})\) are the corresponding variances for
RCT and RWE, respectively. The resulting SMDs will be interpreted such
that with an SMD of 1.00, the effect estimate from the RCT and the RWE
emulation are 1 standard deviation apart. For an \(\alpha\)-level of
0.05, the null hypothesis of no difference would be rejected whenever
\(|Z| > 1.96\).

For the secondary endpoints of interest (e.g., median survival time or
survival probabilities) only the SMD agreement metric will be
applicable.

\subsection{Study design and statistical
analysis}\label{study-design-and-statistical-analysis}

The study design for each trial emulation will be visualized as part of
the protocol using a graphical depiction of the exact measurement
windows of eligibility criteria, washout periods and covariates relative
to the cohort entry time.\textsuperscript{31}

\subsubsection{Missing data}\label{missing-data}

To establish an analytic cohort, key eligibility criteria will be
applied in which patients with missing values in eligibility criteria
are considered eligible in the respective attrition steps to allow for
thorough missing data investigations. These missing data investigations
will empirically assess assumptions on potential underlying missingness
mechanisms according to Rubin's classification of missing data (i.e.,
missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
missing not at random {[}MNAR{]}).\textsuperscript{32} To that end, we
will adopt a principled process on missing data that was developed as
part of a FDA Sentinel Innovation Center causal inference workstream
that empirically evaluates different aspects across partially observed
covariates based on three group diagnostics.\textsuperscript{33, 34} In
brief, these diagnostics cover (1) comparisons of patients
characteristics with and without an observed level of the partially
observed covariate, (2) ability to predict missingness given observed
data, and (3) assessments if outcomes between patients with a missing
value are systematically different. Together with expert domain
knowledge and assumptions about the underlying missing data structure
through canonical causal diagrams,\textsuperscript{35} this will inform
decisions regarding the in- or exclusion of patients with missing values
in key eligibility criteria and potential sensitivity analyses to assess
the robustness of these decisions.\textsuperscript{36}

While the MAR assumption is a strong assumption to hold across all
considered covariates, it was shown that especially in the context of
partially observed covariate data (as opposed to missing exposure and
outcome data), only mechanisms in which a covariate causes its own
missingness leads to critical bias (MNAR).\textsuperscript{35} Hence,
methodologies which retain patients and give the potential to adjust for
a broader set of prognostic factors (e.g., multiple
imputation\textsuperscript{37} or doubly robust
methods\textsuperscript{38}) may be preferred over complete case
analyses.

\subsubsection{Endpoints and propensity score
analyses}\label{endpoints-and-propensity-score-analyses}

Due to its ubiquity in oncology trials, the primary parameter of
interest in ENCORE will be defined as the marginal hazard ratio (HR)
coefficient for the treatment comparison for time to all-cause mortality
(OS).\textsuperscript{39} However, the HR has many limitations that can
make the comparison of results across varying follow-up times between
RCT and emulations challenging, including its non-collapsibility,
in-built selection bias and the requirement of proportional
hazards.\textsuperscript{40, 41} For this reason, we will also consider
alternative endpoints on an absolute risk scale such as median survival
times, survival probabilities at pre-defined time points during
follow-up\textsuperscript{42} and restricted mean survival times as
secondary endpoints of interest.

For the estimation of marginal treatment effects, we will employ
propensity score methods to adjust for measured confounding between
treatment arms. The selection of important prognostic covariates will be
based on expert clinical knowledge and published literature on
prognostic scores in oncology.\textsuperscript{43} The implementation of
propensity scores in combination with multiple imputation will follow
the `\emph{within}' methodology as described by Leyrat et
al.\textsuperscript{44, 45} That is, propensity score matching or
weighting will be applied to each imputed dataset. The marginal
treatment effect will then be estimated in each imputed and matched or
weighted dataset separately and pooled into a final estimate following
Rubin's rule.\textsuperscript{46, 47} This approach has been shown to
lead to unbiased estimates across different simulated scenarios with a
sufficient estimation of the variance.\textsuperscript{44}

To asses the balance of pre-exposure covariates after matching or
weighting on each imputed dataset, the average SMD and corresponding
minimum and maximum SMD range will be visualized (see example in
Figure~\ref{fig-balance}). Covariate balance is typically considered at
a SMD \textless{} 0.1.\textsuperscript{48} Further, we will compute the
average post-matching or post-weighting
C-statistics.\textsuperscript{49} In addition, we will use a published
prognostic score for OS\textsuperscript{43} as a balance measure to
visually assess if the prognostic score is balanced between treatment
arms after propensity score matching or weighting as this approach was
described to show the highest correlations with bias compared with other
balance measures and not affected by model
misspecification.\textsuperscript{50}

Similarly, survival probabilities for individual time points will be
estimated in each imputed and propensity score matched or weighted
dataset according to the Kaplan-Meier method.\textsuperscript{42} Since
survival probabilities typically do not follow normal distributions
which are required to apply Rubin's rule, these will be transformed
through a complementary log-log transformation \(log(-log(1-pr(surv)))\)
with \(pr(surv)\) denoting the survival probability at a given time
during follow-up.\textsuperscript{51, 52} The transformed survival
probabilities are then pooled across imputed datasets and individual
time points following Rubin's rule and back-transformed via
\(1-exp(-exp(qbar))\) with \(qbar\) denoting the pooled survival
probability. The median survival time can be finally determined by
extracting the time point during follow-up at which the survival
probability drops below 0.5 for the first time.

\subsubsection{Sensitivity analyses}\label{sensitivity-analyses}

Given that a large number of factors could influence a difference
between RCT results and emulated trial results, a broad range of
sensitivity analyses will be conducted. This can comprise decisions on
the considered databases, calendar time period, covariate measurement
(e.g., measurement windows or trade-offs on sensitivity versus
specificity in measurements), approaches to missing data, selection of
covariates for imputation and propensity score models and decisions on
when to censor patients. All sensitivity analyses will be pre-specified
in the study protocol and reported using appropriate visualizations such
as forest plots.

\subsubsection{Reproducibility and consistency across
emulation}\label{reproducibility-and-consistency-across-emulation}

To ensure a transparent, reproducible and consistent way of deriving
analytic cohorts and performing statistical analyses across trial
emulations, we developed an internal R package \texttt{encore.io} with
parameterized functions. Detailed documentation can be found in the
Supplementary Material.

\newpage{}

\section{Discussion}\label{discussion}

Building on an established approach based on the RCT DUPLICATE
project\textsuperscript{9, 30}, the ENCORE project aims to emulate 12
oncology trials using multiple EHR data sources to inform the potential
use of RWE for regulatory science in oncology. The project focuses on
four common cancers and will evaluate the agreement of treatment effect
estimates between RCTs and their respective emulations. Historically,
administrative health claims databases have been the backbone of most
research in RWE. With increasing access to EHR data and a maturing set
of methodological approaches to draw causal inferences from such data,
the ENCORE project is one of the first of its kind to evaluate when and
how RWD can be used to deliver similar causal conclusions compared to
RCTs in the field of oncology.

The RCT DUPLICATE project\textsuperscript{9, 30} identified multiple
emulation challenges which we expect to also encounter in ENCORE. One
particular aspect that we may not always be able to emulate is the exact
distribution of patient characteristics of the trial population. This
can be due to the lack of data granularity and comprehensiveness to
emulate relevant eligibility criteria or the fact that large pivotal
trials in oncology are typically conducted in multiple countries
worldwide, which we will not be able to mirror given that all considered
databases reflect the US only. This may be a critical factor especially
given that the pathophysiology, prognosis and factors that drive
heterogeneous treatment outcomes of certain cancers differ between
countries. For example, some cancer indications (e.g., GI cancers) or
some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are
much more prevalent in Asian countries compared to the US or Europe.

Another common challenge in the emulation of oncology trials is the
estimation of an ``intention-to-treat'' (ITT) analogous estimand which
is usually the primary estimand reported in oncology RCTs. Due to
intercurrent events, such as non-adherence, crossover of a high
proportion of patients from the control to the intervention arm, or
differences in subsequent therapy lines there may be estimand
differences between trial and emulation.\textsuperscript{53} Although
this is also a common challenge in the analysis of
RCTs,\textsuperscript{27} treatment protocols in routine care are often
observed to be less stringent compared to trials.\textsuperscript{8} In
order to derive comparable estimands it is therefore crucial to
understand and contextualize the proportion and timing of treatment
switching and discontinuation in both the trial and its emulation.

This issue may be even augmented with more complex treatment protocols
like combination regimens (e.g., CheckMate9LA) as compared to
monotherapies (e.g., CheckMate057) since the ascertainment of the
exposure needs to happen over a pre-defined time window which is often
difficult to calibrate in RWD. As a result, exposure misclassification
(due to overly ascertainment windows) and selection bias (due to overly
long ascertainment windows) are common trade-offs in such scenarios.
Alternative analytic approaches which target a per-protocol estimand,
such as the clone-censor-weight design,\textsuperscript{54} may be
viable options if the ITT estimand cannot be estimated due to the
aforementioned parameters, provided that necessary covariate
measurements are available to account for the artificial censoring
introduced with these methods.

\subsection{Conclusions}\label{conclusions}

RWE based on fit-for-purpose data and principled, well-designed and
reproducible studies may complement evidence coming from RCTs. Through a
systematic benchmarking approach, the ENCORE project will provide
insights as to how measurement, design and analytic decisions influence
bias and validity in oncological RWE studies.

\newpage{}

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{0}{1}
\bibitem[\citeproctext]{ref-RWEFDA}
\textbf{1}. Framework for FDA's real-world evidence program (last
accessed 11/28/2024) {[}Internet{]}, 2018Available from:
\url{https://www.fda.gov/media/120060/download?attachment}

\bibitem[\citeproctext]{ref-purpura2022role}
\textbf{2}. Purpura CA, Garry EM, Honig N, et al: The role of real-world
evidence in FDA-approved new drug and biologics license applications.
Clinical Pharmacology \& Therapeutics 111:135--144, 2022

\bibitem[\citeproctext]{ref-senior2024fresh}
\textbf{3}. Senior M: Fresh from the biotech pipeline: Record-breaking
FDA approvals. Nature Biotechnology, 2024

\bibitem[\citeproctext]{ref-Merola2022}
\textbf{4}. Merola D, Schneeweiss S, Schrag D, et al: An algorithm to
predict data completeness in oncology electronic medical records for
comparative effectiveness research {[}Internet{]}. Annals of
Epidemiology 76:143--149, 2022Available from:
\url{http://dx.doi.org/10.1016/j.annepidem.2022.07.007}

\bibitem[\citeproctext]{ref-joshua2022longitudinal}
\textbf{5}. Joshua Lin K, Jin Y, Gagne J, et al: Longitudinal data
discontinuity in electronic health records and consequences for
medication effectiveness studies. Clinical Pharmacology \& Therapeutics
111:243--251, 2022

\bibitem[\citeproctext]{ref-rider2024emulations}
\textbf{6}. Rider JR, Wasserman A, Slipski L, et al: Emulations of
oncology trials using real-world data: A systematic literature review.
American journal of epidemiology kwae346, 2024

\bibitem[\citeproctext]{ref-merola2023aetion}
\textbf{7}. Merola D, Campbell U, Gautam N, et al: The aetion coalition
to advance real-world evidence through randomized controlled trial
emulation initiative: oncology. Clinical Pharmacology \& Therapeutics
113:1217--1222, 2023

\bibitem[\citeproctext]{ref-merola2024calibrating}
\textbf{8}. Merola D, Campbell U, Lenis D, et al: Calibrating
observational health record data against a randomized trial. JAMA
Network Open 7:e2436535--e2436535, 2024

\bibitem[\citeproctext]{ref-wang2023emulation}
\textbf{9}. Wang SV, Schneeweiss S, Franklin JM, et al:
\href{https://doi.org/10.1001/jama.2023.4221}{Emulation of randomized
clinical trials with nonrandomized database analyses: Results of 32
clinical trials}. Jama 329:1376--1385, 2023

\bibitem[\citeproctext]{ref-heyard2024design}
\textbf{10}. Heyard R, Held L, Schneeweiss S, et al: Design differences
and variation in results between randomised trials and non-randomised
emulations: Meta-analysis of RCT-DUPLICATE data {[}Internet{]}. BMJ
medicine 3, 2024Available from:
\url{https://doi.org/10.1136/bmjmed-2023-000709}

\bibitem[\citeproctext]{ref-encoreFDA}
\textbf{11}. Calibrating real-world evidence studies in oncology against
randomized trials: ENCORE (last accessed 11/28/2024) {[}Internet{]},
2024Available from:
\url{https://www.fda.gov/about-fda/oncology-center-excellence/calibrating-real-world-evidence-studies-oncology-against-randomized-trials-encore}

\bibitem[\citeproctext]{ref-rivera2024oncology}
\textbf{12}. Rivera DR, Eckert JC, Rodriguez-Watson C, et al: The
oncology QCARD initiative: Fostering efficient evaluation of initial
real-world data proposals. Pharmacoepidemiology and Drug Safety
33:e5818, 2024

\bibitem[\citeproctext]{ref-gatto2022structured}
\textbf{13}. Gatto NM, Campbell UB, Rubinstein E, et al: The structured
process to identify fit-for-purpose data: A data feasibility assessment
framework. Clinical Pharmacology \& Therapeutics 111:122--134, 2022

\bibitem[\citeproctext]{ref-tasneem2012database}
\textbf{14}. Tasneem A, Aberle L, Ananth H, et al: The database for
aggregate analysis of ClinicalTrials. Gov (AACT) and subsequent
regrouping by clinical specialty. PloS one 7:e33677, 2012

\bibitem[\citeproctext]{ref-eisenhauer2009new}
\textbf{15}. Eisenhauer EA, Therasse P, Bogaerts J, et al: New response
evaluation criteria in solid tumours: Revised RECIST guideline (version
1.1). European journal of cancer 45:228--247, 2009

\bibitem[\citeproctext]{ref-ThanCCR}
\textbf{16}. Ton TGN, Pal N, Trinh H, et al: Replication of overall
survival, progression-free survival, and overall response in
chemotherapy arms of nonâsmall cell lung cancer trials using real-world
data {[}Internet{]}. Clinical Cancer Research 28:2844--2853,
2022Available from: \url{https://doi.org/10.1158/1078-0432.CCR-22-0471}

\bibitem[\citeproctext]{ref-mckelvey2024evaluation}
\textbf{17}. McKelvey BA, Garrett-Mayer E, Rivera DR, et al: Evaluation
of real-world tumor response derived from electronic health record data
sources: A feasibility analysis in patients with metastatic non--small
cell lung cancer treated with chemotherapy. JCO Clinical Cancer
Informatics 8:e2400091, 2024

\bibitem[\citeproctext]{ref-rwdRECIST}
\textbf{18}. Chen L, Davis R, Lee J, et al: Comparison of response from
RECIST1.1 and abstraction in real-world patients with lung cancer.
{[}Internet{]}. Journal of Clinical Oncology 41:e21194--e21194,
2023Available from:
\url{https://ascopubs.org/doi/abs/10.1200/JCO.2023.41.16_suppl.e21194}

\bibitem[\citeproctext]{ref-rivera2022friends}
\textbf{19}. Rivera DR, Henk HJ, Garrett-Mayer E, et al: The friends of
cancer research real-world data collaboration pilot 2.0: Methodological
recommendations from oncology case studies. Clinical Pharmacology \&
Therapeutics 111:283--292, 2022

\bibitem[\citeproctext]{ref-ackerman2024measurement}
\textbf{20}. Ackerman B, Gan RW, Meyer CS, et al: Measurement error and
bias in real-world oncology endpoints when constructing external control
arms. Frontiers in Drug Safety and Regulation 4:1423493, 2024

\bibitem[\citeproctext]{ref-nikolakopoulou2014interpret}
\textbf{21}. Nikolakopoulou A, Mavridis D, Salanti G: How to interpret
meta-analysis models: Fixed effect and random effects meta-analyses. BMJ
Ment Health 17:64--64, 2014

\bibitem[\citeproctext]{ref-wang2022harmonized}
\textbf{22}. Wang SV, Pottegård A, Crown W, et al: HARmonized protocol
template to enhance reproducibility of hypothesis evaluating real-world
evidence studies on treatment effects: A good practices report of a
joint ISPE/ISPOR task force {[}Internet{]}. Value in Health
25:1663--1672, 2022Available from:
\url{https://doi.org/10.1016/j.jval.2022.09.001}

\bibitem[\citeproctext]{ref-guidelinegeneral}
\textbf{23}. ICH guideline. General principles on plan, design and
analysis of pharmacoepidemiological studies that utilize real-world data
for safety assessment of medicines M14. Available from:
Https://database.ich.org/sites/default/files/ICH\_M14\_Step3\_DraftGuideline\_2024\_0521.pdf
(last accessed 01/07/2025)

\bibitem[\citeproctext]{ref-hernan2022target}
\textbf{24}. Hernán MA, Wang W, Leaf DE: Target trial emulation: A
framework for causal inference from observational data. Jama
328:2446--2447, 2022

\bibitem[\citeproctext]{ref-hernan2016specifying}
\textbf{25}. Hernán MA, Sauer BC, Hernández-Dı́az S, et al:
\href{https://doi.org/10.1016/j.jclinepi.2016.04.014}{Specifying a
target trial prevents immortal time bias and other self-inflicted
injuries in observational analyses}. Journal of clinical epidemiology
79:70--75, 2016

\bibitem[\citeproctext]{ref-meyer2020open}
\textbf{26}. Meyer A-M, Davies J, Taylor M, et al: Open cohorts and
ghost-time bias in real world data, in PHARMACOEPIDEMIOLOGY AND DRUG
SAFETY. WILEY 111 RIVER ST, HOBOKEN 07030-5774, NJ USA, 2020, pp
426--426

\bibitem[\citeproctext]{ref-rufibach2018}
\textbf{27}. Rufibach K: Treatment effect quantification for
time{-}to{-}event endpoints{\textendash}Estimands, analysis strategies,
and beyond {[}Internet{]}. Pharmaceutical Statistics 18:145--165,
2018Available from: \url{http://dx.doi.org/10.1002/pst.1917}

\bibitem[\citeproctext]{ref-hernan2022causal}
\textbf{28}. Hernán MA: Causal analyses of existing databases: No power
calculations required. Journal of clinical epidemiology 144:203--205,
2022

\bibitem[\citeproctext]{ref-schoenfeld1981asymptotic}
\textbf{29}. Schoenfeld D: The asymptotic properties of nonparametric
tests for comparing survival distributions. Biometrika 68:316--319, 1981

\bibitem[\citeproctext]{ref-franklin2020nonrandomized}
\textbf{30}. Franklin JM, Pawar A, Martin D, et al: Nonrandomized
real-world evidence to support regulatory decision making: Process for a
randomized trial replication project. Clinical Pharmacology \&
Therapeutics 107:817--826, 2020

\bibitem[\citeproctext]{ref-schneeweiss2019graphical}
\textbf{31}. Schneeweiss S, Rassen JA, Brown JS, et al: Graphical
depiction of longitudinal study designs in health care databases. Annals
of internal medicine 170:398--406, 2019

\bibitem[\citeproctext]{ref-rubin1976inference}
\textbf{32}. Rubin DB: Inference and missing data. Biometrika
63:581--592, 1976

\bibitem[\citeproctext]{ref-weberpals2024smdi}
\textbf{33}. Weberpals J, Raman SR, Shaw PA, et al: Smdi: An r package
to perform structural missing data investigations on partially observed
confounders in real-world evidence studies. JAMIA open 7:ooae008, 2024

\bibitem[\citeproctext]{ref-weberpals2024}
\textbf{34}. Weberpals J, Raman SR, Shaw PA, et al: A principled
approach to characterize and analyze partially observed confounder data
from electronic health records {[}Internet{]}. Clinical Epidemiology
16:329--343, 2024Available
from:\href{\%20\%0A\%20\%20\%20\%20\%0A\%20\%20\%20\%20\%0A\%20\%20\%20\%20\%20\%20\%20\%20https://www.tandfonline.com/doi/abs/10.2147/CLEP.S436131\%0A\%20\%20\%20\%20\%0A\%0A}{
https://www.tandfonline.com/doi/abs/10.2147/CLEP.S436131 }

\bibitem[\citeproctext]{ref-moreno2018canonical}
\textbf{35}. Moreno-Betancur M, Lee KJ, Leacy FP, et al: Canonical
causal diagrams to guide the treatment of missing data in epidemiologic
studies. American journal of epidemiology 187:2705--2715, 2018

\bibitem[\citeproctext]{ref-tompsett2018use}
\textbf{36}. Tompsett DM, Leacy F, Moreno-Betancur M, et al: On the use
of the not-at-random fully conditional specification (NARFCS) procedure
in practice. Statistics in medicine 37:2338--2353, 2018

\bibitem[\citeproctext]{ref-weberpals2024hdmi}
\textbf{37}. Weberpals J, Shaw PA, Lin KJ, et al: High-dimensional
multiple imputation (HDMI) for partially observed confounders including
natural language processing-derived auxiliary covariates {[}Internet{]}.
American Journal of Epidemiology, 2025Available from:
\url{https://arxiv.org/abs/2405.10925}

\bibitem[\citeproctext]{ref-Shaw2024}
\textbf{38}. Shaw CK P., Williamson BD: Assessing treatment effects in
observational data with missing confounders: A comparative study of
practical doubly-robust and traditional missing data methods
{[}Internet{]}. GitHub repository, 2024Available from:
\url{https://github.com/PamelaShaw/Missing-Confounders-Methods}

\bibitem[\citeproctext]{ref-cox1972regression}
\textbf{39}. Cox DR: Regression models and life-tables. Journal of the
royal statistical society. Series B (Methodological) 34:187--220, 1972

\bibitem[\citeproctext]{ref-hernan2010hazards}
\textbf{40}. Hernán MA: The hazards of hazard ratios. Epidemiology
21:13--15, 2010

\bibitem[\citeproctext]{ref-stensrud2020test}
\textbf{41}. Stensrud MJ, Hernán MA:
\href{https://doi.org/10.1001/jama.2020.1267}{Why test for proportional
hazards?} Jama 323:1401--1402, 2020

\bibitem[\citeproctext]{ref-kaplan1958nonparametric}
\textbf{42}. Kaplan EL, Meier P: Nonparametric estimation from
incomplete observations. Journal of the American statistical association
53:457--481, 1958

\bibitem[\citeproctext]{ref-becker2020enhanced}
\textbf{43}. Becker T, Weberpals J, Jegg A, et al: An enhanced
prognostic score for overall survival of patients with cancer derived
from a large real-world cohort. Annals of Oncology 31:1561--1568, 2020

\bibitem[\citeproctext]{ref-leyrat2019propensity}
\textbf{44}. Leyrat C, Seaman SR, White IR, et al: Propensity score
analysis with partially observed covariates: How should multiple
imputation be used? Statistical methods in medical research 28:3--19,
2019

\bibitem[\citeproctext]{ref-pishgar2020matchthem}
\textbf{45}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
and weighting after multiple imputation. arXiv preprint arXiv:200911772,
2020

\bibitem[\citeproctext]{ref-rubin2018multiple}
\textbf{46}. Rubin DB: Multiple imputation, in Flexible imputation of
missing data, second edition. Chapman; Hall/CRC, 2018, pp 29--62

\bibitem[\citeproctext]{ref-van2011mice}
\textbf{47}. Van Buuren S, Groothuis-Oudshoorn K: Mice: Multivariate
imputation by chained equations in r. Journal of statistical software
45:1--67, 2011

\bibitem[\citeproctext]{ref-austin2009balance}
\textbf{48}. Austin PC: Balance diagnostics for comparing the
distribution of baseline covariates between treatment groups in
propensity-score matched samples. Statistics in medicine 28:3083--3107,
2009

\bibitem[\citeproctext]{ref-franklin2014metrics}
\textbf{49}. Franklin JM, Rassen JA, Ackermann D, et al: Metrics for
covariate balance in cohort studies of causal effects. Statistics in
medicine 33:1685--1699, 2014

\bibitem[\citeproctext]{ref-stuart2013prognostic}
\textbf{50}. Stuart EA, Lee BK, Leacy FP: Prognostic score--based
balance measures can be a useful diagnostic for propensity score methods
in comparative effectiveness research. Journal of clinical epidemiology
66:S84--S90, 2013

\bibitem[\citeproctext]{ref-warwick28685}
\textbf{51}. Marshall A(Andrea), Billingham LJ, Bryan S: Can we afford
to ignore missing data in cost-effectiveness analyses? {[}Internet{]}.
European Journal of Health Economics Vol.10:1--3, 2009Available from:
\url{http://dx.doi.org/10.1007/s10198-008-0129-y}

\bibitem[\citeproctext]{ref-morisot2015prostate}
\textbf{52}. Morisot A, Bessaoud F, Landais P, et al: Prostate cancer:
Net survival and cause-specific survival rates after multiple
imputation. BMC medical research methodology 15:1--14, 2015

\bibitem[\citeproctext]{ref-manitz2022estimands}
\textbf{53}. Manitz J, Kan-Dobrosky N, Buchner H, et al: Estimands for
overall survival in clinical trials with treatment switching in
oncology. Pharmaceutical Statistics 21:150--162, 2022

\bibitem[\citeproctext]{ref-gaber2024mystifying}
\textbf{54}. Gaber CE, Ghazarian AA, Strassle PD, et al: De-mystifying
the clone-censor-weight method for causal research using observational
data: A primer for cancer researchers. Cancer Medicine 13:e70461, 2024

\end{CSLReferences}

\newpage{}

\section*{Tables}\label{tables}
\addcontentsline{toc}{section}{Tables}

\begin{table}[h]

\caption{\label{tbl-criteria}Criteria to select eligible trials for
emulation in ENCORE.}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{../tables/Table_1_trial_eligibility.png}}

}

\end{table}%

\newpage{}

\begin{table}[h]

\caption{\label{tbl-rcts}Tentative list of randomized controlled trials
(RCTs) considered for emulation.}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{../tables/Table_2_trial_selection.png}}

}

\end{table}%

\newpage{}

\begin{table}[h]

\caption{\label{tbl-metrics}Visualization of agreement metrics example.}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{../tables/Table_3_agreement_metrics.png}}

}

\end{table}%

\newpage{}

\section*{Figures}\label{figures}
\addcontentsline{toc}{section}{Figures}

\begin{figure}[h]

\caption{\label{fig-process}Systematic process to understand
effectiveness claims of oncology trials using real-world evidence.}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{../figures/process.png}}

}

\end{figure}%

\newpage{}

\begin{figure}[h]

\caption{\label{fig-initiators}Example visualization of descriptive drug
utilization analyses displaying a) initiation trends between compared
regimens based on calendar time, b) cumulative rate of patients
switching to another line of treatment.}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{Figure_2_utlization.png}}

}

\end{figure}%

\newpage{}

\begin{figure}[h]

\caption{\label{fig-balance}Assessment of a) covariate balance and b)
distributional balance of a prognostic score for overall survival before
and after propensity score matching or weighting across multiple imputed
datasets.}

\centering{

\pandocbounded{\includegraphics[keepaspectratio]{Figure_3_balance.png}}

}

\end{figure}%

\end{document}

### 2025-02-26 manuscript/manuscript.qmd
decreased size figure 2

@@
 #| fig-cap: "Example visualization of descriptive drug utilization analyses displaying a) initiation trends between compared regimens based on calendar time, b) cumulative rate of patients switching to another line of treatment."
-#| fig-widh: 3
-#| fig-height: 6
+#| fig-widh: 2.5
+#| fig-height: 5
 #| fig-pos: 'h'

### 2025-03-04 manuscript/manuscript.qmd

@@
 ---
-title: "A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)"
+title: "A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies"
 
@@
 Belli^5^, Ching-Kun Wang^5^, Jenna Collins^6^**,** JonathanKish^6^**,** Janet Espirito^7^, Nicholas J Robert^7^, Robert J.
-Glynn^1^ ,Sebastian Schneeweiss^1^, Shirley V.
+Glynn^1^, Sebastian Schneeweiss^1^, Shirley V.
 Wang^1^
@@
 Dr.
-Schneeweiss is participating in investigator-initiated grants to the Brigham and Women's Hospital from Boehringer Ingelheim, Takeda, and UCB unrelated to the topic of this study.
-He owns equity in Aetion Inc., a software manufacturer.
+Schneeweiss is participating in investigator-initiated grants to the Brigham and Women's Hospital from Bayer and UCB unrelated to the topic of this study.
+He consults for and owns equity in Aetion Inc., a software manufacturer.
 He is an advisor to Temedica GmbH, a patient-oriented data generation company.
@@
 Advancements in the curation and accessibility of electronic health record data (EHR) have presented the opportunity to investigate disease domains such as oncology, where administrative healthcare claims databases alone are not fit-for-purpose.
-The RCT DUPLICATE initiative has previously enhanced understanding of when RWE studies can draw causal conclusions by emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas.
-In this process paper, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to oncology.
-This expansion will emulate 12 RCTs in multiple EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
-The objectives of this project are to develop state-of-the-art methodological approaches and to apply them in a demonstration project designed to create insights for guiding the use of RWE in oncology.
-The project will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to analysis.
+The RCT DUPLICATE initiative has previously enhanced understanding of when RWE studies come to the same causal conclusions by emulating trials in non-oncology indications.
+Here, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to oncology.
+ENCORE will emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
+It will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to analysis.
 Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria.
-Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence bias and validity in oncological RWE studies.
+Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence the validity of oncological RWE studies.
 
@@
 
-Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products.
-With the advent of the 21^st^ Century Cures Act directive,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023.[@senior2024fresh] RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology, where potential use cases include the assessment of effectiveness in rare patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment to a randomized trial may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
+Randomized controlled trials (RCTs) have been the go-to methodology for establishing the efficacy and safety of medical products.
+Under the 21^st^ Century Cures Act directive,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023.[@senior2024fresh] RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology, where potential use cases include the assessment of effectiveness in specific patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment to a randomized trial may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
 
-However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors.
-Frequently referenced limitations include missing data, small sample sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.
+However, the validity and transportability of results derived between RWE studies and RCTs depends on multiple factors.
+Frequently referenced limitations include, lack of baseline randomization, missing data, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.
 
 The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
-In settings where the RCT designs could be emulated well, RWE studies came to the same conclusions.[@heyard2024design] However, prior work from RCT-DUPLICATE has focused primarily on emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas using claims databases.
+In settings where the RCT designs could be emulated well, RWE studies came to the same conclusions.[@heyard2024design] However, prior work from RCT-DUPLICATE has focused primarily on emulating trials in non-oncology settings using claims databases.
 
-The *Emulation of Comparative Oncology Trials with Real-world Evidence* (ENCORE) project [@encoreFDA] aims to extend this work to the field of oncology.
+The *Emulation of Comparative Oncology Trials with Real-world Evidence* (ENCORE) project [@encoreFDA] aims to extend this work to oncology.
 Studies in oncology come with their own unique set of challenges which must be systematically explored and understood.
@@
 
-In this process paper, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development and pre-registration, study design and statistical analysis, and pre-specified agreement metrics to evaluate the concordance between RCTs and emulations.
+Here, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development and pre-registration, study design and statistical analysis, and pre-specified agreement metrics to evaluate the concordance between RCTs and emulations.
 
@@
 To that end, the emphasis of the project is on trials of therapies for the most common cancers for which there has been substantial therapeutic development in recent years.
-After careful review and exchange with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma.
+After review and discussion with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma.
 For each cancer type, we aim to conduct three trial emulations using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).
@@
 Due to the rapid adoption of new breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in current clinical practice.
-We also did not include trials with results published too recently.
-This decision was in order to allow for enough data and follow-up time accrual in the databases used for this project.
-Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] In addition, the timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
+Conversely, we excluded trials with results published too recently.
+To allow for enough data and follow-up time accrual in the databases used for this project.
+Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
 Therefore, we include only trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.
@@
 This criterion involved extensive human review.
-The critical points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary sample size counts are reasonable.
+The critical points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary study size counts are reasonable.
 Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.
@@
 A key objective that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD.
-If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the sample size is not sufficient, then runner-up candidates will be considered instead.
+If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidates will be considered instead.
 
@@
 
-For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] recommended for regulatory submissions of RWE studies[@guidelinegeneral] and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel.
+For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] recommended for regulatory submissions of RWE studies[@guidelinegeneral] and will be registered on ClinicalTrials.gov after review by a clinical and FDA regulatory expert panel.
 Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics).
@@
 There will be special emphasis on how exposure, in context of their respective disease and line of therapy settings, and the OS outcome will be emulated.
-For all considered databases, the OS endpoint is typically a composite that is derived from different sources comprising EHR documentation, social security death index, obituary and other linkages.
+For all considered databases, the OS endpoint is typically a composite that, depending on the underlying database, can be derived from different sources comprising EHR documentation, social security death index, obituary and other linkages.
 Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open]
@@
 
-#### **Statistical power**
+#### **Statistical power considerations**
 
@@
 To establish an analytic cohort, key eligibility criteria will be applied in which patients with missing values in eligibility criteria are considered eligible in the respective attrition steps to allow for thorough missing data investigations.
-These missing data investigations will empirically assess assumptions on potential underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that was developed as part of a FDA Sentinel Innovation Center causal inference workstream that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] In brief, these diagnostics cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different.
+These missing data investigations will empirically assess assumptions on potential underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] These diagnostics cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different.
 Together with expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams,[@moreno2018canonical] this will inform decisions regarding the in- or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]
@@
 
-Due to its ubiquity in oncology trials, the primary parameter of interest in ENCORE will be defined as the marginal hazard ratio (HR) coefficient for the treatment comparison for time to all-cause mortality (OS).[@cox1972regression] However, the HR has many limitations that can make the comparison of results across varying follow-up times between RCT and emulations challenging, including its non-collapsibility, in-built selection bias and the requirement of proportional hazards.[@hernan2010hazards; @stensrud2020test] For this reason, we will also consider alternative endpoints on an absolute risk scale such as median survival times, survival probabilities at pre-defined time points during follow-up[@kaplan1958nonparametric] and restricted mean survival times as secondary endpoints of interest.
+Due to its ubiquity in oncology trials, the primary parameter of interest in ENCORE will be defined as the marginal hazard ratio (HR) coefficient for the treatment comparison for time to all-cause mortality (OS).[@cox1972regression] We will also consider alternative endpoints on an absolute risk scale such as median survival times, survival probabilities at pre-defined time points during follow-up[@kaplan1958nonparametric] and restricted mean survival times as secondary endpoints of interest.
 
@@
 
-Given that a large number of factors could influence a difference between RCT results and emulated trial results, a broad range of sensitivity analyses will be conducted.
+With the goal to better understand which factors could influence a difference between RCT results and emulated trial results, a range of sensitivity analyses will be conducted.
 This can comprise decisions on the considered databases, calendar time period, covariate measurement (e.g., measurement windows or trade-offs on sensitivity versus specificity in measurements), approaches to missing data, selection of covariates for imputation and propensity score models and decisions on when to censor patients.
@@
 
-Building on an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to inform the potential use of RWE for regulatory science in oncology.
+Building on an established approach developed by the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using four oncology EHR data sources to inform the potential use of RWE for regulatory science in oncology.
 The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations.
@@
 
-The RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges which we expect to also encounter in ENCORE.
+The RCT DUPLICATE initiative[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges which we expect to also encounter in ENCORE.
 One particular aspect that we may not always be able to emulate is the exact distribution of patient characteristics of the trial population.

### 2025-03-04 manuscript/manuscript.tex
Manuscript updated

@@
 \hypersetup{
-  pdftitle={A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)},
+  pdftitle={A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies},
   colorlinks=true,
@@
 
-\title{A Process for the Emulation of Comparative Oncology Trials with
-Real-world Evidence (ENCORE)}
+\title{A Process for Emulating Comparative Oncology Trials with
+Real-world Evidence Studies}
 \author{}
@@
 Espirito\textsuperscript{7}, Nicholas J Robert\textsuperscript{7},
-Robert J. Glynn\textsuperscript{1} ,Sebastian
+Robert J. Glynn\textsuperscript{1}, Sebastian
 Schneeweiss\textsuperscript{1}, Shirley V. Wang\textsuperscript{1}
@@
 investigator-initiated grants to the Brigham and Women's Hospital from
-Boehringer Ingelheim, Takeda, and UCB unrelated to the topic of this
-study. He owns equity in Aetion Inc., a software manufacturer. He is an
-advisor to Temedica GmbH, a patient-oriented data generation company.
-His interests were declared, reviewed, and approved by the Brigham and
-Women's Hospital in accordance with their institutional compliance
-policies.
+Bayer and UCB unrelated to the topic of this study. He consults for and
+owns equity in Aetion Inc., a software manufacturer. He is an advisor to
+Temedica GmbH, a patient-oriented data generation company. His interests
+were declared, reviewed, and approved by the Brigham and Women's
+Hospital in accordance with their institutional compliance policies.
 
@@
 
-\emph{Manuscript last updated: 2025-02-26 21:44:25.38931}
+\emph{Manuscript last updated: 2025-03-04 18:43:42.372054}
 
@@
 DUPLICATE initiative has previously enhanced understanding of when RWE
-studies can draw causal conclusions by emulating trials in the
-cardio-metabolic, renal, and pulmonary clinical areas. In this process
-paper, we present the Emulation of Comparative Oncology Trials with
-Real-world Evidence (ENCORE) project, which aims to extend this work to
-oncology. This expansion will emulate 12 RCTs in multiple EHR databases
-across four different cancer indications, specifically non-small cell
-lung cancer, breast cancer, colorectal cancer, and multiple myeloma. The
-objectives of this project are to develop state-of-the-art
-methodological approaches and to apply them in a demonstration project
-designed to create insights for guiding the use of RWE in oncology. The
-project will place a special emphasis on systematic evaluation of
-fitness of data in relation to the study design and statistical analysis
-for a particular research question, and pre-registration of study
-protocols prior to analysis. Agreement of treatment effect estimates
-between RCTs and their respective emulations will be assessed using
-pre-specified criteria. Through extensive sensitivity analyses
-benchmarked against RCT results, the ENCORE project will inform
-understanding of how measurement, design, and analytic decisions
-influence bias and validity in oncological RWE studies.
+studies come to the same causal conclusions by emulating trials in
+non-oncology indications. Here, we present the Emulation of Comparative
+Oncology Trials with Real-world Evidence (ENCORE) project, which aims to
+extend this work to oncology. ENCORE will emulate 12 RCTs in four
+oncology-specialized EHR databases across four different cancer
+indications, specifically non-small cell lung cancer, breast cancer,
+colorectal cancer, and multiple myeloma. It will place a special
+emphasis on systematic evaluation of fitness of data in relation to the
+study design and statistical analysis for a particular research
+question, and pre-registration of study protocols prior to analysis.
+Agreement of treatment effect estimates between RCTs and their
+respective emulations will be assessed using pre-specified criteria.
+Through extensive sensitivity analyses benchmarked against RCT results,
+the ENCORE project will inform understanding of how measurement, design,
+and analytic decisions influence the validity of oncological RWE
+studies.
 
@@
 
-Randomized controlled trials (RCTs) have been the gold standard for
-establishing the efficacy and safety of medical products. With the
-advent of the 21\textsuperscript{st} Century Cures Act
-directive,\textsuperscript{1} the Food and Drug Administration (FDA)
-established a framework to increasingly consider real-world evidence
-(RWE) generated from routine-care health data such as electronic health
-records (EHR) to evaluate and contextualize the comparative safety and
-effectiveness of novel cancer therapies.\textsuperscript{2} Accounting
-for 21\% of all approvals, oncology was the disease area with the most
-FDA drug approvals in 2023.\textsuperscript{3} RWE has particularly
-important potential to complement evidence coming from RCTs in the field
-of precision oncology, where potential use cases include the assessment
-of effectiveness in rare patient populations that are underrepresented
+Randomized controlled trials (RCTs) have been the go-to methodology for
+establishing the efficacy and safety of medical products. Under the
+21\textsuperscript{st} Century Cures Act directive,\textsuperscript{1}
+the Food and Drug Administration (FDA) established a framework to
+increasingly consider real-world evidence (RWE) generated from
+routine-care health data such as electronic health records (EHR) to
+evaluate and contextualize the comparative safety and effectiveness of
+novel cancer therapies.\textsuperscript{2} Accounting for 21\% of all
+approvals, oncology was the disease area with the most FDA drug
+approvals in 2023.\textsuperscript{3} RWE has particularly important
+potential to complement evidence coming from RCTs in the field of
+precision oncology, where potential use cases include the assessment of
+effectiveness in specific patient populations that are underrepresented
 in RCTs, the construction of external control arms in single-arm trials
@@
 However, the validity and transportability of results derived between
-RWE studies and RCTs can depend on many factors. Frequently referenced
-limitations include missing data, small sample sizes, data
-discontinuity,\textsuperscript{4, 5} adoption of changes in guidelines
-in real-world care and the inability to measure and emulate common
-eligibility criteria, including prognostic factors, and standardized
-response assessments in real-world data (RWD).\textsuperscript{6} While
-there are already published examples of oncology trial
-emulations,\textsuperscript{6--8} a systematic and scaled approach to
-emulate a diverse set of different oncology trials with multiple
-heterogeneous databases is necessary to gain confidence in the validity
-of RWE studies and to provide context as to which questions can be
-validly answered.
+RWE studies and RCTs depends on multiple factors. Frequently referenced
+limitations include, lack of baseline randomization, missing data, small
+study sizes, data discontinuity,\textsuperscript{4, 5} adoption of
+changes in guidelines in real-world care and the inability to measure
+and emulate common eligibility criteria, including prognostic factors,
+and standardized response assessments in real-world data
+(RWD).\textsuperscript{6} While there are already published examples of
+oncology trial emulations,\textsuperscript{6--8} a systematic and scaled
+approach to emulate a diverse set of different oncology trials with
+multiple heterogeneous databases is necessary to gain confidence in the
+validity of RWE studies and to provide context as to which questions can
+be validly answered.
 
@@
 to the same conclusions.\textsuperscript{10} However, prior work from
-RCT-DUPLICATE has focused primarily on emulating trials in the
-cardio-metabolic, renal, and pulmonary clinical areas using claims
-databases.
+RCT-DUPLICATE has focused primarily on emulating trials in non-oncology
+settings using claims databases.
 
@@
 Evidence} (ENCORE) project\textsuperscript{11} aims to extend this work
-to the field of oncology. Studies in oncology come with their own unique
-set of challenges which must be systematically explored and understood.
+to oncology. Studies in oncology come with their own unique set of
+challenges which must be systematically explored and understood.
 Building on a process co-developed with the FDA through RCT
@@
 
-In this process paper, we describe the design and process for the
-selection of the 12 oncology RCTs, the assessment of the database
-quality and selection, protocol development and pre-registration, study
-design and statistical analysis, and pre-specified agreement metrics to
-evaluate the concordance between RCTs and emulations.
+Here, we describe the design and process for the selection of the 12
+oncology RCTs, the assessment of the database quality and selection,
+protocol development and pre-registration, study design and statistical
+analysis, and pre-specified agreement metrics to evaluate the
+concordance between RCTs and emulations.
 
@@
 common cancers for which there has been substantial therapeutic
-development in recent years. After careful review and exchange with
-clinical and regulatory experts, four cancer indications were
-identified: non-small cell lung cancer, breast cancer, colorectal cancer
-and multiple myeloma. For each cancer type, we aim to conduct three
-trial emulations using multiple databases accessible for the scope of
-this project (i.e., the total number of emulations will equal 12 trials
-x \emph{n} databases which are found fit-for-purpose for each trial).
+development in recent years. After review and discussion with clinical
+and regulatory experts, four cancer indications were identified:
+non-small cell lung cancer, breast cancer, colorectal cancer and
+multiple myeloma. For each cancer type, we aim to conduct three trial
+emulations using multiple databases accessible for the scope of this
+project (i.e., the total number of emulations will equal 12 trials x
+\emph{n} databases which are found fit-for-purpose for each trial).
 
@@
 still treated with outdated treatment regimens in current clinical
-practice. We also did not include trials with results published too
-recently. This decision was in order to allow for enough data and
-follow-up time accrual in the databases used for this project. Although
-there have been substantial methodological advancements to increase our
-understanding on the emulation and comparison of real-world
-progression-free survival (PFS) and objective response rates (ORR) to a
+practice. Conversely, we excluded trials with results published too
+recently. To allow for enough data and follow-up time accrual in the
+databases used for this project. Although there have been substantial
+methodological advancements to increase our understanding on the
+emulation and comparison of real-world progression-free survival (PFS)
+and objective response rates (ORR) to a
 RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
@@
 chart-abstracted assessments of a patient's progression in routine
-care.\textsuperscript{18, 19} In addition, the timing and cadence of
-intervals between progression assessments can differ between RCTs and
-routine care which may result in measurement error and
-bias.\textsuperscript{20} Given this, and the large number of other
-methodological challenges like missing data, small sample sizes, data
-discontinuity and rapidly changing guideline treatments, we focus on the
-emulation of overall survival (OS) as the endpoint of interest.
-Therefore, we include only trials that have reported overall survival
-(OS) as one of the pre-specified endpoints in the protocol.
+care.\textsuperscript{18, 19} The timing and cadence of intervals
+between progression assessments can differ between RCTs and routine care
+which may result in measurement error and bias.\textsuperscript{20}
+Given this, and the large number of other methodological challenges like
+missing data, small sample sizes, data discontinuity and rapidly
+changing guideline treatments, we focus on the emulation of overall
+survival (OS) as the endpoint of interest. Therefore, we include only
+trials that have reported overall survival (OS) as one of the
+pre-specified endpoints in the protocol.
 
@@
 biomarker status) and prognostic factors (e.g., ECOG performance score)
-are measurable and whether preliminary sample size counts are
-reasonable. Lastly, trial candidates were ranked and shortlisted into
-primary and runner-up candidates based on their clinical and regulatory
-relevance.
+are measurable and whether preliminary study size counts are reasonable.
+Lastly, trial candidates were ranked and shortlisted into primary and
+runner-up candidates based on their clinical and regulatory relevance.
 
@@
 from mis-measurement of key study parameters or residual confounding
-remains high, or that the sample size is not sufficient, then runner-up
+remains high, or that the study size is not sufficient, then runner-up
 candidates will be considered instead.
@@
 regulatory submissions of RWE studies\textsuperscript{23} and will be
-registered on ClinicalTrials.gov after careful review by a clinical and
-FDA regulatory expert panel. Following the target trial emulation
-framework, we will provide an explicit statement and rationale on how
-each element will be emulated including database selection, covariate
-measurement, operationalization of key eligibility criteria, study
-design, data analysis and causal contrasts of
-interest.\textsuperscript{24, 25} Since it is common that oncology RCTs
-update OS estimates periodically based on accrued follow-up time, the
-protocol will give a brief summary of each emulated RCT and specify
-which target OS estimates will be used to compare agreement metrics to
-(see Section~\ref{sec-agreement-metrics}). All eligibility criteria will
-be extracted based on publications, publicly available protocols and
-statistical analysis plans of the selected RCT.
+registered on ClinicalTrials.gov after review by a clinical and FDA
+regulatory expert panel. Following the target trial emulation framework,
+we will provide an explicit statement and rationale on how each element
+will be emulated including database selection, covariate measurement,
+operationalization of key eligibility criteria, study design, data
+analysis and causal contrasts of interest.\textsuperscript{24, 25} Since
+it is common that oncology RCTs update OS estimates periodically based
+on accrued follow-up time, the protocol will give a brief summary of
+each emulated RCT and specify which target OS estimates will be used to
+compare agreement metrics to (see Section~\ref{sec-agreement-metrics}).
+All eligibility criteria will be extracted based on publications,
+publicly available protocols and statistical analysis plans of the
+selected RCT.
 
@@
 outcome will be emulated. For all considered databases, the OS endpoint
-is typically a composite that is derived from different sources
-comprising EHR documentation, social security death index, obituary and
-other linkages. Given that not all relevant sources that provide
-mortality data are synchronized and updated uniformly, sensitivity
-analyses with more conservative (i.e., earlier) censoring dates will be
-considered for each trial emulation to mitigate the potential impact of
-ghost-time bias.\textsuperscript{26}
+is typically a composite that, depending on the underlying database, can
+be derived from different sources comprising EHR documentation, social
+security death index, obituary and other linkages. Given that not all
+relevant sources that provide mortality data are synchronized and
+updated uniformly, sensitivity analyses with more conservative (i.e.,
+earlier) censoring dates will be considered for each trial emulation to
+mitigate the potential impact of ghost-time bias.\textsuperscript{26}
 
@@
 
-\paragraph{\texorpdfstring{\textbf{Statistical
-power}}{Statistical power}}\label{statistical-power}
+\paragraph{\texorpdfstring{\textbf{Statistical power
+considerations}}{Statistical power considerations}}\label{statistical-power-considerations}
 
@@
 missing not at random {[}MNAR{]}).\textsuperscript{32} To that end, we
-will adopt a principled process on missing data that was developed as
-part of a FDA Sentinel Innovation Center causal inference workstream
-that empirically evaluates different aspects across partially observed
-covariates based on three group diagnostics.\textsuperscript{33, 34} In
-brief, these diagnostics cover (1) comparisons of patients
-characteristics with and without an observed level of the partially
-observed covariate, (2) ability to predict missingness given observed
-data, and (3) assessments if outcomes between patients with a missing
-value are systematically different. Together with expert domain
-knowledge and assumptions about the underlying missing data structure
-through canonical causal diagrams,\textsuperscript{35} this will inform
-decisions regarding the in- or exclusion of patients with missing values
-in key eligibility criteria and potential sensitivity analyses to assess
-the robustness of these decisions.\textsuperscript{36}
+will adopt a principled process on missing data that empirically
+evaluates different aspects across partially observed covariates based
+on three group diagnostics.\textsuperscript{33, 34} These diagnostics
+cover (1) comparisons of patients characteristics with and without an
+observed level of the partially observed covariate, (2) ability to
+predict missingness given observed data, and (3) assessments if outcomes
+between patients with a missing value are systematically different.
+Together with expert domain knowledge and assumptions about the
+underlying missing data structure through canonical causal
+diagrams,\textsuperscript{35} this will inform decisions regarding the
+in- or exclusion of patients with missing values in key eligibility
+criteria and potential sensitivity analyses to assess the robustness of
+these decisions.\textsuperscript{36}
 
@@
 coefficient for the treatment comparison for time to all-cause mortality
-(OS).\textsuperscript{39} However, the HR has many limitations that can
-make the comparison of results across varying follow-up times between
-RCT and emulations challenging, including its non-collapsibility,
-in-built selection bias and the requirement of proportional
-hazards.\textsuperscript{40, 41} For this reason, we will also consider
-alternative endpoints on an absolute risk scale such as median survival
-times, survival probabilities at pre-defined time points during
-follow-up\textsuperscript{42} and restricted mean survival times as
+(OS).\textsuperscript{39} We will also consider alternative endpoints on
+an absolute risk scale such as median survival times, survival
+probabilities at pre-defined time points during
+follow-up\textsuperscript{40} and restricted mean survival times as
 secondary endpoints of interest.
@@
 based on expert clinical knowledge and published literature on
-prognostic scores in oncology.\textsuperscript{43} The implementation of
+prognostic scores in oncology.\textsuperscript{41} The implementation of
 propensity scores in combination with multiple imputation will follow
 the `\emph{within}' methodology as described by Leyrat et
-al.\textsuperscript{44, 45} That is, propensity score matching or
+al.\textsuperscript{42, 43} That is, propensity score matching or
 weighting will be applied to each imputed dataset. The marginal
@@
 weighted dataset separately and pooled into a final estimate following
-Rubin's rule.\textsuperscript{46, 47} This approach has been shown to
+Rubin's rule.\textsuperscript{44, 45} This approach has been shown to
 lead to unbiased estimates across different simulated scenarios with a
-sufficient estimation of the variance.\textsuperscript{44}
+sufficient estimation of the variance.\textsuperscript{42}
 
@@
 Figure~\ref{fig-balance}). Covariate balance is typically considered at
-a SMD \textless{} 0.1.\textsuperscript{48} Further, we will compute the
+a SMD \textless{} 0.1.\textsuperscript{46} Further, we will compute the
 average post-matching or post-weighting
-C-statistics.\textsuperscript{49} In addition, we will use a published
-prognostic score for OS\textsuperscript{43} as a balance measure to
+C-statistics.\textsuperscript{47} In addition, we will use a published
+prognostic score for OS\textsuperscript{41} as a balance measure to
 visually assess if the prognostic score is balanced between treatment
@@
 balance measures and not affected by model
-misspecification.\textsuperscript{50}
+misspecification.\textsuperscript{48}
 
@@
 estimated in each imputed and propensity score matched or weighted
-dataset according to the Kaplan-Meier method.\textsuperscript{42} Since
+dataset according to the Kaplan-Meier method.\textsuperscript{40} Since
 survival probabilities typically do not follow normal distributions
@@
 with \(pr(surv)\) denoting the survival probability at a given time
-during follow-up.\textsuperscript{51, 52} The transformed survival
+during follow-up.\textsuperscript{49, 50} The transformed survival
 probabilities are then pooled across imputed datasets and individual
@@
 
-Given that a large number of factors could influence a difference
-between RCT results and emulated trial results, a broad range of
+With the goal to better understand which factors could influence a
+difference between RCT results and emulated trial results, a range of
 sensitivity analyses will be conducted. This can comprise decisions on
@@
 
-Building on an established approach based on the RCT DUPLICATE
+Building on an established approach developed by the RCT DUPLICATE
 project\textsuperscript{9, 30}, the ENCORE project aims to emulate 12
-oncology trials using multiple EHR data sources to inform the potential
-use of RWE for regulatory science in oncology. The project focuses on
-four common cancers and will evaluate the agreement of treatment effect
-estimates between RCTs and their respective emulations. Historically,
-administrative health claims databases have been the backbone of most
-research in RWE. With increasing access to EHR data and a maturing set
-of methodological approaches to draw causal inferences from such data,
-the ENCORE project is one of the first of its kind to evaluate when and
-how RWD can be used to deliver similar causal conclusions compared to
-RCTs in the field of oncology.
-
-The RCT DUPLICATE project\textsuperscript{9, 30} identified multiple
+oncology trials using four oncology EHR data sources to inform the
+potential use of RWE for regulatory science in oncology. The project
+focuses on four common cancers and will evaluate the agreement of
+treatment effect estimates between RCTs and their respective emulations.
+Historically, administrative health claims databases have been the
+backbone of most research in RWE. With increasing access to EHR data and
+a maturing set of methodological approaches to draw causal inferences
+from such data, the ENCORE project is one of the first of its kind to
+evaluate when and how RWD can be used to deliver similar causal
+conclusions compared to RCTs in the field of oncology.
+
+The RCT DUPLICATE initiative\textsuperscript{9, 30} identified multiple
 emulation challenges which we expect to also encounter in ENCORE. One
@@
 differences in subsequent therapy lines there may be estimand
-differences between trial and emulation.\textsuperscript{53} Although
+differences between trial and emulation.\textsuperscript{51} Although
 this is also a common challenge in the analysis of
@@
 Alternative analytic approaches which target a per-protocol estimand,
-such as the clone-censor-weight design,\textsuperscript{54} may be
+such as the clone-censor-weight design,\textsuperscript{52} may be
 viable options if the ITT estimand cannot be estimated due to the
@@
 
-\bibitem[\citeproctext]{ref-hernan2010hazards}
-\textbf{40}. Hernán MA: The hazards of hazard ratios. Epidemiology
-21:13--15, 2010
-
-\bibitem[\citeproctext]{ref-stensrud2020test}
-\textbf{41}. Stensrud MJ, Hernán MA:
-\href{https://doi.org/10.1001/jama.2020.1267}{Why test for proportional
-hazards?} Jama 323:1401--1402, 2020
-
 \bibitem[\citeproctext]{ref-kaplan1958nonparametric}
-\textbf{42}. Kaplan EL, Meier P: Nonparametric estimation from
+\textbf{40}. Kaplan EL, Meier P: Nonparametric estimation from
 incomplete observations. Journal of the American statistical association
@@
 \bibitem[\citeproctext]{ref-becker2020enhanced}
-\textbf{43}. Becker T, Weberpals J, Jegg A, et al: An enhanced
+\textbf{41}. Becker T, Weberpals J, Jegg A, et al: An enhanced
 prognostic score for overall survival of patients with cancer derived
@@
 \bibitem[\citeproctext]{ref-leyrat2019propensity}
-\textbf{44}. Leyrat C, Seaman SR, White IR, et al: Propensity score
+\textbf{42}. Leyrat C, Seaman SR, White IR, et al: Propensity score
 analysis with partially observed covariates: How should multiple
@@
 \bibitem[\citeproctext]{ref-pishgar2020matchthem}
-\textbf{45}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
+\textbf{43}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
 and weighting after multiple imputation. arXiv preprint arXiv:200911772,
@@
 \bibitem[\citeproctext]{ref-rubin2018multiple}
-\textbf{46}. Rubin DB: Multiple imputation, in Flexible imputation of
+\textbf{44}. Rubin DB: Multiple imputation, in Flexible imputation of
 missing data, second edition. Chapman; Hall/CRC, 2018, pp 29--62
@@
 \bibitem[\citeproctext]{ref-van2011mice}
-\textbf{47}. Van Buuren S, Groothuis-Oudshoorn K: Mice: Multivariate
+\textbf{45}. Van Buuren S, Groothuis-Oudshoorn K: Mice: Multivariate
 imputation by chained equations in r. Journal of statistical software
@@
 \bibitem[\citeproctext]{ref-austin2009balance}
-\textbf{48}. Austin PC: Balance diagnostics for comparing the
+\textbf{46}. Austin PC: Balance diagnostics for comparing the
 distribution of baseline covariates between treatment groups in
@@
 \bibitem[\citeproctext]{ref-franklin2014metrics}
-\textbf{49}. Franklin JM, Rassen JA, Ackermann D, et al: Metrics for
+\textbf{47}. Franklin JM, Rassen JA, Ackermann D, et al: Metrics for
 covariate balance in cohort studies of causal effects. Statistics in
@@
 \bibitem[\citeproctext]{ref-stuart2013prognostic}
-\textbf{50}. Stuart EA, Lee BK, Leacy FP: Prognostic score--based
+\textbf{48}. Stuart EA, Lee BK, Leacy FP: Prognostic score--based
 balance measures can be a useful diagnostic for propensity score methods
@@
 \bibitem[\citeproctext]{ref-warwick28685}
-\textbf{51}. Marshall A(Andrea), Billingham LJ, Bryan S: Can we afford
+\textbf{49}. Marshall A(Andrea), Billingham LJ, Bryan S: Can we afford
 to ignore missing data in cost-effectiveness analyses? {[}Internet{]}.
@@
 \bibitem[\citeproctext]{ref-morisot2015prostate}
-\textbf{52}. Morisot A, Bessaoud F, Landais P, et al: Prostate cancer:
+\textbf{50}. Morisot A, Bessaoud F, Landais P, et al: Prostate cancer:
 Net survival and cause-specific survival rates after multiple
@@
 \bibitem[\citeproctext]{ref-manitz2022estimands}
-\textbf{53}. Manitz J, Kan-Dobrosky N, Buchner H, et al: Estimands for
+\textbf{51}. Manitz J, Kan-Dobrosky N, Buchner H, et al: Estimands for
 overall survival in clinical trials with treatment switching in
@@
 \bibitem[\citeproctext]{ref-gaber2024mystifying}
-\textbf{54}. Gaber CE, Ghazarian AA, Strassle PD, et al: De-mystifying
+\textbf{52}. Gaber CE, Ghazarian AA, Strassle PD, et al: De-mystifying
 the clone-censor-weight method for causal research using observational

### 2025-03-30 manuscript/manuscript.qmd

@@
 ---
-title: "A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies"
+title: "A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data"
 
@@
 Kehl^2^, Donna R.
-Rivera^3^, Pallavi Mishra-Kalyani^3^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^4^, Andrew J.
-Belli^5^, Ching-Kun Wang^5^, Jenna Collins^6^**,** JonathanKish^6^**,** Janet Espirito^7^, Nicholas J Robert^7^, Robert J.
+Rivera^3^, Pallavi Mishra-Kalyani^3^, Catherine C.
+Lerro^3^, Erin Larkins^4^, Sonia Singh^4^, Preeti Narayan^4^, Richard Curley^4^, Paul Kluetz^4^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^5^, Andrew J.
+Belli^6^, Ching-Kun Wang^6^, Jenna Collins^7^**,** Jonathan Kish^7^**,** Janet Espirito^8^, Nicholas J Robert^8^, Robert J.
 Glynn^1^, Sebastian Schneeweiss^1^, Shirley V.
@@
 
-^4^ ConcertAI, Cambridge, MA, USA
+^4^ Center for Drug Evaluation and Research, US Food and Drug Administration, Silver Spring, MD, USA
 
-^5^ COTA, Inc., New York, NY, USA
+^5^ ConcertAI, Cambridge, MA, USA
 
-^6^ Flatiron Health, Inc., New York, NY, USA
+^6^ COTA, Inc., New York, NY, USA
 
-^7^ Ontada, Boston, MA, USA
+^7^ Flatiron Health, Inc., New York, NY, USA
+
+^8^ Ontada, Boston, MA, USA
 
@@
 
-Real-world evidence (RWE) studies are increasingly used to complement evidence from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in clinical practice.
-Advancements in the curation and accessibility of electronic health record data (EHR) have presented the opportunity to investigate disease domains such as oncology, where administrative healthcare claims databases alone are not fit-for-purpose.
-The RCT DUPLICATE initiative has previously enhanced understanding of when RWE studies come to the same causal conclusions by emulating trials in non-oncology indications.
+Real-world evidence (RWE) is increasingly used to complement evidence from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in routine clinical practice.
+Advancements in the curation and accessibility of electronic health record data (EHR) have presented the opportunity to utilize real-world data (RWD) to investigate therapeutic areas such as oncology, where administrative healthcare claims databases alone are often not fit-for-purpose.
+The RCT DUPLICATE initiative has previously enhanced understanding of when RWE can most appropriately draw causal conclusions by emulating trials in non-oncology indications.
 Here, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to oncology.
-ENCORE will emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
-It will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to analysis.
+ENCORE is designed to emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
+It will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to initiation and analysis.
 Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria.
@@
 
-Randomized controlled trials (RCTs) have been the go-to methodology for establishing the efficacy and safety of medical products.
-Under the 21^st^ Century Cures Act directive,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023.[@senior2024fresh] RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology, where potential use cases include the assessment of effectiveness in specific patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment to a randomized trial may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
+Randomized controlled trials (RCTs) are the go-to methodology for establishing the efficacy and safety of medical products.
+Under the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor consideration of real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 30% of new drug approvals, oncology was the disease area with the most FDA drug approvals in 2024[@senior2024fresh] and has many areas of high unmet medical need to treat serious conditions.
+Therefore, RWE has particularly important potential to complement evidence coming from RCTs in the field of oncology.
+Potential use cases include the assessment of effectiveness in specific patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials that have feasibility or equipoise challenges, or the precision oncology-focused discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
 
 However, the validity and transportability of results derived between RWE studies and RCTs depends on multiple factors.
-Frequently referenced limitations include, lack of baseline randomization, missing data, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.
+Frequently referenced limitations include, lack of baseline randomization, missing data, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.
 
 The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
-In settings where the RCT designs could be emulated well, RWE studies came to the same conclusions.[@heyard2024design] However, prior work from RCT-DUPLICATE has focused primarily on emulating trials in non-oncology settings using claims databases.
+In settings where the RCT designs could be emulated well, RWE studies were able to reach the same conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.
 
 The *Emulation of Comparative Oncology Trials with Real-world Evidence* (ENCORE) project [@encoreFDA] aims to extend this work to oncology.
-Studies in oncology come with their own unique set of challenges which must be systematically explored and understood.
+Studies in oncology come with their own unique set of methodological challenges which must be systematically explored and understood.
 Building on a process co-developed with the FDA through RCT DUPLICATE,[@wang2023emulation] this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources.
-The process will emphasize transparency and include documented data fitness assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.
+The process will emphasize transparency and include documented data fit-for-purpose assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.
 
-The objectives of this project are to 1) develop state-of-the-art methodological approaches and 2) to apply them to create insights that may provide guidance on the potential use of RWE for regulatory science in oncology.
-To achieve these objectives, this demonstration project will systematically emulate 12 oncology trials across four cancers and assess the agreement of treatment effect estimates between RCTs and their respective emulations.
+The objectives of this project are to 1) develop state-of-the-art methodological approaches and 2) to apply them to create insights into the potential use of RWE to enhance regulatory science in oncology.
+To achieve these objectives, this demonstration project will systematically emulate 12 oncology trials across four cancer types and assess the agreement of treatment effect estimates between RCTs and their respective emulations.
 
@@
 
-The focus of ENCORE is to maximize potential learnings on when RWE studies can or cannot yield similar results compared to RCTs.
-To that end, the emphasis of the project is on trials of therapies for the most common cancers for which there has been substantial therapeutic development in recent years.
-After review and discussion with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma.
-For each cancer type, we aim to conduct three trial emulations using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).
+The focus of ENCORE is to efficiently explore in what contexts RWE studies can or cannot yield similar results compared to RCTs.
+Therefore, the project emphasizes trials of therapies for common cancers for which there has been substantial therapeutic development in recent years.
+After review and collaboration with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma.
+For each cancer type, we aim to conduct three trial emulations using multiple accessible databases (i.e., the total equaling 12 trial emulations x *n* databases which are found fit-for-purpose for each trial).
 
-We used a semi-automated process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs for each cancer type.
-The search was conducted using the *Aggregated Analysis of ClinicalTrials.gov* (AACT) database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains all information (protocol and result data elements) about every study registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *Medical Subject Headings **(***MeSH) term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.
+We used a systematic process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs for each cancer type.
+The search was conducted using the *Aggregated Analysis of ClinicalTrials.gov* (AACT) database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains information (protocol and result data elements) about studies registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *Medical Subject Headings **(***MeSH) term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.
 
@@
 
--   Primary purpose was to study treatment effects
+-   Primary purpose was study treatment efficacy
 
@@
 The rationale and operationalization of each criterion is listed in detail in @tbl-criteria.
-We will mainly consider pivotal interventional, randomized trials after 2011 because treatment guidelines among included cancer indications have undergone significant changes in recent years.
-Due to the rapid adoption of new breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in current clinical practice.
+We will mainly consider pivotal large late-phase interventional RCTs after 2011 because treatment guidelines among included cancer indications have undergone significant changes in recent years.
+Due to the rapid adoption of breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in current clinical practice.
 Conversely, we excluded trials with results published too recently.
-To allow for enough data and follow-up time accrual in the databases used for this project.
-Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not be necessarily reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges like missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
+To allow for enough data and follow-up time accrual in the real-world databases used for this project.
+Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not is often not reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges including missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
 Therefore, we include only trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.
@@
 This criterion involved extensive human review.
-The critical points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) and prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary study size counts are reasonable.
+The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary study size is reasonable.
 Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.
@@
 A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
-Naturally, the majority of trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts start in these settings.
-A key objective that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD.
-If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidates will be considered instead.
+Naturally, the majority of trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts have recently focused on these settings.
+A key objective that we aim to explore with the selected trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD.
+If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.
 
@@
 
-The ENCORE project will use data from four oncology-specific electronic health records (EHR)-derived data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, and Ontada/McKesson.
-All available databases draw from a comprehensive national sample of patients with cancer in the US with the detailed EHR-based data necessary to study medication effectiveness in oncology.
-A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol.
+The ENCORE project will use data from four oncology-specific electronic health records (EHR)-derived RWD data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, and Ontada/McKesson.
+The included databases comprise the necessary information to study medical product effectiveness in oncology derived from US-based oncology EHRs.
+A detailed description and sampling methodology will be provided with each trial emulation protocol.
 For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (the numbering does not coincide with the above order of mention of the databases).
@@
 
-For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] recommended for regulatory submissions of RWE studies[@guidelinegeneral] and will be registered on ClinicalTrials.gov after review by a clinical and FDA regulatory expert panel.
-Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics).
-All eligibility criteria will be extracted based on publications, publicly available protocols and statistical analysis plans of the selected RCT.
+For each selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] recommended for regulatory submissions of RWE studies[@guidelinegeneral] and will be registered on ClinicalTrials.gov after review by a clinical and FDA regulatory expert panel.
+Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used for agreement metric evaluation (see @sec-agreement-metrics).
+All eligibility criteria will be extracted based on publications, publicly available protocols, and statistical analysis plans of the selected RCT.
 
@@
 
-Real-world data fitness and emulation feasibility for each shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care.
-Since many oncological RCTs in recent years have focused on selected, biomarker-defined populations, subtleties in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population.
+Real-world data fitness and emulation feasibility for each shortlisted candidate trial will be assessed in multiple steps based on direction from of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure, line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care.
+Since many oncological RCTs in recent years have focused on selected, biomarker-defined populations, nuances in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population.
 For example, immune checkpoint inhibitors have significantly changed the cancer treatment landscape since the first approvals of the CTLA-4 inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and nivolumab in 2014 in the United States.
-With many trials that have followed thereafter, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has also evolved, and PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.
+However, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has evolved, and as such PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.
 
-According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step will outline tables that describe how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database.
-As there are general eligibility criteria in oncology trials which either will not be possible to emulate (e.g., physician-assessed survival prognosis of x months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.
+According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step outlines how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database.
+As there are general eligibility criteria in oncology clinical trials which either are infeasible to emulate (e.g., physician-assessed survival prognosis of x months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.
 
-We will additionally provide a definition on how exposures, outcomes and covariates are exactly defined and operationalized in each respective database.
-There will be special emphasis on how exposure, in context of their respective disease and line of therapy settings, and the OS outcome will be emulated.
-For all considered databases, the OS endpoint is typically a composite that, depending on the underlying database, can be derived from different sources comprising EHR documentation, social security death index, obituary and other linkages.
+Additionally, we will provide conceptual and operationalized definitions on how exposures, outcomes and covariates are defined in each respective database.
+There will be special emphasis on how exposure, in context of disease and line of therapy settings, and the OS outcome are emulated.
+For all considered databases, OS is typically a composite measure that, depending on the underlying database, can be derived from different mortality sources (i.e., EHR documentation, social security death index, obituary and other linkages).
 Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open]
@@
 
-Critical aspects when emulating oncology trials are the choice and estimation of the appropriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the (historical) control cohort, the adoption rate of the novel intervention in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null.
+One important aspect when emulating oncology trials is the choice and estimation of the appropriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the control cohort, the adoption rate of the novel medical product in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null.
 To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial.
@@
 
-The distribution of patient characteristics, stratified by exposure status, will be examined in Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT.
+The distribution of patient characteristics, by exposure status, will be examined using Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT.
 Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, exposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting.
-At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design and analytic choices based on known outcome information.
+At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design or analytic choices based on known outcome information.
 
@@
 
-Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power as RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results.
+Causal analyses of observational data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results.
 Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
@@
 
-To establish an analytic cohort, key eligibility criteria will be applied in which patients with missing values in eligibility criteria are considered eligible in the respective attrition steps to allow for thorough missing data investigations.
-These missing data investigations will empirically assess assumptions on potential underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] These diagnostics cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different.
-Together with expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams,[@moreno2018canonical] this will inform decisions regarding the in- or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]
+To establish an analytic cohort, missingness will be assessed in a first iteration across patients who meet all eligibility criteria, including those with missing values.
+These missing data investigations will empirically assess assumptions on potential underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] We will adopt a principled process on missing data that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] These diagnostics cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different.
+Expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams[@moreno2018canonical] will additionally inform decisions regarding the inclusion or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]
 
-While the MAR assumption is a strong assumption to hold across all considered covariates, it was shown that especially in the context of partially observed covariate data (as opposed to missing exposure and outcome data), only mechanisms in which a covariate causes its own missingness leads to critical bias (MNAR).[@moreno2018canonical] Hence, methodologies which retain patients and give the potential to adjust for a broader set of prognostic factors (e.g., multiple imputation[@weberpals2024hdmi] or doubly robust methods[@Shaw2024]) may be preferred over complete case analyses.
+While the MAR assumption is a strong assumption to hold across all considered covariates, it has been shown that especially in the context of partially observed covariate data (as opposed to missing exposure and outcome data), only mechanisms in which a covariate causes its own missingness lead to critical bias (MNAR).[@moreno2018canonical] Hence, methodologies which retain patients and give the potential to adjust for a broader set of prognostic factors (e.g., multiple imputation[@weberpals2024hdmi] or doubly robust methods[@Shaw2024]) may be preferred over complete case analyses.
 
-### Endpoints and propensity score analyses
+### Outcome and propensity score analyses
 
-Due to its ubiquity in oncology trials, the primary parameter of interest in ENCORE will be defined as the marginal hazard ratio (HR) coefficient for the treatment comparison for time to all-cause mortality (OS).[@cox1972regression] We will also consider alternative endpoints on an absolute risk scale such as median survival times, survival probabilities at pre-defined time points during follow-up[@kaplan1958nonparametric] and restricted mean survival times as secondary endpoints of interest.
+Due to its ubiquity in oncology trials, the primary parameter of interest in ENCORE will be defined as the marginal hazard ratio (HR) coefficient for the treatment comparison for OS.[@cox1972regression] We will also consider alternative endpoints on an absolute risk scale such as median survival times, survival probabilities at pre-defined time points during follow-up[@kaplan1958nonparametric] and restricted mean survival times as secondary endpoints of interest.
 
@@
 To asses the balance of pre-exposure covariates after matching or weighting on each imputed dataset, the average SMD and corresponding minimum and maximum SMD range will be visualized (see example in @fig-balance).
-Covariate balance is typically considered at a SMD \< 0.1.[@austin2009balance] Further, we will compute the average post-matching or post-weighting C-statistics.[@franklin2014metrics] In addition, we will use a published prognostic score for OS[@becker2020enhanced] as a balance measure to visually assess if the prognostic score is balanced between treatment arms after propensity score matching or weighting as this approach was described to show the highest correlations with bias compared with other balance measures and not affected by model misspecification.[@stuart2013prognostic]
+Covariate balance is often reasonably considered at a SMD \< 0.1.[@austin2009balance] Further, we will compute the average post-matching or post-weighting C-statistics.[@franklin2014metrics] In addition, we will use a published prognostic score for OS[@becker2020enhanced] as a balance measure to visually assess if the prognostic score is balanced between treatment arms after propensity score matching or weighting as this approach was described to show the highest correlations with bias compared with other balance measures and not affected by model misspecification.[@stuart2013prognostic]
 
@@
 
-With the goal to better understand which factors could influence a difference between RCT results and emulated trial results, a range of sensitivity analyses will be conducted.
-This can comprise decisions on the considered databases, calendar time period, covariate measurement (e.g., measurement windows or trade-offs on sensitivity versus specificity in measurements), approaches to missing data, selection of covariates for imputation and propensity score models and decisions on when to censor patients.
+With the goal to better understand which factors could influence a difference between RCT results and emulated trial results, it is appropriate to conduct a range of sensitivity analyses.
+This can comprise decisions on the considered databases, calendar time period, covariate measurement (e.g., measurement windows or trade-offs on sensitivity versus specificity in measurements), approaches to missing data, selection of covariates for imputation and propensity score models and censoring decisions.
 All sensitivity analyses will be pre-specified in the study protocol and reported using appropriate visualizations such as forest plots.
@@
 
-Building on an established approach developed by the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using four oncology EHR data sources to inform the potential use of RWE for regulatory science in oncology.
+Building on an established approach developed by the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using four oncology EHR data sources to inform the potential use of RWD for regulatory purposes in oncology.
 The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations.
-Historically, administrative health claims databases have been the backbone of most research in RWE.
-With increasing access to EHR data and a maturing set of methodological approaches to draw causal inferences from such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to deliver similar causal conclusions compared to RCTs in the field of oncology.
+Historically, administrative claims databases have been the predominant data source used to evaluate medical product safety and effectiveness in the postmarketing setting.
+With increasing access to EHR data, which are more fit for use in oncology than claims due to availability of clinical data elements, and a maturing set of methodological approaches for causal inference using such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to evaluate similar causal conclusions compared to RCTs in the field of oncology.
 
-The RCT DUPLICATE initiative[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges which we expect to also encounter in ENCORE.
-One particular aspect that we may not always be able to emulate is the exact distribution of patient characteristics of the trial population.
-This can be due to the lack of data granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide, which we will not be able to mirror given that all considered databases reflect the US only.
-This may be a critical factor especially given that the pathophysiology, prognosis and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries.
-For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are much more prevalent in Asian countries compared to the US or Europe.
+The RCT DUPLICATE initiative[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges that may be similar in ENCORE.
+One particular aspect that we may not always be able to emulate is the exact distribution of patient characteristics of the trial population, or that the comparability between these two populations may not be known.
+This can be due to the lack of RWD granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide (all considered databases reflect the US only).
+This may be a factor especially given that the pathophysiology, prognosis, and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries.
+For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are much more prevalent in certain geographic regions compared to the US.
 
 Another common challenge in the emulation of oncology trials is the estimation of an "intention-to-treat" (ITT) analogous estimand which is usually the primary estimand reported in oncology RCTs.
-Due to intercurrent events, such as non-adherence, crossover of a high proportion of patients from the control to the intervention arm, or differences in subsequent therapy lines there may be estimand differences between trial and emulation.[@manitz2022estimands] Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment protocols in routine care are often observed to be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
+Due to intercurrent events, such as non-adherence, crossover of a high proportion of patients from the control to the intervention arm, or differences in subsequent therapy lines there may be estimand differences between trial and emulation.[@manitz2022estimands] Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment practice in routine clinical might be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
 
-This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD.
+This issue may be amplified with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD.
 As a result, exposure misclassification (due to overly ascertainment windows) and selection bias (due to overly long ascertainment windows) are common trade-offs in such scenarios.
@@
 
-RWE based on fit-for-purpose data and principled, well-designed and reproducible studies may complement evidence coming from RCTs.
-Through a systematic benchmarking approach, the ENCORE project will provide insights as to how measurement, design and analytic decisions influence bias and validity in oncological RWE studies.
+RWE based on fit-for-purpose data and principled, well-designed and reproducible studies may complement evidence from RCTs.
+Through a systematic benchmarking approach, the ENCORE project will provide insights as to how measurement, design and analytic decisions influence bias and validity in oncological non-interventional studies.
 
@@
 #| label: tbl-metrics
-#| tbl-cap: "Visualization of agreement metrics example."
+#| tbl-cap: "Example visualization of agreement metrics."
 #| fig-pos: 'h'

### 2025-03-30 manuscript/manuscript.tex
Manuscript updated

@@
 \hypersetup{
-  pdftitle={A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies},
+  pdftitle={A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data},
   colorlinks=true,
@@
 \title{A Process for Emulating Comparative Oncology Trials with
-Real-world Evidence Studies}
+Real-world Evidence Studies (ENCORE): Process Development and
+Methodological Considerations for Oncology Real-World Data}
 \author{}
@@
 Kehl\textsuperscript{2}, Donna R. Rivera\textsuperscript{3}, Pallavi
-Mishra-Kalyani\textsuperscript{3}, Georg Hahn\textsuperscript{1},
-Priyanka Anand\textsuperscript{1}, Yanina Natanzon\textsuperscript{4},
-Andrew J. Belli\textsuperscript{5}, Ching-Kun Wang\textsuperscript{5},
-Jenna Collins\textsuperscript{6}\textbf{,}
-JonathanKish\textsuperscript{6}\textbf{,} Janet
-Espirito\textsuperscript{7}, Nicholas J Robert\textsuperscript{7},
-Robert J. Glynn\textsuperscript{1}, Sebastian
-Schneeweiss\textsuperscript{1}, Shirley V. Wang\textsuperscript{1}
+Mishra-Kalyani\textsuperscript{3}, Catherine C.
+Lerro\textsuperscript{3}, Erin Larkins\textsuperscript{4}, Sonia
+Singh\textsuperscript{4}, Preeti Narayan\textsuperscript{4}, Richard
+Curley\textsuperscript{4}, Paul Kluetz\textsuperscript{4}, Georg
+Hahn\textsuperscript{1}, Priyanka Anand\textsuperscript{1}, Yanina
+Natanzon\textsuperscript{5}, Andrew J. Belli\textsuperscript{6},
+Ching-Kun Wang\textsuperscript{6}, Jenna
+Collins\textsuperscript{7}\textbf{,} Jonathan
+Kish\textsuperscript{7}\textbf{,} Janet Espirito\textsuperscript{8},
+Nicholas J Robert\textsuperscript{8}, Robert J.
+Glynn\textsuperscript{1}, Sebastian Schneeweiss\textsuperscript{1},
+Shirley V. Wang\textsuperscript{1}
 
@@
 
-\textsuperscript{4} ConcertAI, Cambridge, MA, USA
-
-\textsuperscript{5} COTA, Inc., New York, NY, USA
-
-\textsuperscript{6} Flatiron Health, Inc., New York, NY, USA
-
-\textsuperscript{7} Ontada, Boston, MA, USA
+\textsuperscript{4} Center for Drug Evaluation and Research, US Food and
+Drug Administration, Silver Spring, MD, USA
+
+\textsuperscript{5} ConcertAI, Cambridge, MA, USA
+
+\textsuperscript{6} COTA, Inc., New York, NY, USA
+
+\textsuperscript{7} Flatiron Health, Inc., New York, NY, USA
+
+\textsuperscript{8} Ontada, Boston, MA, USA
 
@@
 
-\emph{Manuscript last updated: 2025-03-04 18:43:42.372054}
+\emph{Manuscript last updated: 2025-03-30 15:14:45.088413}
 
@@
 
-Real-world evidence (RWE) studies are increasingly used to complement
-evidence from randomized controlled trials (RCTs), contextualizing the
+Real-world evidence (RWE) is increasingly used to complement evidence
+from randomized controlled trials (RCTs), contextualizing the
 effectiveness and safety of medical interventions as delivered in
-clinical practice. Advancements in the curation and accessibility of
-electronic health record data (EHR) have presented the opportunity to
-investigate disease domains such as oncology, where administrative
-healthcare claims databases alone are not fit-for-purpose. The RCT
-DUPLICATE initiative has previously enhanced understanding of when RWE
-studies come to the same causal conclusions by emulating trials in
-non-oncology indications. Here, we present the Emulation of Comparative
-Oncology Trials with Real-world Evidence (ENCORE) project, which aims to
-extend this work to oncology. ENCORE will emulate 12 RCTs in four
+routine clinical practice. Advancements in the curation and
+accessibility of electronic health record data (EHR) have presented the
+opportunity to utilize real-world data (RWD) to investigate therapeutic
+areas such as oncology, where administrative healthcare claims databases
+alone are often not fit-for-purpose. The RCT DUPLICATE initiative has
+previously enhanced understanding of when RWE can most appropriately
+draw causal conclusions by emulating trials in non-oncology indications.
+Here, we present the Emulation of Comparative Oncology Trials with
+Real-world Evidence (ENCORE) project, which aims to extend this work to
+oncology. ENCORE is designed to emulate 12 RCTs in four
 oncology-specialized EHR databases across four different cancer
@@
 study design and statistical analysis for a particular research
-question, and pre-registration of study protocols prior to analysis.
-Agreement of treatment effect estimates between RCTs and their
-respective emulations will be assessed using pre-specified criteria.
-Through extensive sensitivity analyses benchmarked against RCT results,
-the ENCORE project will inform understanding of how measurement, design,
-and analytic decisions influence the validity of oncological RWE
-studies.
+question, and pre-registration of study protocols prior to initiation
+and analysis. Agreement of treatment effect estimates between RCTs and
+their respective emulations will be assessed using pre-specified
+criteria. Through extensive sensitivity analyses benchmarked against RCT
+results, the ENCORE project will inform understanding of how
+measurement, design, and analytic decisions influence the validity of
+oncological RWE studies.
 
@@
 
-Randomized controlled trials (RCTs) have been the go-to methodology for
+Randomized controlled trials (RCTs) are the go-to methodology for
 establishing the efficacy and safety of medical products. Under the
-21\textsuperscript{st} Century Cures Act directive,\textsuperscript{1}
-the Food and Drug Administration (FDA) established a framework to
-increasingly consider real-world evidence (RWE) generated from
-routine-care health data such as electronic health records (EHR) to
-evaluate and contextualize the comparative safety and effectiveness of
-novel cancer therapies.\textsuperscript{2} Accounting for 21\% of all
-approvals, oncology was the disease area with the most FDA drug
-approvals in 2023.\textsuperscript{3} RWE has particularly important
+21\textsuperscript{st} Century Cures Act,\textsuperscript{1} the Food
+and Drug Administration (FDA) established a framework to factor
+consideration of real-world evidence (RWE) generated from routine-care
+health data such as electronic health records (EHR) to evaluate and
+contextualize the safety and effectiveness of novel cancer
+therapies.\textsuperscript{2} Accounting for 30\% of new drug approvals,
+oncology was the disease area with the most FDA drug approvals in
+2024\textsuperscript{3} and has many areas of high unmet medical need to
+treat serious conditions. Therefore, RWE has particularly important
 potential to complement evidence coming from RCTs in the field of
-precision oncology, where potential use cases include the assessment of
-effectiveness in specific patient populations that are underrepresented
-in RCTs, the construction of external control arms in single-arm trials
-where active recruitment to a randomized trial may not be feasible, or
-the discovery of biomarkers among pan-tumor populations that harbor
-specific genomic and immuno-pathological signatures.
+oncology. Potential use cases include the assessment of effectiveness in
+specific patient populations that are underrepresented in RCTs, the
+construction of external control arms in single-arm trials that have
+feasibility or equipoise challenges, or the precision oncology-focused
+discovery of biomarkers among pan-tumor populations that harbor specific
+genomic and immuno-pathological signatures.
 
@@
 and standardized response assessments in real-world data
-(RWD).\textsuperscript{6} While there are already published examples of
-oncology trial emulations,\textsuperscript{6--8} a systematic and scaled
+(RWD).\textsuperscript{6} While examples of oncology trial emulations
+have been published,\textsuperscript{6--8} a systematic and scaled
 approach to emulate a diverse set of different oncology trials with
 multiple heterogeneous databases is necessary to gain confidence in the
-validity of RWE studies and to provide context as to which questions can
-be validly answered.
+validity of RWE studies, evaluate regulatory considerations, and to
+provide context as to which questions can be validly answered.
 
@@
 assumption that each RCT finding reflects a causal treatment effect. In
-settings where the RCT designs could be emulated well, RWE studies came
-to the same conclusions.\textsuperscript{10} However, prior work from
-RCT-DUPLICATE has focused primarily on emulating trials in non-oncology
-settings using claims databases.
+settings where the RCT designs could be emulated well, RWE studies were
+able to reach the same conclusions.\textsuperscript{10} However, this
+prior work from RCT-DUPLICATE focused primarily on emulating trials in
+non-oncology settings using claims databases.
 
@@
 to oncology. Studies in oncology come with their own unique set of
-challenges which must be systematically explored and understood.
-Building on a process co-developed with the FDA through RCT
+methodological challenges which must be systematically explored and
+understood. Building on a process co-developed with the FDA through RCT
 DUPLICATE,\textsuperscript{9} this expansion to oncology will emulate 12
@@
 sources. The process will emphasize transparency and include documented
-data fitness assessment\textsuperscript{12, 13} for each RWD source with
-respect to each trial emulation as well as extensive sensitivity
-analyses to assess robustness of findings.
+data fit-for-purpose assessment\textsuperscript{12, 13} for each RWD
+source with respect to each trial emulation as well as extensive
+sensitivity analyses to assess robustness of findings.
 
 The objectives of this project are to 1) develop state-of-the-art
-methodological approaches and 2) to apply them to create insights that
-may provide guidance on the potential use of RWE for regulatory science
-in oncology. To achieve these objectives, this demonstration project
-will systematically emulate 12 oncology trials across four cancers and
-assess the agreement of treatment effect estimates between RCTs and
-their respective emulations.
+methodological approaches and 2) to apply them to create insights into
+the potential use of RWE to enhance regulatory science in oncology. To
+achieve these objectives, this demonstration project will systematically
+emulate 12 oncology trials across four cancer types and assess the
+agreement of treatment effect estimates between RCTs and their
+respective emulations.
 
@@
 
-The focus of ENCORE is to maximize potential learnings on when RWE
-studies can or cannot yield similar results compared to RCTs. To that
-end, the emphasis of the project is on trials of therapies for the most
-common cancers for which there has been substantial therapeutic
-development in recent years. After review and discussion with clinical
-and regulatory experts, four cancer indications were identified:
-non-small cell lung cancer, breast cancer, colorectal cancer and
-multiple myeloma. For each cancer type, we aim to conduct three trial
-emulations using multiple databases accessible for the scope of this
-project (i.e., the total number of emulations will equal 12 trials x
-\emph{n} databases which are found fit-for-purpose for each trial).
-
-We used a semi-automated process for trial selection where the
-eligibility criteria are documented in CONSORT diagrams showing reasons
-for excluding RCTs for each cancer type. The search was conducted using
-the \emph{Aggregated Analysis of ClinicalTrials.gov} (AACT) database
-which is a publicly available relational database developed and
-maintained by the Clinical Trials Transformation Initiative (CTTI) that
-contains all information (protocol and result data elements) about every
-study registered on ClinicalTrials.gov.\textsuperscript{14} To identify
-eligible trials, we used a combined search query strategy of the
-National Library of Medicine (NLM)-controlled \emph{Medical Subject
-Headings \textbf{(}}MeSH) term and a free keyword search for the
-respective cancer indication in the \emph{conditions}, \emph{studies}
-and \emph{detailed\_descriptions} fields of each trial entry on
+The focus of ENCORE is to efficiently explore in what contexts RWE
+studies can or cannot yield similar results compared to RCTs. Therefore,
+the project emphasizes trials of therapies for common cancers for which
+there has been substantial therapeutic development in recent years.
+After review and collaboration with clinical and regulatory experts,
+four cancer indications were identified: non-small cell lung cancer,
+breast cancer, colorectal cancer and multiple myeloma. For each cancer
+type, we aim to conduct three trial emulations using multiple accessible
+databases (i.e., the total equaling 12 trial emulations x \emph{n}
+databases which are found fit-for-purpose for each trial).
+
+We used a systematic process for trial selection where the eligibility
+criteria are documented in CONSORT diagrams showing reasons for
+excluding RCTs for each cancer type. The search was conducted using the
+\emph{Aggregated Analysis of ClinicalTrials.gov} (AACT) database which
+is a publicly available relational database developed and maintained by
+the Clinical Trials Transformation Initiative (CTTI) that contains
+information (protocol and result data elements) about studies registered
+on ClinicalTrials.gov.\textsuperscript{14} To identify eligible trials,
+we used a combined search query strategy of the National Library of
+Medicine (NLM)-controlled \emph{Medical Subject Headings
+\textbf{(}}MeSH) term and a free keyword search for the respective
+cancer indication in the \emph{conditions}, \emph{studies} and
+\emph{detailed\_descriptions} fields of each trial entry on
 ClinicalTrials.gov.
@@
 \item
-  Primary purpose was to study treatment effects
+  Primary purpose was study treatment efficacy
 \item
@@
 detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
-interventional, randomized trials after 2011 because treatment
+large late-phase interventional RCTs after 2011 because treatment
 guidelines among included cancer indications have undergone significant
-changes in recent years. Due to the rapid adoption of new breakthrough
+changes in recent years. Due to the rapid adoption of breakthrough
 therapies in routine care, it is unlikely to find patients who may be
@@
 recently. To allow for enough data and follow-up time accrual in the
-databases used for this project. Although there have been substantial
-methodological advancements to increase our understanding on the
-emulation and comparison of real-world progression-free survival (PFS)
-and objective response rates (ORR) to a
+real-world databases used for this project. Although there have been
+substantial methodological advancements to increase our understanding on
+the emulation and comparison of real-world progression-free survival
+(PFS) and objective response rates (ORR) to a
 RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
 RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
-level of granularity which may not be necessarily reflected in
+level of granularity which may not is often not reflected in
 chart-abstracted assessments of a patient's progression in routine
@@
 which may result in measurement error and bias.\textsuperscript{20}
-Given this, and the large number of other methodological challenges like
-missing data, small sample sizes, data discontinuity and rapidly
-changing guideline treatments, we focus on the emulation of overall
-survival (OS) as the endpoint of interest. Therefore, we include only
-trials that have reported overall survival (OS) as one of the
+Given this, and the large number of other methodological challenges
+including missing data, small sample sizes, data discontinuity and
+rapidly changing guideline treatments, we focus on the emulation of
+overall survival (OS) as the endpoint of interest. Therefore, we include
+only trials that have reported overall survival (OS) as one of the
 pre-specified endpoints in the protocol.
@@
 feasibility and clinical relevance. This criterion involved extensive
-human review. The critical points considered in this step include an
+human review. The fundamental points considered in this step include an
 initial feasibility assessment of the data fitness,\textsuperscript{19}
 including assessment of whether critical eligibility criteria (e.g.,
-biomarker status) and prognostic factors (e.g., ECOG performance score)
-are measurable and whether preliminary study size counts are reasonable.
+biomarker status) including prognostic factors (e.g., ECOG performance
+score) are measurable and whether preliminary study size is reasonable.
 Lastly, trial candidates were ranked and shortlisted into primary and
@@
 inoperable, recurrent/progressive disease) or metastatic cancer
-populations because a large proportion of drug development efforts start
-in these settings. A key objective that we aim to foster with the
-shortlisted trials is to achieve a better understanding how different
-disease settings (early, late), line settings ({[}neo{]}adjuvant, first
-line, advanced lines of therapy), therapy protocols (monotherapy,
-combination therapy) and population characteristics (simple versus
-complex genetic or immunological signatures) can be emulated using RWD.
-If more thorough feasibility assessments suggest that the threat of bias
-from mis-measurement of key study parameters or residual confounding
-remains high, or that the study size is not sufficient, then runner-up
-candidates will be considered instead.
+populations because a large proportion of drug development efforts have
+recently focused on these settings. A key objective that we aim to
+explore with the selected trials is to achieve a better understanding
+how different disease settings (early, late), line settings
+({[}neo{]}adjuvant, first line, advanced lines of therapy), therapy
+protocols (monotherapy, combination therapy) and clinical
+characteristics (all-comer versus complex genetic or immunological
+signatures) can be emulated using RWD. If more thorough feasibility
+assessments suggest that the threat of bias from mis-measurement of key
+study parameters or residual confounding remains high, or that the study
+size is not sufficient, then runner-up candidate trials will be
+considered instead.
 
@@
 The ENCORE project will use data from four oncology-specific electronic
-health records (EHR)-derived data sources (in alphabetical order):
-ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. All available
-databases draw from a comprehensive national sample of patients with
-cancer in the US with the detailed EHR-based data necessary to study
-medication effectiveness in oncology. A detailed description and
-methodology on how patients are sampled will be provided with each trial
-emulation protocol. For ENCORE, not all databases will be available for
-each cancer indication and the names of the databases will be blinded
-and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final
-reporting of results (the numbering does not coincide with the above
-order of mention of the databases). If more than one database is
+health records (EHR)-derived RWD data sources (in alphabetical order):
+ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. The included
+databases comprise the necessary information to study medical product
+effectiveness in oncology derived from US-based oncology EHRs. A
+detailed description and sampling methodology will be provided with each
+trial emulation protocol. For ENCORE, not all databases will be
+available for each cancer indication and the names of the databases will
+be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for
+the final reporting of results (the numbering does not coincide with the
+above order of mention of the databases). If more than one database is
 considered fit-for-purpose for a respective trial emulation, the best
@@
 
-For each shortlisted and selected RCT, a detailed protocol,
-pre-specifying key elements of the trial emulation, will be developed
-using the HARPER protocol template\textsuperscript{22} recommended for
-regulatory submissions of RWE studies\textsuperscript{23} and will be
-registered on ClinicalTrials.gov after review by a clinical and FDA
-regulatory expert panel. Following the target trial emulation framework,
-we will provide an explicit statement and rationale on how each element
-will be emulated including database selection, covariate measurement,
-operationalization of key eligibility criteria, study design, data
-analysis and causal contrasts of interest.\textsuperscript{24, 25} Since
-it is common that oncology RCTs update OS estimates periodically based
-on accrued follow-up time, the protocol will give a brief summary of
-each emulated RCT and specify which target OS estimates will be used to
-compare agreement metrics to (see Section~\ref{sec-agreement-metrics}).
-All eligibility criteria will be extracted based on publications,
-publicly available protocols and statistical analysis plans of the
-selected RCT.
+For each selected RCT, a detailed protocol, pre-specifying key elements
+of the trial emulation, will be developed using the HARPER protocol
+template\textsuperscript{22} recommended for regulatory submissions of
+RWE studies\textsuperscript{23} and will be registered on
+ClinicalTrials.gov after review by a clinical and FDA regulatory expert
+panel. Following the target trial emulation framework, we will provide
+an explicit statement and rationale on how each element will be emulated
+including database selection, covariate measurement, operationalization
+of key eligibility criteria, study design, data analysis and causal
+contrasts of interest.\textsuperscript{24, 25} Since it is common that
+oncology RCTs update OS estimates periodically based on accrued
+follow-up time, the protocol will give a brief summary of each emulated
+RCT and specify which target OS estimates will be used for agreement
+metric evaluation (see Section~\ref{sec-agreement-metrics}). All
+eligibility criteria will be extracted based on publications, publicly
+available protocols, and statistical analysis plans of the selected RCT.
 
@@
 Real-world data fitness and emulation feasibility for each shortlisted
-candidate trial will be assessed in multiple steps based on guidance of
-the oncology quality, characterization, and assessment of real-world
-data (Oncology QCARD) Initiative.\textsuperscript{19} The first step
-assesses if relevant variables like exposure/line of therapy, outcomes,
-and covariates are generally available, measured and operationalizable
-in routine-care. Since many oncological RCTs in recent years have
-focused on selected, biomarker-defined populations, subtleties in
-measurement and operationalizability of specific biomarkers must be
+candidate trial will be assessed in multiple steps based on direction
+from of the oncology quality, characterization, and assessment of
+real-world data (Oncology QCARD) Initiative.\textsuperscript{19} The
+first step assesses if relevant variables like exposure, line of
+therapy, outcomes, and covariates are generally available, measured and
+operationalizable in routine-care. Since many oncological RCTs in recent
+years have focused on selected, biomarker-defined populations, nuances
+in measurement and operationalizability of specific biomarkers must be
 reflected to ensure a representative and large enough study population.
@@
 inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and
-nivolumab in 2014 in the United States. With many trials that have
-followed thereafter, the operationalization of the expression of the
-PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion
-score or combined positive score) has also evolved, and PD-L1
-'\emph{positivity'} may have different definitions across calendar years
-based on different cut-off values.
+nivolumab in 2014 in the United States. However, the operationalization
+of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent
+staining, tumor proportion score or combined positive score) has
+evolved, and as such PD-L1 '\emph{positivity'} may have different
+definitions across calendar years based on different cut-off values.
 
 According to the \emph{Structured Process to Identify Fit-For-Purpose
-Data} (SPIFD) framework\textsuperscript{13}, the next step will outline
-tables that describe how eligibility criteria will be ascertained using
-a color-coded heatmap that will indicate the level of confidence on how
-well each criterion can be emulated in each selected database. As there
-are general eligibility criteria in oncology trials which either will
-not be possible to emulate (e.g., physician-assessed survival prognosis
-of x months) or that are clinically not relevant for the emulation of
-the trial (e.g., male patients should be willing to use barrier
+Data} (SPIFD) framework\textsuperscript{13}, the next step outlines how
+eligibility criteria will be ascertained using a color-coded heatmap
+that will indicate the level of confidence on how well each criterion
+can be emulated in each selected database. As there are general
+eligibility criteria in oncology clinical trials which either are
+infeasible to emulate (e.g., physician-assessed survival prognosis of x
+months) or that are clinically not relevant for the emulation of the
+trial (e.g., male patients should be willing to use barrier
 contraception), the study team will decide on key eligibility criteria
@@
 
-We will additionally provide a definition on how exposures, outcomes and
-covariates are exactly defined and operationalized in each respective
+Additionally, we will provide conceptual and operationalized definitions
+on how exposures, outcomes and covariates are defined in each respective
 database. There will be special emphasis on how exposure, in context of
-their respective disease and line of therapy settings, and the OS
-outcome will be emulated. For all considered databases, the OS endpoint
-is typically a composite that, depending on the underlying database, can
-be derived from different sources comprising EHR documentation, social
-security death index, obituary and other linkages. Given that not all
-relevant sources that provide mortality data are synchronized and
-updated uniformly, sensitivity analyses with more conservative (i.e.,
-earlier) censoring dates will be considered for each trial emulation to
-mitigate the potential impact of ghost-time bias.\textsuperscript{26}
+disease and line of therapy settings, and the OS outcome are emulated.
+For all considered databases, OS is typically a composite measure that,
+depending on the underlying database, can be derived from different
+mortality sources (i.e., EHR documentation, social security death index,
+obituary and other linkages). Given that not all relevant sources that
+provide mortality data are synchronized and updated uniformly,
+sensitivity analyses with more conservative (i.e., earlier) censoring
+dates will be considered for each trial emulation to mitigate the
+potential impact of ghost-time bias.\textsuperscript{26}
 
@@
 
-Critical aspects when emulating oncology trials are the choice and
+One important aspect when emulating oncology trials is the choice and
 estimation of the appropriate estimand of interest.\textsuperscript{27}
@@
 treatments, multiple aspects need to be considered such as the
-contemporaneity of the (historical) control cohort, the adoption rate of
-the novel intervention in routine care, the magnitude of the clinical
-treatment benefit and the rate in which (particularly patients in the
-control arm) discontinue or cross-over to the interventional treatment,
-which could lastly bias emulated treatment effects towards the null. To
-that end, comprehensive data explorations will be performed as part of
-the protocol development to contextualize these parameters and (if
-reported) draw comparisons to the emulated trial. Examples for such
-standard diagnostics are visualized in Figure~\ref{fig-initiators}. All
+contemporaneity of the control cohort, the adoption rate of the novel
+medical product in routine care, the magnitude of the clinical treatment
+benefit and the rate in which (particularly patients in the control arm)
+discontinue or cross-over to the interventional treatment, which could
+lastly bias emulated treatment effects towards the null. To that end,
+comprehensive data explorations will be performed as part of the
+protocol development to contextualize these parameters and (if reported)
+draw comparisons to the emulated trial. Examples for such standard
+diagnostics are visualized in Figure~\ref{fig-initiators}. All
 exploratory analyses will be conducted blinded towards the outcome to
@@
 
-The distribution of patient characteristics, stratified by exposure
-status, will be examined in Table 1's before and after applying
-eligibility criteria and contrasted with the distributions of patient
-characteristics of the original RCT. Initial propensity score matching
-or weighting methods will be applied to ensure that measured
-pre-exposure covariates can be balanced, exposure cohorts are
-conditionally exchangeable at baseline and resulting sample sizes are
-still sufficient after matching or weighting. At this stage, all
-exploratory analyses will be conducted blinded towards the outcome to
-not bias any study design and analytic choices based on known outcome
-information.
+The distribution of patient characteristics, by exposure status, will be
+examined using Table 1's before and after applying eligibility criteria
+and contrasted with the distributions of patient characteristics of the
+original RCT. Initial propensity score matching or weighting methods
+will be applied to ensure that measured pre-exposure covariates can be
+balanced, exposure cohorts are conditionally exchangeable at baseline
+and resulting sample sizes are still sufficient after matching or
+weighting. At this stage, all exploratory analyses will be conducted
+blinded towards the outcome to not bias any study design or analytic
+choices based on known outcome information.
 
@@
 
-Causal analyses of observational data may not have the same
-pre-requisites in terms of formal hypothesis testing and statistical
-power as RCTs since the number of `recruited' patients is given and
-cannot be influenced.\textsuperscript{28} For this project, however, the
-emulations in EHR data must have at least equal power to the relevant
-trial for interpretation of the pre-specified agreements metrics between
-RCT and RWD results. Since the main outcome of interest is defined as
-time to all-cause mortality (OS), the estimation of the statistical
-power is driven by the number of events rather than the number of
-patients. To assess if the overall number of events, unstratified by
-exposure, is sufficient such that a significant difference can be
-detected based on the original RCT-reported hazard ratio (HR), the
-statistical power (1-\(\beta\)) will be estimated using Schoenfeld's
-sample-size formula for the proportional-hazards regression
+Causal analyses of observational data are often not designed with
+respect to formal hypothesis testing and statistical power in the same
+manner as RCTs since the number of `recruited' patients is limited by
+available data and cannot be influenced.\textsuperscript{28} For this
+project, however, the emulations in EHR data must have at least equal
+power to the relevant trial for interpretation of the pre-specified
+agreements metrics between RCT and RWD results. Since the main outcome
+of interest is defined as time to all-cause mortality (OS), the
+estimation of the statistical power is driven by the number of events
+rather than the number of patients. To assess if the overall number of
+events, unstratified by exposure, is sufficient such that a significant
+difference can be detected based on the original RCT-reported hazard
+ratio (HR), the statistical power (1-\(\beta\)) will be estimated using
+Schoenfeld's sample-size formula for the proportional-hazards regression
 model.\textsuperscript{29}
@@
 
-To establish an analytic cohort, key eligibility criteria will be
-applied in which patients with missing values in eligibility criteria
-are considered eligible in the respective attrition steps to allow for
-thorough missing data investigations. These missing data investigations
-will empirically assess assumptions on potential underlying missingness
+To establish an analytic cohort, missingness will be assessed in a first
+iteration across patients who meet all eligibility criteria, including
+those with missing values. These missing data investigations will
+empirically assess assumptions on potential underlying missingness
 mechanisms according to Rubin's classification of missing data (i.e.,
 missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
-missing not at random {[}MNAR{]}).\textsuperscript{32} To that end, we
-will adopt a principled process on missing data that empirically
-evaluates different aspects across partially observed covariates based
-on three group diagnostics.\textsuperscript{33, 34} These diagnostics
-cover (1) comparisons of patients characteristics with and without an
-observed level of the partially observed covariate, (2) ability to
-predict missingness given observed data, and (3) assessments if outcomes
-between patients with a missing value are systematically different.
-Together with expert domain knowledge and assumptions about the
-underlying missing data structure through canonical causal
-diagrams,\textsuperscript{35} this will inform decisions regarding the
-in- or exclusion of patients with missing values in key eligibility
-criteria and potential sensitivity analyses to assess the robustness of
-these decisions.\textsuperscript{36}
+missing not at random {[}MNAR{]}).\textsuperscript{32} We will adopt a
+principled process on missing data that empirically evaluates different
+aspects across partially observed covariates based on three group
+diagnostics.\textsuperscript{33, 34} These diagnostics cover (1)
+comparisons of patients characteristics with and without an observed
+level of the partially observed covariate, (2) ability to predict
+missingness given observed data, and (3) assessments if outcomes between
+patients with a missing value are systematically different. Expert
+domain knowledge and assumptions about the underlying missing data
+structure through canonical causal diagrams\textsuperscript{35} will
+additionally inform decisions regarding the inclusion or exclusion of
+patients with missing values in key eligibility criteria and potential
+sensitivity analyses to assess the robustness of these
+decisions.\textsuperscript{36}
 
 While the MAR assumption is a strong assumption to hold across all
-considered covariates, it was shown that especially in the context of
-partially observed covariate data (as opposed to missing exposure and
+considered covariates, it has been shown that especially in the context
+of partially observed covariate data (as opposed to missing exposure and
 outcome data), only mechanisms in which a covariate causes its own
-missingness leads to critical bias (MNAR).\textsuperscript{35} Hence,
+missingness lead to critical bias (MNAR).\textsuperscript{35} Hence,
 methodologies which retain patients and give the potential to adjust for
@@
 
-\subsubsection{Endpoints and propensity score
-analyses}\label{endpoints-and-propensity-score-analyses}
+\subsubsection{Outcome and propensity score
+analyses}\label{outcome-and-propensity-score-analyses}
 
@@
 interest in ENCORE will be defined as the marginal hazard ratio (HR)
-coefficient for the treatment comparison for time to all-cause mortality
-(OS).\textsuperscript{39} We will also consider alternative endpoints on
-an absolute risk scale such as median survival times, survival
-probabilities at pre-defined time points during
-follow-up\textsuperscript{40} and restricted mean survival times as
-secondary endpoints of interest.
+coefficient for the treatment comparison for OS.\textsuperscript{39} We
+will also consider alternative endpoints on an absolute risk scale such
+as median survival times, survival probabilities at pre-defined time
+points during follow-up\textsuperscript{40} and restricted mean survival
+times as secondary endpoints of interest.
 
@@
 minimum and maximum SMD range will be visualized (see example in
-Figure~\ref{fig-balance}). Covariate balance is typically considered at
-a SMD \textless{} 0.1.\textsuperscript{46} Further, we will compute the
-average post-matching or post-weighting
+Figure~\ref{fig-balance}). Covariate balance is often reasonably
+considered at a SMD \textless{} 0.1.\textsuperscript{46} Further, we
+will compute the average post-matching or post-weighting
 C-statistics.\textsuperscript{47} In addition, we will use a published
@@
 With the goal to better understand which factors could influence a
-difference between RCT results and emulated trial results, a range of
-sensitivity analyses will be conducted. This can comprise decisions on
-the considered databases, calendar time period, covariate measurement
-(e.g., measurement windows or trade-offs on sensitivity versus
-specificity in measurements), approaches to missing data, selection of
-covariates for imputation and propensity score models and decisions on
-when to censor patients. All sensitivity analyses will be pre-specified
+difference between RCT results and emulated trial results, it is
+appropriate to conduct a range of sensitivity analyses. This can
+comprise decisions on the considered databases, calendar time period,
+covariate measurement (e.g., measurement windows or trade-offs on
+sensitivity versus specificity in measurements), approaches to missing
+data, selection of covariates for imputation and propensity score models
+and censoring decisions. All sensitivity analyses will be pre-specified
 in the study protocol and reported using appropriate visualizations such
@@
 oncology trials using four oncology EHR data sources to inform the
-potential use of RWE for regulatory science in oncology. The project
+potential use of RWD for regulatory purposes in oncology. The project
 focuses on four common cancers and will evaluate the agreement of
 treatment effect estimates between RCTs and their respective emulations.
-Historically, administrative health claims databases have been the
-backbone of most research in RWE. With increasing access to EHR data and
-a maturing set of methodological approaches to draw causal inferences
-from such data, the ENCORE project is one of the first of its kind to
-evaluate when and how RWD can be used to deliver similar causal
-conclusions compared to RCTs in the field of oncology.
+Historically, administrative claims databases have been the predominant
+data source used to evaluate medical product safety and effectiveness in
+the postmarketing setting. With increasing access to EHR data, which are
+more fit for use in oncology than claims due to availability of clinical
+data elements, and a maturing set of methodological approaches for
+causal inference using such data, the ENCORE project is one of the first
+of its kind to evaluate when and how RWD can be used to evaluate similar
+causal conclusions compared to RCTs in the field of oncology.
 
 The RCT DUPLICATE initiative\textsuperscript{9, 30} identified multiple
-emulation challenges which we expect to also encounter in ENCORE. One
-particular aspect that we may not always be able to emulate is the exact
-distribution of patient characteristics of the trial population. This
-can be due to the lack of data granularity and comprehensiveness to
+emulation challenges that may be similar in ENCORE. One particular
+aspect that we may not always be able to emulate is the exact
+distribution of patient characteristics of the trial population, or that
+the comparability between these two populations may not be known. This
+can be due to the lack of RWD granularity and comprehensiveness to
 emulate relevant eligibility criteria or the fact that large pivotal
 trials in oncology are typically conducted in multiple countries
-worldwide, which we will not be able to mirror given that all considered
-databases reflect the US only. This may be a critical factor especially
-given that the pathophysiology, prognosis and factors that drive
-heterogeneous treatment outcomes of certain cancers differ between
-countries. For example, some cancer indications (e.g., GI cancers) or
-some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are
-much more prevalent in Asian countries compared to the US or Europe.
+worldwide (all considered databases reflect the US only). This may be a
+factor especially given that the pathophysiology, prognosis, and factors
+that drive heterogeneous treatment outcomes of certain cancers differ
+between countries. For example, some cancer indications (e.g., GI
+cancers) or some genetic mutations (e.g., EGFR mutations in lung
+adenocarcinoma) are much more prevalent in certain geographic regions
+compared to the US.
 
@@
 this is also a common challenge in the analysis of
-RCTs,\textsuperscript{27} treatment protocols in routine care are often
-observed to be less stringent compared to trials.\textsuperscript{8} In
-order to derive comparable estimands it is therefore crucial to
-understand and contextualize the proportion and timing of treatment
-switching and discontinuation in both the trial and its emulation.
-
-This issue may be even augmented with more complex treatment protocols
-like combination regimens (e.g., CheckMate9LA) as compared to
-monotherapies (e.g., CheckMate057) since the ascertainment of the
-exposure needs to happen over a pre-defined time window which is often
-difficult to calibrate in RWD. As a result, exposure misclassification
-(due to overly ascertainment windows) and selection bias (due to overly
-long ascertainment windows) are common trade-offs in such scenarios.
+RCTs,\textsuperscript{27} treatment practice in routine clinical might
+be less stringent compared to trials.\textsuperscript{8} In order to
+derive comparable estimands it is therefore crucial to understand and
+contextualize the proportion and timing of treatment switching and
+discontinuation in both the trial and its emulation.
+
+This issue may be amplified with more complex treatment protocols like
+combination regimens (e.g., CheckMate9LA) as compared to monotherapies
+(e.g., CheckMate057) since the ascertainment of the exposure needs to
+happen over a pre-defined time window which is often difficult to
+calibrate in RWD. As a result, exposure misclassification (due to overly
+ascertainment windows) and selection bias (due to overly long
+ascertainment windows) are common trade-offs in such scenarios.
 Alternative analytic approaches which target a per-protocol estimand,
@@
 RWE based on fit-for-purpose data and principled, well-designed and
-reproducible studies may complement evidence coming from RCTs. Through a
+reproducible studies may complement evidence from RCTs. Through a
 systematic benchmarking approach, the ENCORE project will provide
 insights as to how measurement, design and analytic decisions influence
-bias and validity in oncological RWE studies.
+bias and validity in oncological non-interventional studies.
 
@@
 
-\caption{\label{tbl-metrics}Visualization of agreement metrics example.}
+\caption{\label{tbl-metrics}Example visualization of agreement metrics.}
 

### 2025-03-30 manuscript/manuscript.qmd

@@
 Dr.
+Glynn has received support for investigator-initiated grants to the Brigham and Women's Hospital from Amarin, AstraZeneca, Kowa, Novartis, and Pfizer unrelated to the current work.
+Dr.
 Schneeweiss is participating in investigator-initiated grants to the Brigham and Women's Hospital from Bayer and UCB unrelated to the topic of this study.
@@
 
-Real-world evidence (RWE) is increasingly used to complement evidence from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in routine clinical practice.
-Advancements in the curation and accessibility of electronic health record data (EHR) have presented the opportunity to utilize real-world data (RWD) to investigate therapeutic areas such as oncology, where administrative healthcare claims databases alone are often not fit-for-purpose.
+Real-world evidence (RWE) is increasingly used to complement findings from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in routine clinical practice.
+Advancements in the curation and accessibility of electronic health record data (EHR) present the opportunity to utilize real-world data (RWD) to investigate therapeutic areas including oncology, where administrative healthcare claims databases alone are often not fit-for-purpose.
 The RCT DUPLICATE initiative has previously enhanced understanding of when RWE can most appropriately draw causal conclusions by emulating trials in non-oncology indications.
-Here, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to oncology.
+Here, we present the design and trial selection for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which extends this work to oncology.
 ENCORE is designed to emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
@@
 Randomized controlled trials (RCTs) are the go-to methodology for establishing the efficacy and safety of medical products.
-Under the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor consideration of real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 30% of new drug approvals, oncology was the disease area with the most FDA drug approvals in 2024[@senior2024fresh] and has many areas of high unmet medical need to treat serious conditions.
-Therefore, RWE has particularly important potential to complement evidence coming from RCTs in the field of oncology.
+Directed by the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor consideration of real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 30% of new drug approvals, oncology was the disease area with the most FDA drug approvals in 2024[@senior2024fresh] and has many areas of high unmet medical need to treat serious conditions.
+Therefore, RWE has particularly important potential to complement evidence from RCTs in the field of oncology.
 Potential use cases include the assessment of effectiveness in specific patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials that have feasibility or equipoise challenges, or the precision oncology-focused discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
@@
 However, the validity and transportability of results derived between RWE studies and RCTs depends on multiple factors.
-Frequently referenced limitations include, lack of baseline randomization, missing data, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.
+Frequently referenced limitations include lack of baseline randomization, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.
 
 The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
-In settings where the RCT designs could be emulated well, RWE studies were able to reach the same conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.
+In settings where the RCT designs could be emulated well, RWE studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.
 
@@
 
-The rationale and operationalization of each criterion is listed in detail in @tbl-criteria.
+The rationale and operationalization for each criterion is listed in detail in @tbl-criteria.
 We will mainly consider pivotal large late-phase interventional RCTs after 2011 because treatment guidelines among included cancer indications have undergone significant changes in recent years.
-Due to the rapid adoption of breakthrough therapies in routine care, it is unlikely to find patients who may be still treated with outdated treatment regimens in current clinical practice.
-Conversely, we excluded trials with results published too recently.
-To allow for enough data and follow-up time accrual in the real-world databases used for this project.
-Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which may not is often not reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges including missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
+Due to the rapid adoption of breakthrough therapies in routine care, few patients currently initiate outdated treatment regimens in current clinical practice.
+Conversely, we excluded trials with results published too recently to allow for enough data and follow-up time accrual in the real-world databases used for this project.
+We did not define a global cut-off as the requirements for follow-up time are different for each cancer type and population (e.g., advanced NSCLC versus early-stage breast cancer) and the decisions were made on a case-by-case basis.
+
+Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which is not well reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges including missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
 Therefore, we include only trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.

### 2025-03-30 manuscript/manuscript.tex
added updated word count

@@
 funded research center for the Centers for Medicare and Medicaid
-Services on unrelated work. Dr.~Schneeweiss is participating in
+Services on unrelated work. Dr.~Glynn has received support for
 investigator-initiated grants to the Brigham and Women's Hospital from
-Bayer and UCB unrelated to the topic of this study. He consults for and
-owns equity in Aetion Inc., a software manufacturer. He is an advisor to
-Temedica GmbH, a patient-oriented data generation company. His interests
-were declared, reviewed, and approved by the Brigham and Women's
-Hospital in accordance with their institutional compliance policies.
+Amarin, AstraZeneca, Kowa, Novartis, and Pfizer unrelated to the current
+work. Dr.~Schneeweiss is participating in investigator-initiated grants
+to the Brigham and Women's Hospital from Bayer and UCB unrelated to the
+topic of this study. He consults for and owns equity in Aetion Inc., a
+software manufacturer. He is an advisor to Temedica GmbH, a
+patient-oriented data generation company. His interests were declared,
+reviewed, and approved by the Brigham and Women's Hospital in accordance
+with their institutional compliance policies.
 
@@
 
-\emph{Manuscript last updated: 2025-03-30 15:14:45.088413}
+\emph{Manuscript last updated: 2025-03-30 19:11:35.894604}
 
@@
 
-Real-world evidence (RWE) is increasingly used to complement evidence
+Real-world evidence (RWE) is increasingly used to complement findings
 from randomized controlled trials (RCTs), contextualizing the
@@
 routine clinical practice. Advancements in the curation and
-accessibility of electronic health record data (EHR) have presented the
+accessibility of electronic health record data (EHR) present the
 opportunity to utilize real-world data (RWD) to investigate therapeutic
-areas such as oncology, where administrative healthcare claims databases
-alone are often not fit-for-purpose. The RCT DUPLICATE initiative has
-previously enhanced understanding of when RWE can most appropriately
-draw causal conclusions by emulating trials in non-oncology indications.
-Here, we present the Emulation of Comparative Oncology Trials with
-Real-world Evidence (ENCORE) project, which aims to extend this work to
+areas including oncology, where administrative healthcare claims
+databases alone are often not fit-for-purpose. The RCT DUPLICATE
+initiative has previously enhanced understanding of when RWE can most
+appropriately draw causal conclusions by emulating trials in
+non-oncology indications. Here, we present the design and trial
+selection for the Emulation of Comparative Oncology Trials with
+Real-world Evidence (ENCORE) project, which extends this work to
 oncology. ENCORE is designed to emulate 12 RCTs in four
@@
 Randomized controlled trials (RCTs) are the go-to methodology for
-establishing the efficacy and safety of medical products. Under the
-21\textsuperscript{st} Century Cures Act,\textsuperscript{1} the Food
-and Drug Administration (FDA) established a framework to factor
+establishing the efficacy and safety of medical products. Directed by
+the 21\textsuperscript{st} Century Cures Act,\textsuperscript{1} the
+Food and Drug Administration (FDA) established a framework to factor
 consideration of real-world evidence (RWE) generated from routine-care
@@
 treat serious conditions. Therefore, RWE has particularly important
-potential to complement evidence coming from RCTs in the field of
-oncology. Potential use cases include the assessment of effectiveness in
-specific patient populations that are underrepresented in RCTs, the
-construction of external control arms in single-arm trials that have
-feasibility or equipoise challenges, or the precision oncology-focused
-discovery of biomarkers among pan-tumor populations that harbor specific
-genomic and immuno-pathological signatures.
+potential to complement evidence from RCTs in the field of oncology.
+Potential use cases include the assessment of effectiveness in specific
+patient populations that are underrepresented in RCTs, the construction
+of external control arms in single-arm trials that have feasibility or
+equipoise challenges, or the precision oncology-focused discovery of
+biomarkers among pan-tumor populations that harbor specific genomic and
+immuno-pathological signatures.
 
@@
 RWE studies and RCTs depends on multiple factors. Frequently referenced
-limitations include, lack of baseline randomization, missing data, small
-study sizes, data discontinuity,\textsuperscript{4, 5} adoption of
-changes in guidelines in real-world care and the inability to measure
-and emulate common eligibility criteria, including prognostic factors,
-and standardized response assessments in real-world data
-(RWD).\textsuperscript{6} While examples of oncology trial emulations
-have been published,\textsuperscript{6--8} a systematic and scaled
-approach to emulate a diverse set of different oncology trials with
-multiple heterogeneous databases is necessary to gain confidence in the
-validity of RWE studies, evaluate regulatory considerations, and to
-provide context as to which questions can be validly answered.
+limitations include lack of baseline randomization, missing data,
+unmeasured confounding, small study sizes, data
+discontinuity,\textsuperscript{4, 5} adoption of changes in guidelines
+in real-world care and the inability to measure and emulate common
+eligibility criteria, including prognostic factors, and standardized
+response assessments in real-world data (RWD).\textsuperscript{6} While
+examples of oncology trial emulations have been
+published,\textsuperscript{6--8} a systematic and scaled approach to
+emulate a diverse set of oncology trials with multiple heterogeneous
+databases is necessary to gain confidence in the validity of RWE
+studies, evaluate regulatory considerations, and to provide context as
+to which questions can be validly answered.
 
@@
 settings where the RCT designs could be emulated well, RWE studies were
-able to reach the same conclusions.\textsuperscript{10} However, this
+able to reach similar conclusions.\textsuperscript{10} However, this
 prior work from RCT-DUPLICATE focused primarily on emulating trials in
@@
 
-The rationale and operationalization of each criterion is listed in
+The rationale and operationalization for each criterion is listed in
 detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
@@
 changes in recent years. Due to the rapid adoption of breakthrough
-therapies in routine care, it is unlikely to find patients who may be
-still treated with outdated treatment regimens in current clinical
-practice. Conversely, we excluded trials with results published too
-recently. To allow for enough data and follow-up time accrual in the
-real-world databases used for this project. Although there have been
-substantial methodological advancements to increase our understanding on
-the emulation and comparison of real-world progression-free survival
-(PFS) and objective response rates (ORR) to a
+therapies in routine care, few patients currently initiate outdated
+treatment regimens in current clinical practice. Conversely, we excluded
+trials with results published too recently to allow for enough data and
+follow-up time accrual in the real-world databases used for this
+project. We did not define a global cut-off as the requirements for
+follow-up time are different for each cancer type and population (e.g.,
+advanced NSCLC versus early-stage breast cancer) and the decisions were
+made on a case-by-case basis.
+
+Although there have been substantial methodological advancements to
+increase our understanding on the emulation and comparison of real-world
+progression-free survival (PFS) and objective response rates (ORR) to a
 RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
 RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
-level of granularity which may not is often not reflected in
-chart-abstracted assessments of a patient's progression in routine
+level of granularity which is not well reflected in chart-abstracted
+assessments of a patient's progression in routine
 care.\textsuperscript{18, 19} The timing and cadence of intervals

### 2025-03-30 manuscript/manuscript.qmd
typo

@@
 
-[**Manuscript word count:**]{.underline} 4,131 words / 8,000 words
+[**Manuscript word count:**]{.underline} 4,200 words / 8,000 words
 
-[**Abstract word count:**]{.underline} 248 words / 250 words
+[**Abstract word count:**]{.underline} 224 words / 250 words
 
@@
 
--   Primary purpose was study treatment efficacy
+-   Primary purpose was to study treatment efficacy
 

### 2025-05-27 manuscript/manuscript.qmd
Added paragraph emulating oncology RCT estimands as per Bob's comments and Shirley's excellent additions

@@
 
-[**Analytic code sharing statement:**]{.underline} Code to reproduce this manuscript, including all figures and tables, can be found at <https://github.com/janickweberpals/encore-process-manuscript>.
+[**Analytic code sharing statement:**]{.underline} Simulated examples and code to implement analytic workflows described in this manuscript are illustrated at <https://janickweberpals.github.io/imputation-ps-workflows/> and can be reproduced via the `encore.analytics` R package (<https://github.com/janickweberpals/encore.analytics/>).
 
@@
 This criterion involved extensive human review.
-The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., ECOG performance score) are measurable and whether preliminary study size is reasonable.
+The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., ECOG performance score) are measurable, and whether preliminary study size based on number of patients observed in the data with the combination of treatment regimen and line of treatment is reasonable.
 Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.
@@
 A key objective that we aim to explore with the selected trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD.
-If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.
+If more thorough feasibility assessments after applying all relevant eligibility criteria suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.
 
@@
 All eligibility criteria will be extracted based on publications, publicly available protocols, and statistical analysis plans of the selected RCT.
+
+**Emulating the RCT estimand**
+
+One important aspect when emulating oncology trials is the choice and estimation of the appropriate estimand of interest.[@rufibach2018] In prior RCT-DUPLICATE efforts, a “while on treatment” strategy was chosen for database studies that were designed to emulate trials where the primary trial analysis was intention-to-treat (e.g., a “treatment policy” estimand).
+This estimand was chosen to mimic the high adherence typically observed in large cardiovascular clinical trials in the context of low adherence in clinical practice.
+However, a “while on treatment” estimand would be inappropriate when emulating oncology trials, not only because of highly informative censoring, but also because OS outcome analyses in oncology are typically “treatment policy” estimands that allow crossover.[@manitz2022estimands] We plan to focus on “treatment policy” estimands for the primary analyses in our trial emulations.
+However, allowing crossover can dilute treatment effects, moving point estimates toward the null.
+Therefore, differences in cross-over and discontinuation rates between the trial and emulation may still contribute to estimand differences.
+Recognizing these practical challenges, we will carefully characterize the patterns of discontinuation or cross-over in both the trials and the database studies designed to emulate them.
 
@@
 
-One important aspect when emulating oncology trials is the choice and estimation of the appropriate estimand of interest.[@rufibach2018] Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered such as the contemporaneity of the control cohort, the adoption rate of the novel medical product in routine care, the magnitude of the clinical treatment benefit and the rate in which (particularly patients in the control arm) discontinue or cross-over to the interventional treatment, which could lastly bias emulated treatment effects towards the null.
+Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered, such as the contemporaneity of the control cohort, the adoption rate of the novel medical product in routine care, the magnitude of the clinical treatment benefit and the rate in which patients discontinue or cross-over, which could dilute effect estimates.
+
 To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial.
@@
 
-Another common challenge in the emulation of oncology trials is the estimation of an "intention-to-treat" (ITT) analogous estimand which is usually the primary estimand reported in oncology RCTs.
-Due to intercurrent events, such as non-adherence, crossover of a high proportion of patients from the control to the intervention arm, or differences in subsequent therapy lines there may be estimand differences between trial and emulation.[@manitz2022estimands] Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment practice in routine clinical might be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
+Another common challenge in the emulation of oncology trials is the estimation of an "treatment policy"/"intention-to-treat" analogous estimand which is usually the primary estimand reported in oncology RCTs.
+Due to intercurrent events, such as non-adherence, crossover of a high proportion of patients from the control to the intervention arm, or differences in subsequent therapy lines there may be estimand differences between trial and emulation.[@manitz2022estimands] Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment practice in routine clinical practice might be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
 

### 2025-05-27 manuscript/manuscript.qmd
added encore.analytics and more language on reproducibility

@@
 
-To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed an internal R package `encore.io` with parameterized functions.
-Detailed documentation can be found in the Supplementary Material.
+To ensure a transparent, reproducible and consistent way of deriving analytic cohorts and performing statistical analyses across trial emulations, we developed two R packages.
+The first package, `encore.io,` streamlines the query of analytic cohorts and measurement of covariates and outcomes using parameterized functions.
+Due to the blinding of databases used for ENCORE, this package cannot be made public but we provide a comprehensive and detailed documentation for each function in the Supplementary Material.
+A second open-source R package, `encore.analytics`, will then facilitate the complex multi-step workflows of multiple imputation, propensity score matching and weighting and estimating pooled marginal treatment effects and Kaplan-Meier curves along with other established statistical R packages.[@anesrake; @MatchThem; @survival; @encore.analytics; @gtsummary; @ggsurvfit] Simulated examples with code that illustrate the analytic workflows described in this manuscript are available in the online Supplementary Material at <https://janickweberpals.github.io/imputation-ps-workflows/> and can be reproduced via the `encore.analytics` R package (<https://github.com/janickweberpals/encore.analytics/>).
 

### 2025-05-28 manuscript/manuscript.qmd
additions on runner up clinical trial candidates for emulation

@@
 This criterion involved extensive human review.
-The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] including assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., ECOG performance score) are measurable, and whether preliminary study size based on number of patients observed in the data with the combination of treatment regimen and line of treatment is reasonable.
+The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] involving an assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., ECOG performance score) are measurable, and whether preliminary study size based on a rough estimation of the number of patients observed in the data with the combination of treatment regimen and line of treatment is reasonable for shortlisting the trial.
 Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.
 
-A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
-Naturally, the majority of trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts have recently focused on these settings.
+A list of tentative, shortlisted primary candidates is presented in @tbl-rcts.
+The corresponding selection process is illustrated in the CONSORT diagrams (**Supplementary Figures 1-4**) and runner up candidate trials are listed in **Supplementary Table 1**.
+Naturally, the majority of shortliested trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts have recently focused on these settings.
 A key objective that we aim to explore with the selected trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD.

### 2025-05-28 manuscript/supplement.qmd
additions on runner up clinical trial candidates for emulation

@@
 library(knitr)
+library(readxl)
+library(gt)
+library(dplyr)
 
@@
 
-# Supplementary Figures
+# Supplementary Figures and Tables
 
@@
 
+```{r}
+#| label: tbl-runner-ups
+#| tbl-cap: "Oncology trial emulation potential runner up candidates."
+#| tbl-cap-location: top
+
+# read excel blueprint
+runner_ups <- read_excel(
+  path = here("tables", "helper_tables.xlsx"),
+  sheet = "runner_ups"
+  )
+
+# convert to gt table
+runner_ups_gt <- runner_ups |> 
+  group_by(cancer) |> 
+  gt() |> 
+  cols_label(
+    nctid = md("**NCTID**"),
+    acronym = md("**Acronym**"),
+    setting = md("**Clinical setting**"),
+    line = md("**Line of therapy**"),
+    exposures = md("**Treatment comparison**")
+    ) |> 
+  tab_style(
+    style = cell_text(weight = "bold"),
+    locations = cells_row_groups()
+    )
+
+# save image to display within manuscript document
+suppressMessages(runner_ups_gt |> 
+  gtsave(
+    filename = here("tables", "STable_runner_ups.png"),
+    expand = 10, # amount of whitespace,
+    zoom = 5 # higher resolution
+    ))
+
+include_graphics(path = here("tables", "STable_runner_ups.png"))
+```
+
 {{< pagebreak >}}

### 2025-05-28 manuscript/manuscript.tex
Manuscript updated

@@
+% Options for packages loaded elsewhere
 % Options for packages loaded elsewhere
@@
   numbers=noendperiod]{scrartcl}
-
+\usepackage{xcolor}
 \usepackage{amsmath,amssymb}
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
@@
 \makeatother
-\usepackage{xcolor}
-\ifLuaTeX
-  \usepackage{luacolor}
-  \usepackage[soul]{lua-ul}
-\else
-  \usepackage{soul}
-
-\fi
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
 % Make \paragraph and \subparagraph free-standing
@@
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{longtable,booktabs,array}
 \usepackage{calc} % for calculating minipage widths
@@
 \makeatother
+
+\ifLuaTeX
+  \usepackage{luacolor}
+  \usepackage[soul]{lua-ul}
+\else
+  \usepackage{soul}
+\fi
+
 % definitions for citeproc citations
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 
@@
 \makeatother
-
 \usepackage{bookmark}
-
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 \date{}
-
 \begin{document}
@@
 
-\ul{\textbf{Manuscript word count:}} 4,131 words / 8,000 words
-
-\ul{\textbf{Abstract word count:}} 248 words / 250 words
+\ul{\textbf{Manuscript word count:}} 4,509 words / 8,000 words
+
+\ul{\textbf{Abstract word count:}} 224 words / 250 words
 
@@
 
-\ul{\textbf{Supplementary material:}} Supplementary figures and material
+\ul{\textbf{Supplementary material:}} Supplementary figures, tables and
+material
 
@@
 
-\ul{\textbf{Analytic code sharing statement:}} Code to reproduce this
-manuscript, including all figures and tables, can be found at
-\url{https://github.com/janickweberpals/encore-process-manuscript}.
+\ul{\textbf{Analytic code sharing statement:}} Simulated examples and
+code to implement analytic workflows described in this manuscript are
+illustrated at
+\url{https://janickweberpals.github.io/imputation-ps-workflows/} and can
+be reproduced via the \texttt{encore.analytics} R package
+(\url{https://github.com/janickweberpals/encore.analytics/}).
 
@@
 
-\emph{Manuscript last updated: 2025-03-30 19:11:35.894604}
+\emph{Manuscript last updated: 2025-05-28 16:51:13.630823}
 
@@
 \item
-  Primary purpose was study treatment efficacy
+  Primary purpose was to study treatment efficacy
 \item
@@
 initial feasibility assessment of the data fitness,\textsuperscript{19}
-including assessment of whether critical eligibility criteria (e.g.,
+involving an assessment of whether critical eligibility criteria (e.g.,
 biomarker status) including prognostic factors (e.g., ECOG performance
-score) are measurable and whether preliminary study size is reasonable.
-Lastly, trial candidates were ranked and shortlisted into primary and
-runner-up candidates based on their clinical and regulatory relevance.
+score) are measurable, and whether preliminary study size based on a
+rough estimation of the number of patients observed in the data with the
+combination of treatment regimen and line of treatment is reasonable for
+shortlisting the trial. Lastly, trial candidates were ranked and
+shortlisted into primary and runner-up candidates based on their
+clinical and regulatory relevance.
 
 A list of tentative, shortlisted primary candidates is presented in
-Table~\ref{tbl-rcts} and the corresponding selection process is
-illustrated in the CONSORT diagrams (Supplementary Figures 1-4).
-Naturally, the majority of trials cover advanced (locally advanced,
-inoperable, recurrent/progressive disease) or metastatic cancer
-populations because a large proportion of drug development efforts have
-recently focused on these settings. A key objective that we aim to
-explore with the selected trials is to achieve a better understanding
-how different disease settings (early, late), line settings
-({[}neo{]}adjuvant, first line, advanced lines of therapy), therapy
-protocols (monotherapy, combination therapy) and clinical
+Table~\ref{tbl-rcts}. The corresponding selection process is illustrated
+in the CONSORT diagrams (\textbf{Supplementary Figures 1-4}) and runner
+up candidate trials are listed in \textbf{Supplementary Table 1}.
+Naturally, the majority of shortliested trials cover advanced (locally
+advanced, inoperable, recurrent/progressive disease) or metastatic
+cancer populations because a large proportion of drug development
+efforts have recently focused on these settings. A key objective that we
+aim to explore with the selected trials is to achieve a better
+understanding how different disease settings (early, late), line
+settings ({[}neo{]}adjuvant, first line, advanced lines of therapy),
+therapy protocols (monotherapy, combination therapy) and clinical
 characteristics (all-comer versus complex genetic or immunological
 signatures) can be emulated using RWD. If more thorough feasibility
-assessments suggest that the threat of bias from mis-measurement of key
-study parameters or residual confounding remains high, or that the study
-size is not sufficient, then runner-up candidate trials will be
-considered instead.
+assessments after applying all relevant eligibility criteria suggest
+that the threat of bias from mis-measurement of key study parameters or
+residual confounding remains high, or that the study size is not
+sufficient, then runner-up candidate trials will be considered instead.
 
@@
 
+\textbf{Emulating the RCT estimand}
+
+One important aspect when emulating oncology trials is the choice and
+estimation of the appropriate estimand of interest.\textsuperscript{26}
+In prior RCT-DUPLICATE efforts, a ``while on treatment'' strategy was
+chosen for database studies that were designed to emulate trials where
+the primary trial analysis was intention-to-treat (e.g., a ``treatment
+policy'' estimand). This estimand was chosen to mimic the high adherence
+typically observed in large cardiovascular clinical trials in the
+context of low adherence in clinical practice. However, a ``while on
+treatment'' estimand would be inappropriate when emulating oncology
+trials, not only because of highly informative censoring, but also
+because OS outcome analyses in oncology are typically ``treatment
+policy'' estimands that allow crossover.\textsuperscript{27} We plan to
+focus on ``treatment policy'' estimands for the primary analyses in our
+trial emulations. However, allowing crossover can dilute treatment
+effects, moving point estimates toward the null. Therefore, differences
+in cross-over and discontinuation rates between the trial and emulation
+may still contribute to estimand differences. Recognizing these
+practical challenges, we will carefully characterize the patterns of
+discontinuation or cross-over in both the trials and the database
+studies designed to emulate them.
+
 \subsubsection{Emulation feasibility}\label{emulation-feasibility}
@@
 dates will be considered for each trial emulation to mitigate the
-potential impact of ghost-time bias.\textsuperscript{26}
+potential impact of ghost-time bias.\textsuperscript{28}
 
@@
 
-One important aspect when emulating oncology trials is the choice and
-estimation of the appropriate estimand of interest.\textsuperscript{27}
 Particularly when emulating pivotal trials of paradigm-changing
-treatments, multiple aspects need to be considered such as the
+treatments, multiple aspects need to be considered, such as the
 contemporaneity of the control cohort, the adoption rate of the novel
 medical product in routine care, the magnitude of the clinical treatment
-benefit and the rate in which (particularly patients in the control arm)
-discontinue or cross-over to the interventional treatment, which could
-lastly bias emulated treatment effects towards the null. To that end,
-comprehensive data explorations will be performed as part of the
-protocol development to contextualize these parameters and (if reported)
-draw comparisons to the emulated trial. Examples for such standard
-diagnostics are visualized in Figure~\ref{fig-initiators}. All
+benefit and the rate in which patients discontinue or cross-over, which
+could dilute effect estimates.
+
+To that end, comprehensive data explorations will be performed as part
+of the protocol development to contextualize these parameters and (if
+reported) draw comparisons to the emulated trial. Examples for such
+standard diagnostics are visualized in Figure~\ref{fig-initiators}. All
 exploratory analyses will be conducted blinded towards the outcome to
@@
 manner as RCTs since the number of `recruited' patients is limited by
-available data and cannot be influenced.\textsuperscript{28} For this
+available data and cannot be influenced.\textsuperscript{29} For this
 project, however, the emulations in EHR data must have at least equal
@@
 Schoenfeld's sample-size formula for the proportional-hazards regression
-model.\textsuperscript{29}
+model.\textsuperscript{30}
 
@@
 emulations, we will adapt the approach of the RCT-DUPLICATE
-project.\textsuperscript{9, 30} That is, for the primary endpoint of
+project.\textsuperscript{9, 31} That is, for the primary endpoint of
 interest (HR for OS and corresponding 95\% confidence intervals), we
@@
 windows of eligibility criteria, washout periods and covariates relative
-to the cohort entry time.\textsuperscript{31}
+to the cohort entry time.\textsuperscript{32}
 
@@
 missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
-missing not at random {[}MNAR{]}).\textsuperscript{32} We will adopt a
+missing not at random {[}MNAR{]}).\textsuperscript{33} We will adopt a
 principled process on missing data that empirically evaluates different
 aspects across partially observed covariates based on three group
-diagnostics.\textsuperscript{33, 34} These diagnostics cover (1)
+diagnostics.\textsuperscript{34, 35} These diagnostics cover (1)
 comparisons of patients characteristics with and without an observed
@@
 domain knowledge and assumptions about the underlying missing data
-structure through canonical causal diagrams\textsuperscript{35} will
+structure through canonical causal diagrams\textsuperscript{36} will
 additionally inform decisions regarding the inclusion or exclusion of
@@
 sensitivity analyses to assess the robustness of these
-decisions.\textsuperscript{36}
+decisions.\textsuperscript{37}
 
@@
 outcome data), only mechanisms in which a covariate causes its own
-missingness lead to critical bias (MNAR).\textsuperscript{35} Hence,
+missingness lead to critical bias (MNAR).\textsuperscript{36} Hence,
 methodologies which retain patients and give the potential to adjust for
 a broader set of prognostic factors (e.g., multiple
-imputation\textsuperscript{37} or doubly robust
-methods\textsuperscript{38}) may be preferred over complete case
+imputation\textsuperscript{38} or doubly robust
+methods\textsuperscript{39}) may be preferred over complete case
 analyses.
@@
 interest in ENCORE will be defined as the marginal hazard ratio (HR)
-coefficient for the treatment comparison for OS.\textsuperscript{39} We
+coefficient for the treatment comparison for OS.\textsuperscript{40} We
 will also consider alternative endpoints on an absolute risk scale such
 as median survival times, survival probabilities at pre-defined time
-points during follow-up\textsuperscript{40} and restricted mean survival
+points during follow-up\textsuperscript{41} and restricted mean survival
 times as secondary endpoints of interest.
@@
 based on expert clinical knowledge and published literature on
-prognostic scores in oncology.\textsuperscript{41} The implementation of
+prognostic scores in oncology.\textsuperscript{42} The implementation of
 propensity scores in combination with multiple imputation will follow
 the `\emph{within}' methodology as described by Leyrat et
-al.\textsuperscript{42, 43} That is, propensity score matching or
+al.\textsuperscript{43, 44} That is, propensity score matching or
 weighting will be applied to each imputed dataset. The marginal
@@
 weighted dataset separately and pooled into a final estimate following
-Rubin's rule.\textsuperscript{44, 45} This approach has been shown to
+Rubin's rule.\textsuperscript{45, 46} This approach has been shown to
 lead to unbiased estimates across different simulated scenarios with a
-sufficient estimation of the variance.\textsuperscript{42}
+sufficient estimation of the variance.\textsuperscript{43}
 
@@
 Figure~\ref{fig-balance}). Covariate balance is often reasonably
-considered at a SMD \textless{} 0.1.\textsuperscript{46} Further, we
+considered at a SMD \textless{} 0.1.\textsuperscript{47} Further, we
 will compute the average post-matching or post-weighting
-C-statistics.\textsuperscript{47} In addition, we will use a published
-prognostic score for OS\textsuperscript{41} as a balance measure to
+C-statistics.\textsuperscript{48} In addition, we will use a published
+prognostic score for OS\textsuperscript{42} as a balance measure to
 visually assess if the prognostic score is balanced between treatment
@@
 balance measures and not affected by model
-misspecification.\textsuperscript{48}
+misspecification.\textsuperscript{49}
 
@@
 estimated in each imputed and propensity score matched or weighted
-dataset according to the Kaplan-Meier method.\textsuperscript{40} Since
+dataset according to the Kaplan-Meier method.\textsuperscript{41} Since
 survival probabilities typically do not follow normal distributions
@@
 with \(pr(surv)\) denoting the survival probability at a given time
-during follow-up.\textsuperscript{49, 50} The transformed survival
+during follow-up.\textsuperscript{50, 51} The transformed survival
 probabilities are then pooled across imputed datasets and individual
@@
 analytic cohorts and performing statistical analyses across trial
-emulations, we developed an internal R package \texttt{encore.io} with
-parameterized functions. Detailed documentation can be found in the
-Supplementary Material.
+emulations, we developed two R packages. The first package,
+\texttt{encore.io,} streamlines the query of analytic cohorts and
+measurement of covariates and outcomes using parameterized functions.
+Due to the blinding of databases used for ENCORE, this package cannot be
+made public but we provide a comprehensive and detailed documentation
+for each function in the Supplementary Material. A second open-source R
+package, \texttt{encore.analytics}, will then facilitate the complex
+multi-step workflows of multiple imputation, propensity score matching
+and weighting and estimating pooled marginal treatment effects and
+Kaplan-Meier curves along with other established statistical R
+packages.\textsuperscript{52--57} Simulated examples with code that
+illustrate the analytic workflows described in this manuscript are
+available in the online Supplementary Material at
+\url{https://janickweberpals.github.io/imputation-ps-workflows/} and can
+be reproduced via the \texttt{encore.analytics} R package
+(\url{https://github.com/janickweberpals/encore.analytics/}).
 
@@
 Building on an established approach developed by the RCT DUPLICATE
-project\textsuperscript{9, 30}, the ENCORE project aims to emulate 12
+project\textsuperscript{9, 31}, the ENCORE project aims to emulate 12
 oncology trials using four oncology EHR data sources to inform the
@@
 
-The RCT DUPLICATE initiative\textsuperscript{9, 30} identified multiple
+The RCT DUPLICATE initiative\textsuperscript{9, 31} identified multiple
 emulation challenges that may be similar in ENCORE. One particular
@@
 Another common challenge in the emulation of oncology trials is the
-estimation of an ``intention-to-treat'' (ITT) analogous estimand which
-is usually the primary estimand reported in oncology RCTs. Due to
-intercurrent events, such as non-adherence, crossover of a high
-proportion of patients from the control to the intervention arm, or
+estimation of an ``treatment policy''/``intention-to-treat'' analogous
+estimand which is usually the primary estimand reported in oncology
+RCTs. Due to intercurrent events, such as non-adherence, crossover of a
+high proportion of patients from the control to the intervention arm, or
 differences in subsequent therapy lines there may be estimand
-differences between trial and emulation.\textsuperscript{51} Although
+differences between trial and emulation.\textsuperscript{27} Although
 this is also a common challenge in the analysis of
-RCTs,\textsuperscript{27} treatment practice in routine clinical might
-be less stringent compared to trials.\textsuperscript{8} In order to
-derive comparable estimands it is therefore crucial to understand and
-contextualize the proportion and timing of treatment switching and
-discontinuation in both the trial and its emulation.
+RCTs,\textsuperscript{26} treatment practice in routine clinical
+practice might be less stringent compared to trials.\textsuperscript{8}
+In order to derive comparable estimands it is therefore crucial to
+understand and contextualize the proportion and timing of treatment
+switching and discontinuation in both the trial and its emulation.
 
@@
 Alternative analytic approaches which target a per-protocol estimand,
-such as the clone-censor-weight design,\textsuperscript{52} may be
+such as the clone-censor-weight design,\textsuperscript{58} may be
 viable options if the ITT estimand cannot be estimated due to the
@@
 
+\bibitem[\citeproctext]{ref-rufibach2018}
+\textbf{26}. Rufibach K: Treatment effect quantification for
+time{-}to{-}event endpoints{\textendash}Estimands, analysis strategies,
+and beyond {[}Internet{]}. Pharmaceutical Statistics 18:145--165,
+2018Available from: \url{http://dx.doi.org/10.1002/pst.1917}
+
+\bibitem[\citeproctext]{ref-manitz2022estimands}
+\textbf{27}. Manitz J, Kan-Dobrosky N, Buchner H, et al: Estimands for
+overall survival in clinical trials with treatment switching in
+oncology. Pharmaceutical Statistics 21:150--162, 2022
+
 \bibitem[\citeproctext]{ref-meyer2020open}
-\textbf{26}. Meyer A-M, Davies J, Taylor M, et al: Open cohorts and
+\textbf{28}. Meyer A-M, Davies J, Taylor M, et al: Open cohorts and
 ghost-time bias in real world data, in PHARMACOEPIDEMIOLOGY AND DRUG
@@
 
-\bibitem[\citeproctext]{ref-rufibach2018}
-\textbf{27}. Rufibach K: Treatment effect quantification for
-time{-}to{-}event endpoints{\textendash}Estimands, analysis strategies,
-and beyond {[}Internet{]}. Pharmaceutical Statistics 18:145--165,
-2018Available from: \url{http://dx.doi.org/10.1002/pst.1917}
-
 \bibitem[\citeproctext]{ref-hernan2022causal}
-\textbf{28}. Hernán MA: Causal analyses of existing databases: No power
+\textbf{29}. Hernán MA: Causal analyses of existing databases: No power
 calculations required. Journal of clinical epidemiology 144:203--205,
@@
 \bibitem[\citeproctext]{ref-schoenfeld1981asymptotic}
-\textbf{29}. Schoenfeld D: The asymptotic properties of nonparametric
+\textbf{30}. Schoenfeld D: The asymptotic properties of nonparametric
 tests for comparing survival distributions. Biometrika 68:316--319, 1981
@@
 \bibitem[\citeproctext]{ref-franklin2020nonrandomized}
-\textbf{30}. Franklin JM, Pawar A, Martin D, et al: Nonrandomized
+\textbf{31}. Franklin JM, Pawar A, Martin D, et al: Nonrandomized
 real-world evidence to support regulatory decision making: Process for a
@@
 \bibitem[\citeproctext]{ref-schneeweiss2019graphical}
-\textbf{31}. Schneeweiss S, Rassen JA, Brown JS, et al: Graphical
+\textbf{32}. Schneeweiss S, Rassen JA, Brown JS, et al: Graphical
 depiction of longitudinal study designs in health care databases. Annals
@@
 \bibitem[\citeproctext]{ref-rubin1976inference}
-\textbf{32}. Rubin DB: Inference and missing data. Biometrika
+\textbf{33}. Rubin DB: Inference and missing data. Biometrika
 63:581--592, 1976
@@
 \bibitem[\citeproctext]{ref-weberpals2024smdi}
-\textbf{33}. Weberpals J, Raman SR, Shaw PA, et al: Smdi: An r package
+\textbf{34}. Weberpals J, Raman SR, Shaw PA, et al: Smdi: An r package
 to perform structural missing data investigations on partially observed
@@
 \bibitem[\citeproctext]{ref-weberpals2024}
-\textbf{34}. Weberpals J, Raman SR, Shaw PA, et al: A principled
+\textbf{35}. Weberpals J, Raman SR, Shaw PA, et al: A principled
 approach to characterize and analyze partially observed confounder data
@@
 16:329--343, 2024Available
-from:\href{\%20\%0A\%20\%20\%20\%20\%0A\%20\%20\%20\%20\%0A\%20\%20\%20\%20\%20\%20\%20\%20https://www.tandfonline.com/doi/abs/10.2147/CLEP.S436131\%0A\%20\%20\%20\%20\%0A\%0A}{
-https://www.tandfonline.com/doi/abs/10.2147/CLEP.S436131 }
+from:\href{\%20\%0A\%20\%20\%20\%20\%0A\%20\%20\%20\%20\%0A\%20\%20\%20\%20\%20\%20\%20\%20https://www.tandfonline.com/doi/abs/10.2147/CLEP.S436131\%0A\%20\%20\%20\%20\%0A\%0A}{https://www.tandfonline.com/doi/abs/10.2147/CLEP.S436131
+}
 
 \bibitem[\citeproctext]{ref-moreno2018canonical}
-\textbf{35}. Moreno-Betancur M, Lee KJ, Leacy FP, et al: Canonical
+\textbf{36}. Moreno-Betancur M, Lee KJ, Leacy FP, et al: Canonical
 causal diagrams to guide the treatment of missing data in epidemiologic
@@
 \bibitem[\citeproctext]{ref-tompsett2018use}
-\textbf{36}. Tompsett DM, Leacy F, Moreno-Betancur M, et al: On the use
+\textbf{37}. Tompsett DM, Leacy F, Moreno-Betancur M, et al: On the use
 of the not-at-random fully conditional specification (NARFCS) procedure
@@
 \bibitem[\citeproctext]{ref-weberpals2024hdmi}
-\textbf{37}. Weberpals J, Shaw PA, Lin KJ, et al: High-dimensional
+\textbf{38}. Weberpals J, Shaw PA, Lin KJ, et al: High-dimensional
 multiple imputation (HDMI) for partially observed confounders including
@@
 \bibitem[\citeproctext]{ref-Shaw2024}
-\textbf{38}. Shaw CK P., Williamson BD: Assessing treatment effects in
+\textbf{39}. Shaw CK P., Williamson BD: Assessing treatment effects in
 observational data with missing confounders: A comparative study of
@@
 \bibitem[\citeproctext]{ref-cox1972regression}
-\textbf{39}. Cox DR: Regression models and life-tables. Journal of the
+\textbf{40}. Cox DR: Regression models and life-tables. Journal of the
 royal statistical society. Series B (Methodological) 34:187--220, 1972
@@
 \bibitem[\citeproctext]{ref-kaplan1958nonparametric}
-\textbf{40}. Kaplan EL, Meier P: Nonparametric estimation from
+\textbf{41}. Kaplan EL, Meier P: Nonparametric estimation from
 incomplete observations. Journal of the American statistical association
@@
 \bibitem[\citeproctext]{ref-becker2020enhanced}
-\textbf{41}. Becker T, Weberpals J, Jegg A, et al: An enhanced
+\textbf{42}. Becker T, Weberpals J, Jegg A, et al: An enhanced
 prognostic score for overall survival of patients with cancer derived
@@
 \bibitem[\citeproctext]{ref-leyrat2019propensity}
-\textbf{42}. Leyrat C, Seaman SR, White IR, et al: Propensity score
+\textbf{43}. Leyrat C, Seaman SR, White IR, et al: Propensity score
 analysis with partially observed covariates: How should multiple
@@
 \bibitem[\citeproctext]{ref-pishgar2020matchthem}
-\textbf{43}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
+\textbf{44}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
 and weighting after multiple imputation. arXiv preprint arXiv:200911772,
@@
 \bibitem[\citeproctext]{ref-rubin2018multiple}
-\textbf{44}. Rubin DB: Multiple imputation, in Flexible imputation of
+\textbf{45}. Rubin DB: Multiple imputation, in Flexible imputation of
 missing data, second edition. Chapman; Hall/CRC, 2018, pp 29--62
@@
 \bibitem[\citeproctext]{ref-van2011mice}
-\textbf{45}. Van Buuren S, Groothuis-Oudshoorn K: Mice: Multivariate
+\textbf{46}. Van Buuren S, Groothuis-Oudshoorn K: Mice: Multivariate
 imputation by chained equations in r. Journal of statistical software
@@
 \bibitem[\citeproctext]{ref-austin2009balance}
-\textbf{46}. Austin PC: Balance diagnostics for comparing the
+\textbf{47}. Austin PC: Balance diagnostics for comparing the
 distribution of baseline covariates between treatment groups in
@@
 \bibitem[\citeproctext]{ref-franklin2014metrics}
-\textbf{47}. Franklin JM, Rassen JA, Ackermann D, et al: Metrics for
+\textbf{48}. Franklin JM, Rassen JA, Ackermann D, et al: Metrics for
 covariate balance in cohort studies of causal effects. Statistics in
@@
 \bibitem[\citeproctext]{ref-stuart2013prognostic}
-\textbf{48}. Stuart EA, Lee BK, Leacy FP: Prognostic score--based
+\textbf{49}. Stuart EA, Lee BK, Leacy FP: Prognostic score--based
 balance measures can be a useful diagnostic for propensity score methods
@@
 \bibitem[\citeproctext]{ref-warwick28685}
-\textbf{49}. Marshall A(Andrea), Billingham LJ, Bryan S: Can we afford
+\textbf{50}. Marshall A(Andrea), Billingham LJ, Bryan S: Can we afford
 to ignore missing data in cost-effectiveness analyses? {[}Internet{]}.
@@
 \bibitem[\citeproctext]{ref-morisot2015prostate}
-\textbf{50}. Morisot A, Bessaoud F, Landais P, et al: Prostate cancer:
+\textbf{51}. Morisot A, Bessaoud F, Landais P, et al: Prostate cancer:
 Net survival and cause-specific survival rates after multiple
@@
 
-\bibitem[\citeproctext]{ref-manitz2022estimands}
-\textbf{51}. Manitz J, Kan-Dobrosky N, Buchner H, et al: Estimands for
-overall survival in clinical trials with treatment switching in
-oncology. Pharmaceutical Statistics 21:150--162, 2022
+\bibitem[\citeproctext]{ref-anesrake}
+\textbf{52}. Pasek J: Anesrake: ANES raking implementation, 2018
+
+\bibitem[\citeproctext]{ref-MatchThem}
+\textbf{53}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
+and weighting after multiple imputation {[}Internet{]}, 2021Available
+from: \url{https://journal.r-project.org/archive/2021/RJ-2021-073/}
+
+\bibitem[\citeproctext]{ref-survival}
+\textbf{54}. Therneau TM: A package for survival analysis in r
+{[}Internet{]}, 2024Available from:
+\url{https://CRAN.R-project.org/package=survival}
+
+\bibitem[\citeproctext]{ref-encore.analytics}
+\textbf{55}. Weberpals J: Encore.analytics: Functions and wrappers to
+streamline complex analytic workflows in real-world data studies based
+on the ENCORE trial emulation project {[}Internet{]}, 2025Available
+from: \url{https://github.com/janickweberpals/encore.analytics}
+
+\bibitem[\citeproctext]{ref-gtsummary}
+\textbf{56}. Sjoberg DD, Whiting K, Curry M, et al: Reproducible summary
+tables with the gtsummary package {[}Internet{]}13:570--580,
+2021Available from: \url{https://doi.org/10.32614/RJ-2021-053}
+
+\bibitem[\citeproctext]{ref-ggsurvfit}
+\textbf{57}. Sjoberg DD, Baillie M, Fruechtenicht C, et al: Ggsurvfit:
+Flexible time-to-event figures {[}Internet{]}, 2024Available from:
+\url{https://github.com/pharmaverse/ggsurvfit}
 
 \bibitem[\citeproctext]{ref-gaber2024mystifying}
-\textbf{52}. Gaber CE, Ghazarian AA, Strassle PD, et al: De-mystifying
+\textbf{58}. Gaber CE, Ghazarian AA, Strassle PD, et al: De-mystifying
 the clone-censor-weight method for causal research using observational

### 2025-05-28 manuscript/manuscript.qmd
updated word count on title page after test run
typo

@@
 
-[**Manuscript word count:**]{.underline} 4,200 words / 8,000 words
+[**Manuscript word count:**]{.underline} 4,509 words / 8,000 words
 
@@
 
-[**Supplementary material:**]{.underline} Supplementary figures and material
+[**Supplementary material:**]{.underline} Supplementary figures, tables and material
 
@@
 The corresponding selection process is illustrated in the CONSORT diagrams (**Supplementary Figures 1-4**) and runner up candidate trials are listed in **Supplementary Table 1**.
-Naturally, the majority of shortliested trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts have recently focused on these settings.
+Naturally, the majority of shortlisted trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts have recently focused on these settings.
 A key objective that we aim to explore with the selected trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD.

### 2025-05-28 manuscript/manuscript.tex
Manuscript updated

@@
 
-\emph{Manuscript last updated: 2025-05-28 16:51:13.630823}
+\emph{Manuscript last updated: 2025-05-28 17:11:23.084986}
 
@@
 up candidate trials are listed in \textbf{Supplementary Table 1}.
-Naturally, the majority of shortliested trials cover advanced (locally
+Naturally, the majority of shortlisted trials cover advanced (locally
 advanced, inoperable, recurrent/progressive disease) or metastatic

### 2025-07-06 manuscript/manuscript.qmd
moving Sebastian to second author position
COTA review round 2

@@
 
-**Authors**: Janick Weberpals^1^, Kenneth L.
+**Authors**: Janick Weberpals^1^, Sebastian Schneeweiss^1^, Kenneth L.
 Kehl^2^, Donna R.
@@
 Belli^6^, Ching-Kun Wang^6^, Jenna Collins^7^**,** Jonathan Kish^7^**,** Janet Espirito^8^, Nicholas J Robert^8^, Robert J.
-Glynn^1^, Sebastian Schneeweiss^1^, Shirley V.
+Glynn^1^, Shirley V.
 Wang^1^
@@
 To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR for OS and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD).
-Examples are illustrated @tbl-metrics.
+Examples are illustrated in @tbl-metrics.
 
@@
 This may be a factor especially given that the pathophysiology, prognosis, and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries.
-For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are much more prevalent in certain geographic regions compared to the US.
+For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are much more prevalent in certain geographic regions and populations compared to the US.
 

### 2025-07-06 manuscript/manuscript.qmd
added FDA round 2 comments and edits

@@
 ---
-title: "A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data"
+title: "Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data"
 
@@
 Rivera^3^, Pallavi Mishra-Kalyani^3^, Catherine C.
-Lerro^3^, Erin Larkins^4^, Sonia Singh^4^, Preeti Narayan^4^, Richard Curley^4^, Paul Kluetz^4^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^5^, Andrew J.
+Lerro^3^, Erin Larkins^4^, Sonia Singh^4\*^, Preeti Narayan^4^, Richard Curley^4^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^5^, Andrew J.
 Belli^6^, Ching-Kun Wang^6^, Jenna Collins^7^**,** Jonathan Kish^7^**,** Janet Espirito^8^, Nicholas J Robert^8^, Robert J.
@@
 ^8^ Ontada, Boston, MA, USA
+
+\*At the time this research was conducted
 
@@
 Advancements in the curation and accessibility of electronic health record data (EHR) present the opportunity to utilize real-world data (RWD) to investigate therapeutic areas including oncology, where administrative healthcare claims databases alone are often not fit-for-purpose.
-The RCT DUPLICATE initiative has previously enhanced understanding of when RWE can most appropriately draw causal conclusions by emulating trials in non-oncology indications.
+The RCT DUPLICATE initiative has previously enhanced understanding of when RWE can most appropriately draw causal conclusions by emulating trials for non-oncology indications.
 Here, we present the design and trial selection for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which extends this work to oncology.
@@
 Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria.
-Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence the validity of oncological RWE studies.
+Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence the interpretation of results from emulated oncology trials using RWD.
 
@@
 
-Randomized controlled trials (RCTs) are the go-to methodology for establishing the efficacy and safety of medical products.
-Directed by the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor consideration of real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 30% of new drug approvals, oncology was the disease area with the most FDA drug approvals in 2024[@senior2024fresh] and has many areas of high unmet medical need to treat serious conditions.
-Therefore, RWE has particularly important potential to complement evidence from RCTs in the field of oncology.
-Potential use cases include the assessment of effectiveness in specific patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials that have feasibility or equipoise challenges, or the precision oncology-focused discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
+Randomized controlled trials (RCTs) are the evidentiary gold standard methodology for establishing the efficacy and safety of medical products.
+Guided by the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor considerations for use of real-world evidence (RWE) generated from real-world data (RWD) such as electronic health records (EHR) to support approval of new indications or satisfy postmarketing requirements.[@RWEFDA] Accounting for approximately 30% of new FDA drug approvals, oncology was the disease area with the most new approvals in 2024[@mullard20242023] as well as several indication expansions, and has many areas of high unmet medical need to treat serious conditions.
+Therefore, RWE can have a particularly important potential to complement evidence from RCTs in the field of oncology.
+Potential uses include the assessment of effectiveness in specific patient populations that are not adequately represented in RCTs, or the precision oncology-focused discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.
 
-However, the validity and transportability of results derived between RWE studies and RCTs depends on multiple factors.
-Frequently referenced limitations include lack of baseline randomization, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.
+However, the comparability and transportability of results derived between observational studies and RCTs depends on multiple factors.
+Frequently referenced limitations include lack of baseline randomization and the imbalance in prognostic factors resulting thereof, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in treatment guidelines over time in routine clinical care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to increase confidence in the interpretability of observational studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.
 
-The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
-In settings where the RCT designs could be emulated well, RWE studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.
+The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when observational studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
+In settings where the RCT designs could be emulated well, observational studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.
 
 The *Emulation of Comparative Oncology Trials with Real-world Evidence* (ENCORE) project [@encoreFDA] aims to extend this work to oncology.
-Studies in oncology come with their own unique set of methodological challenges which must be systematically explored and understood.
+Clinical studies in oncology come with unique methodological challenges which must be systematically explored and understood.
 Building on a process co-developed with the FDA through RCT DUPLICATE,[@wang2023emulation] this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources.
-The process will emphasize transparency and include documented data fit-for-purpose assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.
+The process will emphasize transparency and include a data fitness-for-use assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.
 
-The objectives of this project are to 1) develop state-of-the-art methodological approaches and 2) to apply them to create insights into the potential use of RWE to enhance regulatory science in oncology.
+The objectives of this project are to 1) develop state-of-the-art methodological approaches and 2) apply these methods to gain insights into the potential use of RWE to enhance regulatory and clinical decision-making in oncology.
 To achieve these objectives, this demonstration project will systematically emulate 12 oncology trials across four cancer types and assess the agreement of treatment effect estimates between RCTs and their respective emulations.
 
-Here, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development and pre-registration, study design and statistical analysis, and pre-specified agreement metrics to evaluate the concordance between RCTs and emulations.
+Here, we describe the design and process for the selection of the 12 oncology RCTs, assessment of the database quality and selection, protocol development and pre-registration, study design and statistical analysis plans, and pre-specified agreement metrics to evaluate the concordance between RCTs and emulations.
 
@@
 
-The focus of ENCORE is to efficiently explore in what contexts RWE studies can or cannot yield similar results compared to RCTs.
+The focus of ENCORE is to efficiently explore in what contexts observational studies can or cannot yield similar results compared to oncology RCTs.
 Therefore, the project emphasizes trials of therapies for common cancers for which there has been substantial therapeutic development in recent years.
@@
 
-We used a systematic process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs for each cancer type.
+We used a systematic process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs by cancer type.
 The search was conducted using the *Aggregated Analysis of ClinicalTrials.gov* (AACT) database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains information (protocol and result data elements) about studies registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *Medical Subject Headings **(***MeSH) term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.
@@
 
--   Intervention model: parallel assignment
+-   Intervention model: Parallel assignment
 
@@
 
--   Trial start in 2011 or later
+-   Trial start: 2011 or later
 
--   Primary purpose was to study treatment efficacy
+-   Primary purpose: Study treatment efficacy
 
--   Overall survival must be one of the endpoints reported (either as hazard ratio or median overall survival time)
+-   Endpoint(s): Overall survival must be one of the endpoints reported (either as hazard ratio or median overall survival time)
 
@@
 The rationale and operationalization for each criterion is listed in detail in @tbl-criteria.
-We will mainly consider pivotal large late-phase interventional RCTs after 2011 because treatment guidelines among included cancer indications have undergone significant changes in recent years.
-Due to the rapid adoption of breakthrough therapies in routine care, few patients currently initiate outdated treatment regimens in current clinical practice.
-Conversely, we excluded trials with results published too recently to allow for enough data and follow-up time accrual in the real-world databases used for this project.
-We did not define a global cut-off as the requirements for follow-up time are different for each cancer type and population (e.g., advanced NSCLC versus early-stage breast cancer) and the decisions were made on a case-by-case basis.
+We mainly considered pivotal large late-phase RCTs that were initiated after 2011 because treatment guidelines for included cancer indications have undergone significant changes in recent years.
+Due to the rapid adoption of newly-approved therapies in routine care, patients are less likely to receive outdated treatment regimens in clinical practice.
+Conversely, we excluded trials with results published too recently to allow for substantive medical product uptake and follow-up time accrual in real-world databases used for this project.
+We did not define a global cut-off as the requirements for follow-up time are different for each cancer type and population (e.g., advanced NSCLC versus early-stage breast cancer) and the decisions were made on a clinically relevant basis.
 
-Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which is not well reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges including missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
-Therefore, we include only trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.
+Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still require a level of granularity which is not well reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges including missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest.
+Therefore, we include only trials that have reported OS as a pre-specified endpoint in the protocol.
 
@@
 This criterion involved extensive human review.
-The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] involving an assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., ECOG performance score) are measurable, and whether preliminary study size based on a rough estimation of the number of patients observed in the data with the combination of treatment regimen and line of treatment is reasonable for shortlisting the trial.
-Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.
+The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] involving an assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., Eastern Cooperative Oncology Group \[ECOG\] performance score) are measurable, and whether preliminary study size based on a rough estimation of the number of patients observed in the data with the combination of treatment regimen and line of therapy is reasonable for shortlisting the trial.
+Lastly, trial candidates were ranked and shortlisted as primary or runner-up based on their clinical and regulatory relevance.
 
@@
 The corresponding selection process is illustrated in the CONSORT diagrams (**Supplementary Figures 1-4**) and runner up candidate trials are listed in **Supplementary Table 1**.
-Naturally, the majority of shortlisted trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts have recently focused on these settings.
-A key objective that we aim to explore with the selected trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD.
+The majority of shortlisted trials represent indications for advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer patients because a large proportion of drug development efforts have recently focused on these settings.
+A key objective that we aim to explore with the selected trials is to evaluate how stage (early, late), line of therapy (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD.
 If more thorough feasibility assessments after applying all relevant eligibility criteria suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.
@@
 
-The ENCORE project will use data from four oncology-specific electronic health records (EHR)-derived RWD data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, and Ontada/McKesson.
-The included databases comprise the necessary information to study medical product effectiveness in oncology derived from US-based oncology EHRs.
+The ENCORE project will use data from four U.S.-based oncology-specific EHR-derived RWD data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, and Ontada/McKesson.
 A detailed description and sampling methodology will be provided with each trial emulation protocol.
-For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (the numbering does not coincide with the above order of mention of the databases).
-If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]
+For ENCORE, not all databases will be available for each cancer indication given the specificity of data elements required and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (in randomly assigned order).
+If more than one database is considered fit-for-use for a respective trial emulation, the most suitable analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]
 
@@
 
-For each selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] recommended for regulatory submissions of RWE studies[@guidelinegeneral] and will be registered on ClinicalTrials.gov after review by a clinical and FDA regulatory expert panel.
-Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used for agreement metric evaluation (see @sec-agreement-metrics).
+For each selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized;@guidelinegeneral] and will be registered on ClinicalTrials.gov after review by an expert panel comprising clinicians and FDA reveiwers.
+Following the target trial emulation framework, we will provide an explicit description and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update survival estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used for agreement metric evaluation (see @sec-agreement-metrics).
 All eligibility criteria will be extracted based on publications, publicly available protocols, and statistical analysis plans of the selected RCT.
@@
 
-One important aspect when emulating oncology trials is the choice and estimation of the appropriate estimand of interest.[@rufibach2018] In prior RCT-DUPLICATE efforts, a “while on treatment” strategy was chosen for database studies that were designed to emulate trials where the primary trial analysis was intention-to-treat (e.g., a “treatment policy” estimand).
+An important aspect when emulating oncology trials is the choice and estimation of the appropriate estimand of interest.[@rufibach2018] In prior RCT-DUPLICATE efforts, a “while on treatment” strategy was chosen for database studies that were designed to emulate trials where the primary trial analysis was intention-to-treat (e.g., a “treatment policy” estimand).
 This estimand was chosen to mimic the high adherence typically observed in large cardiovascular clinical trials in the context of low adherence in clinical practice.
-However, a “while on treatment” estimand would be inappropriate when emulating oncology trials, not only because of highly informative censoring, but also because OS outcome analyses in oncology are typically “treatment policy” estimands that allow crossover.[@manitz2022estimands] We plan to focus on “treatment policy” estimands for the primary analyses in our trial emulations.
+However, a “while on treatment” estimand would be inappropriate when emulating oncology trials, not only because of highly informative censoring, but also because OS outcome analyses in oncology are typically “treatment policy” estimands that allow for crossover or other post-progression antineoplastic therapy.[@manitz2022estimands] We plan to focus on “treatment policy” estimands for the primary analyses in our trial emulations.
 However, allowing crossover can dilute treatment effects, moving point estimates toward the null.
-Therefore, differences in cross-over and discontinuation rates between the trial and emulation may still contribute to estimand differences.
+Therefore, differences in cross-over and discontinuation rates between the trial and emulation may still contribute to differences in observed outcomes.
 Recognizing these practical challenges, we will carefully characterize the patterns of discontinuation or cross-over in both the trials and the database studies designed to emulate them.
@@
 
-Real-world data fitness and emulation feasibility for each shortlisted candidate trial will be assessed in multiple steps based on direction from of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure, line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care.
-Since many oncological RCTs in recent years have focused on selected, biomarker-defined populations, nuances in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population.
+Real-world data fitness and emulation feasibility for each candidate trial will be assessed in multiple steps based on the approach described by the Oncology Quality, Characterization, and Assessment of Real-World Data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure, line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care.
+Since many oncology RCTs in recent years have focused on biomarker-defined populations, nuances in measurement and operationalizability of specific biomarkers must be understood to ensure a representative and adequately sized study population.
 For example, immune checkpoint inhibitors have significantly changed the cancer treatment landscape since the first approvals of the CTLA-4 inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and nivolumab in 2014 in the United States.
-However, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has evolved, and as such PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.
+However, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., percent staining, tumor proportion score, or combined positive score) has evolved, and as such PD-L1 '*positivity'* may have different definitions by year based on different cut-off values.
 
 According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step outlines how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database.
-As there are general eligibility criteria in oncology clinical trials which either are infeasible to emulate (e.g., physician-assessed survival prognosis of x months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.
+As there are eligibility criteria in oncology clinical trials which either are infeasible to emulate (e.g., physician-assessed survival prognosis quantification) or that do not impact clinical care for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will determine key eligibility criteria for the emulation of the trial based on consensus.
 
 Additionally, we will provide conceptual and operationalized definitions on how exposures, outcomes and covariates are defined in each respective database.
-There will be special emphasis on how exposure, in context of disease and line of therapy settings, and the OS outcome are emulated.
-For all considered databases, OS is typically a composite measure that, depending on the underlying database, can be derived from different mortality sources (i.e., EHR documentation, social security death index, obituary and other linkages).
+There will be special emphasis on how exposure, in context of disease and line of therapy settings, and the survival outcomes are emulated.
+For all considered databases, real-world OS (rwOS) is typically a composite measure that, depending on the underlying database, can be derived from different mortality sources (i.e., EHR documentation, Social Security Death Index, and other linkages).
 Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open]
@@
 
-Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered, such as the contemporaneity of the control cohort, the adoption rate of the novel medical product in routine care, the magnitude of the clinical treatment benefit and the rate in which patients discontinue or cross-over, which could dilute effect estimates.
+Particularly when emulating pivotal trials of practice-changing treatments, multiple aspects need to be considered, such as the contemporaneity of the control cohort, the adoption rate of the novel medical product in routine care, the magnitude of the clinical treatment benefit, and the rate in which patients discontinue or cross-over, as they could influence effect estimates.
 
-To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial.
+To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) evaluate comparisons to the emulated trial.
 Examples for such standard diagnostics are visualized in @fig-initiators.
-All exploratory analyses will be conducted blinded towards the outcome to avoid influencing study design and analytic choices.
 
-The distribution of patient characteristics, by exposure status, will be examined using Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT.
-Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, exposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting.
+The distribution of patient characteristics, by exposure status, will be examined before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT.
+Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, exposure cohorts are conditionally exchangeable at baseline, and resulting sample sizes are still sufficient after matching or weighting.
 At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design or analytic choices based on known outcome information.
@@
 
-Causal analyses of observational data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results.
-Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
+Causal analyses of observational data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data .[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreement metrics between RCT and RWD results.
+Since the main outcome of interest is defined as time from index treatment initiation to all-cause mortality (rwOS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
 To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]
@@
 
-To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR for OS and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD).
+To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (hazard ratio \[HR\] for OS and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD).
 Examples are illustrated in @tbl-metrics.
 
-**Statistical significance agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.
+**Statistical significance agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance (nominal in the case of emulated trials).
 
 **Estimate agreement**: agreement that the estimated RWE treatment effect is within the 95% CI of the RCT treatment effect estimate.
-Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no statistical significance agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).
+Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no statistical significance agreement (RCT estimate crossing the null although the treatment effect estimates are overlapping) or vice versa in case the RCT has a larger power than the RWE emulation.
 
@@
 
-The study design for each trial emulation will be visualized as part of the protocol using a graphical depiction of the exact measurement windows of eligibility criteria, washout periods and covariates relative to the cohort entry time.[@schneeweiss2019graphical]
+The study design for each trial emulation will be visualized as part of the protocol using a graphical depiction of the exact measurement windows of eligibility criteria, washout periods, and covariates relative to the cohort entry time.[@schneeweiss2019graphical]
 
@@
 
-To establish an analytic cohort, missingness will be assessed in a first iteration across patients who meet all eligibility criteria, including those with missing values.
+To establish an analytic cohort, missingness will be assessed across patients who meet all eligibility criteria, including those with missing values as a first iteration.
 These missing data investigations will empirically assess assumptions on potential underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] We will adopt a principled process on missing data that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] These diagnostics cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different.
-Expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams[@moreno2018canonical] will additionally inform decisions regarding the inclusion or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]
+Expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams[@moreno2018canonical] will additionally inform decisions regarding the inclusion or exclusion of individuals with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]
 
-While the MAR assumption is a strong assumption to hold across all considered covariates, it has been shown that especially in the context of partially observed covariate data (as opposed to missing exposure and outcome data), only mechanisms in which a covariate causes its own missingness lead to critical bias (MNAR).[@moreno2018canonical] Hence, methodologies which retain patients and give the potential to adjust for a broader set of prognostic factors (e.g., multiple imputation[@weberpals2024hdmi] or doubly robust methods[@Shaw2024]) may be preferred over complete case analyses.
+While the MAR assumption is a strong assumption to hold across all considered covariates, it has been shown that especially in the context of partially observed covariate data (as opposed to missing exposure and outcome data), generally only mechanisms in which a covariate causes its own missingness lead to significant bias (MNAR).[@moreno2018canonical] Hence, methodologies which retain otherwise eligible patients and give the potential to adjust for a broader set of prognostic factors (e.g., multiple imputation[@weberpals2024hdmi] or doubly robust methods[@Shaw2024]) may be preferred over complete case analyses, although such tradeoffs must be carefully evaluated within the clinical and study contexts.
 
@@
 
-Due to its ubiquity in oncology trials, the primary parameter of interest in ENCORE will be defined as the marginal hazard ratio (HR) coefficient for the treatment comparison for OS.[@cox1972regression] We will also consider alternative endpoints on an absolute risk scale such as median survival times, survival probabilities at pre-defined time points during follow-up[@kaplan1958nonparametric] and restricted mean survival times as secondary endpoints of interest.
+Due to its frequency of use in oncology trials, the primary parameter of interest in ENCORE will be defined as the marginal HR coefficient for the treatment comparison for OS.[@cox1972regression] We will also consider alternative endpoints on an absolute risk scale such as median survival times, survival probabilities at pre-defined time points during follow-up[@kaplan1958nonparametric] and restricted mean survival times as secondary endpoints of interest.
 
 For the estimation of marginal treatment effects, we will employ propensity score methods to adjust for measured confounding between treatment arms.
-The selection of important prognostic covariates will be based on expert clinical knowledge and published literature on prognostic scores in oncology.[@becker2020enhanced] The implementation of propensity scores in combination with multiple imputation will follow the '*within*' methodology as described by Leyrat et al.[@leyrat2019propensity; @pishgar2020matchthem] That is, propensity score matching or weighting will be applied to each imputed dataset.
+The selection of relevant prognostic covariates will be based on expert clinical knowledge and published literature on prognostic scores in oncology.[@becker2020enhanced] The implementation of propensity scores in combination with multiple imputation will follow the '*within*' methodology as described by Leyrat et al.[@leyrat2019propensity; @pishgar2020matchthem] As appropriate, propensity score matching or weighting will be applied to each imputed dataset.
 The marginal treatment effect will then be estimated in each imputed and matched or weighted dataset separately and pooled into a final estimate following Rubin's rule.[@rubin2018multiple; @van2011mice] This approach has been shown to lead to unbiased estimates across different simulated scenarios with a sufficient estimation of the variance.[@leyrat2019propensity]
 
-To asses the balance of pre-exposure covariates after matching or weighting on each imputed dataset, the average SMD and corresponding minimum and maximum SMD range will be visualized (see example in @fig-balance).
-Covariate balance is often reasonably considered at a SMD \< 0.1.[@austin2009balance] Further, we will compute the average post-matching or post-weighting C-statistics.[@franklin2014metrics] In addition, we will use a published prognostic score for OS[@becker2020enhanced] as a balance measure to visually assess if the prognostic score is balanced between treatment arms after propensity score matching or weighting as this approach was described to show the highest correlations with bias compared with other balance measures and not affected by model misspecification.[@stuart2013prognostic]
+To assess the balance of pre-exposure covariates after matching or weighting on each imputed dataset, the average SMD and corresponding minimum and maximum SMD range will be visualized (see example in @fig-balance).
+Covariate balance is often considered reasonable at a SMD \< 0.1.[@austin2009balance] Further, we will compute the average post-matching or post-weighting C-statistics.[@franklin2014metrics] In addition, we will use a published prognostic score for OS[@becker2020enhanced] as a balance measure to visually assess if the prognostic score is balanced between treatment arms after propensity score matching or weighting as this approach was described to show the highest correlations with bias compared with other balance measures and not affected by model misspecification.[@stuart2013prognostic]
 
-Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according to the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically do not follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given time during follow-up.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability.
+Similarly, survival probabilities for individual time points will be estimated in each imputed and propensity score matched or weighted dataset according to the Kaplan-Meier method.[@kaplan1958nonparametric] Since survival probabilities typically do not follow normal distributions which are required to apply Rubin's rule, these will be transformed through a complementary log-log transformation $log(-log(1-pr(surv)))$ with $pr(surv)$ denoting the survival probability at a given follow-up time during.[@warwick28685; @morisot2015prostate] The transformed survival probabilities are then pooled across imputed datasets and individual time points following Rubin's rule and back-transformed via $1-exp(-exp(qbar))$ with $qbar$ denoting the pooled survival probability.
 The median survival time can be finally determined by extracting the time point during follow-up at which the survival probability drops below 0.5 for the first time.
@@
 
-With the goal to better understand which factors could influence a difference between RCT results and emulated trial results, it is appropriate to conduct a range of sensitivity analyses.
-This can comprise decisions on the considered databases, calendar time period, covariate measurement (e.g., measurement windows or trade-offs on sensitivity versus specificity in measurements), approaches to missing data, selection of covariates for imputation and propensity score models and censoring decisions.
+With the goal to better understand which factors could influence differences between RCT results and emulated trial results, it is appropriate to conduct a range of sensitivity analyses.
+This can comprise decisions on the considered databases, calendar time period, covariate measurement (e.g., measurement windows or trade-offs on sensitivity versus specificity in measurements), approaches to missing data, selection of covariates for imputation and propensity score models, and censoring decisions.
 All sensitivity analyses will be pre-specified in the study protocol and reported using appropriate visualizations such as forest plots.
@@
 The first package, `encore.io,` streamlines the query of analytic cohorts and measurement of covariates and outcomes using parameterized functions.
-Due to the blinding of databases used for ENCORE, this package cannot be made public but we provide a comprehensive and detailed documentation for each function in the Supplementary Material.
+Due to the blinding of databases used for ENCORE, this package is not able to be publicly available; however a comprehensive and detailed documentation for each function are available in the Supplementary Material.
 A second open-source R package, `encore.analytics`, will then facilitate the complex multi-step workflows of multiple imputation, propensity score matching and weighting and estimating pooled marginal treatment effects and Kaplan-Meier curves along with other established statistical R packages.[@anesrake; @MatchThem; @survival; @encore.analytics; @gtsummary; @ggsurvfit] Simulated examples with code that illustrate the analytic workflows described in this manuscript are available in the online Supplementary Material at <https://janickweberpals.github.io/imputation-ps-workflows/> and can be reproduced via the `encore.analytics` R package (<https://github.com/janickweberpals/encore.analytics/>).
@@
 Historically, administrative claims databases have been the predominant data source used to evaluate medical product safety and effectiveness in the postmarketing setting.
-With increasing access to EHR data, which are more fit for use in oncology than claims due to availability of clinical data elements, and a maturing set of methodological approaches for causal inference using such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to evaluate similar causal conclusions compared to RCTs in the field of oncology.
+With increasing access to EHR data and availability of granular clinical data elements often necessary to gather fit-for-use data in oncology, as well as a maturing set of methodological approaches for causal inference using such data, the ENCORE project is one of the first of its kind to evaluate contexts where RWD may be reliably used to draw similar conclusions compared to RCTs in the field of oncology.
 
 The RCT DUPLICATE initiative[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges that may be similar in ENCORE.
-One particular aspect that we may not always be able to emulate is the exact distribution of patient characteristics of the trial population, or that the comparability between these two populations may not be known.
-This can be due to the lack of RWD granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide (all considered databases reflect the US only).
-This may be a factor especially given that the pathophysiology, prognosis, and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries.
-For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are much more prevalent in certain geographic regions and populations compared to the US.
+One particular aspect that we may not always be able to emulate is the exact distribution of patient characteristics of the trial population.
+This can be due to the lack of data granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide (all considered databases reflect the US only).
+This may impact the results of this trial emulation given that the distribution of genetic alterations and biomarkers, prognosis, and factors that drive heterogeneous treatment outcomes of certain cancers can differ between countries.
+For example, certain cancers (e.g., GI cancers) or certain genetic mutations (e.g., EGFR mutations in lung cancer) are much more prevalent in certain geographic regions and populations compared to the US.
+Additionally, treatment landscapes such as the preference for specific regimens or compounds, their use in earlier or later lines of therapies, and supportive care practices can differ between geographic regions, too.
 
 Another common challenge in the emulation of oncology trials is the estimation of an "treatment policy"/"intention-to-treat" analogous estimand which is usually the primary estimand reported in oncology RCTs.
-Due to intercurrent events, such as non-adherence, crossover of a high proportion of patients from the control to the intervention arm, or differences in subsequent therapy lines there may be estimand differences between trial and emulation.[@manitz2022estimands] Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment practice in routine clinical practice might be less stringent compared to trials.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
+Due to intercurrent events, such as non-adherence, crossover of a high proportion of patients from the control to the intervention arm, or differences in subsequent therapies there may be observed differences in outcomes between trial and emulation.[@manitz2022estimands] Although this is also a common challenge in the analysis of RCTs,[@rufibach2018] treatment in routine clinical practice might be less stringent compared to trials in terms of pre-specified dosing schedules, monitoring and surveillance.[@merola2024calibrating] In order to derive comparable estimands it is therefore crucial to understand and contextualize the proportion and timing of treatment switching and discontinuation in both the trial and its emulation.
 
 This issue may be amplified with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD.
-As a result, exposure misclassification (due to overly ascertainment windows) and selection bias (due to overly long ascertainment windows) are common trade-offs in such scenarios.
+As a result, exposure misclassification (due to overly narrow ascertainment windows) and selection bias (due to overly long ascertainment windows) are common trade-offs in such scenarios.
 Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design,[@gaber2024mystifying] may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters, provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.
@@
 
-RWE based on fit-for-purpose data and principled, well-designed and reproducible studies may complement evidence from RCTs.
-Through a systematic benchmarking approach, the ENCORE project will provide insights as to how measurement, design and analytic decisions influence bias and validity in oncological non-interventional studies.
+Principled, well-designed and reproducible studies using fit-for-purpose RWD to generate RWE, such as through trial emulation, may complement evidence from RCTs.
+Through a systematic benchmarking approach, the ENCORE project will provide insights as to how methodological, design, and analytic decisions influence quantifiable bias and interpretability of observational studies in oncology.
 

### 2025-07-06 manuscript/manuscript.qmd

@@
 
-For each selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized;@guidelinegeneral] and will be registered on ClinicalTrials.gov after review by an expert panel comprising clinicians and FDA reveiwers.
+For each selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized; @guidelinegeneral] and will be registered on ClinicalTrials.gov after review by an expert panel comprising clinicians and FDA reveiwers.
 Following the target trial emulation framework, we will provide an explicit description and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update survival estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used for agreement metric evaluation (see @sec-agreement-metrics).

### 2025-07-06 manuscript/manuscript.tex
Manuscript updated

@@
 \hypersetup{
-  pdftitle={A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data},
+  pdftitle={Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data},
   colorlinks=true,
@@
 
-\title{A Process for Emulating Comparative Oncology Trials with
-Real-world Evidence Studies (ENCORE): Process Development and
-Methodological Considerations for Oncology Real-World Data}
+\title{Emulating Comparative Oncology Trials with Real-world Evidence
+Studies (ENCORE): Process Development and Methodological Considerations
+for Oncology Real-World Data}
 \author{}
@@
 
-\textbf{Authors}: Janick Weberpals\textsuperscript{1}, Kenneth L.
-Kehl\textsuperscript{2}, Donna R. Rivera\textsuperscript{3}, Pallavi
+\textbf{Authors}: Janick Weberpals\textsuperscript{1}, Sebastian
+Schneeweiss\textsuperscript{1}, Kenneth L. Kehl\textsuperscript{2},
+Donna R. Rivera\textsuperscript{3}, Pallavi
 Mishra-Kalyani\textsuperscript{3}, Catherine C.
 Lerro\textsuperscript{3}, Erin Larkins\textsuperscript{4}, Sonia
-Singh\textsuperscript{4}, Preeti Narayan\textsuperscript{4}, Richard
-Curley\textsuperscript{4}, Paul Kluetz\textsuperscript{4}, Georg
-Hahn\textsuperscript{1}, Priyanka Anand\textsuperscript{1}, Yanina
-Natanzon\textsuperscript{5}, Andrew J. Belli\textsuperscript{6},
-Ching-Kun Wang\textsuperscript{6}, Jenna
+Singh\textsuperscript{4*}, Preeti Narayan\textsuperscript{4}, Richard
+Curley\textsuperscript{4}, Georg Hahn\textsuperscript{1}, Priyanka
+Anand\textsuperscript{1}, Yanina Natanzon\textsuperscript{5}, Andrew J.
+Belli\textsuperscript{6}, Ching-Kun Wang\textsuperscript{6}, Jenna
 Collins\textsuperscript{7}\textbf{,} Jonathan
@@
 Nicholas J Robert\textsuperscript{8}, Robert J.
-Glynn\textsuperscript{1}, Sebastian Schneeweiss\textsuperscript{1},
-Shirley V. Wang\textsuperscript{1}
+Glynn\textsuperscript{1}, Shirley V. Wang\textsuperscript{1}
 
@@
 \textsuperscript{8} Ontada, Boston, MA, USA
+
+*At the time this research was conducted
 
@@
 
-\emph{Manuscript last updated: 2025-05-28 17:11:23.084986}
+\emph{Manuscript last updated: 2025-07-06 14:54:57.141696}
 
@@
 initiative has previously enhanced understanding of when RWE can most
-appropriately draw causal conclusions by emulating trials in
+appropriately draw causal conclusions by emulating trials for
 non-oncology indications. Here, we present the design and trial
@@
 results, the ENCORE project will inform understanding of how
-measurement, design, and analytic decisions influence the validity of
-oncological RWE studies.
+measurement, design, and analytic decisions influence the interpretation
+of results from emulated oncology trials using RWD.
 
@@
 
-Randomized controlled trials (RCTs) are the go-to methodology for
-establishing the efficacy and safety of medical products. Directed by
-the 21\textsuperscript{st} Century Cures Act,\textsuperscript{1} the
-Food and Drug Administration (FDA) established a framework to factor
-consideration of real-world evidence (RWE) generated from routine-care
-health data such as electronic health records (EHR) to evaluate and
-contextualize the safety and effectiveness of novel cancer
-therapies.\textsuperscript{2} Accounting for 30\% of new drug approvals,
-oncology was the disease area with the most FDA drug approvals in
-2024\textsuperscript{3} and has many areas of high unmet medical need to
-treat serious conditions. Therefore, RWE has particularly important
-potential to complement evidence from RCTs in the field of oncology.
-Potential use cases include the assessment of effectiveness in specific
-patient populations that are underrepresented in RCTs, the construction
-of external control arms in single-arm trials that have feasibility or
-equipoise challenges, or the precision oncology-focused discovery of
-biomarkers among pan-tumor populations that harbor specific genomic and
+Randomized controlled trials (RCTs) are the evidentiary gold standard
+methodology for establishing the efficacy and safety of medical
+products. Guided by the 21\textsuperscript{st} Century Cures
+Act,\textsuperscript{1} the Food and Drug Administration (FDA)
+established a framework to factor considerations for use of real-world
+evidence (RWE) generated from real-world data (RWD) such as electronic
+health records (EHR) to support approval of new indications or satisfy
+postmarketing requirements.\textsuperscript{1} Accounting for
+approximately 30\% of new FDA drug approvals, oncology was the disease
+area with the most new approvals in 2024\textsuperscript{2} as well as
+several indication expansions, and has many areas of high unmet medical
+need to treat serious conditions. Therefore, RWE can have a particularly
+important potential to complement evidence from RCTs in the field of
+oncology. Potential uses include the assessment of effectiveness in
+specific patient populations that are not adequately represented in
+RCTs, or the precision oncology-focused discovery of biomarkers among
+pan-tumor populations that harbor specific genomic and
 immuno-pathological signatures.
 
-However, the validity and transportability of results derived between
-RWE studies and RCTs depends on multiple factors. Frequently referenced
-limitations include lack of baseline randomization, missing data,
+However, the comparability and transportability of results derived
+between observational studies and RCTs depends on multiple factors.
+Frequently referenced limitations include lack of baseline randomization
+and the imbalance in prognostic factors resulting thereof, missing data,
 unmeasured confounding, small study sizes, data
-discontinuity,\textsuperscript{4, 5} adoption of changes in guidelines
-in real-world care and the inability to measure and emulate common
-eligibility criteria, including prognostic factors, and standardized
-response assessments in real-world data (RWD).\textsuperscript{6} While
-examples of oncology trial emulations have been
-published,\textsuperscript{6--8} a systematic and scaled approach to
-emulate a diverse set of oncology trials with multiple heterogeneous
-databases is necessary to gain confidence in the validity of RWE
-studies, evaluate regulatory considerations, and to provide context as
-to which questions can be validly answered.
-
-The RCT DUPLICATE initiative\textsuperscript{9} increased our
-understanding of when RWE studies can come to causal conclusions on
-treatment effects by benchmarking results against RCTs under the
-assumption that each RCT finding reflects a causal treatment effect. In
-settings where the RCT designs could be emulated well, RWE studies were
-able to reach similar conclusions.\textsuperscript{10} However, this
-prior work from RCT-DUPLICATE focused primarily on emulating trials in
-non-oncology settings using claims databases.
+discontinuity,\textsuperscript{3, 4} adoption of changes in treatment
+guidelines over time in routine clinical care and the inability to
+measure and emulate common eligibility criteria, including prognostic
+factors, and standardized response assessments in real-world data
+(RWD).\textsuperscript{5} While examples of oncology trial emulations
+have been published,\textsuperscript{5--7} a systematic and scaled
+approach to emulate a diverse set of oncology trials with multiple
+heterogeneous databases is necessary to increase confidence in the
+interpretability of observational studies, evaluate regulatory
+considerations, and to provide context as to which questions can be
+validly answered.
+
+The RCT DUPLICATE initiative\textsuperscript{8} increased our
+understanding of when observational studies can come to causal
+conclusions on treatment effects by benchmarking results against RCTs
+under the assumption that each RCT finding reflects a causal treatment
+effect. In settings where the RCT designs could be emulated well,
+observational studies were able to reach similar
+conclusions.\textsuperscript{9} However, this prior work from
+RCT-DUPLICATE focused primarily on emulating trials in non-oncology
+settings using claims databases.
 
 The \emph{Emulation of Comparative Oncology Trials with Real-world
-Evidence} (ENCORE) project\textsuperscript{11} aims to extend this work
-to oncology. Studies in oncology come with their own unique set of
+Evidence} (ENCORE) project\textsuperscript{10} aims to extend this work
+to oncology. Clinical studies in oncology come with unique
 methodological challenges which must be systematically explored and
 understood. Building on a process co-developed with the FDA through RCT
-DUPLICATE,\textsuperscript{9} this expansion to oncology will emulate 12
+DUPLICATE,\textsuperscript{8} this expansion to oncology will emulate 12
 randomized clinical trials using multiple specialty oncology EHR data
-sources. The process will emphasize transparency and include documented
-data fit-for-purpose assessment\textsuperscript{12, 13} for each RWD
-source with respect to each trial emulation as well as extensive
-sensitivity analyses to assess robustness of findings.
+sources. The process will emphasize transparency and include a data
+fitness-for-use assessment\textsuperscript{11, 12} for each RWD source
+with respect to each trial emulation as well as extensive sensitivity
+analyses to assess robustness of findings.
 
 The objectives of this project are to 1) develop state-of-the-art
-methodological approaches and 2) to apply them to create insights into
-the potential use of RWE to enhance regulatory science in oncology. To
-achieve these objectives, this demonstration project will systematically
-emulate 12 oncology trials across four cancer types and assess the
-agreement of treatment effect estimates between RCTs and their
-respective emulations.
+methodological approaches and 2) apply these methods to gain insights
+into the potential use of RWE to enhance regulatory and clinical
+decision-making in oncology. To achieve these objectives, this
+demonstration project will systematically emulate 12 oncology trials
+across four cancer types and assess the agreement of treatment effect
+estimates between RCTs and their respective emulations.
 
 Here, we describe the design and process for the selection of the 12
-oncology RCTs, the assessment of the database quality and selection,
+oncology RCTs, assessment of the database quality and selection,
 protocol development and pre-registration, study design and statistical
-analysis, and pre-specified agreement metrics to evaluate the
+analysis plans, and pre-specified agreement metrics to evaluate the
 concordance between RCTs and emulations.
@@
 
-The focus of ENCORE is to efficiently explore in what contexts RWE
-studies can or cannot yield similar results compared to RCTs. Therefore,
-the project emphasizes trials of therapies for common cancers for which
-there has been substantial therapeutic development in recent years.
-After review and collaboration with clinical and regulatory experts,
-four cancer indications were identified: non-small cell lung cancer,
-breast cancer, colorectal cancer and multiple myeloma. For each cancer
-type, we aim to conduct three trial emulations using multiple accessible
-databases (i.e., the total equaling 12 trial emulations x \emph{n}
-databases which are found fit-for-purpose for each trial).
+The focus of ENCORE is to efficiently explore in what contexts
+observational studies can or cannot yield similar results compared to
+oncology RCTs. Therefore, the project emphasizes trials of therapies for
+common cancers for which there has been substantial therapeutic
+development in recent years. After review and collaboration with
+clinical and regulatory experts, four cancer indications were
+identified: non-small cell lung cancer, breast cancer, colorectal cancer
+and multiple myeloma. For each cancer type, we aim to conduct three
+trial emulations using multiple accessible databases (i.e., the total
+equaling 12 trial emulations x \emph{n} databases which are found
+fit-for-purpose for each trial).
 
@@
 criteria are documented in CONSORT diagrams showing reasons for
-excluding RCTs for each cancer type. The search was conducted using the
+excluding RCTs by cancer type. The search was conducted using the
 \emph{Aggregated Analysis of ClinicalTrials.gov} (AACT) database which
@@
 information (protocol and result data elements) about studies registered
-on ClinicalTrials.gov.\textsuperscript{14} To identify eligible trials,
+on ClinicalTrials.gov.\textsuperscript{13} To identify eligible trials,
 we used a combined search query strategy of the National Library of
@@
 \item
-  Intervention model: parallel assignment
+  Intervention model: Parallel assignment
 \item
@@
 \item
-  Trial start in 2011 or later
+  Trial start: 2011 or later
 \item
-  Primary purpose was to study treatment efficacy
+  Primary purpose: Study treatment efficacy
 \item
-  Overall survival must be one of the endpoints reported (either as
-  hazard ratio or median overall survival time)
+  Endpoint(s): Overall survival must be one of the endpoints reported
+  (either as hazard ratio or median overall survival time)
 \item
@@
 The rationale and operationalization for each criterion is listed in
-detail in Table~\ref{tbl-criteria}. We will mainly consider pivotal
-large late-phase interventional RCTs after 2011 because treatment
-guidelines among included cancer indications have undergone significant
-changes in recent years. Due to the rapid adoption of breakthrough
-therapies in routine care, few patients currently initiate outdated
-treatment regimens in current clinical practice. Conversely, we excluded
-trials with results published too recently to allow for enough data and
-follow-up time accrual in the real-world databases used for this
-project. We did not define a global cut-off as the requirements for
-follow-up time are different for each cancer type and population (e.g.,
-advanced NSCLC versus early-stage breast cancer) and the decisions were
-made on a case-by-case basis.
+detail in Table~\ref{tbl-criteria}. We mainly considered pivotal large
+late-phase RCTs that were initiated after 2011 because treatment
+guidelines for included cancer indications have undergone significant
+changes in recent years. Due to the rapid adoption of newly-approved
+therapies in routine care, patients are less likely to receive outdated
+treatment regimens in clinical practice. Conversely, we excluded trials
+with results published too recently to allow for substantive medical
+product uptake and follow-up time accrual in real-world databases used
+for this project. We did not define a global cut-off as the requirements
+for follow-up time are different for each cancer type and population
+(e.g., advanced NSCLC versus early-stage breast cancer) and the
+decisions were made on a clinically relevant basis.
 
@@
 progression-free survival (PFS) and objective response rates (ORR) to a
-RECISTv1.1\textsuperscript{15}-based PFS and ORR assessment in
-RCTs\textsuperscript{16, 17}, imaging-based evaluations still hold a
+RECISTv1.1\textsuperscript{14}-based PFS and ORR assessment in
+RCTs\textsuperscript{15, 16}, imaging-based evaluations still require a
 level of granularity which is not well reflected in chart-abstracted
 assessments of a patient's progression in routine
-care.\textsuperscript{18, 19} The timing and cadence of intervals
+care.\textsuperscript{17, 18} The timing and cadence of intervals
 between progression assessments can differ between RCTs and routine care
-which may result in measurement error and bias.\textsuperscript{20}
+which may result in measurement error and bias.\textsuperscript{19}
 Given this, and the large number of other methodological challenges
@@
 overall survival (OS) as the endpoint of interest. Therefore, we include
-only trials that have reported overall survival (OS) as one of the
-pre-specified endpoints in the protocol.
+only trials that have reported OS as a pre-specified endpoint in the
+protocol.
 
@@
 human review. The fundamental points considered in this step include an
-initial feasibility assessment of the data fitness,\textsuperscript{19}
+initial feasibility assessment of the data fitness,\textsuperscript{18}
 involving an assessment of whether critical eligibility criteria (e.g.,
-biomarker status) including prognostic factors (e.g., ECOG performance
-score) are measurable, and whether preliminary study size based on a
-rough estimation of the number of patients observed in the data with the
-combination of treatment regimen and line of treatment is reasonable for
-shortlisting the trial. Lastly, trial candidates were ranked and
-shortlisted into primary and runner-up candidates based on their
-clinical and regulatory relevance.
+biomarker status) including prognostic factors (e.g., Eastern
+Cooperative Oncology Group {[}ECOG{]} performance score) are measurable,
+and whether preliminary study size based on a rough estimation of the
+number of patients observed in the data with the combination of
+treatment regimen and line of therapy is reasonable for shortlisting the
+trial. Lastly, trial candidates were ranked and shortlisted as primary
+or runner-up based on their clinical and regulatory relevance.
 
@@
 in the CONSORT diagrams (\textbf{Supplementary Figures 1-4}) and runner
-up candidate trials are listed in \textbf{Supplementary Table 1}.
-Naturally, the majority of shortlisted trials cover advanced (locally
-advanced, inoperable, recurrent/progressive disease) or metastatic
-cancer populations because a large proportion of drug development
-efforts have recently focused on these settings. A key objective that we
-aim to explore with the selected trials is to achieve a better
-understanding how different disease settings (early, late), line
-settings ({[}neo{]}adjuvant, first line, advanced lines of therapy),
-therapy protocols (monotherapy, combination therapy) and clinical
-characteristics (all-comer versus complex genetic or immunological
-signatures) can be emulated using RWD. If more thorough feasibility
-assessments after applying all relevant eligibility criteria suggest
-that the threat of bias from mis-measurement of key study parameters or
-residual confounding remains high, or that the study size is not
-sufficient, then runner-up candidate trials will be considered instead.
+up candidate trials are listed in \textbf{Supplementary Table 1}. The
+majority of shortlisted trials represent indications for advanced
+(locally advanced, inoperable, recurrent/progressive disease) or
+metastatic cancer patients because a large proportion of drug
+development efforts have recently focused on these settings. A key
+objective that we aim to explore with the selected trials is to evaluate
+how stage (early, late), line of therapy ({[}neo{]}adjuvant, first line,
+advanced lines of therapy), therapy protocols (monotherapy, combination
+therapy) and clinical characteristics (all-comer versus complex genetic
+or immunological signatures) can be emulated using RWD. If more thorough
+feasibility assessments after applying all relevant eligibility criteria
+suggest that the threat of bias from mis-measurement of key study
+parameters or residual confounding remains high, or that the study size
+is not sufficient, then runner-up candidate trials will be considered
+instead.
 
@@
 
-The ENCORE project will use data from four oncology-specific electronic
-health records (EHR)-derived RWD data sources (in alphabetical order):
-ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. The included
-databases comprise the necessary information to study medical product
-effectiveness in oncology derived from US-based oncology EHRs. A
-detailed description and sampling methodology will be provided with each
-trial emulation protocol. For ENCORE, not all databases will be
-available for each cancer indication and the names of the databases will
-be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for
-the final reporting of results (the numbering does not coincide with the
-above order of mention of the databases). If more than one database is
-considered fit-for-purpose for a respective trial emulation, the best
-possible analytic model will be employed for each database separately
-and final treatment effect estimates will be pooled using a
-meta-analytic approach using fixed effects (weights reflect the inverse
-of the variance of each database study) and random effects (weights
-reflect the inverse of the variance of each database study plus an
-additional variance term that quantifies the assumed heterogeneity
-between databases) models.\textsuperscript{21}
+The ENCORE project will use data from four U.S.-based oncology-specific
+EHR-derived RWD data sources (in alphabetical order): ConcertAI, COTA,
+Flatiron Health, and Ontada/McKesson. A detailed description and
+sampling methodology will be provided with each trial emulation
+protocol. For ENCORE, not all databases will be available for each
+cancer indication given the specificity of data elements required and
+the names of the databases will be blinded and referred to as ENCORE
+DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (in
+randomly assigned order). If more than one database is considered
+fit-for-use for a respective trial emulation, the most suitable analytic
+model will be employed for each database separately and final treatment
+effect estimates will be pooled using a meta-analytic approach using
+fixed effects (weights reflect the inverse of the variance of each
+database study) and random effects (weights reflect the inverse of the
+variance of each database study plus an additional variance term that
+quantifies the assumed heterogeneity between databases)
+models.\textsuperscript{20}
 
@@
 of the trial emulation, will be developed using the HARPER protocol
-template\textsuperscript{22} recommended for regulatory submissions of
-RWE studies\textsuperscript{23} and will be registered on
-ClinicalTrials.gov after review by a clinical and FDA regulatory expert
-panel. Following the target trial emulation framework, we will provide
-an explicit statement and rationale on how each element will be emulated
-including database selection, covariate measurement, operationalization
-of key eligibility criteria, study design, data analysis and causal
-contrasts of interest.\textsuperscript{24, 25} Since it is common that
-oncology RCTs update OS estimates periodically based on accrued
-follow-up time, the protocol will give a brief summary of each emulated
-RCT and specify which target OS estimates will be used for agreement
-metric evaluation (see Section~\ref{sec-agreement-metrics}). All
-eligibility criteria will be extracted based on publications, publicly
-available protocols, and statistical analysis plans of the selected RCT.
+template\textsuperscript{21, 22} and will be registered on
+ClinicalTrials.gov after review by an expert panel comprising clinicians
+and FDA reveiwers. Following the target trial emulation framework, we
+will provide an explicit description and rationale on how each element
+will be emulated including database selection, covariate measurement,
+operationalization of key eligibility criteria, study design, data
+analysis and causal contrasts of interest.\textsuperscript{23, 24} Since
+it is common that oncology RCTs update survival estimates periodically
+based on accrued follow-up time, the protocol will give a brief summary
+of each emulated RCT and specify which target OS estimates will be used
+for agreement metric evaluation (see
+Section~\ref{sec-agreement-metrics}). All eligibility criteria will be
+extracted based on publications, publicly available protocols, and
+statistical analysis plans of the selected RCT.
 
@@
 
-One important aspect when emulating oncology trials is the choice and
-estimation of the appropriate estimand of interest.\textsuperscript{26}
+An important aspect when emulating oncology trials is the choice and
+estimation of the appropriate estimand of interest.\textsuperscript{25}
 In prior RCT-DUPLICATE efforts, a ``while on treatment'' strategy was
@@
 because OS outcome analyses in oncology are typically ``treatment
-policy'' estimands that allow crossover.\textsuperscript{27} We plan to
-focus on ``treatment policy'' estimands for the primary analyses in our
-trial emulations. However, allowing crossover can dilute treatment
-effects, moving point estimates toward the null. Therefore, differences
-in cross-over and discontinuation rates between the trial and emulation
-may still contribute to estimand differences. Recognizing these
+policy'' estimands that allow for crossover or other post-progression
+antineoplastic therapy.\textsuperscript{26} We plan to focus on
+``treatment policy'' estimands for the primary analyses in our trial
+emulations. However, allowing crossover can dilute treatment effects,
+moving point estimates toward the null. Therefore, differences in
+cross-over and discontinuation rates between the trial and emulation may
+still contribute to differences in observed outcomes. Recognizing these
 practical challenges, we will carefully characterize the patterns of
@@
 
-Real-world data fitness and emulation feasibility for each shortlisted
-candidate trial will be assessed in multiple steps based on direction
-from of the oncology quality, characterization, and assessment of
-real-world data (Oncology QCARD) Initiative.\textsuperscript{19} The
-first step assesses if relevant variables like exposure, line of
-therapy, outcomes, and covariates are generally available, measured and
-operationalizable in routine-care. Since many oncological RCTs in recent
-years have focused on selected, biomarker-defined populations, nuances
-in measurement and operationalizability of specific biomarkers must be
-reflected to ensure a representative and large enough study population.
-For example, immune checkpoint inhibitors have significantly changed the
-cancer treatment landscape since the first approvals of the CTLA-4
-inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and
-nivolumab in 2014 in the United States. However, the operationalization
-of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent
-staining, tumor proportion score or combined positive score) has
-evolved, and as such PD-L1 '\emph{positivity'} may have different
-definitions across calendar years based on different cut-off values.
+Real-world data fitness and emulation feasibility for each candidate
+trial will be assessed in multiple steps based on the approach described
+by the Oncology Quality, Characterization, and Assessment of Real-World
+Data (Oncology QCARD) Initiative.\textsuperscript{18} The first step
+assesses if relevant variables like exposure, line of therapy, outcomes,
+and covariates are generally available, measured and operationalizable
+in routine-care. Since many oncology RCTs in recent years have focused
+on biomarker-defined populations, nuances in measurement and
+operationalizability of specific biomarkers must be understood to ensure
+a representative and adequately sized study population. For example,
+immune checkpoint inhibitors have significantly changed the cancer
+treatment landscape since the first approvals of the CTLA-4 inhibitor
+ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and nivolumab
+in 2014 in the United States. However, the operationalization of the
+expression of the PD-L1 biomarker in RCTs (e.g., percent staining, tumor
+proportion score, or combined positive score) has evolved, and as such
+PD-L1 '\emph{positivity'} may have different definitions by year based
+on different cut-off values.
 
 According to the \emph{Structured Process to Identify Fit-For-Purpose
-Data} (SPIFD) framework\textsuperscript{13}, the next step outlines how
+Data} (SPIFD) framework\textsuperscript{12}, the next step outlines how
 eligibility criteria will be ascertained using a color-coded heatmap
 that will indicate the level of confidence on how well each criterion
-can be emulated in each selected database. As there are general
-eligibility criteria in oncology clinical trials which either are
-infeasible to emulate (e.g., physician-assessed survival prognosis of x
-months) or that are clinically not relevant for the emulation of the
-trial (e.g., male patients should be willing to use barrier
-contraception), the study team will decide on key eligibility criteria
-for the emulation of the trial.
+can be emulated in each selected database. As there are eligibility
+criteria in oncology clinical trials which either are infeasible to
+emulate (e.g., physician-assessed survival prognosis quantification) or
+that do not impact clinical care for the emulation of the trial (e.g.,
+male patients should be willing to use barrier contraception), the study
+team will determine key eligibility criteria for the emulation of the
+trial based on consensus.
 
@@
 database. There will be special emphasis on how exposure, in context of
-disease and line of therapy settings, and the OS outcome are emulated.
-For all considered databases, OS is typically a composite measure that,
-depending on the underlying database, can be derived from different
-mortality sources (i.e., EHR documentation, social security death index,
-obituary and other linkages). Given that not all relevant sources that
-provide mortality data are synchronized and updated uniformly,
-sensitivity analyses with more conservative (i.e., earlier) censoring
-dates will be considered for each trial emulation to mitigate the
-potential impact of ghost-time bias.\textsuperscript{28}
+disease and line of therapy settings, and the survival outcomes are
+emulated. For all considered databases, real-world OS (rwOS) is
+typically a composite measure that, depending on the underlying
+database, can be derived from different mortality sources (i.e., EHR
+documentation, Social Security Death Index, and other linkages). Given
+that not all relevant sources that provide mortality data are
+synchronized and updated uniformly, sensitivity analyses with more
+conservative (i.e., earlier) censoring dates will be considered for each
+trial emulation to mitigate the potential impact of ghost-time
+bias.\textsuperscript{27}
 
@@
 
-Particularly when emulating pivotal trials of paradigm-changing
+Particularly when emulating pivotal trials of practice-changing
 treatments, multiple aspects need to be considered, such as the
@@
 medical product in routine care, the magnitude of the clinical treatment
-benefit and the rate in which patients discontinue or cross-over, which
-could dilute effect estimates.
+benefit, and the rate in which patients discontinue or cross-over, as
+they could influence effect estimates.
 
@@
 of the protocol development to contextualize these parameters and (if
-reported) draw comparisons to the emulated trial. Examples for such
-standard diagnostics are visualized in Figure~\ref{fig-initiators}. All
-exploratory analyses will be conducted blinded towards the outcome to
-avoid influencing study design and analytic choices.
+reported) evaluate comparisons to the emulated trial. Examples for such
+standard diagnostics are visualized in Figure~\ref{fig-initiators}.
 
 The distribution of patient characteristics, by exposure status, will be
-examined using Table 1's before and after applying eligibility criteria
-and contrasted with the distributions of patient characteristics of the
-original RCT. Initial propensity score matching or weighting methods
-will be applied to ensure that measured pre-exposure covariates can be
-balanced, exposure cohorts are conditionally exchangeable at baseline
-and resulting sample sizes are still sufficient after matching or
-weighting. At this stage, all exploratory analyses will be conducted
-blinded towards the outcome to not bias any study design or analytic
-choices based on known outcome information.
+examined before and after applying eligibility criteria and contrasted
+with the distributions of patient characteristics of the original RCT.
+Initial propensity score matching or weighting methods will be applied
+to ensure that measured pre-exposure covariates can be balanced,
+exposure cohorts are conditionally exchangeable at baseline, and
+resulting sample sizes are still sufficient after matching or weighting.
+At this stage, all exploratory analyses will be conducted blinded
+towards the outcome to not bias any study design or analytic choices
+based on known outcome information.
 
@@
 manner as RCTs since the number of `recruited' patients is limited by
-available data and cannot be influenced.\textsuperscript{29} For this
-project, however, the emulations in EHR data must have at least equal
-power to the relevant trial for interpretation of the pre-specified
-agreements metrics between RCT and RWD results. Since the main outcome
-of interest is defined as time to all-cause mortality (OS), the
+available data .\textsuperscript{28} For this project, however, the
+emulations in EHR data must have at least equal power to the relevant
+trial for interpretation of the pre-specified agreement metrics between
+RCT and RWD results. Since the main outcome of interest is defined as
+time from index treatment initiation to all-cause mortality (rwOS), the
 estimation of the statistical power is driven by the number of events
@@
 Schoenfeld's sample-size formula for the proportional-hazards regression
-model.\textsuperscript{30}
+model.\textsuperscript{29}
 
@@
 emulations, we will adapt the approach of the RCT-DUPLICATE
-project.\textsuperscript{9, 31} That is, for the primary endpoint of
-interest (HR for OS and corresponding 95\% confidence intervals), we
-will derive three qualitative agreement metrics: statistical agreement,
-estimate agreement and agreement based on the standardized mean
-difference (SMD). Examples are illustrated Table~\ref{tbl-metrics}.
+project.\textsuperscript{8, 30} That is, for the primary endpoint of
+interest (hazard ratio {[}HR{]} for OS and corresponding 95\% confidence
+intervals), we will derive three qualitative agreement metrics:
+statistical agreement, estimate agreement and agreement based on the
+standardized mean difference (SMD). Examples are illustrated in
+Table~\ref{tbl-metrics}.
 
@@
 emulated trial treatment effect with regards to directionality and
-statistical significance.
+statistical significance (nominal in the case of emulated trials).
 
@@
 larger than that of the original RCT, this could lead to situations
-where there is no statistical significance agreement although the
-treatment effect estimates are highly overlapping but with the RCT
-estimate crossing the null (or vice versa in case the RCT has a larger
-power than the RWE emulation).
+where there is no statistical significance agreement (RCT estimate
+crossing the null although the treatment effect estimates are
+overlapping) or vice versa in case the RCT has a larger power than the
+RWE emulation.
 
@@
 the protocol using a graphical depiction of the exact measurement
-windows of eligibility criteria, washout periods and covariates relative
-to the cohort entry time.\textsuperscript{32}
+windows of eligibility criteria, washout periods, and covariates
+relative to the cohort entry time.\textsuperscript{31}
 
@@
 
-To establish an analytic cohort, missingness will be assessed in a first
-iteration across patients who meet all eligibility criteria, including
-those with missing values. These missing data investigations will
+To establish an analytic cohort, missingness will be assessed across
+patients who meet all eligibility criteria, including those with missing
+values as a first iteration. These missing data investigations will
 empirically assess assumptions on potential underlying missingness
@@
 missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
-missing not at random {[}MNAR{]}).\textsuperscript{33} We will adopt a
+missing not at random {[}MNAR{]}).\textsuperscript{32} We will adopt a
 principled process on missing data that empirically evaluates different
 aspects across partially observed covariates based on three group
-diagnostics.\textsuperscript{34, 35} These diagnostics cover (1)
+diagnostics.\textsuperscript{33, 34} These diagnostics cover (1)
 comparisons of patients characteristics with and without an observed
@@
 domain knowledge and assumptions about the underlying missing data
-structure through canonical causal diagrams\textsuperscript{36} will
+structure through canonical causal diagrams\textsuperscript{35} will
 additionally inform decisions regarding the inclusion or exclusion of
-patients with missing values in key eligibility criteria and potential
-sensitivity analyses to assess the robustness of these
-decisions.\textsuperscript{37}
+individuals with missing values in key eligibility criteria and
+potential sensitivity analyses to assess the robustness of these
+decisions.\textsuperscript{36}
 
@@
 of partially observed covariate data (as opposed to missing exposure and
-outcome data), only mechanisms in which a covariate causes its own
-missingness lead to critical bias (MNAR).\textsuperscript{36} Hence,
-methodologies which retain patients and give the potential to adjust for
-a broader set of prognostic factors (e.g., multiple
-imputation\textsuperscript{38} or doubly robust
-methods\textsuperscript{39}) may be preferred over complete case
-analyses.
+outcome data), generally only mechanisms in which a covariate causes its
+own missingness lead to significant bias (MNAR).\textsuperscript{35}
+Hence, methodologies which retain otherwise eligible patients and give
+the potential to adjust for a broader set of prognostic factors (e.g.,
+multiple imputation\textsuperscript{37} or doubly robust
+methods\textsuperscript{38}) may be preferred over complete case
+analyses, although such tradeoffs must be carefully evaluated within the
+clinical and study contexts.
 
@@
 
-Due to its ubiquity in oncology trials, the primary parameter of
-interest in ENCORE will be defined as the marginal hazard ratio (HR)
-coefficient for the treatment comparison for OS.\textsuperscript{40} We
-will also consider alternative endpoints on an absolute risk scale such
-as median survival times, survival probabilities at pre-defined time
-points during follow-up\textsuperscript{41} and restricted mean survival
-times as secondary endpoints of interest.
+Due to its frequency of use in oncology trials, the primary parameter of
+interest in ENCORE will be defined as the marginal HR coefficient for
+the treatment comparison for OS.\textsuperscript{39} We will also
+consider alternative endpoints on an absolute risk scale such as median
+survival times, survival probabilities at pre-defined time points during
+follow-up\textsuperscript{40} and restricted mean survival times as
+secondary endpoints of interest.
 
@@
 propensity score methods to adjust for measured confounding between
-treatment arms. The selection of important prognostic covariates will be
+treatment arms. The selection of relevant prognostic covariates will be
 based on expert clinical knowledge and published literature on
-prognostic scores in oncology.\textsuperscript{42} The implementation of
+prognostic scores in oncology.\textsuperscript{41} The implementation of
 propensity scores in combination with multiple imputation will follow
 the `\emph{within}' methodology as described by Leyrat et
-al.\textsuperscript{43, 44} That is, propensity score matching or
+al.\textsuperscript{42, 43} As appropriate, propensity score matching or
 weighting will be applied to each imputed dataset. The marginal
@@
 weighted dataset separately and pooled into a final estimate following
-Rubin's rule.\textsuperscript{45, 46} This approach has been shown to
+Rubin's rule.\textsuperscript{44, 45} This approach has been shown to
 lead to unbiased estimates across different simulated scenarios with a
-sufficient estimation of the variance.\textsuperscript{43}
-
-To asses the balance of pre-exposure covariates after matching or
+sufficient estimation of the variance.\textsuperscript{42}
+
+To assess the balance of pre-exposure covariates after matching or
 weighting on each imputed dataset, the average SMD and corresponding
 minimum and maximum SMD range will be visualized (see example in
-Figure~\ref{fig-balance}). Covariate balance is often reasonably
-considered at a SMD \textless{} 0.1.\textsuperscript{47} Further, we
+Figure~\ref{fig-balance}). Covariate balance is often considered
+reasonable at a SMD \textless{} 0.1.\textsuperscript{46} Further, we
 will compute the average post-matching or post-weighting
-C-statistics.\textsuperscript{48} In addition, we will use a published
-prognostic score for OS\textsuperscript{42} as a balance measure to
+C-statistics.\textsuperscript{47} In addition, we will use a published
+prognostic score for OS\textsuperscript{41} as a balance measure to
 visually assess if the prognostic score is balanced between treatment
@@
 balance measures and not affected by model
-misspecification.\textsuperscript{49}
+misspecification.\textsuperscript{48}
 
@@
 estimated in each imputed and propensity score matched or weighted
-dataset according to the Kaplan-Meier method.\textsuperscript{41} Since
+dataset according to the Kaplan-Meier method.\textsuperscript{40} Since
 survival probabilities typically do not follow normal distributions
@@
 through a complementary log-log transformation \(log(-log(1-pr(surv)))\)
-with \(pr(surv)\) denoting the survival probability at a given time
-during follow-up.\textsuperscript{50, 51} The transformed survival
+with \(pr(surv)\) denoting the survival probability at a given follow-up
+time during.\textsuperscript{49, 50} The transformed survival
 probabilities are then pooled across imputed datasets and individual
@@
 
-With the goal to better understand which factors could influence a
-difference between RCT results and emulated trial results, it is
+With the goal to better understand which factors could influence
+differences between RCT results and emulated trial results, it is
 appropriate to conduct a range of sensitivity analyses. This can
@@
 sensitivity versus specificity in measurements), approaches to missing
-data, selection of covariates for imputation and propensity score models
-and censoring decisions. All sensitivity analyses will be pre-specified
-in the study protocol and reported using appropriate visualizations such
-as forest plots.
+data, selection of covariates for imputation and propensity score
+models, and censoring decisions. All sensitivity analyses will be
+pre-specified in the study protocol and reported using appropriate
+visualizations such as forest plots.
 
@@
 measurement of covariates and outcomes using parameterized functions.
-Due to the blinding of databases used for ENCORE, this package cannot be
-made public but we provide a comprehensive and detailed documentation
-for each function in the Supplementary Material. A second open-source R
-package, \texttt{encore.analytics}, will then facilitate the complex
-multi-step workflows of multiple imputation, propensity score matching
-and weighting and estimating pooled marginal treatment effects and
-Kaplan-Meier curves along with other established statistical R
-packages.\textsuperscript{52--57} Simulated examples with code that
-illustrate the analytic workflows described in this manuscript are
-available in the online Supplementary Material at
+Due to the blinding of databases used for ENCORE, this package is not
+able to be publicly available; however a comprehensive and detailed
+documentation for each function are available in the Supplementary
+Material. A second open-source R package, \texttt{encore.analytics},
+will then facilitate the complex multi-step workflows of multiple
+imputation, propensity score matching and weighting and estimating
+pooled marginal treatment effects and Kaplan-Meier curves along with
+other established statistical R packages.\textsuperscript{51--56}
+Simulated examples with code that illustrate the analytic workflows
+described in this manuscript are available in the online Supplementary
+Material at
 \url{https://janickweberpals.github.io/imputation-ps-workflows/} and can
@@
 Building on an established approach developed by the RCT DUPLICATE
-project\textsuperscript{9, 31}, the ENCORE project aims to emulate 12
+project\textsuperscript{8, 30}, the ENCORE project aims to emulate 12
 oncology trials using four oncology EHR data sources to inform the
@@
 data source used to evaluate medical product safety and effectiveness in
-the postmarketing setting. With increasing access to EHR data, which are
-more fit for use in oncology than claims due to availability of clinical
-data elements, and a maturing set of methodological approaches for
-causal inference using such data, the ENCORE project is one of the first
-of its kind to evaluate when and how RWD can be used to evaluate similar
-causal conclusions compared to RCTs in the field of oncology.
-
-The RCT DUPLICATE initiative\textsuperscript{9, 31} identified multiple
+the postmarketing setting. With increasing access to EHR data and
+availability of granular clinical data elements often necessary to
+gather fit-for-use data in oncology, as well as a maturing set of
+methodological approaches for causal inference using such data, the
+ENCORE project is one of the first of its kind to evaluate contexts
+where RWD may be reliably used to draw similar conclusions compared to
+RCTs in the field of oncology.
+
+The RCT DUPLICATE initiative\textsuperscript{8, 30} identified multiple
 emulation challenges that may be similar in ENCORE. One particular
 aspect that we may not always be able to emulate is the exact
-distribution of patient characteristics of the trial population, or that
-the comparability between these two populations may not be known. This
-can be due to the lack of RWD granularity and comprehensiveness to
+distribution of patient characteristics of the trial population. This
+can be due to the lack of data granularity and comprehensiveness to
 emulate relevant eligibility criteria or the fact that large pivotal
 trials in oncology are typically conducted in multiple countries
-worldwide (all considered databases reflect the US only). This may be a
-factor especially given that the pathophysiology, prognosis, and factors
-that drive heterogeneous treatment outcomes of certain cancers differ
-between countries. For example, some cancer indications (e.g., GI
-cancers) or some genetic mutations (e.g., EGFR mutations in lung
-adenocarcinoma) are much more prevalent in certain geographic regions
-compared to the US.
+worldwide (all considered databases reflect the US only). This may
+impact the results of this trial emulation given that the distribution
+of genetic alterations and biomarkers, prognosis, and factors that drive
+heterogeneous treatment outcomes of certain cancers can differ between
+countries. For example, certain cancers (e.g., GI cancers) or certain
+genetic mutations (e.g., EGFR mutations in lung cancer) are much more
+prevalent in certain geographic regions and populations compared to the
+US. Additionally, treatment landscapes such as the preference for
+specific regimens or compounds, their use in earlier or later lines of
+therapies, and supportive care practices can differ between geographic
+regions, too.
 
@@
 high proportion of patients from the control to the intervention arm, or
-differences in subsequent therapy lines there may be estimand
-differences between trial and emulation.\textsuperscript{27} Although
-this is also a common challenge in the analysis of
-RCTs,\textsuperscript{26} treatment practice in routine clinical
-practice might be less stringent compared to trials.\textsuperscript{8}
-In order to derive comparable estimands it is therefore crucial to
-understand and contextualize the proportion and timing of treatment
-switching and discontinuation in both the trial and its emulation.
+differences in subsequent therapies there may be observed differences in
+outcomes between trial and emulation.\textsuperscript{26} Although this
+is also a common challenge in the analysis of RCTs,\textsuperscript{25}
+treatment in routine clinical practice might be less stringent compared
+to trials in terms of pre-specified dosing schedules, monitoring and
+surveillance.\textsuperscript{7} In order to derive comparable estimands
+it is therefore crucial to understand and contextualize the proportion
+and timing of treatment switching and discontinuation in both the trial
+and its emulation.
 
@@
 calibrate in RWD. As a result, exposure misclassification (due to overly
-ascertainment windows) and selection bias (due to overly long
+narrow ascertainment windows) and selection bias (due to overly long
 ascertainment windows) are common trade-offs in such scenarios.
 Alternative analytic approaches which target a per-protocol estimand,
-such as the clone-censor-weight design,\textsuperscript{58} may be
+such as the clone-censor-weight design,\textsuperscript{57} may be
 viable options if the ITT estimand cannot be estimated due to the
@@
 
-RWE based on fit-for-purpose data and principled, well-designed and
-reproducible studies may complement evidence from RCTs. Through a
-systematic benchmarking approach, the ENCORE project will provide
-insights as to how measurement, design and analytic decisions influence
-bias and validity in oncological non-interventional studies.
+Principled, well-designed and reproducible studies using fit-for-purpose
+RWD to generate RWE, such as through trial emulation, may complement
+evidence from RCTs. Through a systematic benchmarking approach, the
+ENCORE project will provide insights as to how methodological, design,
+and analytic decisions influence quantifiable bias and interpretability
+of observational studies in oncology.
 
@@
 
-\bibitem[\citeproctext]{ref-purpura2022role}
-\textbf{2}. Purpura CA, Garry EM, Honig N, et al: The role of real-world
-evidence in FDA-approved new drug and biologics license applications.
-Clinical Pharmacology \& Therapeutics 111:135--144, 2022
-
-\bibitem[\citeproctext]{ref-senior2024fresh}
-\textbf{3}. Senior M: Fresh from the biotech pipeline: Record-breaking
-FDA approvals. Nature Biotechnology, 2024
+\bibitem[\citeproctext]{ref-mullard20242023}
+\textbf{2}. Mullard A: 2023 FDA approvals. Nat Rev Drug Discov
+23:88--95, 2024
 
 \bibitem[\citeproctext]{ref-Merola2022}
-\textbf{4}. Merola D, Schneeweiss S, Schrag D, et al: An algorithm to
+\textbf{3}. Merola D, Schneeweiss S, Schrag D, et al: An algorithm to
 predict data completeness in oncology electronic medical records for
@@
 \bibitem[\citeproctext]{ref-joshua2022longitudinal}
-\textbf{5}. Joshua Lin K, Jin Y, Gagne J, et al: Longitudinal data
+\textbf{4}. Joshua Lin K, Jin Y, Gagne J, et al: Longitudinal data
 discontinuity in electronic health records and consequences for
@@
 \bibitem[\citeproctext]{ref-rider2024emulations}
-\textbf{6}. Rider JR, Wasserman A, Slipski L, et al: Emulations of
+\textbf{5}. Rider JR, Wasserman A, Slipski L, et al: Emulations of
 oncology trials using real-world data: A systematic literature review.
@@
 \bibitem[\citeproctext]{ref-merola2023aetion}
-\textbf{7}. Merola D, Campbell U, Gautam N, et al: The aetion coalition
+\textbf{6}. Merola D, Campbell U, Gautam N, et al: The aetion coalition
 to advance real-world evidence through randomized controlled trial
@@
 \bibitem[\citeproctext]{ref-merola2024calibrating}
-\textbf{8}. Merola D, Campbell U, Lenis D, et al: Calibrating
+\textbf{7}. Merola D, Campbell U, Lenis D, et al: Calibrating
 observational health record data against a randomized trial. JAMA
@@
 \bibitem[\citeproctext]{ref-wang2023emulation}
-\textbf{9}. Wang SV, Schneeweiss S, Franklin JM, et al:
+\textbf{8}. Wang SV, Schneeweiss S, Franklin JM, et al:
 \href{https://doi.org/10.1001/jama.2023.4221}{Emulation of randomized
@@
 \bibitem[\citeproctext]{ref-heyard2024design}
-\textbf{10}. Heyard R, Held L, Schneeweiss S, et al: Design differences
+\textbf{9}. Heyard R, Held L, Schneeweiss S, et al: Design differences
 and variation in results between randomised trials and non-randomised
@@
 \bibitem[\citeproctext]{ref-encoreFDA}
-\textbf{11}. Calibrating real-world evidence studies in oncology against
+\textbf{10}. Calibrating real-world evidence studies in oncology against
 randomized trials: ENCORE (last accessed 11/28/2024) {[}Internet{]},
@@
 \bibitem[\citeproctext]{ref-rivera2024oncology}
-\textbf{12}. Rivera DR, Eckert JC, Rodriguez-Watson C, et al: The
+\textbf{11}. Rivera DR, Eckert JC, Rodriguez-Watson C, et al: The
 oncology QCARD initiative: Fostering efficient evaluation of initial
@@
 \bibitem[\citeproctext]{ref-gatto2022structured}
-\textbf{13}. Gatto NM, Campbell UB, Rubinstein E, et al: The structured
+\textbf{12}. Gatto NM, Campbell UB, Rubinstein E, et al: The structured
 process to identify fit-for-purpose data: A data feasibility assessment
@@
 \bibitem[\citeproctext]{ref-tasneem2012database}
-\textbf{14}. Tasneem A, Aberle L, Ananth H, et al: The database for
+\textbf{13}. Tasneem A, Aberle L, Ananth H, et al: The database for
 aggregate analysis of ClinicalTrials. Gov (AACT) and subsequent
@@
 \bibitem[\citeproctext]{ref-eisenhauer2009new}
-\textbf{15}. Eisenhauer EA, Therasse P, Bogaerts J, et al: New response
+\textbf{14}. Eisenhauer EA, Therasse P, Bogaerts J, et al: New response
 evaluation criteria in solid tumours: Revised RECIST guideline (version
@@
 \bibitem[\citeproctext]{ref-ThanCCR}
-\textbf{16}. Ton TGN, Pal N, Trinh H, et al: Replication of overall
+\textbf{15}. Ton TGN, Pal N, Trinh H, et al: Replication of overall
 survival, progression-free survival, and overall response in
@@
 \bibitem[\citeproctext]{ref-mckelvey2024evaluation}
-\textbf{17}. McKelvey BA, Garrett-Mayer E, Rivera DR, et al: Evaluation
+\textbf{16}. McKelvey BA, Garrett-Mayer E, Rivera DR, et al: Evaluation
 of real-world tumor response derived from electronic health record data
@@
 \bibitem[\citeproctext]{ref-rwdRECIST}
-\textbf{18}. Chen L, Davis R, Lee J, et al: Comparison of response from
+\textbf{17}. Chen L, Davis R, Lee J, et al: Comparison of response from
 RECIST1.1 and abstraction in real-world patients with lung cancer.
@@
 \bibitem[\citeproctext]{ref-rivera2022friends}
-\textbf{19}. Rivera DR, Henk HJ, Garrett-Mayer E, et al: The friends of
+\textbf{18}. Rivera DR, Henk HJ, Garrett-Mayer E, et al: The friends of
 cancer research real-world data collaboration pilot 2.0: Methodological
@@
 \bibitem[\citeproctext]{ref-ackerman2024measurement}
-\textbf{20}. Ackerman B, Gan RW, Meyer CS, et al: Measurement error and
+\textbf{19}. Ackerman B, Gan RW, Meyer CS, et al: Measurement error and
 bias in real-world oncology endpoints when constructing external control
@@
 \bibitem[\citeproctext]{ref-nikolakopoulou2014interpret}
-\textbf{21}. Nikolakopoulou A, Mavridis D, Salanti G: How to interpret
+\textbf{20}. Nikolakopoulou A, Mavridis D, Salanti G: How to interpret
 meta-analysis models: Fixed effect and random effects meta-analyses. BMJ
@@
 \bibitem[\citeproctext]{ref-wang2022harmonized}
-\textbf{22}. Wang SV, Pottegård A, Crown W, et al: HARmonized protocol
+\textbf{21}. Wang SV, Pottegård A, Crown W, et al: HARmonized protocol
 template to enhance reproducibility of hypothesis evaluating real-world
@@
 \bibitem[\citeproctext]{ref-guidelinegeneral}
-\textbf{23}. ICH guideline. General principles on plan, design and
+\textbf{22}. ICH guideline. General principles on plan, design and
 analysis of pharmacoepidemiological studies that utilize real-world data
@@
 \bibitem[\citeproctext]{ref-hernan2022target}
-\textbf{24}. Hernán MA, Wang W, Leaf DE: Target trial emulation: A
+\textbf{23}. Hernán MA, Wang W, Leaf DE: Target trial emulation: A
 framework for causal inference from observational data. Jama
@@
 \bibitem[\citeproctext]{ref-hernan2016specifying}
-\textbf{25}. Hernán MA, Sauer BC, Hernández-Dı́az S, et al:
+\textbf{24}. Hernán MA, Sauer BC, Hernández-Dı́az S, et al:
 \href{https://doi.org/10.1016/j.jclinepi.2016.04.014}{Specifying a
@@
 \bibitem[\citeproctext]{ref-rufibach2018}
-\textbf{26}. Rufibach K: Treatment effect quantification for
+\textbf{25}. Rufibach K: Treatment effect quantification for
 time{-}to{-}event endpoints{\textendash}Estimands, analysis strategies,
@@
 \bibitem[\citeproctext]{ref-manitz2022estimands}
-\textbf{27}. Manitz J, Kan-Dobrosky N, Buchner H, et al: Estimands for
+\textbf{26}. Manitz J, Kan-Dobrosky N, Buchner H, et al: Estimands for
 overall survival in clinical trials with treatment switching in
@@
 \bibitem[\citeproctext]{ref-meyer2020open}
-\textbf{28}. Meyer A-M, Davies J, Taylor M, et al: Open cohorts and
+\textbf{27}. Meyer A-M, Davies J, Taylor M, et al: Open cohorts and
 ghost-time bias in real world data, in PHARMACOEPIDEMIOLOGY AND DRUG
@@
 \bibitem[\citeproctext]{ref-hernan2022causal}
-\textbf{29}. Hernán MA: Causal analyses of existing databases: No power
+\textbf{28}. Hernán MA: Causal analyses of existing databases: No power
 calculations required. Journal of clinical epidemiology 144:203--205,
@@
 \bibitem[\citeproctext]{ref-schoenfeld1981asymptotic}
-\textbf{30}. Schoenfeld D: The asymptotic properties of nonparametric
+\textbf{29}. Schoenfeld D: The asymptotic properties of nonparametric
 tests for comparing survival distributions. Biometrika 68:316--319, 1981
@@
 \bibitem[\citeproctext]{ref-franklin2020nonrandomized}
-\textbf{31}. Franklin JM, Pawar A, Martin D, et al: Nonrandomized
+\textbf{30}. Franklin JM, Pawar A, Martin D, et al: Nonrandomized
 real-world evidence to support regulatory decision making: Process for a
@@
 \bibitem[\citeproctext]{ref-schneeweiss2019graphical}
-\textbf{32}. Schneeweiss S, Rassen JA, Brown JS, et al: Graphical
+\textbf{31}. Schneeweiss S, Rassen JA, Brown JS, et al: Graphical
 depiction of longitudinal study designs in health care databases. Annals
@@
 \bibitem[\citeproctext]{ref-rubin1976inference}
-\textbf{33}. Rubin DB: Inference and missing data. Biometrika
+\textbf{32}. Rubin DB: Inference and missing data. Biometrika
 63:581--592, 1976
@@
 \bibitem[\citeproctext]{ref-weberpals2024smdi}
-\textbf{34}. Weberpals J, Raman SR, Shaw PA, et al: Smdi: An r package
+\textbf{33}. Weberpals J, Raman SR, Shaw PA, et al: Smdi: An r package
 to perform structural missing data investigations on partially observed
@@
 \bibitem[\citeproctext]{ref-weberpals2024}
-\textbf{35}. Weberpals J, Raman SR, Shaw PA, et al: A principled
+\textbf{34}. Weberpals J, Raman SR, Shaw PA, et al: A principled
 approach to characterize and analyze partially observed confounder data
@@
 \bibitem[\citeproctext]{ref-moreno2018canonical}
-\textbf{36}. Moreno-Betancur M, Lee KJ, Leacy FP, et al: Canonical
+\textbf{35}. Moreno-Betancur M, Lee KJ, Leacy FP, et al: Canonical
 causal diagrams to guide the treatment of missing data in epidemiologic
@@
 \bibitem[\citeproctext]{ref-tompsett2018use}
-\textbf{37}. Tompsett DM, Leacy F, Moreno-Betancur M, et al: On the use
+\textbf{36}. Tompsett DM, Leacy F, Moreno-Betancur M, et al: On the use
 of the not-at-random fully conditional specification (NARFCS) procedure
@@
 \bibitem[\citeproctext]{ref-weberpals2024hdmi}
-\textbf{38}. Weberpals J, Shaw PA, Lin KJ, et al: High-dimensional
+\textbf{37}. Weberpals J, Shaw PA, Lin KJ, et al: High-dimensional
 multiple imputation (HDMI) for partially observed confounders including
@@
 \bibitem[\citeproctext]{ref-Shaw2024}
-\textbf{39}. Shaw CK P., Williamson BD: Assessing treatment effects in
+\textbf{38}. Shaw CK P., Williamson BD: Assessing treatment effects in
 observational data with missing confounders: A comparative study of
@@
 \bibitem[\citeproctext]{ref-cox1972regression}
-\textbf{40}. Cox DR: Regression models and life-tables. Journal of the
+\textbf{39}. Cox DR: Regression models and life-tables. Journal of the
 royal statistical society. Series B (Methodological) 34:187--220, 1972
@@
 \bibitem[\citeproctext]{ref-kaplan1958nonparametric}
-\textbf{41}. Kaplan EL, Meier P: Nonparametric estimation from
+\textbf{40}. Kaplan EL, Meier P: Nonparametric estimation from
 incomplete observations. Journal of the American statistical association
@@
 \bibitem[\citeproctext]{ref-becker2020enhanced}
-\textbf{42}. Becker T, Weberpals J, Jegg A, et al: An enhanced
+\textbf{41}. Becker T, Weberpals J, Jegg A, et al: An enhanced
 prognostic score for overall survival of patients with cancer derived
@@
 \bibitem[\citeproctext]{ref-leyrat2019propensity}
-\textbf{43}. Leyrat C, Seaman SR, White IR, et al: Propensity score
+\textbf{42}. Leyrat C, Seaman SR, White IR, et al: Propensity score
 analysis with partially observed covariates: How should multiple
@@
 \bibitem[\citeproctext]{ref-pishgar2020matchthem}
-\textbf{44}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
+\textbf{43}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
 and weighting after multiple imputation. arXiv preprint arXiv:200911772,
@@
 \bibitem[\citeproctext]{ref-rubin2018multiple}
-\textbf{45}. Rubin DB: Multiple imputation, in Flexible imputation of
+\textbf{44}. Rubin DB: Multiple imputation, in Flexible imputation of
 missing data, second edition. Chapman; Hall/CRC, 2018, pp 29--62
@@
 \bibitem[\citeproctext]{ref-van2011mice}
-\textbf{46}. Van Buuren S, Groothuis-Oudshoorn K: Mice: Multivariate
+\textbf{45}. Van Buuren S, Groothuis-Oudshoorn K: Mice: Multivariate
 imputation by chained equations in r. Journal of statistical software
@@
 \bibitem[\citeproctext]{ref-austin2009balance}
-\textbf{47}. Austin PC: Balance diagnostics for comparing the
+\textbf{46}. Austin PC: Balance diagnostics for comparing the
 distribution of baseline covariates between treatment groups in
@@
 \bibitem[\citeproctext]{ref-franklin2014metrics}
-\textbf{48}. Franklin JM, Rassen JA, Ackermann D, et al: Metrics for
+\textbf{47}. Franklin JM, Rassen JA, Ackermann D, et al: Metrics for
 covariate balance in cohort studies of causal effects. Statistics in
@@
 \bibitem[\citeproctext]{ref-stuart2013prognostic}
-\textbf{49}. Stuart EA, Lee BK, Leacy FP: Prognostic score--based
+\textbf{48}. Stuart EA, Lee BK, Leacy FP: Prognostic score--based
 balance measures can be a useful diagnostic for propensity score methods
@@
 \bibitem[\citeproctext]{ref-warwick28685}
-\textbf{50}. Marshall A(Andrea), Billingham LJ, Bryan S: Can we afford
+\textbf{49}. Marshall A(Andrea), Billingham LJ, Bryan S: Can we afford
 to ignore missing data in cost-effectiveness analyses? {[}Internet{]}.
@@
 \bibitem[\citeproctext]{ref-morisot2015prostate}
-\textbf{51}. Morisot A, Bessaoud F, Landais P, et al: Prostate cancer:
+\textbf{50}. Morisot A, Bessaoud F, Landais P, et al: Prostate cancer:
 Net survival and cause-specific survival rates after multiple
@@
 \bibitem[\citeproctext]{ref-anesrake}
-\textbf{52}. Pasek J: Anesrake: ANES raking implementation, 2018
+\textbf{51}. Pasek J: Anesrake: ANES raking implementation, 2018
 
 \bibitem[\citeproctext]{ref-MatchThem}
-\textbf{53}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
+\textbf{52}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
 and weighting after multiple imputation {[}Internet{]}, 2021Available
@@
 \bibitem[\citeproctext]{ref-survival}
-\textbf{54}. Therneau TM: A package for survival analysis in r
+\textbf{53}. Therneau TM: A package for survival analysis in r
 {[}Internet{]}, 2024Available from:
@@
 \bibitem[\citeproctext]{ref-encore.analytics}
-\textbf{55}. Weberpals J: Encore.analytics: Functions and wrappers to
+\textbf{54}. Weberpals J: Encore.analytics: Functions and wrappers to
 streamline complex analytic workflows in real-world data studies based
@@
 \bibitem[\citeproctext]{ref-gtsummary}
-\textbf{56}. Sjoberg DD, Whiting K, Curry M, et al: Reproducible summary
+\textbf{55}. Sjoberg DD, Whiting K, Curry M, et al: Reproducible summary
 tables with the gtsummary package {[}Internet{]}13:570--580,
@@
 \bibitem[\citeproctext]{ref-ggsurvfit}
-\textbf{57}. Sjoberg DD, Baillie M, Fruechtenicht C, et al: Ggsurvfit:
+\textbf{56}. Sjoberg DD, Baillie M, Fruechtenicht C, et al: Ggsurvfit:
 Flexible time-to-event figures {[}Internet{]}, 2024Available from:
@@
 \bibitem[\citeproctext]{ref-gaber2024mystifying}
-\textbf{58}. Gaber CE, Ghazarian AA, Strassle PD, et al: De-mystifying
+\textbf{57}. Gaber CE, Ghazarian AA, Strassle PD, et al: De-mystifying
 the clone-censor-weight method for causal research using observational

### 2025-07-06 manuscript/manuscript.qmd
updated word count

@@
 
-[**Manuscript word count:**]{.underline} 4,509 words / 8,000 words
+[**Manuscript word count:**]{.underline} 4,541 words / 8,000 words
 
-[**Abstract word count:**]{.underline} 224 words / 250 words
+[**Abstract word count:**]{.underline} 228 words / 250 words
 
@@
 [**Analytic code sharing statement:**]{.underline} Simulated examples and code to implement analytic workflows described in this manuscript are illustrated at <https://janickweberpals.github.io/imputation-ps-workflows/> and can be reproduced via the `encore.analytics` R package (<https://github.com/janickweberpals/encore.analytics/>).
-
-[***Proposed target journals:***]{.underline} *CPT =\> JCO CCI =\> ...*
 

### 2025-07-06 manuscript/manuscript.qmd
corrected year according to updated reference

@@
 Randomized controlled trials (RCTs) are the evidentiary gold standard methodology for establishing the efficacy and safety of medical products.
-Guided by the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor considerations for use of real-world evidence (RWE) generated from real-world data (RWD) such as electronic health records (EHR) to support approval of new indications or satisfy postmarketing requirements.[@RWEFDA] Accounting for approximately 30% of new FDA drug approvals, oncology was the disease area with the most new approvals in 2024[@mullard20242023] as well as several indication expansions, and has many areas of high unmet medical need to treat serious conditions.
+Guided by the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor considerations for use of real-world evidence (RWE) generated from real-world data (RWD) such as electronic health records (EHR) to support approval of new indications or satisfy postmarketing requirements.[@RWEFDA] Accounting for approximately 30% of new FDA drug approvals, oncology was the disease area with the most new approvals in 2023[@mullard20242023] as well as several indication expansions, and has many areas of high unmet medical need to treat serious conditions.
 Therefore, RWE can have a particularly important potential to complement evidence from RCTs in the field of oncology.

### 2025-07-06 manuscript/manuscript.tex
Manuscript updated

@@
 
-\ul{\textbf{Manuscript word count:}} 4,509 words / 8,000 words
-
-\ul{\textbf{Abstract word count:}} 224 words / 250 words
+\ul{\textbf{Manuscript word count:}} 4,541 words / 8,000 words
+
+\ul{\textbf{Abstract word count:}} 228 words / 250 words
 
@@
 
-\ul{\textbf{\emph{Proposed target journals:}}} \emph{CPT =\textgreater{}
-JCO CCI =\textgreater{} \ldots{}}
-
-\emph{Manuscript last updated: 2025-07-06 14:54:57.141696}
+\emph{Manuscript last updated: 2025-07-06 16:21:00.006613}
 
@@
 approximately 30\% of new FDA drug approvals, oncology was the disease
-area with the most new approvals in 2024\textsuperscript{2} as well as
+area with the most new approvals in 2023\textsuperscript{2} as well as
 several indication expansions, and has many areas of high unmet medical
@@
 unmeasured confounding, small study sizes, data
-discontinuity,\textsuperscript{3, 4} adoption of changes in treatment
+discontinuity,\textsuperscript{3,4} adoption of changes in treatment
 guidelines over time in routine clinical care and the inability to
@@
 sources. The process will emphasize transparency and include a data
-fitness-for-use assessment\textsuperscript{11, 12} for each RWD source
+fitness-for-use assessment\textsuperscript{11,12} for each RWD source
 with respect to each trial emulation as well as extensive sensitivity
@@
 RECISTv1.1\textsuperscript{14}-based PFS and ORR assessment in
-RCTs\textsuperscript{15, 16}, imaging-based evaluations still require a
+RCTs\textsuperscript{15,16}, imaging-based evaluations still require a
 level of granularity which is not well reflected in chart-abstracted
 assessments of a patient's progression in routine
-care.\textsuperscript{17, 18} The timing and cadence of intervals
-between progression assessments can differ between RCTs and routine care
-which may result in measurement error and bias.\textsuperscript{19}
-Given this, and the large number of other methodological challenges
-including missing data, small sample sizes, data discontinuity and
-rapidly changing guideline treatments, we focus on the emulation of
-overall survival (OS) as the endpoint of interest. Therefore, we include
-only trials that have reported OS as a pre-specified endpoint in the
+care.\textsuperscript{17,18} The timing and cadence of intervals between
+progression assessments can differ between RCTs and routine care which
+may result in measurement error and bias.\textsuperscript{19} Given
+this, and the large number of other methodological challenges including
+missing data, small sample sizes, data discontinuity and rapidly
+changing guideline treatments, we focus on the emulation of overall
+survival (OS) as the endpoint of interest. Therefore, we include only
+trials that have reported OS as a pre-specified endpoint in the
 protocol.
@@
 of the trial emulation, will be developed using the HARPER protocol
-template\textsuperscript{21, 22} and will be registered on
+template\textsuperscript{21,22} and will be registered on
 ClinicalTrials.gov after review by an expert panel comprising clinicians
@@
 operationalization of key eligibility criteria, study design, data
-analysis and causal contrasts of interest.\textsuperscript{23, 24} Since
+analysis and causal contrasts of interest.\textsuperscript{23,24} Since
 it is common that oncology RCTs update survival estimates periodically
@@
 emulations, we will adapt the approach of the RCT-DUPLICATE
-project.\textsuperscript{8, 30} That is, for the primary endpoint of
+project.\textsuperscript{8,30} That is, for the primary endpoint of
 interest (hazard ratio {[}HR{]} for OS and corresponding 95\% confidence
@@
 aspects across partially observed covariates based on three group
-diagnostics.\textsuperscript{33, 34} These diagnostics cover (1)
+diagnostics.\textsuperscript{33,34} These diagnostics cover (1)
 comparisons of patients characteristics with and without an observed
@@
 the `\emph{within}' methodology as described by Leyrat et
-al.\textsuperscript{42, 43} As appropriate, propensity score matching or
+al.\textsuperscript{42,43} As appropriate, propensity score matching or
 weighting will be applied to each imputed dataset. The marginal
@@
 weighted dataset separately and pooled into a final estimate following
-Rubin's rule.\textsuperscript{44, 45} This approach has been shown to
+Rubin's rule.\textsuperscript{44,45} This approach has been shown to
 lead to unbiased estimates across different simulated scenarios with a
@@
 with \(pr(surv)\) denoting the survival probability at a given follow-up
-time during.\textsuperscript{49, 50} The transformed survival
+time during.\textsuperscript{49,50} The transformed survival
 probabilities are then pooled across imputed datasets and individual
@@
 Building on an established approach developed by the RCT DUPLICATE
-project\textsuperscript{8, 30}, the ENCORE project aims to emulate 12
+project\textsuperscript{8,30}, the ENCORE project aims to emulate 12
 oncology trials using four oncology EHR data sources to inform the
@@
 
-The RCT DUPLICATE initiative\textsuperscript{8, 30} identified multiple
+The RCT DUPLICATE initiative\textsuperscript{8,30} identified multiple
 emulation challenges that may be similar in ENCORE. One particular
@@
 \phantomsection\label{refs}
-\begin{CSLReferences}{0}{1}
+\begin{CSLReferences}{0}{0}
 \bibitem[\citeproctext]{ref-RWEFDA}
-\textbf{1}. Framework for FDA's real-world evidence program (last
-accessed 11/28/2024) {[}Internet{]}, 2018Available from:
-\url{https://www.fda.gov/media/120060/download?attachment}
+\CSLLeftMargin{1. }%
+\CSLRightInline{Framework for FDA's real-world evidence program (last
+accessed 11/28/2024). (2018).at
+\textless{}\url{https://www.fda.gov/media/120060/download?attachment}\textgreater{}}
 
 \bibitem[\citeproctext]{ref-mullard20242023}
-\textbf{2}. Mullard A: 2023 FDA approvals. Nat Rev Drug Discov
-23:88--95, 2024
+\CSLLeftMargin{2. }%
+\CSLRightInline{Mullard, A. 2023 FDA approvals. \emph{Nat Rev Drug
+Discov} \textbf{23}, 88--95 (2024).}
 
 \bibitem[\citeproctext]{ref-Merola2022}
-\textbf{3}. Merola D, Schneeweiss S, Schrag D, et al: An algorithm to
-predict data completeness in oncology electronic medical records for
-comparative effectiveness research {[}Internet{]}. Annals of
-Epidemiology 76:143--149, 2022Available from:
-\url{http://dx.doi.org/10.1016/j.annepidem.2022.07.007}
+\CSLLeftMargin{3. }%
+\CSLRightInline{Merola, D., Schneeweiss, S., Schrag, D., Lii, J. \& Lin,
+K. J. \href{https://doi.org/10.1016/j.annepidem.2022.07.007}{An
+algorithm to predict data completeness in oncology electronic medical
+records for comparative effectiveness research}. \emph{Annals of
+Epidemiology} \textbf{76}, 143--149 (2022).}
 
 \bibitem[\citeproctext]{ref-joshua2022longitudinal}
-\textbf{4}. Joshua Lin K, Jin Y, Gagne J, et al: Longitudinal data
+\CSLLeftMargin{4. }%
+\CSLRightInline{Joshua Lin, K. \emph{et al.} Longitudinal data
 discontinuity in electronic health records and consequences for
-medication effectiveness studies. Clinical Pharmacology \& Therapeutics
-111:243--251, 2022
+medication effectiveness studies. \emph{Clinical Pharmacology \&
+Therapeutics} \textbf{111}, 243--251 (2022).}
 
 \bibitem[\citeproctext]{ref-rider2024emulations}
-\textbf{5}. Rider JR, Wasserman A, Slipski L, et al: Emulations of
-oncology trials using real-world data: A systematic literature review.
-American journal of epidemiology kwae346, 2024
+\CSLLeftMargin{5. }%
+\CSLRightInline{Rider, J. R. \emph{et al.} Emulations of oncology trials
+using real-world data: A systematic literature review. \emph{American
+journal of epidemiology} kwae346 (2024).}
 
 \bibitem[\citeproctext]{ref-merola2023aetion}
-\textbf{6}. Merola D, Campbell U, Gautam N, et al: The aetion coalition
-to advance real-world evidence through randomized controlled trial
-emulation initiative: oncology. Clinical Pharmacology \& Therapeutics
-113:1217--1222, 2023
+\CSLLeftMargin{6. }%
+\CSLRightInline{Merola, D. \emph{et al.} The aetion coalition to advance
+real-world evidence through randomized controlled trial emulation
+initiative: oncology. \emph{Clinical Pharmacology \& Therapeutics}
+\textbf{113}, 1217--1222 (2023).}
 
 \bibitem[\citeproctext]{ref-merola2024calibrating}
-\textbf{7}. Merola D, Campbell U, Lenis D, et al: Calibrating
-observational health record data against a randomized trial. JAMA
-Network Open 7:e2436535--e2436535, 2024
+\CSLLeftMargin{7. }%
+\CSLRightInline{Merola, D. \emph{et al.} Calibrating observational
+health record data against a randomized trial. \emph{JAMA Network Open}
+\textbf{7}, e2436535--e2436535 (2024).}
 
 \bibitem[\citeproctext]{ref-wang2023emulation}
-\textbf{8}. Wang SV, Schneeweiss S, Franklin JM, et al:
+\CSLLeftMargin{8. }%
+\CSLRightInline{Wang, S. V. \emph{et al.}
 \href{https://doi.org/10.1001/jama.2023.4221}{Emulation of randomized
 clinical trials with nonrandomized database analyses: Results of 32
-clinical trials}. Jama 329:1376--1385, 2023
+clinical trials}. \emph{Jama} \textbf{329}, 1376--1385 (2023).}
 
 \bibitem[\citeproctext]{ref-heyard2024design}
-\textbf{9}. Heyard R, Held L, Schneeweiss S, et al: Design differences
-and variation in results between randomised trials and non-randomised
-emulations: Meta-analysis of RCT-DUPLICATE data {[}Internet{]}. BMJ
-medicine 3, 2024Available from:
-\url{https://doi.org/10.1136/bmjmed-2023-000709}
+\CSLLeftMargin{9. }%
+\CSLRightInline{Heyard, R., Held, L., Schneeweiss, S. \& Wang, S. V.
+\href{https://doi.org/10.1136/bmjmed-2023-000709}{Design differences and
+variation in results between randomised trials and non-randomised
+emulations: Meta-analysis of RCT-DUPLICATE data}. \emph{BMJ medicine}
+\textbf{3}, (2024).}
 
 \bibitem[\citeproctext]{ref-encoreFDA}
-\textbf{10}. Calibrating real-world evidence studies in oncology against
-randomized trials: ENCORE (last accessed 11/28/2024) {[}Internet{]},
-2024Available from:
-\url{https://www.fda.gov/about-fda/oncology-center-excellence/calibrating-real-world-evidence-studies-oncology-against-randomized-trials-encore}
+\CSLLeftMargin{10. }%
+\CSLRightInline{Calibrating real-world evidence studies in oncology
+against randomized trials: ENCORE (last accessed 11/28/2024). (2024).at
+\textless{}\url{https://www.fda.gov/about-fda/oncology-center-excellence/calibrating-real-world-evidence-studies-oncology-against-randomized-trials-encore}\textgreater{}}
 
 \bibitem[\citeproctext]{ref-rivera2024oncology}
-\textbf{11}. Rivera DR, Eckert JC, Rodriguez-Watson C, et al: The
-oncology QCARD initiative: Fostering efficient evaluation of initial
-real-world data proposals. Pharmacoepidemiology and Drug Safety
-33:e5818, 2024
+\CSLLeftMargin{11. }%
+\CSLRightInline{Rivera, D. R. \emph{et al.} The oncology QCARD
+initiative: Fostering efficient evaluation of initial real-world data
+proposals. \emph{Pharmacoepidemiology and Drug Safety} \textbf{33},
+e5818 (2024).}
 
 \bibitem[\citeproctext]{ref-gatto2022structured}
-\textbf{12}. Gatto NM, Campbell UB, Rubinstein E, et al: The structured
-process to identify fit-for-purpose data: A data feasibility assessment
-framework. Clinical Pharmacology \& Therapeutics 111:122--134, 2022
+\CSLLeftMargin{12. }%
+\CSLRightInline{Gatto, N. M. \emph{et al.} The structured process to
+identify fit-for-purpose data: A data feasibility assessment framework.
+\emph{Clinical Pharmacology \& Therapeutics} \textbf{111}, 122--134
+(2022).}
 
 \bibitem[\citeproctext]{ref-tasneem2012database}
-\textbf{13}. Tasneem A, Aberle L, Ananth H, et al: The database for
-aggregate analysis of ClinicalTrials. Gov (AACT) and subsequent
-regrouping by clinical specialty. PloS one 7:e33677, 2012
+\CSLLeftMargin{13. }%
+\CSLRightInline{Tasneem, A. \emph{et al.} The database for aggregate
+analysis of ClinicalTrials. Gov (AACT) and subsequent regrouping by
+clinical specialty. \emph{PloS one} \textbf{7}, e33677 (2012).}
 
 \bibitem[\citeproctext]{ref-eisenhauer2009new}
-\textbf{14}. Eisenhauer EA, Therasse P, Bogaerts J, et al: New response
-evaluation criteria in solid tumours: Revised RECIST guideline (version
-1.1). European journal of cancer 45:228--247, 2009
+\CSLLeftMargin{14. }%
+\CSLRightInline{Eisenhauer, E. A. \emph{et al.} New response evaluation
+criteria in solid tumours: Revised RECIST guideline (version 1.1).
+\emph{European journal of cancer} \textbf{45}, 228--247 (2009).}
 
 \bibitem[\citeproctext]{ref-ThanCCR}
-\textbf{15}. Ton TGN, Pal N, Trinh H, et al: Replication of overall
-survival, progression-free survival, and overall response in
+\CSLLeftMargin{15. }%
+\CSLRightInline{Ton, T. G. N. \emph{et al.}
+\href{https://doi.org/10.1158/1078-0432.CCR-22-0471}{Replication of
+overall survival, progression-free survival, and overall response in
 chemotherapy arms of nonâsmall cell lung cancer trials using real-world
-data {[}Internet{]}. Clinical Cancer Research 28:2844--2853,
-2022Available from: \url{https://doi.org/10.1158/1078-0432.CCR-22-0471}
+data}. \emph{Clinical Cancer Research} \textbf{28}, 2844--2853 (2022).}
 
 \bibitem[\citeproctext]{ref-mckelvey2024evaluation}
-\textbf{16}. McKelvey BA, Garrett-Mayer E, Rivera DR, et al: Evaluation
-of real-world tumor response derived from electronic health record data
-sources: A feasibility analysis in patients with metastatic non--small
-cell lung cancer treated with chemotherapy. JCO Clinical Cancer
-Informatics 8:e2400091, 2024
+\CSLLeftMargin{16. }%
+\CSLRightInline{McKelvey, B. A. \emph{et al.} Evaluation of real-world
+tumor response derived from electronic health record data sources: A
+feasibility analysis in patients with metastatic non--small cell lung
+cancer treated with chemotherapy. \emph{JCO Clinical Cancer Informatics}
+\textbf{8}, e2400091 (2024).}
 
 \bibitem[\citeproctext]{ref-rwdRECIST}
-\textbf{17}. Chen L, Davis R, Lee J, et al: Comparison of response from
-RECIST1.1 and abstraction in real-world patients with lung cancer.
-{[}Internet{]}. Journal of Clinical Oncology 41:e21194--e21194,
-2023Available from:
-\url{https://ascopubs.org/doi/abs/10.1200/JCO.2023.41.16_suppl.e21194}
+\CSLLeftMargin{17. }%
+\CSLRightInline{Chen, L. \emph{et al.}
+\href{https://doi.org/10.1200/JCO.2023.41.16/_suppl.e21194}{Comparison
+of response from RECIST1.1 and abstraction in real-world patients with
+lung cancer.} \emph{Journal of Clinical Oncology} \textbf{41},
+e21194--e21194 (2023).}
 
 \bibitem[\citeproctext]{ref-rivera2022friends}
-\textbf{18}. Rivera DR, Henk HJ, Garrett-Mayer E, et al: The friends of
-cancer research real-world data collaboration pilot 2.0: Methodological
-recommendations from oncology case studies. Clinical Pharmacology \&
-Therapeutics 111:283--292, 2022
+\CSLLeftMargin{18. }%
+\CSLRightInline{Rivera, D. R. \emph{et al.} The friends of cancer
+research real-world data collaboration pilot 2.0: Methodological
+recommendations from oncology case studies. \emph{Clinical Pharmacology
+\& Therapeutics} \textbf{111}, 283--292 (2022).}
 
 \bibitem[\citeproctext]{ref-ackerman2024measurement}
-\textbf{19}. Ackerman B, Gan RW, Meyer CS, et al: Measurement error and
-bias in real-world oncology endpoints when constructing external control
-arms. Frontiers in Drug Safety and Regulation 4:1423493, 2024
+\CSLLeftMargin{19. }%
+\CSLRightInline{Ackerman, B. \emph{et al.} Measurement error and bias in
+real-world oncology endpoints when constructing external control arms.
+\emph{Frontiers in Drug Safety and Regulation} \textbf{4}, 1423493
+(2024).}
 
 \bibitem[\citeproctext]{ref-nikolakopoulou2014interpret}
-\textbf{20}. Nikolakopoulou A, Mavridis D, Salanti G: How to interpret
-meta-analysis models: Fixed effect and random effects meta-analyses. BMJ
-Ment Health 17:64--64, 2014
+\CSLLeftMargin{20. }%
+\CSLRightInline{Nikolakopoulou, A., Mavridis, D. \& Salanti, G. How to
+interpret meta-analysis models: Fixed effect and random effects
+meta-analyses. \emph{BMJ Ment Health} \textbf{17}, 64--64 (2014).}
 
 \bibitem[\citeproctext]{ref-wang2022harmonized}
-\textbf{21}. Wang SV, Pottegård A, Crown W, et al: HARmonized protocol
+\CSLLeftMargin{21. }%
+\CSLRightInline{Wang, S. V. \emph{et al.}
+\href{https://doi.org/10.1016/j.jval.2022.09.001}{HARmonized protocol
 template to enhance reproducibility of hypothesis evaluating real-world
 evidence studies on treatment effects: A good practices report of a
-joint ISPE/ISPOR task force {[}Internet{]}. Value in Health
-25:1663--1672, 2022Available from:
-\url{https://doi.org/10.1016/j.jval.2022.09.001}
+joint ISPE/ISPOR task force}. \emph{Value in Health} \textbf{25},
+1663--1672 (2022).}
 
 \bibitem[\citeproctext]{ref-guidelinegeneral}
-\textbf{22}. ICH guideline. General principles on plan, design and
+\CSLLeftMargin{22. }%
+\CSLRightInline{ICH guideline. General principles on plan, design and
 analysis of pharmacoepidemiological studies that utilize real-world data
@@
 Https://database.ich.org/sites/default/files/ICH\_M14\_Step3\_DraftGuideline\_2024\_0521.pdf
-(last accessed 01/07/2025)
+(last accessed 01/07/2025). }
 
 \bibitem[\citeproctext]{ref-hernan2022target}
-\textbf{23}. Hernán MA, Wang W, Leaf DE: Target trial emulation: A
-framework for causal inference from observational data. Jama
-328:2446--2447, 2022
+\CSLLeftMargin{23. }%
+\CSLRightInline{Hernán, M. A., Wang, W. \& Leaf, D. E. Target trial
+emulation: A framework for causal inference from observational data.
+\emph{Jama} \textbf{328}, 2446--2447 (2022).}
 
 \bibitem[\citeproctext]{ref-hernan2016specifying}
-\textbf{24}. Hernán MA, Sauer BC, Hernández-Dı́az S, et al:
+\CSLLeftMargin{24. }%
+\CSLRightInline{Hernán, M. A., Sauer, B. C., Hernández-Dı́az, S., Platt,
+R. \& Shrier, I.
 \href{https://doi.org/10.1016/j.jclinepi.2016.04.014}{Specifying a
 target trial prevents immortal time bias and other self-inflicted
-injuries in observational analyses}. Journal of clinical epidemiology
-79:70--75, 2016
+injuries in observational analyses}. \emph{Journal of clinical
+epidemiology} \textbf{79}, 70--75 (2016).}
 
 \bibitem[\citeproctext]{ref-rufibach2018}
-\textbf{25}. Rufibach K: Treatment effect quantification for
-time{-}to{-}event endpoints{\textendash}Estimands, analysis strategies,
-and beyond {[}Internet{]}. Pharmaceutical Statistics 18:145--165,
-2018Available from: \url{http://dx.doi.org/10.1002/pst.1917}
+\CSLLeftMargin{25. }%
+\CSLRightInline{Rufibach, K.
+\href{https://doi.org/10.1002/pst.1917}{Treatment effect quantification
+for time{-}to{-}event endpoints{\textendash}Estimands, analysis
+strategies, and beyond}. \emph{Pharmaceutical Statistics} \textbf{18},
+145--165 (2018).}
 
 \bibitem[\citeproctext]{ref-manitz2022estimands}
-\textbf{26}. Manitz J, Kan-Dobrosky N, Buchner H, et al: Estimands for
-overall survival in clinical trials with treatment switching in
-oncology. Pharmaceutical Statistics 21:150--162, 2022
+\CSLLeftMargin{26. }%
+\CSLRightInline{Manitz, J. \emph{et al.} Estimands for overall survival
+in clinical trials with treatment switching in oncology.
+\emph{Pharmaceutical Statistics} \textbf{21}, 150--162 (2022).}
 
 \bibitem[\citeproctext]{ref-meyer2020open}
-\textbf{27}. Meyer A-M, Davies J, Taylor M, et al: Open cohorts and
-ghost-time bias in real world data, in PHARMACOEPIDEMIOLOGY AND DRUG
-SAFETY. WILEY 111 RIVER ST, HOBOKEN 07030-5774, NJ USA, 2020, pp
-426--426
+\CSLLeftMargin{27. }%
+\CSLRightInline{Meyer, A.-M., Davies, J., Taylor, M. \& Fruechtenicht,
+C. Open cohorts and ghost-time bias in real world data. In
+\emph{PHARMACOEPIDEMIOLOGY AND DRUG SAFETY} \textbf{29}, 426--426 (WILEY
+111 RIVER ST, HOBOKEN 07030-5774, NJ USA, 2020).}
 
 \bibitem[\citeproctext]{ref-hernan2022causal}
-\textbf{28}. Hernán MA: Causal analyses of existing databases: No power
-calculations required. Journal of clinical epidemiology 144:203--205,
-2022
+\CSLLeftMargin{28. }%
+\CSLRightInline{Hernán, M. A. Causal analyses of existing databases: No
+power calculations required. \emph{Journal of clinical epidemiology}
+\textbf{144}, 203--205 (2022).}
 
 \bibitem[\citeproctext]{ref-schoenfeld1981asymptotic}
-\textbf{29}. Schoenfeld D: The asymptotic properties of nonparametric
-tests for comparing survival distributions. Biometrika 68:316--319, 1981
+\CSLLeftMargin{29. }%
+\CSLRightInline{Schoenfeld, D. The asymptotic properties of
+nonparametric tests for comparing survival distributions.
+\emph{Biometrika} \textbf{68}, 316--319 (1981).}
 
 \bibitem[\citeproctext]{ref-franklin2020nonrandomized}
-\textbf{30}. Franklin JM, Pawar A, Martin D, et al: Nonrandomized
-real-world evidence to support regulatory decision making: Process for a
-randomized trial replication project. Clinical Pharmacology \&
-Therapeutics 107:817--826, 2020
+\CSLLeftMargin{30. }%
+\CSLRightInline{Franklin, J. M. \emph{et al.} Nonrandomized real-world
+evidence to support regulatory decision making: Process for a randomized
+trial replication project. \emph{Clinical Pharmacology \& Therapeutics}
+\textbf{107}, 817--826 (2020).}
 
 \bibitem[\citeproctext]{ref-schneeweiss2019graphical}
-\textbf{31}. Schneeweiss S, Rassen JA, Brown JS, et al: Graphical
-depiction of longitudinal study designs in health care databases. Annals
-of internal medicine 170:398--406, 2019
+\CSLLeftMargin{31. }%
+\CSLRightInline{Schneeweiss, S. \emph{et al.} Graphical depiction of
+longitudinal study designs in health care databases. \emph{Annals of
+internal medicine} \textbf{170}, 398--406 (2019).}
 
 \bibitem[\citeproctext]{ref-rubin1976inference}
-\textbf{32}. Rubin DB: Inference and missing data. Biometrika
-63:581--592, 1976
+\CSLLeftMargin{32. }%
+\CSLRightInline{Rubin, D. B. Inference and missing data.
+\emph{Biometrika} \textbf{63}, 581--592 (1976).}
 
 \bibitem[\citeproctext]{ref-weberpals2024smdi}
-\textbf{33}. Weberpals J, Raman SR, Shaw PA, et al: Smdi: An r package
-to perform structural missing data investigations on partially observed
-confounders in real-world evidence studies. JAMIA open 7:ooae008, 2024
+\CSLLeftMargin{33. }%
+\CSLRightInline{Weberpals, J. \emph{et al.} Smdi: An r package to
+perform structural missing data investigations on partially observed
+confounders in real-world evidence studies. \emph{JAMIA open}
+\textbf{7}, ooae008 (2024).}
 
 \bibitem[\citeproctext]{ref-weberpals2024}
-\textbf{34}. Weberpals J, Raman SR, Shaw PA, et al: A principled
-approach to characterize and analyze partially observed confounder data
-from electronic health records {[}Internet{]}. Clinical Epidemiology
-16:329--343, 2024Available
-from:\href{\%20\%0A\%20\%20\%20\%20\%0A\%20\%20\%20\%20\%0A\%20\%20\%20\%20\%20\%20\%20\%20https://www.tandfonline.com/doi/abs/10.2147/CLEP.S436131\%0A\%20\%20\%20\%20\%0A\%0A}{https://www.tandfonline.com/doi/abs/10.2147/CLEP.S436131
-}
+\CSLLeftMargin{34. }%
+\CSLRightInline{Weberpals, J. \emph{et al.}
+\href{https://doi.org/10.2147/CLEP.S436131}{A principled approach to
+characterize and analyze partially observed confounder data from
+electronic health records}. \emph{Clinical Epidemiology} \textbf{16},
+329--343 (2024).}
 
 \bibitem[\citeproctext]{ref-moreno2018canonical}
-\textbf{35}. Moreno-Betancur M, Lee KJ, Leacy FP, et al: Canonical
-causal diagrams to guide the treatment of missing data in epidemiologic
-studies. American journal of epidemiology 187:2705--2715, 2018
+\CSLLeftMargin{35. }%
+\CSLRightInline{Moreno-Betancur, M. \emph{et al.} Canonical causal
+diagrams to guide the treatment of missing data in epidemiologic
+studies. \emph{American journal of epidemiology} \textbf{187},
+2705--2715 (2018).}
 
 \bibitem[\citeproctext]{ref-tompsett2018use}
-\textbf{36}. Tompsett DM, Leacy F, Moreno-Betancur M, et al: On the use
-of the not-at-random fully conditional specification (NARFCS) procedure
-in practice. Statistics in medicine 37:2338--2353, 2018
+\CSLLeftMargin{36. }%
+\CSLRightInline{Tompsett, D. M., Leacy, F., Moreno-Betancur, M., Heron,
+J. \& White, I. R. On the use of the not-at-random fully conditional
+specification (NARFCS) procedure in practice. \emph{Statistics in
+medicine} \textbf{37}, 2338--2353 (2018).}
 
 \bibitem[\citeproctext]{ref-weberpals2024hdmi}
-\textbf{37}. Weberpals J, Shaw PA, Lin KJ, et al: High-dimensional
-multiple imputation (HDMI) for partially observed confounders including
-natural language processing-derived auxiliary covariates {[}Internet{]}.
-American Journal of Epidemiology, 2025Available from:
-\url{https://arxiv.org/abs/2405.10925}
+\CSLLeftMargin{37. }%
+\CSLRightInline{Weberpals, J. \emph{et al.} High-dimensional multiple
+imputation (HDMI) for partially observed confounders including natural
+language processing-derived auxiliary covariates. \emph{American Journal
+of Epidemiology} (2025).at
+\textless{}\url{https://arxiv.org/abs/2405.10925}\textgreater{}}
 
 \bibitem[\citeproctext]{ref-Shaw2024}
-\textbf{38}. Shaw CK P., Williamson BD: Assessing treatment effects in
-observational data with missing confounders: A comparative study of
-practical doubly-robust and traditional missing data methods
-{[}Internet{]}. GitHub repository, 2024Available from:
-\url{https://github.com/PamelaShaw/Missing-Confounders-Methods}
+\CSLLeftMargin{38. }%
+\CSLRightInline{Shaw, C. K., P. \& Williamson, B. D. Assessing treatment
+effects in observational data with missing confounders: A comparative
+study of practical doubly-robust and traditional missing data methods.
+\emph{GitHub repository} (2024).at
+\textless{}\url{https://github.com/PamelaShaw/Missing-Confounders-Methods}\textgreater{}}
 
 \bibitem[\citeproctext]{ref-cox1972regression}
-\textbf{39}. Cox DR: Regression models and life-tables. Journal of the
-royal statistical society. Series B (Methodological) 34:187--220, 1972
+\CSLLeftMargin{39. }%
+\CSLRightInline{Cox, D. R. Regression models and life-tables. Journal of
+the royal statistical society. \emph{Series B (Methodological)}
+\textbf{34}, 187--220 (1972).}
 
 \bibitem[\citeproctext]{ref-kaplan1958nonparametric}
-\textbf{40}. Kaplan EL, Meier P: Nonparametric estimation from
-incomplete observations. Journal of the American statistical association
-53:457--481, 1958
+\CSLLeftMargin{40. }%
+\CSLRightInline{Kaplan, E. L. \& Meier, P. Nonparametric estimation from
+incomplete observations. \emph{Journal of the American statistical
+association} \textbf{53}, 457--481 (1958).}
 
 \bibitem[\citeproctext]{ref-becker2020enhanced}
-\textbf{41}. Becker T, Weberpals J, Jegg A, et al: An enhanced
-prognostic score for overall survival of patients with cancer derived
-from a large real-world cohort. Annals of Oncology 31:1561--1568, 2020
+\CSLLeftMargin{41. }%
+\CSLRightInline{Becker, T. \emph{et al.} An enhanced prognostic score
+for overall survival of patients with cancer derived from a large
+real-world cohort. \emph{Annals of Oncology} \textbf{31}, 1561--1568
+(2020).}
 
 \bibitem[\citeproctext]{ref-leyrat2019propensity}
-\textbf{42}. Leyrat C, Seaman SR, White IR, et al: Propensity score
-analysis with partially observed covariates: How should multiple
-imputation be used? Statistical methods in medical research 28:3--19,
-2019
+\CSLLeftMargin{42. }%
+\CSLRightInline{Leyrat, C. \emph{et al.} Propensity score analysis with
+partially observed covariates: How should multiple imputation be used?
+\emph{Statistical methods in medical research} \textbf{28}, 3--19
+(2019).}
 
 \bibitem[\citeproctext]{ref-pishgar2020matchthem}
-\textbf{43}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
-and weighting after multiple imputation. arXiv preprint arXiv:200911772,
-2020
+\CSLLeftMargin{43. }%
+\CSLRightInline{Pishgar, F., Greifer, N., Leyrat, C. \& Stuart, E.
+MatchThem:: Matching and weighting after multiple imputation.
+\emph{arXiv preprint arXiv:2009.11772} (2020).}
 
 \bibitem[\citeproctext]{ref-rubin2018multiple}
-\textbf{44}. Rubin DB: Multiple imputation, in Flexible imputation of
-missing data, second edition. Chapman; Hall/CRC, 2018, pp 29--62
+\CSLLeftMargin{44. }%
+\CSLRightInline{Rubin, D. B. Multiple imputation. In \emph{Flexible
+imputation of missing data, second edition} 29--62 (Chapman; Hall/CRC,
+2018).}
 
 \bibitem[\citeproctext]{ref-van2011mice}
-\textbf{45}. Van Buuren S, Groothuis-Oudshoorn K: Mice: Multivariate
-imputation by chained equations in r. Journal of statistical software
-45:1--67, 2011
+\CSLLeftMargin{45. }%
+\CSLRightInline{Van Buuren, S. \& Groothuis-Oudshoorn, K. Mice:
+Multivariate imputation by chained equations in r. \emph{Journal of
+statistical software} \textbf{45}, 1--67 (2011).}
 
 \bibitem[\citeproctext]{ref-austin2009balance}
-\textbf{46}. Austin PC: Balance diagnostics for comparing the
+\CSLLeftMargin{46. }%
+\CSLRightInline{Austin, P. C. Balance diagnostics for comparing the
 distribution of baseline covariates between treatment groups in
-propensity-score matched samples. Statistics in medicine 28:3083--3107,
-2009
+propensity-score matched samples. \emph{Statistics in medicine}
+\textbf{28}, 3083--3107 (2009).}
 
 \bibitem[\citeproctext]{ref-franklin2014metrics}
-\textbf{47}. Franklin JM, Rassen JA, Ackermann D, et al: Metrics for
-covariate balance in cohort studies of causal effects. Statistics in
-medicine 33:1685--1699, 2014
+\CSLLeftMargin{47. }%
+\CSLRightInline{Franklin, J. M., Rassen, J. A., Ackermann, D., Bartels,
+D. B. \& Schneeweiss, S. Metrics for covariate balance in cohort studies
+of causal effects. \emph{Statistics in medicine} \textbf{33}, 1685--1699
+(2014).}
 
 \bibitem[\citeproctext]{ref-stuart2013prognostic}
-\textbf{48}. Stuart EA, Lee BK, Leacy FP: Prognostic score--based
-balance measures can be a useful diagnostic for propensity score methods
-in comparative effectiveness research. Journal of clinical epidemiology
-66:S84--S90, 2013
+\CSLLeftMargin{48. }%
+\CSLRightInline{Stuart, E. A., Lee, B. K. \& Leacy, F. P. Prognostic
+score--based balance measures can be a useful diagnostic for propensity
+score methods in comparative effectiveness research. \emph{Journal of
+clinical epidemiology} \textbf{66}, S84--S90 (2013).}
 
 \bibitem[\citeproctext]{ref-warwick28685}
-\textbf{49}. Marshall A(Andrea), Billingham LJ, Bryan S: Can we afford
-to ignore missing data in cost-effectiveness analyses? {[}Internet{]}.
-European Journal of Health Economics Vol.10:1--3, 2009Available from:
-\url{http://dx.doi.org/10.1007/s10198-008-0129-y}
+\CSLLeftMargin{49. }%
+\CSLRightInline{Marshall, A. (Andrea)., Billingham, L. J. \& Bryan, S.
+\href{https://doi.org/10.1007/s10198-008-0129-y\%0A}{Can we afford to
+ignore missing data in cost-effectiveness analyses?} \emph{European
+Journal of Health Economics} \textbf{Vol.10}, 1--3 (2009).}
 
 \bibitem[\citeproctext]{ref-morisot2015prostate}
-\textbf{50}. Morisot A, Bessaoud F, Landais P, et al: Prostate cancer:
-Net survival and cause-specific survival rates after multiple
-imputation. BMC medical research methodology 15:1--14, 2015
+\CSLLeftMargin{50. }%
+\CSLRightInline{Morisot, A. \emph{et al.} Prostate cancer: Net survival
+and cause-specific survival rates after multiple imputation. \emph{BMC
+medical research methodology} \textbf{15}, 1--14 (2015).}
 
 \bibitem[\citeproctext]{ref-anesrake}
-\textbf{51}. Pasek J: Anesrake: ANES raking implementation, 2018
+\CSLLeftMargin{51. }%
+\CSLRightInline{Pasek, J. Anesrake: ANES raking implementation. (2018).}
 
 \bibitem[\citeproctext]{ref-MatchThem}
-\textbf{52}. Pishgar F, Greifer N, Leyrat C, et al: MatchThem:: Matching
-and weighting after multiple imputation {[}Internet{]}, 2021Available
-from: \url{https://journal.r-project.org/archive/2021/RJ-2021-073/}
+\CSLLeftMargin{52. }%
+\CSLRightInline{Pishgar, F., Greifer, N., Leyrat, C. \& Stuart, E.
+MatchThem:: Matching and weighting after multiple imputation.
+(2021).doi:\href{https://doi.org/10.32614/RJ-2021-073}{10.32614/RJ-2021-073}}
 
 \bibitem[\citeproctext]{ref-survival}
-\textbf{53}. Therneau TM: A package for survival analysis in r
-{[}Internet{]}, 2024Available from:
-\url{https://CRAN.R-project.org/package=survival}
+\CSLLeftMargin{53. }%
+\CSLRightInline{Therneau, T. M. A package for survival analysis in r.
+(2024).at
+\textless{}\url{https://CRAN.R-project.org/package=survival}\textgreater{}}
 
 \bibitem[\citeproctext]{ref-encore.analytics}
-\textbf{54}. Weberpals J: Encore.analytics: Functions and wrappers to
-streamline complex analytic workflows in real-world data studies based
-on the ENCORE trial emulation project {[}Internet{]}, 2025Available
-from: \url{https://github.com/janickweberpals/encore.analytics}
+\CSLLeftMargin{54. }%
+\CSLRightInline{Weberpals, J. Encore.analytics: Functions and wrappers
+to streamline complex analytic workflows in real-world data studies
+based on the ENCORE trial emulation project. (2025).at
+\textless{}\url{https://github.com/janickweberpals/encore.analytics}\textgreater{}}
 
 \bibitem[\citeproctext]{ref-gtsummary}
-\textbf{55}. Sjoberg DD, Whiting K, Curry M, et al: Reproducible summary
-tables with the gtsummary package {[}Internet{]}13:570--580,
-2021Available from: \url{https://doi.org/10.32614/RJ-2021-053}
+\CSLLeftMargin{55. }%
+\CSLRightInline{Sjoberg, D. D., Whiting, K., Curry, M., Lavery, J. A. \&
+Larmarange, J. \href{https://doi.org/10.32614/RJ-2021-053}{Reproducible
+summary tables with the gtsummary package}. \textbf{13}, 570--580
+(2021).}
 
 \bibitem[\citeproctext]{ref-ggsurvfit}
-\textbf{56}. Sjoberg DD, Baillie M, Fruechtenicht C, et al: Ggsurvfit:
-Flexible time-to-event figures {[}Internet{]}, 2024Available from:
-\url{https://github.com/pharmaverse/ggsurvfit}
+\CSLLeftMargin{56. }%
+\CSLRightInline{Sjoberg, D. D., Baillie, M., Fruechtenicht, C.,
+Haesendonckx, S. \& Treis, T. Ggsurvfit: Flexible time-to-event figures.
+(2024).at
+\textless{}\url{https://github.com/pharmaverse/ggsurvfit}\textgreater{}}
 
 \bibitem[\citeproctext]{ref-gaber2024mystifying}
-\textbf{57}. Gaber CE, Ghazarian AA, Strassle PD, et al: De-mystifying
-the clone-censor-weight method for causal research using observational
-data: A primer for cancer researchers. Cancer Medicine 13:e70461, 2024
+\CSLLeftMargin{57. }%
+\CSLRightInline{Gaber, C. E. \emph{et al.} De-mystifying the
+clone-censor-weight method for causal research using observational data:
+A primer for cancer researchers. \emph{Cancer Medicine} \textbf{13},
+e70461 (2024).}
 

### 2025-07-08 manuscript/manuscript.qmd
changed referencing style to CPT style
replaced observational with non-interventional

@@
 bibliography: references.bib
-csl: journal-of-clinical-oncology.csl
+csl: clinical-pharmacology-and-therapeutics.csl
 
@@
 
-However, the comparability and transportability of results derived between observational studies and RCTs depends on multiple factors.
-Frequently referenced limitations include lack of baseline randomization and the imbalance in prognostic factors resulting thereof, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in treatment guidelines over time in routine clinical care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to increase confidence in the interpretability of observational studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.
+However, the comparability and transportability of results derived between non-interventional studies and RCTs depends on multiple factors.
+Frequently referenced limitations include lack of baseline randomization and the imbalance in prognostic factors resulting thereof, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in treatment guidelines over time in routine clinical care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to increase confidence in the interpretability of non-interventional studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.
 
-The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when observational studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
-In settings where the RCT designs could be emulated well, observational studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.
+The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when non-interventional studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
+In settings where the RCT designs could be emulated well, non-interventional studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.
 
@@
 
-The focus of ENCORE is to efficiently explore in what contexts observational studies can or cannot yield similar results compared to oncology RCTs.
+The focus of ENCORE is to efficiently explore in what contexts non-interventional studies can or cannot yield similar results compared to oncology RCTs.
 Therefore, the project emphasizes trials of therapies for common cancers for which there has been substantial therapeutic development in recent years.
@@
 
-Causal analyses of observational data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data .[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreement metrics between RCT and RWD results.
+Causal analyses of non-interventional data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data .[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreement metrics between RCT and RWD results.
 Since the main outcome of interest is defined as time from index treatment initiation to all-cause mortality (rwOS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
@@
 Principled, well-designed and reproducible studies using fit-for-purpose RWD to generate RWE, such as through trial emulation, may complement evidence from RCTs.
-Through a systematic benchmarking approach, the ENCORE project will provide insights as to how methodological, design, and analytic decisions influence quantifiable bias and interpretability of observational studies in oncology.
+Through a systematic benchmarking approach, the ENCORE project will provide insights as to how methodological, design, and analytic decisions influence quantifiable bias and interpretability of non-interventional studies in oncology.
 

### 2025-07-08 manuscript/manuscript.tex
Manuscript updated

@@
 
-\emph{Manuscript last updated: 2025-07-06 16:21:00.006613}
+\emph{Manuscript last updated: 2025-07-08 18:15:17.700176}
 
@@
 However, the comparability and transportability of results derived
-between observational studies and RCTs depends on multiple factors.
+between non-interventional studies and RCTs depends on multiple factors.
 Frequently referenced limitations include lack of baseline randomization
@@
 heterogeneous databases is necessary to increase confidence in the
-interpretability of observational studies, evaluate regulatory
+interpretability of non-interventional studies, evaluate regulatory
 considerations, and to provide context as to which questions can be
@@
 The RCT DUPLICATE initiative\textsuperscript{8} increased our
-understanding of when observational studies can come to causal
+understanding of when non-interventional studies can come to causal
 conclusions on treatment effects by benchmarking results against RCTs
@@
 effect. In settings where the RCT designs could be emulated well,
-observational studies were able to reach similar
+non-interventional studies were able to reach similar
 conclusions.\textsuperscript{9} However, this prior work from
@@
 The focus of ENCORE is to efficiently explore in what contexts
-observational studies can or cannot yield similar results compared to
-oncology RCTs. Therefore, the project emphasizes trials of therapies for
-common cancers for which there has been substantial therapeutic
+non-interventional studies can or cannot yield similar results compared
+to oncology RCTs. Therefore, the project emphasizes trials of therapies
+for common cancers for which there has been substantial therapeutic
 development in recent years. After review and collaboration with
@@
 
-Causal analyses of observational data are often not designed with
+Causal analyses of non-interventional data are often not designed with
 respect to formal hypothesis testing and statistical power in the same
@@
 and analytic decisions influence quantifiable bias and interpretability
-of observational studies in oncology.
+of non-interventional studies in oncology.
 

### 2025-07-26 manuscript/manuscript.qmd
included Bob's edits and comments

@@
 Real-world evidence (RWE) is increasingly used to complement findings from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in routine clinical practice.
-Advancements in the curation and accessibility of electronic health record data (EHR) present the opportunity to utilize real-world data (RWD) to investigate therapeutic areas including oncology, where administrative healthcare claims databases alone are often not fit-for-purpose.
-The RCT DUPLICATE initiative has previously enhanced understanding of when RWE can most appropriately draw causal conclusions by emulating trials for non-oncology indications.
+Advances in the curation and accessibility of electronic health record data (EHR) present the opportunity to utilize real-world data (RWD) to investigate therapeutic areas including oncology, where administrative healthcare claims databases alone are often not fit-for-purpose.
+The RCT DUPLICATE initiative has previously evaluated when RWE can most appropriately draw causal conclusions by emulating trials for non-oncology indications.
 Here, we present the design and trial selection for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which extends this work to oncology.
 ENCORE is designed to emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma.
-It will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to initiation and analysis.
-Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria.
-Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence the interpretation of results from emulated oncology trials using RWD.
+It will place special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to initiation and analysis.
+Pre-specified criteria will assess agreement of treatment effect estimates between RCTs and their respective emulations.
+Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project aims to inform understanding of how measurement, design, and analytic decisions influence the interpretation of results from emulated oncology trials using RWD.
 
@@
 However, the comparability and transportability of results derived between non-interventional studies and RCTs depends on multiple factors.
-Frequently referenced limitations include lack of baseline randomization and the imbalance in prognostic factors resulting thereof, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in treatment guidelines over time in routine clinical care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to increase confidence in the interpretability of non-interventional studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.
+Frequently referenced limitations include lack of baseline randomization and the imbalance in prognostic factors resulting thereof, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in treatment guidelines over time in routine clinical care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized endpoint assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases would enhance increase confidence in the interpretability of non-interventional studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.
 
-The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when non-interventional studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect.
-In settings where the RCT designs could be emulated well, non-interventional studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.
+The RCT DUPLICATE initiative[@wang2023emulation] evaluated when non-interventional studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that well-designed and conducted RCT findings reflect causal treatment effects.
+In settings where the RCT designs could be emulated well, non-interventional studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused on emulating trials in non-oncology settings using claims databases.
 
@@
 
-The focus of ENCORE is to efficiently explore in what contexts non-interventional studies can or cannot yield similar results compared to oncology RCTs.
+The focus of ENCORE is to clarify when non-interventional studies can or cannot yield similar results compared to oncology RCTs.
 Therefore, the project emphasizes trials of therapies for common cancers for which there has been substantial therapeutic development in recent years.
-After review and collaboration with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma.
+After review and collaboration with clinical and regulatory experts, four cancer types were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma.
 For each cancer type, we aim to conduct three trial emulations using multiple accessible databases (i.e., the total equaling 12 trial emulations x *n* databases which are found fit-for-purpose for each trial).
@@
 
--   Feasibility and clinical relevance (latter was defined as treatment or paradigm-changing trials or such that challenged existing treatment policies)
+-   Feasibility and clinical relevance (latter was defined as treatment or paradigm-changing trials or trials that challenged existing treatment policies)
 
-The rationale and operationalization for each criterion is listed in detail in @tbl-criteria.
+The operationalization for each criterion is listed in detail in @tbl-criteria.
 We mainly considered pivotal large late-phase RCTs that were initiated after 2011 because treatment guidelines for included cancer indications have undergone significant changes in recent years.
@@
 A key objective that we aim to explore with the selected trials is to evaluate how stage (early, late), line of therapy (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD.
-If more thorough feasibility assessments after applying all relevant eligibility criteria suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.
+If more thorough feasibility assessments after applying all relevant eligibility criteria (including formal power calculations and assessment of sufficient balance after propensity score matching or weighting) suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.
 
@@
 For ENCORE, not all databases will be available for each cancer indication given the specificity of data elements required and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (in randomly assigned order).
-If more than one database is considered fit-for-use for a respective trial emulation, the most suitable analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]
+If more than one database is considered fit-for-use for a respective trial emulation, the most suitable analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using random effects (primary; weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) and fixed effects (secondary; weights reflect the inverse of the variance of each database study) models.[@nikolakopoulou2014interpret]
 
@@
 The distribution of patient characteristics, by exposure status, will be examined before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT.
-Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, exposure cohorts are conditionally exchangeable at baseline, and resulting sample sizes are still sufficient after matching or weighting.
+Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced[@austin2009balance], exposure cohorts are conditionally exchangeable at baseline, and resulting sample sizes are still sufficient after matching or weighting.
 At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design or analytic choices based on known outcome information.
@@
 Since the main outcome of interest is defined as time from index treatment initiation to all-cause mortality (rwOS), the estimation of the statistical power is driven by the number of events rather than the number of patients.
-To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]
+To assess whether the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$ with a 2-sided $\alpha$ of 0.05) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]
 
@@
 
-where $\hat{\theta}_{RCT}$ and $\hat{\theta}_{RWE}$ are the treatment effect estimates (hazard ratios or median survival time differences) and $\text{Var}(\hat{\theta}_{RCT})$ and $\text{Var}(\hat{\theta}_{RWE})$ are the corresponding variances for RCT and RWE, respectively.
+where $\hat{\theta}_{RCT}$ and $\hat{\theta}_{RWE}$ are the treatment effect estimates (log hazard ratios or median survival time differences) and $\text{Var}(\hat{\theta}_{RCT})$ and $\text{Var}(\hat{\theta}_{RWE})$ are the corresponding variances for RCT and RWE, respectively.
 The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart.
@@
 
-While the MAR assumption is a strong assumption to hold across all considered covariates, it has been shown that especially in the context of partially observed covariate data (as opposed to missing exposure and outcome data), generally only mechanisms in which a covariate causes its own missingness lead to significant bias (MNAR).[@moreno2018canonical] Hence, methodologies which retain otherwise eligible patients and give the potential to adjust for a broader set of prognostic factors (e.g., multiple imputation[@weberpals2024hdmi] or doubly robust methods[@Shaw2024]) may be preferred over complete case analyses, although such tradeoffs must be carefully evaluated within the clinical and study contexts.
+While MAR is a strong assumption to hold across all considered covariates, it has been shown that especially in the context of partially observed covariate data (as opposed to missing exposure and outcome data), generally only mechanisms in which a covariate causes its own missingness lead to significant bias (MNAR).[@moreno2018canonical] Hence, methodologies which retain otherwise eligible patients and give the potential to adjust for a broader set of prognostic factors (e.g., multiple imputation[@weberpals2024hdmi] or doubly robust methods[@Shaw2024]) may be preferred over complete case analyses, although such tradeoffs must be carefully evaluated within the clinical and study contexts.
 

### 2025-07-26 manuscript/manuscript.tex
Manuscript updated

@@
 
-\emph{Manuscript last updated: 2025-07-08 18:15:17.700176}
+\emph{Manuscript last updated: 2025-07-26 11:39:19.945045}
 
@@
 effectiveness and safety of medical interventions as delivered in
-routine clinical practice. Advancements in the curation and
-accessibility of electronic health record data (EHR) present the
-opportunity to utilize real-world data (RWD) to investigate therapeutic
-areas including oncology, where administrative healthcare claims
-databases alone are often not fit-for-purpose. The RCT DUPLICATE
-initiative has previously enhanced understanding of when RWE can most
-appropriately draw causal conclusions by emulating trials for
-non-oncology indications. Here, we present the design and trial
-selection for the Emulation of Comparative Oncology Trials with
-Real-world Evidence (ENCORE) project, which extends this work to
-oncology. ENCORE is designed to emulate 12 RCTs in four
+routine clinical practice. Advances in the curation and accessibility of
+electronic health record data (EHR) present the opportunity to utilize
+real-world data (RWD) to investigate therapeutic areas including
+oncology, where administrative healthcare claims databases alone are
+often not fit-for-purpose. The RCT DUPLICATE initiative has previously
+evaluated when RWE can most appropriately draw causal conclusions by
+emulating trials for non-oncology indications. Here, we present the
+design and trial selection for the Emulation of Comparative Oncology
+Trials with Real-world Evidence (ENCORE) project, which extends this
+work to oncology. ENCORE is designed to emulate 12 RCTs in four
 oncology-specialized EHR databases across four different cancer
 indications, specifically non-small cell lung cancer, breast cancer,
-colorectal cancer, and multiple myeloma. It will place a special
-emphasis on systematic evaluation of fitness of data in relation to the
-study design and statistical analysis for a particular research
-question, and pre-registration of study protocols prior to initiation
-and analysis. Agreement of treatment effect estimates between RCTs and
-their respective emulations will be assessed using pre-specified
-criteria. Through extensive sensitivity analyses benchmarked against RCT
-results, the ENCORE project will inform understanding of how
-measurement, design, and analytic decisions influence the interpretation
-of results from emulated oncology trials using RWD.
+colorectal cancer, and multiple myeloma. It will place special emphasis
+on systematic evaluation of fitness of data in relation to the study
+design and statistical analysis for a particular research question, and
+pre-registration of study protocols prior to initiation and analysis.
+Pre-specified criteria will assess agreement of treatment effect
+estimates between RCTs and their respective emulations. Through
+extensive sensitivity analyses benchmarked against RCT results, the
+ENCORE project aims to inform understanding of how measurement, design,
+and analytic decisions influence the interpretation of results from
+emulated oncology trials using RWD.
 
@@
 measure and emulate common eligibility criteria, including prognostic
-factors, and standardized response assessments in real-world data
+factors, and standardized endpoint assessments in real-world data
 (RWD).\textsuperscript{5} While examples of oncology trial emulations
@@
 approach to emulate a diverse set of oncology trials with multiple
-heterogeneous databases is necessary to increase confidence in the
+heterogeneous databases would enhance increase confidence in the
 interpretability of non-interventional studies, evaluate regulatory
@@
 
-The RCT DUPLICATE initiative\textsuperscript{8} increased our
-understanding of when non-interventional studies can come to causal
-conclusions on treatment effects by benchmarking results against RCTs
-under the assumption that each RCT finding reflects a causal treatment
-effect. In settings where the RCT designs could be emulated well,
+The RCT DUPLICATE initiative\textsuperscript{8} evaluated when
+non-interventional studies can come to causal conclusions on treatment
+effects by benchmarking results against RCTs under the assumption that
+well-designed and conducted RCT findings reflect causal treatment
+effects. In settings where the RCT designs could be emulated well,
 non-interventional studies were able to reach similar
 conclusions.\textsuperscript{9} However, this prior work from
-RCT-DUPLICATE focused primarily on emulating trials in non-oncology
-settings using claims databases.
+RCT-DUPLICATE focused on emulating trials in non-oncology settings using
+claims databases.
 
@@
 
-The focus of ENCORE is to efficiently explore in what contexts
-non-interventional studies can or cannot yield similar results compared
-to oncology RCTs. Therefore, the project emphasizes trials of therapies
-for common cancers for which there has been substantial therapeutic
-development in recent years. After review and collaboration with
-clinical and regulatory experts, four cancer indications were
-identified: non-small cell lung cancer, breast cancer, colorectal cancer
-and multiple myeloma. For each cancer type, we aim to conduct three
-trial emulations using multiple accessible databases (i.e., the total
-equaling 12 trial emulations x \emph{n} databases which are found
-fit-for-purpose for each trial).
+The focus of ENCORE is to clarify when non-interventional studies can or
+cannot yield similar results compared to oncology RCTs. Therefore, the
+project emphasizes trials of therapies for common cancers for which
+there has been substantial therapeutic development in recent years.
+After review and collaboration with clinical and regulatory experts,
+four cancer types were identified: non-small cell lung cancer, breast
+cancer, colorectal cancer and multiple myeloma. For each cancer type, we
+aim to conduct three trial emulations using multiple accessible
+databases (i.e., the total equaling 12 trial emulations x \emph{n}
+databases which are found fit-for-purpose for each trial).
 
@@
   Feasibility and clinical relevance (latter was defined as treatment or
-  paradigm-changing trials or such that challenged existing treatment
+  paradigm-changing trials or trials that challenged existing treatment
   policies)
@@
 
-The rationale and operationalization for each criterion is listed in
-detail in Table~\ref{tbl-criteria}. We mainly considered pivotal large
-late-phase RCTs that were initiated after 2011 because treatment
-guidelines for included cancer indications have undergone significant
-changes in recent years. Due to the rapid adoption of newly-approved
-therapies in routine care, patients are less likely to receive outdated
-treatment regimens in clinical practice. Conversely, we excluded trials
-with results published too recently to allow for substantive medical
-product uptake and follow-up time accrual in real-world databases used
-for this project. We did not define a global cut-off as the requirements
-for follow-up time are different for each cancer type and population
-(e.g., advanced NSCLC versus early-stage breast cancer) and the
-decisions were made on a clinically relevant basis.
+The operationalization for each criterion is listed in detail in
+Table~\ref{tbl-criteria}. We mainly considered pivotal large late-phase
+RCTs that were initiated after 2011 because treatment guidelines for
+included cancer indications have undergone significant changes in recent
+years. Due to the rapid adoption of newly-approved therapies in routine
+care, patients are less likely to receive outdated treatment regimens in
+clinical practice. Conversely, we excluded trials with results published
+too recently to allow for substantive medical product uptake and
+follow-up time accrual in real-world databases used for this project. We
+did not define a global cut-off as the requirements for follow-up time
+are different for each cancer type and population (e.g., advanced NSCLC
+versus early-stage breast cancer) and the decisions were made on a
+clinically relevant basis.
 
@@
 feasibility assessments after applying all relevant eligibility criteria
-suggest that the threat of bias from mis-measurement of key study
-parameters or residual confounding remains high, or that the study size
-is not sufficient, then runner-up candidate trials will be considered
-instead.
+(including formal power calculations and assessment of sufficient
+balance after propensity score matching or weighting) suggest that the
+threat of bias from mis-measurement of key study parameters or residual
+confounding remains high, or that the study size is not sufficient, then
+runner-up candidate trials will be considered instead.
 
@@
 effect estimates will be pooled using a meta-analytic approach using
-fixed effects (weights reflect the inverse of the variance of each
-database study) and random effects (weights reflect the inverse of the
-variance of each database study plus an additional variance term that
-quantifies the assumed heterogeneity between databases)
+random effects (primary; weights reflect the inverse of the variance of
+each database study plus an additional variance term that quantifies the
+assumed heterogeneity between databases) and fixed effects (secondary;
+weights reflect the inverse of the variance of each database study)
 models.\textsuperscript{20}
@@
 Initial propensity score matching or weighting methods will be applied
-to ensure that measured pre-exposure covariates can be balanced,
-exposure cohorts are conditionally exchangeable at baseline, and
-resulting sample sizes are still sufficient after matching or weighting.
-At this stage, all exploratory analyses will be conducted blinded
-towards the outcome to not bias any study design or analytic choices
-based on known outcome information.
+to ensure that measured pre-exposure covariates can be
+balanced\textsuperscript{28}, exposure cohorts are conditionally
+exchangeable at baseline, and resulting sample sizes are still
+sufficient after matching or weighting. At this stage, all exploratory
+analyses will be conducted blinded towards the outcome to not bias any
+study design or analytic choices based on known outcome information.
 
@@
 manner as RCTs since the number of `recruited' patients is limited by
-available data .\textsuperscript{28} For this project, however, the
+available data .\textsuperscript{29} For this project, however, the
 emulations in EHR data must have at least equal power to the relevant
@@
 estimation of the statistical power is driven by the number of events
-rather than the number of patients. To assess if the overall number of
-events, unstratified by exposure, is sufficient such that a significant
-difference can be detected based on the original RCT-reported hazard
-ratio (HR), the statistical power (1-\(\beta\)) will be estimated using
-Schoenfeld's sample-size formula for the proportional-hazards regression
-model.\textsuperscript{29}
+rather than the number of patients. To assess whether the overall number
+of events, unstratified by exposure, is sufficient such that a
+significant difference can be detected based on the original
+RCT-reported hazard ratio (HR), the statistical power (1-\(\beta\) with
+a 2-sided \(\alpha\) of 0.05) will be estimated using Schoenfeld's
+sample-size formula for the proportional-hazards regression
+model.\textsuperscript{30}
 
@@
 emulations, we will adapt the approach of the RCT-DUPLICATE
-project.\textsuperscript{8,30} That is, for the primary endpoint of
+project.\textsuperscript{8,31} That is, for the primary endpoint of
 interest (hazard ratio {[}HR{]} for OS and corresponding 95\% confidence
@@
 where \(\hat{\theta}_{RCT}\) and \(\hat{\theta}_{RWE}\) are the
-treatment effect estimates (hazard ratios or median survival time
+treatment effect estimates (log hazard ratios or median survival time
 differences) and \(\text{Var}(\hat{\theta}_{RCT})\) and
@@
 windows of eligibility criteria, washout periods, and covariates
-relative to the cohort entry time.\textsuperscript{31}
+relative to the cohort entry time.\textsuperscript{32}
 
@@
 missing completely at random {[}MCAR{]}, missing at random {[}MAR{]} and
-missing not at random {[}MNAR{]}).\textsuperscript{32} We will adopt a
+missing not at random {[}MNAR{]}).\textsuperscript{33} We will adopt a
 principled process on missing data that empirically evaluates different
 aspects across partially observed covariates based on three group
-diagnostics.\textsuperscript{33,34} These diagnostics cover (1)
+diagnostics.\textsuperscript{34,35} These diagnostics cover (1)
 comparisons of patients characteristics with and without an observed
@@
 potential sensitivity analyses to assess the robustness of these
-decisions.\textsuperscript{36}
-
-While the MAR assumption is a strong assumption to hold across all
-considered covariates, it has been shown that especially in the context
-of partially observed covariate data (as opposed to missing exposure and
+decisions.\textsuperscript{37}
+
+While MAR is a strong assumption to hold across all considered
+covariates, it has been shown that especially in the context of
+partially observed covariate data (as opposed to missing exposure and
 outcome data), generally only mechanisms in which a covariate causes its
-own missingness lead to significant bias (MNAR).\textsuperscript{35}
+own missingness lead to significant bias (MNAR).\textsuperscript{36}
 Hence, methodologies which retain otherwise eligible patients and give
 the potential to adjust for a broader set of prognostic factors (e.g.,
-multiple imputation\textsuperscript{37} or doubly robust
-methods\textsuperscript{38}) may be preferred over complete case
+multiple imputation\textsuperscript{38} or doubly robust
+methods\textsuperscript{39}) may be preferred over complete case
 analyses, although such tradeoffs must be carefully evaluated within the
@@
 interest in ENCORE will be defined as the marginal HR coefficient for
-the treatment comparison for OS.\textsuperscript{39} We will also
+the treatment comparison for OS.\textsuperscript{40} We will also
 consider alternative endpoints on an absolute risk scale such as median
 survival times, survival probabilities at pre-defined time points during
-follow-up\textsuperscript{40} and restricted mean survival times as
+follow-up\textsuperscript{41} and restricted mean survival times as
 secondary endpoints of interest.
@@
 based on expert clinical knowledge and published literature on
-prognostic scores in oncology.\textsuperscript{41} The implementation of
+prognostic scores in oncology.\textsuperscript{42} The implementation of
 propensity scores in combination with multiple imputation will follow
 the `\emph{within}' methodology as described by Leyrat et
-al.\textsuperscript{42,43} As appropriate, propensity score matching or
+al.\textsuperscript{43,44} As appropriate, propensity score matching or
 weighting will be applied to each imputed dataset. The marginal
@@
 weighted dataset separately and pooled into a final estimate following
-Rubin's rule.\textsuperscript{44,45} This approach has been shown to
+Rubin's rule.\textsuperscript{45,46} This approach has been shown to
 lead to unbiased estimates across different simulated scenarios with a
-sufficient estimation of the variance.\textsuperscript{42}
+sufficient estimation of the variance.\textsuperscript{43}
 
@@
 Figure~\ref{fig-balance}). Covariate balance is often considered
-reasonable at a SMD \textless{} 0.1.\textsuperscript{46} Further, we
+reasonable at a SMD \textless{} 0.1.\textsuperscript{28} Further, we
 will compute the average post-matching or post-weighting
 C-statistics.\textsuperscript{47} In addition, we will use a published
-prognostic score for OS\textsuperscript{41} as a balance measure to
+prognostic score for OS\textsuperscript{42} as a balance measure to
 visually assess if the prognostic score is balanced between treatment
@@
 Building on an established approach developed by the RCT DUPLICATE
-project\textsuperscript{8,30}, the ENCORE project aims to emulate 12
+project\textsuperscript{8,31}, the ENCORE project aims to emulate 12
 oncology trials using four oncology EHR data sources to inform the
@@
 
-The RCT DUPLICATE initiative\textsuperscript{8,30} identified multiple
+The RCT DUPLICATE initiative\textsuperscript{8,31} identified multiple
 emulation challenges that may be similar in ENCORE. One particular
@@
 
+\bibitem[\citeproctext]{ref-austin2009balance}
+\CSLLeftMargin{28. }%
+\CSLRightInline{Austin, P. C. Balance diagnostics for comparing the
+distribution of baseline covariates between treatment groups in
+propensity-score matched samples. \emph{Statistics in medicine}
+\textbf{28}, 3083--3107 (2009).}
+
 \bibitem[\citeproctext]{ref-hernan2022causal}
-\CSLLeftMargin{28. }%
+\CSLLeftMargin{29. }%
 \CSLRightInline{Hernán, M. A. Causal analyses of existing databases: No
@@
 \bibitem[\citeproctext]{ref-schoenfeld1981asymptotic}
-\CSLLeftMargin{29. }%
+\CSLLeftMargin{30. }%
 \CSLRightInline{Schoenfeld, D. The asymptotic properties of
@@
 \bibitem[\citeproctext]{ref-franklin2020nonrandomized}
-\CSLLeftMargin{30. }%
+\CSLLeftMargin{31. }%
 \CSLRightInline{Franklin, J. M. \emph{et al.} Nonrandomized real-world
@@
 \bibitem[\citeproctext]{ref-schneeweiss2019graphical}
-\CSLLeftMargin{31. }%
+\CSLLeftMargin{32. }%
 \CSLRightInline{Schneeweiss, S. \emph{et al.} Graphical depiction of
@@
 \bibitem[\citeproctext]{ref-rubin1976inference}
-\CSLLeftMargin{32. }%
+\CSLLeftMargin{33. }%
 \CSLRightInline{Rubin, D. B. Inference and missing data.
@@
 \bibitem[\citeproctext]{ref-weberpals2024smdi}
-\CSLLeftMargin{33. }%
+\CSLLeftMargin{34. }%
 \CSLRightInline{Weberpals, J. \emph{et al.} Smdi: An r package to
@@
 \bibitem[\citeproctext]{ref-weberpals2024}
-\CSLLeftMargin{34. }%
+\CSLLeftMargin{35. }%
 \CSLRightInline{Weberpals, J. \emph{et al.}
@@
 \bibitem[\citeproctext]{ref-moreno2018canonical}
-\CSLLeftMargin{35. }%
+\CSLLeftMargin{36. }%
 \CSLRightInline{Moreno-Betancur, M. \emph{et al.} Canonical causal
@@
 \bibitem[\citeproctext]{ref-tompsett2018use}
-\CSLLeftMargin{36. }%
+\CSLLeftMargin{37. }%
 \CSLRightInline{Tompsett, D. M., Leacy, F., Moreno-Betancur, M., Heron,
@@
 \bibitem[\citeproctext]{ref-weberpals2024hdmi}
-\CSLLeftMargin{37. }%
+\CSLLeftMargin{38. }%
 \CSLRightInline{Weberpals, J. \emph{et al.} High-dimensional multiple
@@
 \bibitem[\citeproctext]{ref-Shaw2024}
-\CSLLeftMargin{38. }%
+\CSLLeftMargin{39. }%
 \CSLRightInline{Shaw, C. K., P. \& Williamson, B. D. Assessing treatment
@@
 \bibitem[\citeproctext]{ref-cox1972regression}
-\CSLLeftMargin{39. }%
+\CSLLeftMargin{40. }%
 \CSLRightInline{Cox, D. R. Regression models and life-tables. Journal of
@@
 \bibitem[\citeproctext]{ref-kaplan1958nonparametric}
-\CSLLeftMargin{40. }%
+\CSLLeftMargin{41. }%
 \CSLRightInline{Kaplan, E. L. \& Meier, P. Nonparametric estimation from
@@
 \bibitem[\citeproctext]{ref-becker2020enhanced}
-\CSLLeftMargin{41. }%
+\CSLLeftMargin{42. }%
 \CSLRightInline{Becker, T. \emph{et al.} An enhanced prognostic score
@@
 \bibitem[\citeproctext]{ref-leyrat2019propensity}
-\CSLLeftMargin{42. }%
+\CSLLeftMargin{43. }%
 \CSLRightInline{Leyrat, C. \emph{et al.} Propensity score analysis with
@@
 \bibitem[\citeproctext]{ref-pishgar2020matchthem}
-\CSLLeftMargin{43. }%
+\CSLLeftMargin{44. }%
 \CSLRightInline{Pishgar, F., Greifer, N., Leyrat, C. \& Stuart, E.
@@
 \bibitem[\citeproctext]{ref-rubin2018multiple}
-\CSLLeftMargin{44. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{Rubin, D. B. Multiple imputation. In \emph{Flexible
@@
 \bibitem[\citeproctext]{ref-van2011mice}
-\CSLLeftMargin{45. }%
+\CSLLeftMargin{46. }%
 \CSLRightInline{Van Buuren, S. \& Groothuis-Oudshoorn, K. Mice:
@@
 statistical software} \textbf{45}, 1--67 (2011).}
-
-\bibitem[\citeproctext]{ref-austin2009balance}
-\CSLLeftMargin{46. }%
-\CSLRightInline{Austin, P. C. Balance diagnostics for comparing the
-distribution of baseline covariates between treatment groups in
-propensity-score matched samples. \emph{Statistics in medicine}
-\textbf{28}, 3083--3107 (2009).}
 

### 2025-10-22 manuscript/manuscript.qmd
updated word count
implemented changes that were introduced in the final FDA cleared version

@@
 Kehl^2^, Donna R.
-Rivera^3^, Pallavi Mishra-Kalyani^3^, Catherine C.
-Lerro^3^, Erin Larkins^4^, Sonia Singh^4\*^, Preeti Narayan^4^, Richard Curley^4^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^5^, Andrew J.
+Rivera^3\*^, Pallavi Mishra-Kalyani^3^, Catherine C.
+Lerro^3\*^, Erin Larkins^4^, Preeti Narayan^4^, Richard Curley^3^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^5^, Andrew J.
 Belli^6^, Ching-Kun Wang^6^, Jenna Collins^7^**,** Jonathan Kish^7^**,** Janet Espirito^8^, Nicholas J Robert^8^, Robert J.
@@
 
-[**Abstract word count:**]{.underline} 228 words / 250 words
+[**Abstract word count:**]{.underline} 224 words / 250 words
 
@@
 
-[**Funding Statement:**]{.underline} This project was supported by an FDA BAA contract with contract number 75F40122C00181.
+[**FDA Disclaimer:**]{.underline}This article reflects the views of the authors and should not be construed to represent FDA’s views or policies.
+
+[**Funding Statement:**]{.underline} This project was supported by the Oncology Center of Excellence, Food and Drug Administration (FDA) of the U.S.
+Department of Health and Human Services (HHS) as part of a contract \[75F40122C00181\].
+The contents are those of the author(s) and do not necessarily represent the official views of, nor an endorsement by FDA/HHS, or the U.S.
+Government.
 

### 2025-10-22 manuscript/manuscript.tex
Manuscript updated

@@
 Schneeweiss\textsuperscript{1}, Kenneth L. Kehl\textsuperscript{2},
-Donna R. Rivera\textsuperscript{3}, Pallavi
+Donna R. Rivera\textsuperscript{3*}, Pallavi
 Mishra-Kalyani\textsuperscript{3}, Catherine C.
-Lerro\textsuperscript{3}, Erin Larkins\textsuperscript{4}, Sonia
-Singh\textsuperscript{4*}, Preeti Narayan\textsuperscript{4}, Richard
-Curley\textsuperscript{4}, Georg Hahn\textsuperscript{1}, Priyanka
-Anand\textsuperscript{1}, Yanina Natanzon\textsuperscript{5}, Andrew J.
-Belli\textsuperscript{6}, Ching-Kun Wang\textsuperscript{6}, Jenna
+Lerro\textsuperscript{3*}, Erin Larkins\textsuperscript{4}, Preeti
+Narayan\textsuperscript{4}, Richard Curley\textsuperscript{3}, Georg
+Hahn\textsuperscript{1}, Priyanka Anand\textsuperscript{1}, Yanina
+Natanzon\textsuperscript{5}, Andrew J. Belli\textsuperscript{6},
+Ching-Kun Wang\textsuperscript{6}, Jenna
 Collins\textsuperscript{7}\textbf{,} Jonathan
@@
 
-\ul{\textbf{Abstract word count:}} 228 words / 250 words
+\ul{\textbf{Abstract word count:}} 224 words / 250 words
 
@@
 
-\ul{\textbf{Funding Statement:}} This project was supported by an FDA
-BAA contract with contract number 75F40122C00181.
+\ul{\textbf{FDA Disclaimer:}}This article reflects the views of the
+authors and should not be construed to represent FDA's views or
+policies.
+
+\ul{\textbf{Funding Statement:}} This project was supported by the
+Oncology Center of Excellence, Food and Drug Administration (FDA) of the
+U.S. Department of Health and Human Services (HHS) as part of a contract
+{[}75F40122C00181{]}. The contents are those of the author(s) and do not
+necessarily represent the official views of, nor an endorsement by
+FDA/HHS, or the U.S. Government.
 
@@
 
-\emph{Manuscript last updated: 2025-07-26 11:39:19.945045}
+\emph{Manuscript last updated: 2025-10-22 19:30:06.576264}
 
